### 2023-03-17 result_sandbox.Rmd
Drafting figures for manuscript based on results

---
title: "MS_metrics results"
author: "William Kumler"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
FT2040_features <- read_csv("made_data_FT2040/features_extracted.csv")
MS3000_features <- read_csv("made_data_MS3000/features_extracted.csv")
```

Can we simply train up a model on one dataset and apply it to another, i.e. how much does our performance suffer when switching to a different dataset?

Test with MESOSCOPE and Falkor, two fully-labeled feature-based datasets. Train up with Falkor and test on MESOSCOPE, train up on MESOSCOPE and test on Falkor, compare model outputs (confusion matrices and model estimates).

```{r Confusion matrices and figure}
dataset_versions <- c("FT2040_features", "MS3000_features")
traintest_gp <- lapply(dataset_versions, function(train_set){
  full_model <- get(train_set) %>% 
    select(-feat_id) %>%
    filter(feat_class%in%c("Good", "Bad")) %>%
    mutate(feat_class=feat_class=="Good") %>%
    glm(formula=feat_class~., family = binomial)
  lapply(dataset_versions, function(test_set){
    get(test_set) %>%
      select(-feat_id) %>%
      filter(feat_class%in%c("Good", "Bad")) %>%
      mutate(pred_prob=predict(object=full_model,newdata = ., 
                               type = "response")) %>%
      mutate(pred_class_50=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
      mutate(pred_class_99=ifelse(pred_prob>0.99, "Good", "Bad")) %>%
      mutate(pred_class_01=ifelse(pred_prob>0.01, "Good", "Bad")) %>%
      select(feat_class, starts_with("pred_class")) %>%
      mutate(train_test=paste(train_set, test_set, sep = "-")) %>%
      mutate(train_test=str_remove_all(train_test, "_features")) %>%
      mutate(cross_type=ifelse(train_set==test_set, "Within", "Cross"))
  }) %>% bind_rows()
}) %>%
  bind_rows() %>%
  pivot_longer(cols = starts_with("pred_class"), 
               names_to = "threshold", 
               values_to = "pred_class") %>%
  mutate(error_type=case_when(
    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
    feat_class=="Bad" & pred_class=="Good" ~ "FP",
    feat_class=="Good" & pred_class=="Bad" ~ "FN",
    feat_class=="Good" & pred_class=="Good" ~ "TP"
  )) %>%
  group_by(cross_type, train_test, threshold) %>%
  count(error_type) %>%
  mutate(perc=round(n/sum(n)*100)) %>%
  ungroup() %>%
  complete(error_type, train_test, threshold, fill = list(n=0, perc=0)) %>%
  mutate(cross_type=ifelse(is.na(cross_type), "Within", cross_type)) %>%
  # Rename things to look nice in ggplot
  mutate(error_type=paste0("%", error_type)) %>%
  mutate(error_type=factor(error_type, paste0("%", c("TN", "TP", "FN", "FP")))) %>%
  mutate(train_test=str_replace_all(train_test, "FT2040", "Falkor")) %>%
  mutate(train_test=str_replace_all(train_test, "MS3000", "MESO")) %>%
  mutate(threshold=as.numeric(str_extract(threshold, "\\d+"))) %>%
  mutate(threshold=paste("Threshold:", threshold/100)) %>%
  mutate(cross_type=paste(cross_type, "dataset", sep = "-")) %>% 
  mutate(cross_type=factor(cross_type, levels=c("Within-dataset", "Cross-dataset"))) %>%
  arrange(train_test) %>% 
  mutate(train_test=factor(train_test, levels=unique(train_test))) %>%
  # Render the ggplot
  ggplot(aes(x=error_type, y=perc)) +
  geom_col(aes(fill=train_test), position = position_dodge()) +
  geom_text(aes(label=n, color=train_test), 
            position = position_dodge(width = 1), vjust=-0.2) +
  facet_grid(threshold~cross_type, scales = "free_x") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_color_discrete("Train-test", aesthetics=c("color", "fill")) +
  theme_bw() +
  theme(axis.title = element_blank())
xcms_gp <- list(Falkor=FT2040_features, MESO=MS3000_features) %>%
  bind_rows(.id = "cruise") %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  group_by(cruise) %>%
  count(feat_class) %>%
  mutate(perc=round(n/sum(n)*100)) %>%
  ungroup() %>%
  add_row(cruise=c("Falkor", "MESO"), feat_class="Bad", perc=NA) %>%
  add_row(cruise=c("Falkor", "MESO"), feat_class="Good", perc=NA) %>%
  arrange(cruise, feat_class, n) %>%
  mutate(pred_class=rep(c("Good", "Bad"), length.out=8)) %>%
  mutate(error_type=case_when(
    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
    feat_class=="Bad" & pred_class=="Good" ~ "FP",
    feat_class=="Good" & pred_class=="Bad" ~ "FN",
    feat_class=="Good" & pred_class=="Good" ~ "TP"
  )) %>%
  mutate(error_type=paste0("%", error_type)) %>%
  mutate(error_type=factor(error_type, paste0("%", c("TN", "TP", "FN", "FP")))) %>%
  ggplot(aes(x=error_type, y=perc)) +
  geom_col(aes(fill=cruise), position = "dodge") +
  geom_text(aes(label=n, color=cruise), 
            position = position_dodge(width = 1), vjust=-0.2) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_fill_discrete("Dataset", aesthetics=c("color", "fill")) +
  theme_bw() +
  theme(legend.background = element_rect(color="black"),
        axis.title = element_blank()) +
  # theme(legend.position = c(0.5, 1),
  #       legend.justification = c(0.5, 1), 
  #       legend.direction = "horizontal") +
  theme(legend.position = c(0, 1),
        legend.justification = c(0, 1),
        legend.direction = "vertical")
plot(gridExtra::arrangeGrob(
  xcms_gp + ggtitle("A. XCMS output"), 
  traintest_gp + ggtitle("B. Performance of logistic model"), 
  layout_matrix = matrix(c(1,2,2,2), nrow = 1)))
```

```{r Confusion matrices and figure, v2 clean}
dataset_versions <- c("FT2040_features", "MS3000_features")
confusion_data <- lapply(dataset_versions, function(train_set){
  full_model <- get(train_set) %>% 
    select(-feat_id) %>%
    filter(feat_class%in%c("Good", "Bad")) %>%
    mutate(feat_class=feat_class=="Good") %>%
    glm(formula=feat_class~., family = binomial)
  lapply(dataset_versions, function(test_set){
    get(test_set) %>%
      select(-feat_id) %>%
      filter(feat_class%in%c("Good", "Bad")) %>%
      mutate(pred_prob=predict(object=full_model,newdata = ., 
                               type = "response")) %>%
      mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
      select(feat_class, starts_with("pred_class")) %>%
      mutate(train_test=paste(train_set, test_set, sep = "-")) %>%
      mutate(train_test=str_remove_all(train_test, "_features")) %>%
      mutate(cross_type=ifelse(train_set==test_set, "Within", "Cross"))
  }) %>% bind_rows()
}) %>%
  bind_rows() %>%
  mutate(error_type=case_when(
    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
    feat_class=="Bad" & pred_class=="Good" ~ "FP",
    feat_class=="Good" & pred_class=="Bad" ~ "FN",
    feat_class=="Good" & pred_class=="Good" ~ "TP"
  )) %>%
  group_by(cross_type, train_test) %>%
  count(error_type) %>%
  mutate(perc=round(n/sum(n)*100)) %>%
  ungroup() %>%
  # Rename things to look nice in ggplot
  mutate(error_type=paste0("%", error_type)) %>%
  mutate(error_type=factor(error_type, paste0("%", c("TN", "TP", "FN", "FP")))) %>%
  mutate(train_test=str_replace_all(train_test, "FT2040", "Falkor")) %>%
  mutate(train_test=str_replace_all(train_test, "MS3000", "MESO")) %>%
  mutate(cross_type=paste(cross_type, "dataset", sep = "-")) %>% 
  mutate(cross_type=factor(cross_type, levels=c("Within-dataset", "Cross-dataset"))) %>%
  arrange(train_test) %>% 
  mutate(train_test=factor(train_test, levels=unique(train_test)))

within_gp <- confusion_data %>%
  filter(cross_type=="Within-dataset") %>%
  ggplot(aes(x=error_type, y=perc)) +
  geom_col(aes(fill=train_test), position = position_dodge()) +
  geom_text(aes(label=n, color=train_test), 
            position = position_dodge(width = 1), vjust=-0.2) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_color_discrete("Train-test", aesthetics=c("color", "fill")) +
  theme_bw() +
  theme(axis.title = element_blank(),
      legend.background = element_rect(color="black"),
      legend.position = c(1, 1),
      legend.justification = c(1, 1),
      legend.direction = "vertical",
      axis.text.y = element_blank())
cross_gp <- confusion_data %>%
  filter(cross_type=="Cross-dataset") %>%
  ggplot(aes(x=error_type, y=perc)) +
  geom_col(aes(fill=train_test), position = position_dodge()) +
  geom_text(aes(label=n, color=train_test), 
            position = position_dodge(width = 1), vjust=-0.2) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_color_discrete("Train-test", aesthetics=c("color", "fill")) +
  theme_bw() +
  theme(axis.title = element_blank(),
      legend.background = element_rect(color="black"),
      legend.position = c(1, 1),
      legend.justification = c(1, 1),
      legend.direction = "vertical",
      axis.text.y = element_blank())
xcms_gp <- list(Falkor=FT2040_features, MESO=MS3000_features) %>%
  bind_rows(.id = "cruise") %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  group_by(cruise) %>%
  count(feat_class) %>%
  mutate(perc=round(n/sum(n)*100)) %>%
  ungroup() %>%
  add_row(cruise=c("Falkor", "MESO"), feat_class="Bad", perc=NA) %>%
  add_row(cruise=c("Falkor", "MESO"), feat_class="Good", perc=NA) %>%
  arrange(cruise, feat_class, n) %>%
  mutate(pred_class=rep(c("Good", "Bad"), length.out=8)) %>%
  mutate(error_type=case_when(
    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
    feat_class=="Bad" & pred_class=="Good" ~ "FP",
    feat_class=="Good" & pred_class=="Bad" ~ "FN",
    feat_class=="Good" & pred_class=="Good" ~ "TP"
  )) %>%
  mutate(error_type=paste0("%", error_type)) %>%
  mutate(error_type=factor(error_type, paste0("%", c("TN", "TP", "FN", "FP")))) %>%
  ggplot(aes(x=error_type, y=perc)) +
  geom_col(aes(fill=cruise), position = "dodge") +
  geom_text(aes(label=n, color=cruise), 
            position = position_dodge(width = 1), vjust=-0.2) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_fill_discrete("Dataset", aesthetics=c("color", "fill")) +
  theme_bw() +
  theme(axis.title = element_blank(),
        legend.background = element_rect(color="black"),
        legend.position = c(0, 1),
        legend.justification = c(0, 1),
        legend.direction = "vertical")
plot(gridExtra::arrangeGrob(
  xcms_gp + ggtitle("A. Default XCMS output\n    (No model fitting)"), 
  within_gp + ggtitle("B. Performance within\n     a given dataset"), 
  cross_gp + ggtitle("C. Performance across\n     datasets"), 
  layout_matrix = matrix(c(1,2,3), nrow = 1)))
```

```{r Thresholding demo}
meso_full_model <- MS3000_features %>% 
  select(-feat_id) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)

plot_breaks <- c(0.01, 0.1, 0.5, 0.9, 0.99)
base_plot <- MS3000_features %>%
  mutate(pred_prob=predict(object=meso_full_model,newdata = ., type = "response")) %>%
  mutate(cut_prob=cut(pred_prob, breaks = c(0, plot_breaks, 1),include.lowest = TRUE)) %>%
  mutate(feat_class=factor(feat_class, levels=c("Bad", "Stans only", "Meh", "Good"))) %>%
  ggplot() +
  geom_bar(aes(x=cut_prob, fill=feat_class), 
           position =position_dodge(preserve = "single")) +
  facet_wrap(~cut_prob, nrow=1, scales = "free_x") +
  scale_y_log10() +
  scale_fill_discrete("Manual feature classification") +
  labs(x="Predicted probability (binned) of being a good peak", y="Count") +
  theme_bw() +
  theme(legend.position = "top", legend.justification = "left") +
  guides(fill=guide_legend(title.position="top"))

threshold_plot <- lapply(plot_breaks, function(threshold){
  MS3000_features %>%
    filter(feat_class%in%c("Good", "Bad")) %>%
    mutate(pred_prob=predict(object=meso_full_model,newdata = ., type = "response")) %>%
    mutate(pred_class=ifelse(pred_prob>threshold, "Good", "Bad")) %>%
    mutate(error_type=case_when(
      feat_class=="Bad" & pred_class=="Bad" ~ "TN",
      feat_class=="Bad" & pred_class=="Good" ~ "FP",
      feat_class=="Good" & pred_class=="Bad" ~ "FN",
      feat_class=="Good" & pred_class=="Good" ~ "TP"
    )) %>%
    count(error_type) %>%
    mutate(perc=round(n/sum(n)*100)) %>%
    ungroup() %>%
    mutate(threshold=threshold)
}) %>%
  bind_rows() %>%
  mutate(error_type=paste0("%", error_type)) %>%
  mutate(error_type=factor(error_type, paste0("%", c("TN", "TP", "FN", "FP")))) %>%
  mutate(threshold=paste("Threshold:", threshold)) %>%
  ggplot(aes(x=error_type, y=perc)) +
  geom_col() +
  geom_text(aes(label=n), vjust=-0.5) +
  facet_wrap(~threshold, nrow = 1) +
  theme_bw() +
  theme(axis.title = element_blank()) +
  scale_y_continuous(limits = c(0, 100))

layout_mat <- matrix(c(c(1,1,1,1,1,1,1,1,1,1),c(NA,2,2,2,2,2,2,2,2,NA)), nrow = 2, byrow = TRUE)
plot(gridExtra::arrangeGrob(base_plot, threshold_plot, layout_matrix = layout_mat))
```

```{r Backup confusion matrix visuals, eval=FALSE}
full_cmat <- rbind(
  rbind(
    data.frame(f2f_cmat) %>% mutate(tested_on="falkor"),
    data.frame(f2m_cmat) %>% mutate(tested_on="mesoscope")
  ) %>% mutate(trained_on="falkor"),
  rbind(
    data.frame(m2m_cmat) %>% mutate(tested_on="mesoscope"),
    data.frame(m2f_cmat) %>% mutate(tested_on="falkor")
  ) %>% mutate(trained_on="mesoscope")
)

library(ggplot2)
full_cmat %>%
  mutate(predicted=paste("Predicted to\nbe", predicted)) %>%
  mutate(predicted=factor(predicted, levels=c("Predicted to\nbe Good", "Predicted to\nbe Bad"))) %>%
  mutate(actual=paste("Actually", actual)) %>%
  mutate(tested_on=paste("Tested on", tested_on)) %>%
  mutate(trained_on=paste("Trained on", trained_on)) %>%
  ggplot() +
  geom_text(aes(x=actual, y=predicted, label=Freq)) +
  facet_grid(tested_on~trained_on, switch = "y") +
  scale_x_discrete(position = "top") + 
  theme_bw() +
  theme(strip.placement = "outside", axis.title = element_blank(),
        axis.text.y = element_text(angle=90, hjust = 0.5, vjust = 0),
        strip.background = element_rect(fill = "white"))

library(flextable)
full_cmat %>%
  arrange(trained_on, tested_on, predicted, actual) %>%
  select(`Trained on`=trained_on, `Tested on`=tested_on, Predicted=predicted, 
         Actual=actual, Count=Freq) %>%
  flextable() %>%
  merge_v(j = "Trained on") %>%
  merge_v(j = "Tested on") %>%
  merge_v(j = "Predicted") %>%
  align(align = "center") %>%
  rotate(i = 1:16, j=1:2, rotation = "btlr") %>%
  theme_box()

library(gt)
cmat_to_df <- function(cmat){
  outmat <- matrix(c(
    "", "Bad", "Good",
    "Bad", cmat[1,1], cmat[1,2],
    "Good", cmat[2,1], cmat[2,2]
  ), nrow = 3, ncol = 3)
  colnames(outmat) <- c("Predicted", "Actual", "Actual2")
  outmat
}
as.data.frame(rbind(
  rbind(cmat_to_df(f2f_cmat), cmat_to_df(f2m_cmat)),
  rbind(cmat_to_df(m2m_cmat), cmat_to_df(m2f_cmat)
  ))) %>%
  gt() %>%
  tab_row_group(
    label="Trained on Mesoscope, tested on Falkor", rows = 10:12
  ) %>%
  tab_row_group(
    label="Trained on Mesoscope, tested on Mesoscope", rows = 7:9
  ) %>%
  tab_row_group(
    label="Trained on Falkor, tested on Mesoscope", rows = 4:6
  ) %>%
  tab_row_group(
    label="Trained on Falkor, tested on Falkor", rows = 1:3
  ) %>%
  tab_options(column_labels.hidden = TRUE)
```

Model estimates:

```{r Model comparison}
falkor_full_model <- FT2040_features %>% 
  select(-feat_id) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)
meso_full_model <- MS3000_features %>% 
  select(-feat_id) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)

good_params <-list(falkor=broom::tidy(falkor_full_model),
                   meso=broom::tidy(meso_full_model)) %>%
  bind_rows(.id="dataset") %>%
  group_by(term) %>%
  mutate(keep_term=all(p.value<0.05)) %>%
  filter(keep_term)

good_params %>%
  ggplot(aes(x=term, color=dataset)) +
  geom_hline(yintercept = 0, color="black") +
  geom_point(aes(y=estimate), position = position_dodge(width = 0.5), size=3) +
  geom_errorbar(aes(ymin=estimate-2*std.error, ymax=estimate+2*std.error), 
                position = position_dodge(width = 0.5), linewidth=1) +
  facet_wrap(~term, scales = "free") +
  labs(x=NULL, y="Model parameter estimate") +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

bind_rows(falkor=FT2040_features, meso=MS3000_features, .id = "dataset") %>%
  select(dataset, all_of(unique(good_params$term)), feat_class) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=ifelse(feat_class=="Good", 1, 0)) %>%
  mutate(feat_class_jitter=feat_class+rnorm(nrow(.), sd = 0.02)) %>%
  pivot_longer(-c(dataset, feat_class, feat_class_jitter), names_to = "param") %>%
  slice_sample(prop = 1) %>%
  ggplot(aes(x=value, y=feat_class)) +
  geom_point(aes(y=feat_class_jitter, color=dataset), alpha=0.1) +
  geom_smooth(aes(group=dataset), formula="y~x", method = "glm",
              method.args=list(family=binomial),
              se=FALSE, linewidth=2, color="white") +
  geom_smooth(aes(color=dataset), formula="y~x", method = "glm",
              method.args=list(family=binomial)) +
  facet_wrap(~param, scales = "free_x")

bind_rows(list(Falkor=FT2040_features, MESOSCOPE=MS3000_features), .id = "cruise") %>%
  ggplot() + 
  geom_histogram(aes(x=med_cor, fill=feat_class), bins=40) +
  facet_wrap(~cruise, ncol = 1, scales = "free_y")
```

Repeat the above with a reduced set of predictors (just the "best" ones). Maybe also with just XCMS parameters?

```{r feature selection and selection testing}
raw_data_params <- c("med_cor", "med_SNR")
xcms_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", 
                 "sd_pw", "log_mean_height", "log_sd_height", "sn", 
                 "f", "scale", "lmin", "feat_npeaks", "n_found", 
                 "samps_found", "stans_found")
all_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", "sd_pw",
                "log_mean_height", "log_sd_height", "sn", "f", "scale", 
                "lmin", "feat_npeaks", "n_found", "samps_found", "stans_found", 
                "med_cor", "med_SNR", "med_missed_scans", "smp_to_blk", 
                "smp_to_std", "shape_cor", "area_cor", "feat_class")
lst(all_params, raw_data_params, xcms_params) %>%
  imap(function(param_selection, param_name){
    lapply(dataset_versions, function(train_set){
      full_model <- get(train_set) %>% 
        select(feat_class, all_of(param_selection)) %>%
        filter(feat_class%in%c("Good", "Bad")) %>%
        mutate(feat_class=feat_class=="Good") %>%
        glm(formula=feat_class~., family = binomial)
      lapply(dataset_versions, function(test_set){
        get(test_set) %>%
          filter(feat_class%in%c("Good", "Bad")) %>%
          mutate(pred_prob=predict(object=full_model,newdata = ., 
                                   type = "response")) %>%
          mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
          select(feat_class, starts_with("pred_class")) %>%
          mutate(train_test=paste(train_set, test_set, sep = "-")) %>%
          mutate(train_test=str_remove_all(train_test, "_features")) %>%
          mutate(cross_type=ifelse(train_set==test_set, "Within", "Cross"))
      }) %>% bind_rows()
    }) %>% bind_rows() %>% mutate(which_params=param_name)
  }) %>% bind_rows() %>%
  mutate(error_type=case_when(
    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
    feat_class=="Bad" & pred_class=="Good" ~ "FP",
    feat_class=="Good" & pred_class=="Bad" ~ "FN",
    feat_class=="Good" & pred_class=="Good" ~ "TP"
  )) %>%
  group_by(cross_type, train_test, which_params) %>%
  count(error_type) %>%
  mutate(perc=round(n/sum(n)*100)) %>%
  ungroup() %>%
  # Rename things to look nice in ggplot
  mutate(error_type=paste0("%", error_type)) %>%
  mutate(error_type=factor(error_type, paste0("%", c("TN", "TP", "FN", "FP")))) %>%
  mutate(train_test=str_replace_all(train_test, "FT2040", "Falkor")) %>%
  mutate(train_test=str_replace_all(train_test, "MS3000", "MESO")) %>%
  mutate(cross_type=paste(cross_type, "dataset", sep = "-")) %>% 
  mutate(cross_type=factor(cross_type, levels=c("Within-dataset", "Cross-dataset"))) %>%
  arrange(train_test) %>% 
  mutate(train_test=factor(train_test, levels=unique(train_test))) %>%
  # Render the ggplot
  ggplot(aes(x=error_type, y=perc)) +
  geom_col(aes(fill=train_test), position = position_dodge()) +
  geom_text(aes(label=n, color=train_test), 
            position = position_dodge(width = 1), vjust=-0.2) +
  facet_grid(which_params~cross_type, scales = "free_x") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_color_discrete("Train-test", aesthetics=c("color", "fill")) +
  theme_bw() +
  theme(axis.title = element_blank())
```

Should we include the subsetting analysis I did where we run the models on a subset of the data to see how large of a sample size we need to reach a stable model?

```{r model stability sample size subsetting}
# Full model
meso_full_model <- MS3000_features %>% 
  select(-feat_id) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial) %>%
  broom::tidy()
set.seed(123)
meso_reps <- replicate(20, {
  meso_full_model <- MS3000_features %>% 
    slice_sample(prop = 0.5) %>%
    select(-feat_id) %>%
    filter(feat_class%in%c("Good", "Bad")) %>%
    mutate(feat_class=feat_class=="Good") %>%
    glm(formula=feat_class~., family = binomial) %>%
    broom::tidy()
}, simplify = FALSE) %>%
  bind_rows()
ggplot() +
  geom_vline(xintercept = 0) +
  geom_point(aes(x=estimate, y=term), data = meso_full_model, size=3) +
  geom_errorbar(aes(xmin=estimate-std.error*2, xmax=estimate+std.error*2, y=term), 
                data = meso_full_model, linewidth=1) +
  geom_point(aes(x=estimate, y=term), data = meso_reps, color="red") +
  facet_wrap(~term, scales = "free") +
  theme(axis.text.y=element_blank())

# Reduced model
meso_min_model <- MS3000_features %>% 
  select(feat_class, med_cor, med_SNR) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial) %>%
  broom::tidy()
set.seed(123)
meso_reps <- replicate(20, {
  meso_min_model <- MS3000_features %>% 
    slice_sample(prop = 0.5) %>%
    select(feat_class, med_cor, med_SNR) %>%
    filter(feat_class%in%c("Good", "Bad")) %>%
    mutate(feat_class=feat_class=="Good") %>%
    glm(formula=feat_class~., family = binomial) %>%
    broom::tidy()
}, simplify = FALSE) %>%
  bind_rows()
ggplot() +
  geom_vline(xintercept = 0) +
  geom_point(aes(x=estimate, y=term), data = meso_min_model, size=3) +
  geom_errorbar(aes(xmin=estimate-std.error*2, xmax=estimate+std.error*2, y=term), 
                data = meso_min_model, linewidth=1) +
  geom_point(aes(x=estimate, y=term), data = meso_reps, color="red") +
  facet_wrap(~term, scales = "free") +
  theme(axis.text.y=element_blank())
```

Clearly, we can't just hand over a model if we want a level of accuracy beyond this. However, labeling things manually one-at-a-time sucks. Can we group/aggregate peaks to label them in clusters?

Yes, using PCA pixelpicking things.

```{r visualize PCA pixelpicking}
library(RaMS)
library(xcms)
dataset_version <- "FT2040"
output_folder <- paste0("made_data_", dataset_version, "/")
msnexp_filled <- readRDS(paste0(output_folder, "msnexp_filled.rds"))
feature_centers <- featureDefinitions(msnexp_filled) %>%
  as.data.frame() %>%
  select(mzmed, rtmed) %>%
  rownames_to_column("feat_id") %>%
  mutate(rtmed=rtmed/60) %>%
  filter(mzmed%between%c(117, 120))
msdata <- readRDS(paste0(output_folder, "msdata.rds"))
pixeldemo_msdata <- msdata$EIC2[mz%between%c(117, 120)]
interp_dt <- mapply(function(mzmed_i, rtmed_i, feat_id_i){
  interp_range <- rtmed_i+c(-0.5, 0.5)
  interp_points <- seq(interp_range[1], interp_range[2], length.out=50)
  pixeldemo_msdata[mz%between%pmppm(mzmed_i)] %>%
    split(.$filename) %>%
    lapply(function(eic_file){
      if(nrow(eic_file)>2){
        setNames(approx(eic_file$rt, eic_file$int, xout=interp_points), c("rt", "int"))
      } else {
        data.frame(rt=numeric(), int=numeric())
      }
    }) %>%
    bind_rows(.id="filename") %>%
    mutate(feature=feat_id_i)
  }, feature_centers$mzmed, feature_centers$rtmed, feature_centers$feat_id, 
  SIMPLIFY = FALSE) %>%
  bind_rows()
interp_dt %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, group=filename)) +
  facet_wrap(~feature, scales="free") +
  theme(axis.text.y = element_blank(), axis.ticks.y=element_blank())

interp_scaled <- interp_dt %>%
  group_by(feature, filename) %>%
  mutate(int=int/max(int)) %>%
  mutate(rt=rank(rt)) %>%
  ungroup() 
  # complete(feature, filename, rt, fill = list(int=0))
interp_scaled %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, color=int, group=filename)) +
  facet_wrap(~feature, scales="free") +
  scale_color_viridis_c()
interp_scaled %>%
  ggplot() +
  geom_tile(aes(x=rt, y=filename, fill=int)) +
  facet_wrap(~feature) +
  scale_fill_viridis_c() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
pcaoutput <- interp_scaled %>%
  ungroup() %>%
  pivot_wider(names_from=feature, values_from = int) %>%
  select(which(colSums(is.na(.))==0)) %>%
  arrange(filename, rt) %>%
  select(-rt, -filename) %>%
  data.matrix() %>%
  prcomp()

library(ggrepel)
pcaoutput$rotation %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  ggplot() +
  geom_text_repel(aes(x=PC1, y=PC2, label=feature), max.overlaps = Inf, 
                  min.segment.length = Inf) +
  theme_bw()

pcapercs <- pcaoutput$sdev^2*100
ggplot() + 
  geom_col(aes(x=seq_along(pcapercs), y=pcapercs)) +
  labs(y="% variance explained", x="PC #") +
  theme_bw() +
  scale_y_continuous(expand = c(0.01, 0.02)) +
  scale_x_continuous(breaks = seq_along(pcapercs), expand = c(0.02, 0))

cbind(distinct(interp_scaled, filename, rt), -pcaoutput$x[,1:4]) %>% 
  pivot_longer(starts_with("PC"), names_to = "PC", values_to = "int") %>%
  ggplot() +
  geom_tile(aes(x=rt, y=filename, fill=int)) +
  facet_wrap(~PC) +
  scale_fill_viridis_c() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        axis.title = element_blank())
```

```{r full PCA pixelpicking}
dataset_version <- "FT2040"
output_folder <- paste0("made_data_", dataset_version, "/")
# msnexp_filled <- readRDS(paste0(output_folder, "msnexp_filled.rds"))
# feature_centers <- featureDefinitions(msnexp_filled) %>%
#   as.data.frame() %>%
#   select(mzmed, rtmed) %>%
#   rownames_to_column("feat_id") %>%
#   mutate(rtmed=rtmed/60)
# msdata <- readRDS(paste0(output_folder, "msdata.rds"))
# interp_dt <- mapply(function(mzmed_i, rtmed_i, feat_id_i){
#   interp_range <- rtmed_i+c(-0.5, 0.5)
#   interp_points <- seq(interp_range[1], interp_range[2], length.out=50)
#   msdata$EIC2[mz%between%pmppm(mzmed_i)] %>%
#     split(.$filename) %>%
#     lapply(function(eic_file){
#       if(nrow(eic_file)>2){
#         setNames(approx(eic_file$rt, eic_file$int, xout=interp_points), c("rt", "int"))
#       } else {
#         data.frame(rt=numeric(), int=numeric())
#       }
#     }) %>%
#     bind_rows(.id="filename") %>%
#     mutate(feature=feat_id_i)
#   }, feature_centers$mzmed, feature_centers$rtmed, feature_centers$feat_id, 
#   SIMPLIFY = FALSE) %>%
#   bind_rows()
# saveRDS(interp_dt, file = paste0(output_folder, "interp_dt.rds"))
interp_dt <- readRDS(file = paste0(output_folder, "interp_dt.rds"))
interp_scaled <- interp_dt %>%
  group_by(feature, filename) %>%
  mutate(int=int/max(int)) %>%
  mutate(rt=rank(rt)) %>%
  ungroup()
pcaoutput <- interp_scaled %>%
  ungroup() %>%
  pivot_wider(names_from=feature, values_from = int) %>%
  select(which(colSums(is.na(.))==0)) %>%
  arrange(filename, rt) %>%
  select(-rt, -filename) %>%
  data.matrix() %>%
  prcomp()
pcaoutput$rotation %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  ggplot() +
  geom_text(aes(x=PC1, y=PC2, label=feature)) +
  theme_bw() +
  coord_fixed()

pc_features <- FT2040_features %>%
  select(feature=feat_id, feat_class) %>%
  filter(feature%in%rownames(pcaoutput$rotation)) %>%
  cbind(pcaoutput$rotation[,c("PC1", "PC2")])

ggplot(pc_features) +
  geom_text(aes(x=PC1, y=PC2, color=feat_class, label=feature)) +
  theme_bw() +
  theme(legend.position = "none") +
  coord_fixed()

pc_circled <- pc_features %>%
  mutate(quick_class=ifelse(sqrt(PC1^2+PC2^2)<0.02, "Bad", "Unclassified")) %>%
  mutate(quick_class=ifelse(sqrt((PC1-0.08)^2+(PC2-0.05)^2)<0.03, 
                           "Good", quick_class))
pc_circled %>%
  mutate(plot_color=ifelse(quick_class=="Unclassified", NA, feat_class)) %>%
  ggplot() +
  geom_point(aes(x=PC1, y=PC2, color=plot_color)) +
  ggforce::geom_circle(aes(x0=0, y0=0, r=0.02), n = 36, linewidth=1,
                       color=scales::hue_pal()(4)[1]) +
  ggforce::geom_circle(aes(x0=0.08, y0=0.05, r=0.03), n = 36, linewidth=1,
                       color=scales::hue_pal()(4)[2]) +
  theme_bw() +
  theme(legend.position = "none") +
  coord_fixed()
```

Prove this with confusion matrices showing very few false pos/neg:

```{r PCA pixelpicking confusion matrix}
table(pc_circled$feat_class, pc_circled$quick_class)
```

How does a model trained on just the best/worst subset perform compared to the full model? Model explosion when complete separation happens? Need to train up a central subset? Show confusion matrices and model parameter estimates.

```{r}
quick_classes <- pc_circled %>%
  select(feature, quick_class) %>%
  filter(quick_class!="Unclassified") %>%
  mutate(quick_class=quick_class=="Good")
summary(quick_classes)

full_quick_model <- FT2040_features %>% 
  filter(feat_id%in%quick_classes$feature) %>%
  select(-feat_class) %>%
  left_join(quick_classes, by=c(feat_id="feature")) %>%
  select(-feat_id) %>%
  glm(formula=quick_class~., family = binomial)

mini_quick_model <- FT2040_features %>% 
  filter(feat_id%in%quick_classes$feature) %>%
  select(-feat_class) %>%
  left_join(quick_classes, by=c(feat_id="feature")) %>%
  select(quick_class, med_cor, med_SNR) %>%
  glm(formula=quick_class~., family = binomial)

FT2040_features %>%
  mutate(pred_prob=predict(object=mini_quick_model, newdata=., type="response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
  with(table(feat_class, pred_class))
```

### 2023-03-18 feature_extraction.R
Refactoring almost all feature extraction code

@@ -31,9 +31,9 @@ trapz <- function(x, y) {
   return(0.5*(p1-p2))
 }
 
-# dataset_version <- "FT2040"
+dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
-dataset_version <- "CultureData"
+# dataset_version <- "CultureData"
 # dataset_version <- "Pttime"
 output_folder <- paste0("made_data_", dataset_version, "/")
 
@@ -54,8 +54,10 @@ peak_data <- msnexp_filled %>%
   rownames_to_column("id") %>%
   unnest_longer(peakidx) %>%
   rename_with(~paste0("feat_", .x), .cols = -peakidx) %>%
+  rename(feature="feat_id") %>%
   left_join(peak_data_long) %>%
-  mutate(filename=file_data$filename[sample])
+  mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
+  complete(feature, filename)
 
 
 
@@ -64,17 +66,23 @@ n_files <- nrow(file_data)
 n_samps <- sum(file_data$samp_type=="Smp")
 n_stans <- sum(file_data$samp_type=="Std")
 simple_feats <- peak_data %>%
-  group_by(feat_id) %>%
-  mutate(sn=log10(sn)) %>%
-  summarise(mean_mz=unique(feat_mzmed), sd_ppm=sd(mz/feat_mzmed),
-            mean_rt=unique(feat_rtmed), sd_rt=sd(rt),
-            mean_pw=mean(rtmax-rtmin), sd_pw=sd(rtmax-rtmin),
-            log_mean_height=log10(mean(maxo)), log_sd_height=log10(sd(maxo)),
-            across(c(sn, f, scale, lmin), function(x)mean(x, na.rm=TRUE)),
-            feat_npeaks=unique(feat_npeaks)/n_files,
-            n_found=(sum(is.na(intb))-n_files)/n_files,
-            samps_found=1-sum(is.na(intb) & str_detect(filename, "Smp"))/n_samps,
-            stans_found=1-sum(is.na(intb) & str_detect(filename, "Std"))/n_stans
+  left_join(file_data) %>%
+  group_by(feature) %>%
+  fill(starts_with("feat"), .direction = "downup") %>%
+  summarise(mean_mz=unique(feat_mzmed), 
+            sd_ppm=sd(mz/feat_mzmed*1e6, na.rm=TRUE),
+            mean_rt=unique(feat_rtmed), 
+            sd_rt=sd(rt, na.rm=TRUE),
[…]

### 2023-03-18 peakpicking_and_prep.R
No longer correcting CultureData RTs
It seemed to only make things worse and the RTs are already pretty darn consistent across all the peaks/files I looked at

@@ -14,9 +14,9 @@ library(xcms)
 library(RaMS)
 options(pillar.sigfig=7)
 
-dataset_version <- "FT2040"
+# dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
-# dataset_version <- "CultureData"
+dataset_version <- "CultureData"
 # dataset_version <- "Pttime"
 
 output_folder <- paste0("made_data_", dataset_version, "/")
@@ -97,13 +97,10 @@ if(dataset_version%in%c("FT2040", "MS3000", "Pttime")){
     response = 1, 
     distFun = "cor_opt"
   )
+  msnexp_rtcor <- adjustRtime(msnexp_withpeaks, obp)
+} else {
+  msnexp_rtcor <- msnexp_withpeaks
 }
-msnexp_rtcor <- adjustRtime(msnexp_withpeaks, obp)
-
-if(dataset_version=="CultureData"){
-  msnexp_rtcor <- dropAdjustedRtime(msnexp_rtcor)
-}
-
 
 if(dataset_version%in%c("FT2040", "MS3000", "Pttime")){
   pdp <- PeakDensityParam(

### 2023-03-19 feature_extraction.R
Finalized feature extraction
Updated to handle Pttime, dropping first 30 seconds and 20+ minutes

@@ -31,10 +31,10 @@ trapz <- function(x, y) {
   return(0.5*(p1-p2))
 }
 
-dataset_version <- "FT2040"
+# dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
 # dataset_version <- "CultureData"
-# dataset_version <- "Pttime"
+dataset_version <- "Pttime"
 output_folder <- paste0("made_data_", dataset_version, "/")
 
 file_data <- read_csv(paste0(output_folder, "file_data.csv")) %>%
@@ -54,7 +54,7 @@ peak_data <- msnexp_filled %>%
   rownames_to_column("id") %>%
   unnest_longer(peakidx) %>%
   rename_with(~paste0("feat_", .x), .cols = -peakidx) %>%
-  rename(feature="feat_id") %>%
+  dplyr::rename(feature="feat_id") %>%
   left_join(peak_data_long) %>%
   mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
   complete(feature, filename)
@@ -150,6 +150,9 @@ peakshape_mets <- peakshape_rawdata %>%
             max_SNR=max(SNR, na.rm = TRUE),
             max_cor=max(peak_cor, na.rm = TRUE),
             medtop3_cor=second_largest(peak_cor)) %>%
+  right_join(distinct(peak_data, feature)) %>%
+  mutate(across(-feature, ~ifelse(is.na(.x), 0, .x))) %>%
+  mutate(across(-feature, ~ifelse(is.infinite(.x), 0, .x))) %>%
   mutate(log_med_cor=log10(1-med_cor))
 write.csv(peakshape_mets, paste0(output_folder, "peakshape_mets.csv"), 
           row.names = FALSE)
@@ -162,7 +165,9 @@ scan_time <- msdata$EIC2 %>%
   summarise(mean_diff=mean(diff(rt))) %>%
   summarise(grand_mean_diff=mean(mean_diff)) %>%
   pull(grand_mean_diff)
-n_scans <- eic_dt[, .N, .(filename, feature)]
+n_scans <- eic_dt[, .N, .(filename, feature)] %>%
+  right_join(distinct(peak_data, feature, filename)) %>%
+  mutate(N=ifelse(is.na(N), 0, N))
 med_missed_scans <- peak_bounds %>%
   mutate(filename=basename(filename)) %>%
[…]

### 2023-03-19 model_validation.R
Updated to use new features_extracted
Changed feat_id to feature universally

@@ -2,16 +2,13 @@
 # Setup ----
 library(tidyverse)
 
-FT2040_features <- read_csv("made_data_FT2040/features_extracted.csv") %>%
-  mutate(sn=ifelse(is.infinite(sn), 0, sn)) %>%
-  select(-blank_found)
-MS3000_features <- read_csv("made_data_MS3000/features_extracted.csv") %>%
-  mutate(sn=ifelse(is.infinite(sn), 0, sn)) %>%
-  select(-blank_found)
+FT2040_features <- read_csv("made_data_FT2040/features_extracted.csv")
+MS3000_features <- read_csv("made_data_MS3000/features_extracted.csv")
+
 # Visualize a single metric across both datasets
 bind_rows(list(Falkor=FT2040_features, MESOSCOPE=MS3000_features), .id = "cruise") %>%
   ggplot() + 
-  geom_histogram(aes(x=sd_rt, fill=feat_class), bins=40) +
+  geom_histogram(aes(x=log_med_cor, fill=feat_class), bins=40) +
   facet_wrap(~cruise, ncol = 1, scales = "free_y")
 library(ggh4x)
 bind_rows(list(Falkor=FT2040_features, MESOSCOPE=MS3000_features), .id = "cruise") %>%
@@ -21,12 +18,12 @@ bind_rows(list(Falkor=FT2040_features, MESOSCOPE=MS3000_features), .id = "cruise
 
 # Testing Falkor full model fit on MESOSCOPE data ----
 falkor_full_model <- FT2040_features %>% 
-  select(-feat_id) %>%
+  select(-feature) %>%
   filter(feat_class%in%c("Good", "Bad")) %>%
   mutate(feat_class=feat_class=="Good") %>%
   glm(formula=feat_class~., family = binomial)
 FT2040_preds <- FT2040_features %>%
-  select(-feat_id) %>%
+  select(-feature) %>%
   filter(feat_class%in%c("Good", "Bad")) %>%
   mutate(pred_prob=predict(object=falkor_full_model,newdata = ., type = "response")) %>%
   mutate(pred_class=pred_prob>0.5) %>%
@@ -49,24 +46,24 @@ MS3000_preds %>%
 
 # Testing Falkor minimal model on MESOSCOPE ----
 falkor_min_model <- FT2040_features %>% 
-  select(feat_class, med_cor, med_SNR, sd_rt) %>%
+  select(feat_class, med_cor, med_SNR) %>%
[…]

### 2023-03-19 regression_attempts.R

@@ -7,8 +7,7 @@ dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
 output_folder <- paste0("made_data_", dataset_version, "/")
 
-features_extracted <- read_csv(paste0(output_folder, "features_extracted.csv")) %>%
-  mutate(sn=ifelse(is.infinite(sn), 0, sn))
+features_extracted <- read_csv(paste0(output_folder, "features_extracted.csv"))
 
 set.seed(123)
 traintestlist <- features_extracted %>%
@@ -32,7 +31,7 @@ traintestlist$test %>%
 
 # All-feature regression ----
 full_model <- traintestlist$train %>% 
-  select(-feat_id, -blank_found, -shape_cor, -area_cor) %>%
+  select(-feature, -blank_found, -shape_cor, -area_cor) %>%
   glm(formula=feat_class~., family = binomial)
 
 summary(full_model)
@@ -45,7 +44,7 @@ traintestlist$test %>%
 
 
 # Stepwise AIC optimization using MASS ----
-all_model_feats <- select(traintestlist$train, -feat_id, -blank_found)
+all_model_feats <- select(traintestlist$train, -feature, -blank_found)
 full_model <- glm(formula=feat_class~., family = binomial, data = all_model_feats)
 step_model <- MASS::stepAIC(full_model, direction = "both")
 
@@ -132,7 +131,7 @@ traintestlist$test %>%
 
 # Visualization (Full model, excludes ggplot fit) ----
 full_model <- traintestlist$train %>% 
-  select(-feat_id, -blank_found) %>%
+  select(-feature, -blank_found) %>%
   glm(formula=feat_class~log_peak_height, family = binomial)
 init_gp <- traintestlist$test %>%
   cbind(pred_prob=predict(full_model, ., type="response")) %>%

### 2023-04-06 feature_extraction.R
Added new feature classes for Pttime and CultureData

@@ -33,8 +33,8 @@ trapz <- function(x, y) {
 
 # dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
-# dataset_version <- "CultureData"
-dataset_version <- "Pttime"
+dataset_version <- "CultureData"
+# dataset_version <- "Pttime"
 output_folder <- paste0("made_data_", dataset_version, "/")
 
 file_data <- read_csv(paste0(output_folder, "file_data.csv")) %>%
@@ -343,7 +343,11 @@ if(file.exists(paste0(output_folder, "classified_feats.csv"))){
   classified_feats <- read_csv(paste0(output_folder, "classified_feats.csv")) %>%
     select(feature, feat_class)
 } else {
-  classified_feats <- data.frame(feature=simple_feats$feature, feat_class="Unclassified")
+  classified_feats <- read_csv(paste0(output_folder, "classified_feats.csv")) %>%
+    select(feature, feat_class) %>%
+    right_join(data.frame(feature=simple_feats$feature)) %>%
+    mutate(feat_class=ifelse(is.na(feat_class), "Unclassified", feat_class)) %>%
+    arrange(feature)
 }
 
 features_extracted <- simple_feats %>%

### 2023-04-06 regression_attempts.R

@@ -19,7 +19,7 @@ traintestlist <- features_extracted %>%
   setNames(c("train", "test"))
 
 # Single-(ML)-feature regression ----
-single_model <- glm(feat_class~log_mean_height, family = binomial, data=traintestlist$train)
+single_model <- glm(feat_class~log_mean_area, family = binomial, data=traintestlist$train)
 summary(single_model)
 traintestlist$test %>%
   cbind(pred_class=predict(single_model, ., type="response")>0.5) %>%
@@ -31,7 +31,7 @@ traintestlist$test %>%
 
 # All-feature regression ----
 full_model <- traintestlist$train %>% 
-  select(-feature, -blank_found, -shape_cor, -area_cor) %>%
+  select(-feature) %>%
   glm(formula=feat_class~., family = binomial)
 
 summary(full_model)
@@ -44,7 +44,7 @@ traintestlist$test %>%
 
 
 # Stepwise AIC optimization using MASS ----
-all_model_feats <- select(traintestlist$train, -feature, -blank_found)
+all_model_feats <- select(traintestlist$train, -feature)
 full_model <- glm(formula=feat_class~., family = binomial, data = all_model_feats)
 step_model <- MASS::stepAIC(full_model, direction = "both")
 
@@ -64,7 +64,7 @@ traintestlist$test %>%
 
 
 # Visualization (Single-metric, includes ggplot fit) ----
-pred_feat <- "mean_mz"
+pred_feat <- "med_cor"
 init_gp <- traintestlist$train %>%
   mutate(pred_cut=cut(get(pred_feat), pretty(get(pred_feat), n=15), include.lowest=TRUE)) %>%
   mutate(feat_class=factor(feat_class, labels = c("Bad", "Good"), levels=c(FALSE, TRUE))) %>%
@@ -115,6 +115,7 @@ raw_curve_data <- traintestlist$train %>%
   mutate(med_SNR=med_SNR*length(SNR_range)) %>%
   mutate(med_cor=((med_cor-min(med_cor))/(max(med_cor)-min(med_cor)))) %>%
   mutate(med_cor=med_cor*length(cor_range))
+library(plotly)
 plot_ly() %>%
   add_trace(x=~med_cor, y=~med_SNR, z=~feat_class, opacity=0.5,
             mode="markers", type="scatter3d", 
@@ -131,8 +132,8 @@ traintestlist$test %>%
 
 # Visualization (Full model, excludes ggplot fit) ----
[…]

### 2023-04-06 model_validation.R
Beginning validation with all 4 datasets

@@ -4,17 +4,24 @@ library(tidyverse)
 
 FT2040_features <- read_csv("made_data_FT2040/features_extracted.csv")
 MS3000_features <- read_csv("made_data_MS3000/features_extracted.csv")
+Pttime_features <- read_csv("made_data_Pttime/features_extracted.csv")
+CultureData_features <- read_csv("made_data_CultureData/features_extracted.csv")
 
-# Visualize a single metric across both datasets
-bind_rows(list(Falkor=FT2040_features, MESOSCOPE=MS3000_features), .id = "cruise") %>%
+# Visualize a single metric across all datasets
+bind_rows(list(Falkor=FT2040_features, MESOSCOPE=MS3000_features,
+               CultureData=CultureData_features, Pttime=Pttime_features), 
+          .id = "cruise") %>%
   ggplot() + 
-  geom_histogram(aes(x=log_med_cor, fill=feat_class), bins=40) +
+  geom_histogram(aes(x=med_cor, fill=feat_class), bins=40) +
   facet_wrap(~cruise, ncol = 1, scales = "free_y")
+
 library(ggh4x)
 bind_rows(list(Falkor=FT2040_features, MESOSCOPE=MS3000_features), .id = "cruise") %>%
   ggplot() + 
   geom_histogram(aes(x=med_SNR, fill=feat_class), bins=40) +
   facet_grid2(cruise~feat_class, scales = "free_y", independent = "y")
+
+
 
 # Testing Falkor full model fit on MESOSCOPE data ----
 falkor_full_model <- FT2040_features %>% 
@@ -151,3 +158,21 @@ list(falkor=falkor_min_model, meso=meso_min_model, both=FM_min_model) %>%
                      y=model, color=model)) +
   facet_wrap(~term, ncol=1, scales = "free")
 
+
+
+# Using combined model to visualize predictions ----
+FT2040_features <- read_csv("made_data_FT2040/features_extracted.csv")
+MS3000_features <- read_csv("made_data_MS3000/features_extracted.csv")
+both_min_model <- rbind(FT2040_features, MS3000_features) %>%
+  select(feat_class, med_cor, med_SNR) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=feat_class=="Good") %>%
+  glm(formula=feat_class~., family = binomial)
+Pttime_features %>%
[…]

### 2023-04-06 peakpicking_and_prep.R
Re-coloring Pttime and CultureData for labeling

@@ -16,8 +16,8 @@ options(pillar.sigfig=7)
 
 # dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
-dataset_version <- "CultureData"
-# dataset_version <- "Pttime"
+# dataset_version <- "CultureData"
+dataset_version <- "Pttime"
 
 output_folder <- paste0("made_data_", dataset_version, "/")
 mzML_files <- list.files(paste0(output_folder, "mzMLs/"), full.names=TRUE)
@@ -54,7 +54,8 @@ if(dataset_version%in%c("FT2040", "MS3000")){
       organism%in%c("116-1", "2090", "371", "Ca", "Cr", "Cs", "P3", "P5-5") ~ "Haptos",
       TRUE~"UNKNOWN" 
     )) %>%
-    mutate(samp_group=str_remove(basename(filename), "_?(St[1-4]_pos|Exp[1-3]_pos|MediaBlk.*|[A-I]{1,2}_pos|\\d)\\.mzML$"))
+    mutate(samp_group=str_remove(basename(filename), "_?(St[1-4]_pos|Exp[1-3]_pos|MediaBlk.*|[A-I]{1,2}_pos|\\d)\\.mzML$")) %>%
+    mutate(col=c("red", "blue", "black", "#008080")[factor(samp_type, levels=c("Blk", "Smp", "Std", "Poo"))], lwd=1)
 } else if(dataset_version=="Pttime"){
   file_data <- data.frame(filename=mzML_files) %>%
     mutate(samp_id=str_remove(basename(filename), "20190429_JJ_VB_BioSFA_Pttime_1_QE144_Ag68377-924_USHXG01160_POS_MSMS-v2_")) %>%
@@ -62,7 +63,8 @@ if(dataset_version%in%c("FT2040", "MS3000")){
     mutate(timepoint=str_extract(samp_id, "\\d+d")) %>%
     mutate(timepoint=ifelse(is.na(timepoint), "ctrl", timepoint)) %>%
     mutate(samp_group=paste(samp_type, timepoint, sep = "-")) %>%
-    arrange(samp_group)
+    arrange(samp_group) %>%
+    mutate(col=scales::hue_pal()(3)[factor(samp_type, levels=c("extr", "exudate", "pellet"))], lwd=1)
 } else {
   stop("Dataset not recognized!")
 }

### 2023-04-06 pixelpicking_attempts.R
Mmmmm not sure, looks like mostly cleanup

@@ -4,9 +4,9 @@ library(tidyverse)
 library(RaMS)
 library(xcms)
 library(data.table)
-dataset_version <- "FT2040"
+# dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
-# dataset_version <- "CultureData"
+dataset_version <- "CultureData"
 # dataset_version <- "Pttime"
 output_folder <- paste0("made_data_", dataset_version, "/")
 msnexp_filled <- readRDS(paste0(output_folder, "msnexp_filled.rds"))
@@ -16,6 +16,7 @@ feature_centers <- featureDefinitions(msnexp_filled) %>%
   rownames_to_column("feat_id") %>%
   mutate(rtmed=rtmed/60)
 msdata <- readRDS(paste0(output_folder, "msdata.rds"))
+# msdata <- grabMSdata(list.files(paste0(output_folder, "mzMLs"), pattern = "Poo", full.names = TRUE))
 
 
 
@@ -44,11 +45,10 @@ interp_dt <- pbapply::pbmapply(function(mzmed_i, rtmed_i, feat_id_i){
 
 
 # Perform the PCA and check variance explained ----
-interp_complete <- interp_dt[, .(rt=rank(rt), int), by=c("feature", "filename")] %>%
-  complete(feature, filename, rt, fill=list(int=-1))
+interp_complete <- interp_dt[, .(rt=rank(rt), int), by=c("feature", "filename")]
 interp_mat <- interp_complete %>%
   pivot_wider(names_from=feature, values_from = int) %>%
-  # select(which(colSums(is.na(.))==0)) %>%
+  select(which(colSums(is.na(.))==0)) %>%
   arrange(filename, rt) %>%
   select(-rt, -filename) %>%
   data.matrix()
@@ -71,16 +71,16 @@ pcaoutput$x[,"PC1"] %>%
   ggplot() +
   geom_tile(aes(x=rownum, y=rt, fill=int))
 
-library(plotly)
-pcaoutput$rotation %>%
-  as.data.frame() %>%
-  rownames_to_column("feature") %>%
-  plot_ly(x=~PC1, y=~PC2, z=~PC3, mode="markers", type="scatter3d", opacity=0.5)
+# library(plotly)
+# pcaoutput$rotation %>%
+#   as.data.frame() %>%
+#   rownames_to_column("feature") %>%
[…]

### 2023-04-06 regression_attempts.R
Added CultureData to versioning

@@ -5,6 +5,7 @@ library(tidyverse)
 # dataset_version <- "FT350"
 dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
+dataset_version <- "CultureData"
 output_folder <- paste0("made_data_", dataset_version, "/")
 
 features_extracted <- read_csv(paste0(output_folder, "features_extracted.csv"))

### 2023-04-06 training_tool.R
Added feature extraction for Pttime and CultureData and labeled things

@@ -12,8 +12,10 @@ library(tidyverse)
 library(RaMS)
 options(pillar.sigfig=7)
 
-dataset_version <- "FT2040"
+# dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
+dataset_version <- "CultureData"
+# dataset_version <- "Pttime"
 output_folder <- paste0("made_data_", dataset_version, "/")
 
 file_data <- read_csv(paste0(output_folder, "file_data.csv")) %>%
@@ -24,13 +26,43 @@ rt_corrections <- read_csv(paste0(output_folder, "rt_corrections.csv"))
 
 
 # Extract feature data from raw mzML files ----
-feature_df <- peak_bounds %>%
-  group_by(feature) %>%
-  summarize(min_mz=min(mzmin), max_mz=max(mzmax), 
-            min_rt=min(rtmin), max_rt=max(rtmax)) %>%
-  mutate(mean_mz=(min_mz+max_mz)/2)
-msdata <- readRDS(paste0(output_folder, "msdata.rds"))
+if(dataset_version%in%c("FT2040", "MS3000")){
+  feature_df <- peak_bounds %>%
+    group_by(feature) %>%
+    summarize(min_mz=min(mzmin), max_mz=max(mzmax), 
+              min_rt=min(rtmin), max_rt=max(rtmax)) %>%
+    mutate(mean_mz=(min_mz+max_mz)/2)
+} else {
+  FT2040_features <- read_csv("made_data_FT2040/features_extracted.csv")
+  MS3000_features <- read_csv("made_data_MS3000/features_extracted.csv")
+  both_min_model <- rbind(FT2040_features, MS3000_features) %>%
+    select(feat_class, med_cor, med_SNR) %>%
+    filter(feat_class%in%c("Good", "Bad")) %>%
+    mutate(feat_class=feat_class=="Good") %>%
+    glm(formula=feat_class~., family = binomial)
+  good_features <- read_csv(paste0(output_folder, "features_extracted.csv")) %>%
+    mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
+    filter(pred_prob>0.9)
+  feature_df <- read_csv(paste0(output_folder, "peak_bounds.csv")) %>%
+    group_by(feature) %>%
+    summarize(min_mz=min(mzmin), max_mz=max(mzmax), 
+              min_rt=min(rtmin), max_rt=max(rtmax)) %>%
[…]

### 2023-04-06 result_sandbox.Rmd
Switching from feat_id to feature and height to area during cleanup

@@
   full_model <- get(train_set) %>% 
-    select(-feat_id) %>%
+    select(-feature) %>%
     filter(feat_class%in%c("Good", "Bad")) %>%
@@
     get(test_set) %>%
-      select(-feat_id) %>%
+      select(-feature) %>%
       filter(feat_class%in%c("Good", "Bad")) %>%
@@
 meso_full_model <- MS3000_features %>% 
-  select(-feat_id) %>%
+  select(-feature) %>%
   filter(feat_class%in%c("Good", "Bad")) %>%
@@
 falkor_full_model <- FT2040_features %>% 
-  select(-feat_id) %>%
+  select(-feature) %>%
   filter(feat_class%in%c("Good", "Bad")) %>%
@@
 xcms_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", 
-                 "sd_pw", "log_mean_height", "log_sd_height", "sn", 
+                 "sd_pw", "log_mean_area", "log_sd_area", "sn", 
                  "f", "scale", "lmin", "feat_npeaks", "n_found", 
@@
 all_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", "sd_pw",
-                "log_mean_height", "log_sd_height", "sn", "f", "scale", 
+                "log_mean_area", "log_sd_area", "sn", "f", "scale", 
                 "lmin", "feat_npeaks", "n_found", "samps_found", "stans_found", 
@@
     slice_sample(prop = 0.5) %>%
-    select(-feat_id) %>%
+    select(-feature) %>%
     filter(feat_class%in%c("Good", "Bad")) %>%
@@
   select(mzmed, rtmed) %>%
-  rownames_to_column("feat_id") %>%
+  rownames_to_column("feature") %>%
   mutate(rtmed=rtmed/60) %>%
@@
 pixeldemo_msdata <- msdata$EIC2[mz%between%c(117, 120)]
-interp_dt <- mapply(function(mzmed_i, rtmed_i, feat_id_i){
+interp_dt <- mapply(function(mzmed_i, rtmed_i, feature_i){
   interp_range <- rtmed_i+c(-0.5, 0.5)
@@
     bind_rows(.id="filename") %>%
-    mutate(feature=feat_id_i)
-  }, feature_centers$mzmed, feature_centers$rtmed, feature_centers$feat_id, 
+    mutate(feature=feature_i)
+  }, feature_centers$mzmed, feature_centers$rtmed, feature_centers$feature, 
   SIMPLIFY = FALSE) %>%
@@
 #   select(mzmed, rtmed) %>%
-#   rownames_to_column("feat_id") %>%
+#   rownames_to_column("feature") %>%
 #   mutate(rtmed=rtmed/60)
 # msdata <- readRDS(paste0(output_folder, "msdata.rds"))
-# interp_dt <- mapply(function(mzmed_i, rtmed_i, feat_id_i){
+# interp_dt <- mapply(function(mzmed_i, rtmed_i, feature_i){
 #   interp_range <- rtmed_i+c(-0.5, 0.5)
@@
 #     bind_rows(.id="filename") %>%
-#     mutate(feature=feat_id_i)
-#   }, feature_centers$mzmed, feature_centers$rtmed, feature_centers$feat_id, 
+#     mutate(feature=feature_i)
+#   }, feature_centers$mzmed, feature_centers$rtmed, feature_centers$feature, 
 #   SIMPLIFY = FALSE) %>%
@@
 
+# Currently broken?
 pc_features <- FT2040_features %>%
-  select(feature=feat_id, feat_class) %>%
+  select(feature, feat_class) %>%
   filter(feature%in%rownames(pcaoutput$rotation)) %>%
@@
 full_quick_model <- FT2040_features %>% 
-  filter(feat_id%in%quick_classes$feature) %>%
+  filter(feature%in%quick_classes$feature) %>%
   select(-feat_class) %>%
-  left_join(quick_classes, by=c(feat_id="feature")) %>%
-  select(-feat_id) %>%
+  left_join(quick_classes, by="feature") %>%
+  select(-feature) %>%
   glm(formula=quick_class~., family = binomial)
@@
 mini_quick_model <- FT2040_features %>% 
-  filter(feat_id%in%quick_classes$feature) %>%
+  filter(feature%in%quick_classes$feature) %>%
   select(-feat_class) %>%
-  left_join(quick_classes, by=c(feat_id="feature")) %>%
+  left_join(quick_classes, by="feature") %>%
   select(quick_class, med_cor, med_SNR) %>%

### 2023-04-06 feature_extraction.R
Added handling for partially classified datasets

@@ -33,8 +33,8 @@ trapz <- function(x, y) {
 
 # dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
-dataset_version <- "CultureData"
-# dataset_version <- "Pttime"
+# dataset_version <- "CultureData"
+dataset_version <- "Pttime"
 output_folder <- paste0("made_data_", dataset_version, "/")
 
 file_data <- read_csv(paste0(output_folder, "file_data.csv")) %>%
@@ -339,16 +339,11 @@ depth_diffs <- read_csv(paste0(output_folder, "depth_diffs.csv"))
 blank_diffs <- read_csv(paste0(output_folder, "blank_diffs.csv"))
 stan_diffs <- read_csv(paste0(output_folder, "stan_diffs.csv"))
 feat_isodata <- read_csv(paste0(output_folder, "feat_isodata.csv"))
-if(file.exists(paste0(output_folder, "classified_feats.csv"))){
-  classified_feats <- read_csv(paste0(output_folder, "classified_feats.csv")) %>%
-    select(feature, feat_class)
-} else {
-  classified_feats <- read_csv(paste0(output_folder, "classified_feats.csv")) %>%
-    select(feature, feat_class) %>%
-    right_join(data.frame(feature=simple_feats$feature)) %>%
-    mutate(feat_class=ifelse(is.na(feat_class), "Unclassified", feat_class)) %>%
-    arrange(feature)
-}
+classified_feats <- read_csv(paste0(output_folder, "classified_feats.csv")) %>%
+  select(feature, feat_class) %>%
+  right_join(data.frame(feature=simple_feats$feature)) %>%
+  mutate(feat_class=ifelse(is.na(feat_class), "Unclassified", feat_class)) %>%
+  arrange(feature)
 
 features_extracted <- simple_feats %>%
   left_join(peakshape_mets) %>% 

### 2023-04-13 model_validation.R
Adding combined model comparisons (old)

@@ -13,7 +13,9 @@ bind_rows(list(Falkor=FT2040_features, MESOSCOPE=MS3000_features,
           .id = "cruise") %>%
   ggplot() + 
   geom_histogram(aes(x=med_cor, fill=feat_class), bins=40) +
-  facet_wrap(~cruise, ncol = 1, scales = "free_y")
+  facet_wrap(~cruise, ncol = 1, scales = "free_y") +
+  scale_fill_manual(breaks = c("Good", "Bad", "Meh", "Stans only", "Unclassified"),
+                    values = c("#f4bb23", "#a41118", "#0b505c", "#028e34", "grey50"))
 
 library(ggh4x)
 bind_rows(list(Falkor=FT2040_features, MESOSCOPE=MS3000_features), .id = "cruise") %>%
@@ -160,7 +162,7 @@ list(falkor=falkor_min_model, meso=meso_min_model, both=FM_min_model) %>%
 
 
 
-# Using combined model to visualize predictions ----
+# Using combined model to visualize predictions in all 4 datasets ----
 FT2040_features <- read_csv("made_data_FT2040/features_extracted.csv")
 MS3000_features <- read_csv("made_data_MS3000/features_extracted.csv")
 both_min_model <- rbind(FT2040_features, MS3000_features) %>%
@@ -168,11 +170,68 @@ both_min_model <- rbind(FT2040_features, MS3000_features) %>%
   filter(feat_class%in%c("Good", "Bad")) %>%
   mutate(feat_class=feat_class=="Good") %>%
   glm(formula=feat_class~., family = binomial)
-Pttime_features %>%
+bind_rows(list(Falkor=FT2040_features, MESOSCOPE=MS3000_features,
+               CultureData=CultureData_features, Pttime=Pttime_features), 
+          .id = "cruise")
+
+pred_probs_all <- bind_rows(lst(Falkor=FT2040_features, MESOSCOPE=MS3000_features,
+                                CultureData=CultureData_features, Pttime=Pttime_features), 
+                            .id = "dataset") %>%
   mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
+  mutate(pred_class=ifelse(pred_prob>0.9, "Good", "Bad")) %>%
[…]

### 2023-04-13 manuscript/outline.Rmd
Creating manuscript outline RMarkdown doc
Contains lots of intro thoughts and figures for the results section

---
title: "Outline"
author: "William Kumler"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = '..')
options(readr.show_col_types = FALSE)
options(tidyverse.quiet = TRUE)
```

```{r load data}
library(tidyverse)

file_data <- read_csv("made_data_MS3000/file_data.csv") %>%
  mutate(depth=factor(depth, levels=c("15m", "DCM", "175m"))) %>%
  mutate(filename=basename(filename))
FT2040_features <- read_csv("made_data_FT2040/features_extracted.csv")
MS3000_features <- read_csv("made_data_MS3000/features_extracted.csv")
Pttime_features <- read_csv("made_data_Pttime/features_extracted.csv")
CultureData_features <- read_csv("made_data_CultureData/features_extracted.csv")
```

## Abstract

## Introduction

LC/MS techniques are neat

  - Fast sample processing
  - Low detection limits
  - Can characterize new molecules

Most of the data produced by them goes unused

  - Targeted workflows look at only a fraction of the data
  - Untargeted algorithms are still fairly new

Peakpicking is hard

  - Smooth transition from "good" into "meh" into "bad" across the limit of detection
  - Fundamental tradeoff between false positives and false negatives
  - Lots of false positive peaks because attempting to minimize false negative
  - Requires manual curation of peak quality which doesn't scale

Algorithms don't provide likelihood of "goodness"

  - Typical output: m/z, rt, area, maybe SN
  - We want/need a probability that this peak is just noise
  - Want to control the false positive/negative rate depending on project goals
  - We don't need new algorithms as much as we need a way to quantify success on the existing ones

Multiple advantages to a robust assessment of peak quality

  - Reduces manual labor in sorting out good peaks from bad
  - Improves statistical power
    - Reduces the number of hypotheses tested for univariate analyses
    - Increases relative power of significant peaks for multivariate analyses
  - Allows optimization of chromatography for untargeted approaches
    - More peaks isn't better if the additional peaks are just noise
  - Allows optimization of peakpicking parameters for untargeted approaches
  - "Good" peaks should be independent of software choices while noise is noisy
  - Enables more consistent quality control across labs
  - Focuses calculation time/energy on best peaks rather than all equally

A single comprehensive quality parameter is more useful than many independent ones

  - Enables sensible threshold selection (50% likely to be a peak)
    - Rather than current arbitrary approaches (SNR > 20, peak height > 10e7)
  - Single metric has multivariate advantage over multiple univariate thresholds
    - Good peak can fall be weak in one or two metrics and still be safe
  - Relative power of different metrics has never been computed/compared
    - More important to have an isotope? Large area? Good shape? Differ between treatments?

Can be used to check improvements in peakpicking/chromatography

  - Traditional method: a few random peaks are checked after making a change to params
  - Total number of peaks used as proxy, not total number of GOOD peaks
  - Did this change improve overall peak quality?
  - Currently impossible to optimize these for untargeted peakpicking

## Methods

### Sample collection

### Sample processing

### LC conditions

### MS conditions

### Peakpicking with XCMS

### Manual inspection and classification
Gold standard: manual inspection by a human expert

  - Decision to view the entire feature rather than a single trace
  - Decision to color by treatments

```{r fig:some good some bad features}

```

### Feature extraction
Feature extraction - many possible metrics, calculated across features

  - Some trivial to calculate
    - Mean m/z, mean RT, number found
  - Some require access to the raw data
    - Peak shape, peak SNR, isotope existence, number of missed scans

```{r how are raw peak shape metrics calculated}

```

## Results

```{r colored histograms on all 4 datasets}
# Train on MESOSCOPE/Falkor combined model
# Raw metrics only
# Color by feature class
both_min_model <- rbind(FT2040_features, MS3000_features) %>%
  select(feat_class, med_cor, med_SNR) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)
bind_rows(lst(
  Falkor=FT2040_features, MESOSCOPE=MS3000_features,
  CultureData=CultureData_features, Pttime=Pttime_features), 
  .id = "dataset") %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.9, "Good", "Bad")) %>%
  mutate(error_type=case_when(
    feat_class=="Good" & pred_class=="Good" ~ "%TP",
    feat_class=="Good" & pred_class=="Bad" ~ "%FN",
    feat_class=="Bad" & pred_class=="Good" ~ "%FP",
    feat_class=="Bad" & pred_class=="Bad" ~ "%TN",
    TRUE ~ "Incomparable"
  )) %>%
  mutate(dataset=factor(dataset, levels=c("MESOSCOPE", "Falkor", "CultureData", "Pttime"))) %>%
  ggplot() +
  geom_histogram(aes(x=pred_prob, fill=feat_class), breaks=seq(0, 1, 0.02)) +
  facet_wrap(~dataset, ncol = 1, scales="free_y") +
  coord_cartesian(ylim = c(0, 200)) +
  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
  scale_fill_manual(breaks = c("Good", "Bad", "Meh", "Stans only", "Unclassified"),
                    values = c("#f4bb23", "#a41118", "#0b505c", "#028e34", "grey50"))
# Add broken y-axis
# Fix axis titles
# Format nicely
```

```{r FDR vs GPF cross-train triangle figure feature selection}
# Dark/light green/blue triangle plot
# Include best-fit lines for easier slope comparison
# but don't extend beyond data
dataset_versions <- c("FT2040_features", "MS3000_features")

raw_data_params <- c("med_cor", "med_SNR")
xcms_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", 
                 "sd_pw", "log_mean_area", "log_sd_area", "sn", 
                 "f", "scale", "lmin", "feat_npeaks", "n_found", 
                 "samps_found", "stans_found")
all_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", "sd_pw",
                "log_mean_area", "log_sd_area", "sn", "f", "scale", 
                "lmin", "feat_npeaks", "n_found", "samps_found", "stans_found", 
                "med_cor", "med_SNR", "med_missed_scans", "smp_to_blk", 
                "smp_to_std", "shape_cor", "area_cor", "feat_class")

lst(all_params, raw_data_params, xcms_params) %>%
  imap(function(param_selection, param_name){
    lapply(dataset_versions, function(train_set){
      full_model <- get(train_set) %>% 
        filter(feat_class%in%c("Good", "Bad")) %>%
        select(feat_class, all_of(param_selection)) %>%
        mutate(feat_class=feat_class=="Good") %>%
        glm(formula=feat_class~., family = binomial)
      lapply(dataset_versions, function(test_set){
        get(test_set) %>%
          mutate(pred_prob=predict(object=full_model,newdata = ., type = "response")) %>%
          mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
          select(feat_class, starts_with("pred_class")) %>%
          mutate(train_test=paste(train_set, test_set, sep = "-")) %>%
          mutate(train_test=str_remove_all(train_test, "_features")) %>%
          mutate(cross_type=ifelse(train_set==test_set, "Same", "Different"))
      }) %>% bind_rows()
    }) %>% bind_rows() %>% mutate(which_params=param_name)
  }) %>% bind_rows() %>%
  mutate(error_type=case_when(
    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
    feat_class=="Bad" & pred_class=="Good" ~ "FP",
    feat_class=="Good" & pred_class=="Bad" ~ "FN",
    feat_class=="Good" & pred_class=="Good" ~ "TP"
  )) %>%
  group_by(train_test, which_params, cross_type) %>%
  count(error_type) %>%
  ungroup() %>%
  mutate(cross_type=factor(cross_type, levels=c("Same", "Different"))) %>%
  pivot_wider(names_from=error_type, values_from = n) %>%
  mutate(FDR=round(FP/(FP+TP)*100)) %>%
  mutate(GPF=round(TP/(FN+TP)*100)) %>%
  pivot_longer(c(FDR, GPF)) %>%
  mutate(name=paste0("%", name)) %>%
  mutate(which_params=factor(which_params, 
                             labels=c("All\nparams", "Default\nonly", "Raw\ndata params"),
                             levels=c("all_params", "xcms_params", "raw_data_params"))) %>%
  mutate(cross_type_num=as.numeric(cross_type)) %>%
  ggplot() +
  geom_point(aes(x=cross_type, y=value, color=train_test, fill=train_test, shape=name)) +
  geom_smooth(aes(x=cross_type_num, y=value, group=name), 
              method = "lm", formula="y~x", color="black", se=FALSE) +
  facet_grid(~which_params) +
  scale_x_discrete(labels = c("Same\ndataset\n", "Different\ndataset\n"),
                   breaks = c("Same", "Different")) +
  scale_color_manual(values = c("#0b505c", "#0d8299", "#184d1f", "#028e34"), 
                     breaks = c("MS3000-MS3000", "MS3000-FT2040", "FT2040-FT2040", "FT2040-MS3000"),
                     aesthetics = c("color", "fill")) +
  scale_shape_manual(breaks = c("%FDR", "%GPF"), values = c(25, 17)) +
  coord_cartesian(ylim = c(0, 100)) +
  theme_bw()
```

```{r red/gold rank dotplot step-down method}
# Make sure to include overplotting somehow?
dataset_versions <- c("FT2040_features", "MS3000_features")
raw_data_params <- c("med_cor", "med_SNR")
all_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", "sd_pw",
                "log_mean_area", "log_sd_area", "sn", "f", "scale", 
                "lmin", "feat_npeaks", "n_found", "samps_found", "stans_found", 
                "med_cor", "med_SNR", "med_missed_scans", "smp_to_blk", 
                "smp_to_std", "shape_cor", "area_cor", "feat_class")
dotplot_stepdown_data <- lst(all_params, raw_data_params) %>%
  imap(function(param_selection, param_name){
    lapply(dataset_versions, function(train_set){
      model_i <- get(train_set) %>% 
        filter(feat_class%in%c("Good", "Bad")) %>%
        select(feat_class, all_of(param_selection)) %>%
        mutate(feat_class=feat_class=="Good") %>%
        glm(formula=feat_class~., family = binomial)
      MS3000_features %>%
        mutate(pred_prob=predict(model_i, ., type="response")) %>%
        select(feature, feat_class, pred_prob) %>%
        mutate(which_params=param_name) %>%
        mutate(training_set=train_set)
    }) %>% bind_rows()
  }) %>% bind_rows()
raw_prob_space_gp <- dotplot_stepdown_data %>%
  pivot_wider(names_from = training_set, values_from = pred_prob) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  ggplot() +
  geom_abline(slope = 1, intercept = 0, color="black", lwd=1) +
  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class)) +
  facet_wrap(~which_params, nrow = 1) +
  scale_fill_manual(values = c("#a41118", "#f4bb23"), breaks = c("Bad", "Good"), 
                    aesthetics = c("color", "fill")) +
  scale_x_continuous(breaks = c(0, 0.5, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(x="Falkor-trained model", y="MESO-trained model")
rank_prob_space_gp <- dotplot_stepdown_data %>%
  group_by(training_set, which_params) %>%
  mutate(pred_prob=rank(pred_prob)) %>%
  pivot_wider(names_from = training_set, values_from = pred_prob) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  ggplot() +
  geom_abline(slope = 1, intercept = 0, color="black", lwd=1) +
  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class)) +
  facet_wrap(~which_params, nrow = 1) +
  scale_fill_manual(values = c("#a41118", "#f4bb23"), breaks = c("Bad", "Good"), 
                    aesthetics = c("color", "fill")) +
  scale_x_continuous(breaks = c(0, nrow(MS3000_features)), 
                     labels = c("Min rank", "Max rank")) +
  scale_y_continuous(breaks = c(0, nrow(MS3000_features)),
                     labels = c("Min\nrank", "Max\nrank")) +
  labs(x="Falkor-trained model", y="MESO-trained model") +
  theme(axis.text.x = element_text(hjust=c(0, 1)))
plot(gridExtra::arrangeGrob(raw_prob_space_gp, rank_prob_space_gp))
```

```{r FDR vs GPF with additional datasets}
# Currently a fairly simple figure, may need additional data?
both_min_model <- rbind(FT2040_features, MS3000_features) %>%
  select(feat_class, med_cor, med_SNR) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)
pred_probs_all <- bind_rows(lst(
  Falkor=FT2040_features, MESOSCOPE=MS3000_features,
  CultureData=CultureData_features, Pttime=Pttime_features), 
  .id = "dataset") %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.9, "Good", "Bad")) %>%
  mutate(error_type=case_when(
    feat_class=="Good" & pred_class=="Good" ~ "%TP",
    feat_class=="Good" & pred_class=="Bad" ~ "%FN",
    feat_class=="Bad" & pred_class=="Good" ~ "%FP",
    feat_class=="Bad" & pred_class=="Bad" ~ "%TN",
    TRUE ~ "Incomparable"
  )) %>%
  mutate(dataset=factor(dataset, levels=c("MESOSCOPE", "Falkor", "CultureData", "Pttime"))) %>%
  filter(error_type!="Incomparable") %>%
  group_by(dataset, error_type) %>%
  summarise(n=n(), .groups = 'drop') %>%
  complete(dataset, error_type, fill=list(n=0))
pred_probs_all %>%
  pivot_wider(names_from=error_type, values_from = n) %>%
  mutate(`%FDR`=`%FP`/(`%FP`+`%TP`)*100) %>%
  mutate(`%GPF`=`%TP`/(`%FN`+`%TP`)*100) %>%
  mutate(`%GPF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `%GPF`)) %>%
  pivot_longer(c(`%FDR`, `%GPF`)) %>%
  mutate(desc=case_when(
    is.na(value) ~ NA,
    name=="%FDR"~ paste0("frac(",`%FP`,",",`%FP`+`%TP`,")"),
    name=="%GPF"~ paste0("frac(",`%TP`,",",`%FN`+`%TP`,")")
  )) %>%
  mutate(vjust_custom=case_when(
    is.na(value) ~ NA,
    name=="%FDR" & dataset=="Pttime" ~ 1.3,
    name=="%FDR" ~ -0.3,
    name=="%GPF" & dataset=="Falkor" ~ 1.3,
    name=="%GPF" ~ -0.3,
  )) %>%
  drop_na() %>% 
  mutate(name=factor(name, levels=c("%GPF", "%FDR"))) %>%
  mutate(dataset=factor(dataset, levels = c("Falkor", "MESOSCOPE", "CultureData", "Pttime"))) %>%
  ggplot(aes(x=dataset, y=value)) +
  geom_point(size=3) +
  geom_text(aes(label=desc, vjust=vjust_custom), parse=TRUE) +
  facet_wrap(~name, scales = "free_y", ncol = 1) +
  scale_y_continuous(limits = c(0, NA))

# Confusion matrices for confirmation
# pred_probs_all
# pred_probs_all %>%
#   ggplot(aes(x=error_type, y=n)) +
#   geom_col(color="#0b505c", fill="#0b505c") +
#   geom_text(aes(label=n), vjust=-0.2, color="#0b505c") +
#   facet_wrap(~dataset, nrow=1) +
#   theme_bw() +
#   theme(axis.title = element_blank())
```

```{r clockplot BMIS}
if(!file.exists("manuscript/MS3000_norm_areas.csv")){
  source("manuscript/BMIS_peakprep.R")
}
```

```{r clockplot dataset munging}
norm_area_data <- read_csv("manuscript/MS3000_norm_areas.csv")

both_min_model <- rbind(FT2040_features, MS3000_features) %>%
  select(feat_class, med_cor, med_SNR) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)
best_feats <- MS3000_features %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
  select(feature, feat_class, pred_class)
clockplot_data <- norm_area_data %>%
  right_join(best_feats, by = join_by(feature)) %>%
  left_join(file_data %>% select(filename, depth), by = join_by(filename))
```

```{r clockplot kw_dunn_tests}
kw_pvals_all <- clockplot_data %>%
  nest(data=-c(feature, pred_class)) %>%
  mutate(tidyaov=map(
    data, ~ broom::tidy(kruskal.test(.x$norm_area~.x$depth)))
  ) %>%
  unnest(tidyaov) %>%
  select(-parameter, -method) %>%
  mutate(cor_pval_all=p.adjust(p.value, method="fdr")) %>%
  group_by(pred_class) %>%
  mutate(cor_pval_byclass=p.adjust(p.value, method="fdr")) %>%
  ungroup() %>%
  select(feature, cor_pval_byclass)
resp_class_levels <- c("All equal", 
    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
    "Hard to tell")
kw_classes_all <- clockplot_data %>%
  group_by(feature) %>%
  mutate(norm_area=rank(norm_area)) %>%
  filter(var(norm_area)>0) %>%
  rstatix::dunn_test(norm_area~depth, p.adjust.method = "none") %>%
  mutate(contrast=paste(group2, group1, sep = "_")) %>%
  select(feature, contrast, est=statistic, pval=p.adj) %>%
  pivot_wider(names_from = contrast, values_from = c(pval, est)) %>%
  mutate(resp_class=case_when(
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m<0 & est_175m_DCM<0 ~ "15m > DCM > 175m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m<0 & est_175m_DCM<0 ~ "DCM > 15m > 175m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m>0 & est_175m_DCM>0 ~ "DCM < 15m < 175m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m>0 & est_175m_DCM>0 ~ "15m < DCM < 175m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m>0 & est_175m_DCM<0 ~ "DCM > 175m > 15m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m<0 & est_175m_DCM>0 ~ "DCM < 175m < 15m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM>0.05 & est_DCM_15m<0 & est_175m_15m<0 ~ "15m > DCM = 175m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM>0.05 & est_DCM_15m>0 & est_175m_15m>0 ~ "15m < DCM = 175m",
    pval_DCM_15m>0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_175m_15m<0 & est_175m_DCM<0 ~ "15m = DCM > 175m",
    pval_DCM_15m>0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_175m_15m>0 & est_175m_DCM>0 ~ "15m = DCM < 175m",
    pval_DCM_15m<0.05 & pval_175m_15m>0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_DCM<0 ~ "DCM > 15m = 175m",
    pval_DCM_15m<0.05 & pval_175m_15m>0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_DCM>0 ~ "DCM < 15m = 175m",
    pval_DCM_15m>0.05 & pval_175m_15m>0.05 & pval_175m_DCM>0.05 ~ "All equal",
    TRUE ~ "Hard to tell"
  )) %>%
  select(feature, resp_class) %>%
  mutate(resp_class=factor(resp_class, levels=resp_class_levels))

clockplot_xy_spots <- clockplot_data %>%
  group_by(feature) %>%
  mutate(norm_area=rank(norm_area)) %>%
  group_by(feature, depth) %>%
  summarise(mean_fold_change=mean(norm_area/50)) %>% # div by 50 bc 99 files
  pivot_wider(names_from=depth, values_from=mean_fold_change) %>%
  mutate(x=DCM-1) %>%           # cursed but works??
  mutate(y=`15m`+DCM/2-1.5) %>% # super fucking cursed
  select(feature, x, y)

clockplot_complete_df <- clockplot_data %>%
  distinct(feature, feat_class, pred_class) %>%
  left_join(clockplot_xy_spots, by = join_by(feature)) %>%
  left_join(kw_pvals_all, by = join_by(feature)) %>%
  left_join(kw_classes_all, by = join_by(feature)) %>%
  mutate(error_type=case_when(
    pred_class=="Good" & feat_class=="Good" ~ "TP",
    pred_class=="Good" & feat_class=="Bad" ~ "FP",
    pred_class=="Bad" & feat_class=="Good" ~ "FN",
    pred_class=="Bad" & feat_class=="Bad" ~ "TN"
  )) %>%
  mutate(error_type=factor(error_type, levels=c("TP", "FP", "FN", "TN"),
                           labels=c("Good peak found", "Bad peak included",
                                    "Good peak missed", "Bad peak avoided")))
```

```{r clockplot create circular colormap}
cmap_name <- "infinity"
cmap_url <- paste0("https://raw.githubusercontent.com/1313e/CMasher/master/",
                   "cmasher/colormaps/", cmap_name, "/", cmap_name, "_norm.txt")
cmtable <- read.table(cmap_url)
cmscale <- rgb(cmtable[,1], cmtable[,2], cmtable[,3])
color_scale <- c("grey50", head(cmscale[seq(1, length(cmscale), length.out=13)], 12), "grey80")
names(color_scale) <- resp_class_levels
```

```{r clockplot}
clock_scale <- 0.66
axis_xends <- c(-1/2, 1.1, -1/2)*clock_scale
axis_yends <- c(sqrt(3)/2, 0, -sqrt(3)/2)*clock_scale
axis_y_dodge <- c(2, 2, -2)*clock_scale/25
clockplot_base_gp <- ggplot() +
  annotate("segment", x=c(0,0,0), y=c(0,0,0), 
           xend=axis_xends, yend=axis_yends,
           arrow=arrow(length=unit(0.02, "npc"), type = "closed")) +
  annotate("segment", x=c(0,0,0), y=c(0,0,0),
           xend=-axis_xends, yend=-axis_yends,
           lty = 2) +
  annotate("label", x=axis_xends, y=axis_yends+axis_y_dodge,
           label=c("15m axis", "DCM axis", "175m axis"), 
           fill="#FFFFFFAA", label.size=0) +
  coord_fixed(xlim = c(-0.65, 1.3)*clock_scale, 
              ylim = c(-1.2, 1.2)*clock_scale, 
              expand=TRUE) +
  scale_color_manual(values = color_scale, breaks = names(color_scale), drop=FALSE) +
  theme_void() +
  theme(legend.position = "right", legend.title = element_blank()) +
  guides(colour = guide_legend(override.aes = list(size=4))) +
  aes(x=x, y=y, color=resp_class, label=feature)

# clockplot_base_gp +
#   geom_point(alpha=0.8, data = clockplot_complete_df)

clockplot_base_gp +
  geom_point(aes(shape=error_type), alpha=0.8, size=2,
             data = clockplot_complete_df) +
  scale_shape_manual(breaks = c("Good peak found", "Bad peak included",
                                "Good peak missed", "Bad peak avoided"), 
                     values = c(19, 3, 4, 1))
```

```{r manual check on clockplot, eval=FALSE}
library(RaMS)
# msdata <- readRDS("made_data_MS3000/msdata.rds")
feat_i <- "FT0828"
row_data <- MS3000_features[MS3000_features$feature==feat_i,]
eic_gp <- msdata$EIC2[mz%between%pmppm(row_data$mean_mz, 5)] %>%
  .[rt%between%(row_data$mean_rt/60+c(-1, 1))] %>%
  .[str_detect(filename, "_Smp_")] %>%
  .[str_detect(filename, "180821")] %>%
  left_join(file_data) %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, group=filename, color=depth))
bxp_gp <- peak_data %>% 
  select(feature, filename, into) %>%
  filter(feature==feat_i) %>%
  filter(str_detect(filename, "_Smp_")) %>%
  filter(str_detect(filename, "180821")) %>%
  left_join(file_data, by = join_by(filename)) %>%
  ggplot() +
  geom_boxplot(aes(x=depth, y=into))
plot(gridExtra::arrangeGrob(eic_gp, bxp_gp))
```

## Discussion

## Citations

2020 review and demand for more robust peakpicking and description of noise sources
https://www.sciencedirect.com/science/article/pii/S0165993620302922

2014 Comparative evaluation of preprocessing freeware on chromatography/mass spectrometry data for signature discovery
https://www.sciencedirect.com/science/article/pii/S0021967314010450?via%3Dihub
  - XCMS is the most cited preprocessing tool currently in the metabolomics literature
  - Table comparing parameters across XCMS, MzMine, MetAlign
  - Selection by appearance in multiple replicates
  - Visual inspection of peaks for quality
  - Reports percentage of false positives as between 90% and 50%
  - Most pressing improvement needed for all the tested data analysis tools was to reduce the percentage of false peaks

2003 the multivariate advantage
https://www.sciencedirect.com/science/article/pii/S0003267003006810?via%3Dihub

2017 One Step Forward for Reducing False Positive and False Negative Compound Identifications

## Unstructured thoughts
Pleas:
  - Report peak boundaries! If not max/min m/z then at least max/min RT
    - Helps with visualization and raw data feature extraction
  - Run a pooled/aggregate sample

Brags:
  - Built the classifier tool in R, very fast, uses RaMS
  - Did the PCA things, sometimes successful

In a Bayesian framework, the base rate (likelihood of being a real peak) is critically important for any downstream analysis.

## Supplement

Literally ALL the confusion matrices (as a table?)

Include data from other classifiers
  - Clustering(?)
  - Random forest
  - Lasso/ridge/elastic net regression

Figure: feature selection / model selection barplot from powerpoint
  - Too many numbers for the main text but useful for power users
  - Colored histograms (Results Fig 1) but for other param options
  - Red/gold barplot at 0.1, 1, 50, 99, 99.9 thresholds

Figure: model stability under smaller training set?
  - Both for all params and raw data params?

Figure: model estimate stability errorbar plot?

### 2023-04-13 manuscript/IS_integrations/is_combine.R
Adding Skyline reports and combined IS file plus creation script

@@ -1,0 +1,47 @@
+library(dplyr)
+
+# Check that SkylineRunner.exe is installed
+if(system("manuscript/IS_integrations/SkylineRunner.exe", show.output.on.console = FALSE)){
+  # Website = https://skyline.ms/wiki/home/software/Skyline/page.view?name=SkylineInstall_64_21-1&submit=false
+  stop("Please grab a copy of SkylineRunner from the website")
+}
+
+sky_neg <- paste0(
+  "manuscript/IS_integrations/SkylineRunner.exe",
+  " --in=manuscript/IS_integrations/HILIC_QE_NEGATIVE_Mesotransect.sky",
+  ' --report-name="Ingalls_Lab_QE_Transition Results"', 
+  ' --report-file="manuscript/IS_integrations/negative_mode.csv"'
+)
+sky_pos <- paste0(
+  "manuscript/IS_integrations/SkylineRunner.exe",
+  " --in=manuscript/IS_integrations/HILIC_QE_POSITIVE_Mesotransect.sky",
+  ' --report-name="Ingalls_Lab_QE_Transition Results"', 
+  ' --report-file="manuscript/IS_integrations/positive_mode.csv"'
+)
+system(sky_neg)
+system(sky_pos)
+
+IS_pos <- read.csv("manuscript/IS_integrations/positive_mode.csv")
+IS_neg <- read.csv("manuscript/IS_integrations/negative_mode.csv")
+IS_all <- bind_rows(IS_pos, IS_neg)
+
+# Fix sample 180821_Smp_MS6C315m_C which didn't get extraction standards added
+# Current fix is to replace with median Area of all other sample files
+ext_stans_list <- c(
+  "Isethionic Acid, 13C2",
+  "L-Cysteic Acid, D3",
+  "Sulfoacetic Acid, 13C2",
+  "Sulfolactic Acid, 13C3",
+  "Taurine, D4"
+)
+update_df <- IS_all %>%
+  filter(grepl("Smp", Replicate.Name)) %>%
+  filter(Precursor.Ion.Name%in%ext_stans_list) %>%
+  mutate(Area=as.numeric(Area)) %>%
+  group_by(Precursor.Ion.Name) %>%
+  summarise(Area=median(Area)) %>%
+  mutate(Replicate.Name="180821_Smp_MS6C315m_C")
+IS_all <- IS_all %>%
[…]

### 2023-04-13 manuscript/BMIS_peakprep.R
Adding BMIS script

@@ -1,0 +1,90 @@
+
+
+# Setup ----
+library(tidyverse)
+library(xcms)
+library(data.table)
+file_data <- read_csv("made_data_MS3000/file_data.csv") %>%
+  mutate(filename=basename(filename))
+msnexp_filled <- readRDS("made_data_MS3000/msnexp_filled.rds")
+
+# Reorganize peak data from XCMS into clean format ----
+peak_data_long <- msnexp_filled %>%
+  chromPeaks() %>%
+  as.data.frame() %>%
+  rownames_to_column() %>%
+  mutate(peakidx=row_number()) %>%
+  select(sample, peakidx, into)
+peak_data <- msnexp_filled %>%
+  featureDefinitions() %>%
+  as.data.frame() %>%
+  filter(rtmed%between%c(30, 1200)) %>%
+  # filter(mzmed%between%pmppm(116.071154, 5))
+  select(peakidx) %>%
+  rownames_to_column("feature") %>%
+  unnest_longer(peakidx) %>%
+  left_join(peak_data_long) %>%
+  mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
+  complete(feature, filename, fill = list(into=0)) %>%
+  left_join(file_data) %>%
+  filter(str_detect(filename, "180821"))
+
+# Perform B-MIS ----
+source("manuscript/IS_integrations/is_combine.R")
+IS_areas <- read_csv("manuscript/IS_integrations/all_IS.csv") %>%
+  mutate(filename=paste0(filename, ".mzML")) %>%
+  rename(area="IS_area") %>%
+  filter(!str_detect(filename, "QC")) %>%
+  pivot_wider(names_from=IS_name, values_from = "IS_area") %>%
+  mutate("Inj_vol"=ifelse(str_detect(filename, "Poo.*Half"), 0.5, 1)) %>%
+  pivot_longer(-filename, names_to = "IS_name", values_to = "IS_area")
+pooled_IS_areas <- IS_areas %>%
+  filter(str_detect(filename, "_Poo_"))
+
+# Check using Proline
+peak_data %>%
+  select(feature, filename, into, depth) %>%
+  filter(str_detect(filename, "_Poo_")) %>%
+  filter(feature=="FT0532") %>%
+  full_join(pooled_IS_areas, multiple = "all") %>%
+  group_by(IS_name) %>%
[…]

### 2023-04-13 manuscript/outline.Rmd
Constructing BMISed areas for clockplot

@@
   group_by(feature, depth) %>%
-  summarise(mean_fold_change=mean(norm_area/50)) %>% # div by 50 bc 99 files
+  summarise(mean_fold_change=mean(norm_area/50), .groups = "drop") %>% 
+  # div by 50 bc 99 files and median 1:99 is 50
   pivot_wider(names_from=depth, values_from=mean_fold_change) %>%

### 2023-04-13 manuscript/BMIS_peakprep.R

@@ -23,10 +23,10 @@ peak_data <- msnexp_filled %>%
   select(peakidx) %>%
   rownames_to_column("feature") %>%
   unnest_longer(peakidx) %>%
-  left_join(peak_data_long) %>%
+  left_join(peak_data_long, by = join_by(peakidx)) %>%
   mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
   complete(feature, filename, fill = list(into=0)) %>%
-  left_join(file_data) %>%
+  left_join(file_data, by = join_by(filename)) %>%
   filter(str_detect(filename, "180821"))
 
 # Perform B-MIS ----
@@ -46,7 +46,7 @@ peak_data %>%
   select(feature, filename, into, depth) %>%
   filter(str_detect(filename, "_Poo_")) %>%
   filter(feature=="FT0532") %>%
-  full_join(pooled_IS_areas, multiple = "all") %>%
+  full_join(pooled_IS_areas, multiple = "all", by = join_by(filename)) %>%
   group_by(IS_name) %>%
   mutate(plot_area=into/IS_area*mean(IS_area)) %>%
   ggplot() +
@@ -60,7 +60,7 @@ peak_data %>%
   
 BMIS <- peak_data %>%
   filter(str_detect(filename, "_Poo_")) %>%
-  full_join(pooled_IS_areas, multiple = "all") %>%
+  full_join(pooled_IS_areas, multiple = "all", by = join_by(filename)) %>%
   group_by(feature, IS_name) %>%
   mutate(plot_area=into/IS_area*mean(IS_area)) %>%
   summarise(cv=sd(plot_area)/mean(plot_area)) %>%

### 2023-04-13 msdial_prep_script.R
Removing MSDIAL efforts no longer relevant

@@ -1,19 +1,0 @@
-
-msdial_dir <- "msdial"
-
-if(dir.exists(msdial_dir)){
-  unlink(msdial_dir, recursive = TRUE, force = TRUE)
-}
-
-R.utils::copyDirectory("mzMLs", msdial_dir)
-file.copy("msdialparams.txt", paste0(msdial_dir, "/msdialparams.txt"), overwrite = TRUE)
-
-msdial_cmd <- paste(
-  r"("C:\Program Files\MSDIAL ver.4.9.221218 Windowsx64\MsdialConsoleApp.exe" lcmsdda)",
-  paste0("-i .\\", msdial_dir, "\\"),
-  paste0("-o .\\", msdial_dir, "\\"),
-  paste0("-m .\\", msdial_dir, "\\msdialparams.txt"),
-  "-p"
-)
-
-system(msdial_cmd)

### 2023-04-19 manuscript/outline.Rmd
Completed first draft of manuscript figures

@@
 date: "`r Sys.Date()`"
-output: html_document
+output: word_document
 ---
@@
 options(tidyverse.quiet = TRUE)
+library(tidyverse)
+library(RaMS)
 ```
@@
 ```{r load data}
-library(tidyverse)
-
 file_data <- read_csv("made_data_MS3000/file_data.csv") %>%
@@
 ## Abstract
+
+TBD
 
@@
     - Increases relative power of significant peaks for multivariate analyses
-  - Allows optimization of chromatography for untargeted approaches
-    - More peaks isn't better if the additional peaks are just noise
   - Allows optimization of peakpicking parameters for untargeted approaches
@@
   - Focuses calculation time/energy on best peaks rather than all equally
+  - Provides consistent, reproducible results via scripting instead of independent expert assessment
 
@@
 
+TBD
+
 ### Sample processing
+
+TBD
 
@@
 
+Get from Laura and MW
+
 ### MS conditions
 
+Get from Laura and MW
+
 ### Peakpicking with XCMS
+
+Fill from the peakpicking_and_prep script or the msnexp_filled object
 
@@
   - Decision to view the entire feature rather than a single trace
+    - 100x less work
+    - More representative of typical MS efforts
   - Decision to color by treatments
+  - Decision to drop missed scans entirely
+
+```{r grab demo msdata, eval=!file.exists("manuscript/minEIC_df.csv")}
+msdata <- readRDS("made_data_MS3000/msdata.rds")
+
+set.seed(123)
+chosen_feats <- MS3000_features %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  group_by(feat_class) %>%
+  slice_sample(n=5) %>%
+  arrange(feat_class) %>%
+  mutate(feat_id=row_number())
+minEIC_df <- pmap(chosen_feats, function(...){
+  with(list(...), {
+    msdata$EIC[mz%between%pmppm(mean_mz, ppm=5)] %>%
+      filter(rt%between%(mean_rt/60+c(-1, 1))) %>%
+      mutate(feature, feat_class, feat_id)
+  })
+}) %>% 
+  bind_rows() %>%
+  filter(str_detect(filename, "180821")) %>%
+  left_join(file_data, by="filename") %>%
+  filter(samp_type=="Smp")
+write_csv(minEIC_df, "manuscript/minEIC_df.csv")
+```
 
 ```{r fig:some good some bad features}
+minEIC_df <- read_csv("manuscript/minEIC_df.csv")
 
+good_chroms_gp <- minEIC_df %>%
+  filter(feat_class=="Good") %>%
+  ggplot() +
+  geom_line(aes(x=rt, y=int, group=filename, color=depth)) +
+  facet_wrap(~feature, nrow = 1, scales = "free") +
+  scale_color_manual(breaks = c("15m", "DCM", "175m"), values = c("#f4bb23", "#028e34", "#0b505c")) +
+  theme_bw() +
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        axis.title = element_blank(),
+        legend.position = "none", 
+        panel.grid = element_blank()) +
+  guides(color=guide_legend(override.aes = list(linewidth=2), title = "Sample depth:  "))
+bad_chroms_gp <- minEIC_df %>%
+  filter(feat_class=="Bad") %>%
+  ggplot() +
+  geom_line(aes(x=rt, y=int, group=filename, color=depth)) +
+  facet_wrap(~feature, nrow = 1, scales = "free") +
+  scale_color_manual(breaks = c("15m", "DCM", "175m"), values = c("#f4bb23", "#028e34", "#0b505c")) +
+  theme_bw() +
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        axis.title = element_blank(),
+        legend.position = "none",
+        panel.grid = element_blank())
+chroms_legend <- cowplot::get_legend(good_chroms_gp + theme(legend.position = "bottom"))
+
+comb_cowplot <- cowplot::plot_grid(good_chroms_gp, bad_chroms_gp, chroms_legend, 
+                                   ncol = 1, rel_heights = c(1, 1, 0.15))
+ggsave(plot = comb_cowplot, 
+       filename = "somegood_somebad_depthcolor_chroms.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 3, units = "in", dpi = 300)
 ```
+
+![Figure 1: Randomly selected ion chromatograms from both "Good" (top row) and "Bad" (bottom row) manual classifications, colored by the depth from which the biological sample was taken.
+DCM = deep chlorophyll maximum, approximately 110 meters.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
 
@@
 ```{r how are raw peak shape metrics calculated}
+minEIC_df <- read_csv("manuscript/minEIC_df.csv")
 
+single_EIC <- minEIC_df %>%
+  filter(feature%in%c("FT0588", "FT0280")) %>%
+  filter(filename=="180821_Smp_MS10C215m_A.mzML")
+
+maybe_skews <- c(2.5, 3, 4, 5)
+skew_names <- c("Large skew", "Small skew", "Minimal skew", "Zero skew")
+names(maybe_skews) <- skew_names
+perf_peak_data <- single_EIC %>%
+  group_by(feature) %>%
+  select(feature, rt, int) %>%
+  mutate(rt=(rt-min(rt))/(max(rt)-min(rt))) %>%
+  mutate(int=(int-min(int))/(max(int)-min(int))) %>%
+  cbind(sapply(maybe_skews, function(x){
+    perf_peak <- dbeta(.$rt, shape1=x, shape2=5)
+    perf_peak/max(perf_peak)
+  })) %>%
+  rename(`Raw data`="int") %>%
+  mutate(feature=factor(feature, levels=c("FT0588", "FT0280"), 
+                        labels=c("Good peak", "Bad peak")))
+summary_stats <- perf_peak_data %>%
+  pivot_longer(-c(feature, rt, `Raw data`)) %>%
+  mutate(new_val=value-`Raw data`) %>%
+  group_by(feature, name) %>%
+  summarise(single_cor=round(cor(`Raw data`, value), 2), 
+            single_sd=round(sd(new_val), 2), 
+            .groups = "drop_last") %>%
+  mutate(name=factor(name, levels=skew_names)) %>%
+  mutate(x_coord=c(0.125+(as.numeric(name)-1)/4))
+
+peak_gp <- perf_peak_data %>%
+  pivot_longer(-c(rt, feature)) %>%
+  mutate(name=factor(name, levels=c("Raw data", skew_names))) %>%
+  ggplot() +
+  geom_line(aes(x=rt, y=value, color=name), lwd=1) +
+  geom_rect(xmin=-Inf, xmax=Inf, ymax=-Inf, ymin=-.05, fill="white", color="grey90") +
+  annotate("text", x=0.5, y=-Inf, vjust=-2, label="Pearson's r:") +
+  facet_wrap(~feature) +
+  geom_text(aes(x=x_coord, y=-Inf, label=single_cor, color=name),
+             data = summary_stats, vjust=-.5, show.legend = FALSE) +
+  scale_color_manual(breaks = c("Raw data", skew_names),
+                     values = c("black", viridisLite::plasma(6)[2:5])) +
+  scale_y_continuous(limits = c(-0.3, 1), breaks = c(0, 0.5, 1.0)) +
+  labs(y="Normalized intensity", color=NULL) +
+  theme_bw() +
+  theme(axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        axis.ticks.x = element_blank(),
+        # strip.background = element_rect(fill="white"),
+        legend.position="none",
+        strip.text = element_text(size = 11))
+residuals_gp <- perf_peak_data %>%
+  pivot_longer(-c(rt, feature, `Raw data`)) %>%
+  mutate(new_val=value-`Raw data`) %>%
+  ggplot() +
+  geom_rect(xmin=-Inf, xmax=Inf, ymax=Inf, ymin=0.85, fill="white", color="grey90") +
+  geom_hline(yintercept = 0, lwd=1) +
+  geom_line(aes(x=rt, y=new_val, color=name), lwd=1) +
+  geom_text(aes(x=x_coord, y=Inf, label=single_sd, color=name),
+             data = summary_stats, vjust=3) +
+  annotate("text", x=0.5, y=Inf, vjust=1.5, label="Residual SD:") +
+  facet_grid(~feature) +
+  scale_color_manual(breaks = c("Raw data", skew_names),
+                     values = c("black", viridisLite::plasma(6)[2:5])) +
+  scale_y_continuous(breaks = c(-1, 0, 1), limits = c(-1.1, 1.3)) +
+  labs(x="Normalized retention time", y="Residuals", color=NULL) +
+  theme_bw() +
+  theme(strip.background = element_blank(),
+        strip.text = element_blank(),
+        legend.position = "none")
+
+fig_legend <- cowplot::get_legend(peak_gp + theme(legend.position="right") + 
+                                    guides(color=guide_legend(override.aes = list(lwd=2)), text="none"))
+
+comb_cowplot <- egg::ggarrange(peak_gp, residuals_gp, ncol = 1)
+
+comb_peakmetrics_gp <- cowplot::plot_grid(
+  comb_cowplot, fig_legend, rel_widths = c(1, 0.25)
+) + theme(plot.background = element_rect(fill = "white", color=NA))
+
+ggsave(plot = comb_peakmetrics_gp, 
+       filename = "peakmetrics_singlechrom.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 4, units = "in", dpi = 300)
 ```
+
+![Figure 2: Method used to calculate the metrics for the two-parameter model from the raw data via comparison to an idealized pseudo-Gaussian peak for both manually identified "Good" and "Bad" peaks.
+Normalization was performed by linearly scaling the raw values into the 0-1 range by subtracting the minimum value and dividing by the maximum.
+Peak shape similarity was measured with Pearson's correlation coefficient and the noise level is estimated as the standard deviation of the residuals after the raw data is subtracted from the idealized peak.](figures/peakmetrics_singlechrom.png){width=70%}
 
@@
 
+### Predicted probability distributions
+
 ```{r colored histograms on all 4 datasets}
 # Train on MESOSCOPE/Falkor combined model
-# Raw metrics only
+# Raw data metrics only
 # Color by feature class
@@
   glm(formula=feat_class~., family = binomial)
-bind_rows(lst(
+pred_prob_class_color_hists_df <- bind_rows(lst(
   Falkor=FT2040_features, MESOSCOPE=MS3000_features,
@@
   )) %>%
-  mutate(dataset=factor(dataset, levels=c("MESOSCOPE", "Falkor", "CultureData", "Pttime"))) %>%
-  ggplot() +
+  mutate(dataset=factor(dataset, levels=c("MESOSCOPE", "Falkor", "CultureData", "Pttime")))
+
+pred_prob_class_color_hists_gp <- ggplot(pred_prob_class_color_hists_df) +
   geom_histogram(aes(x=pred_prob, fill=feat_class), breaks=seq(0, 1, 0.02)) +
@@
   scale_fill_manual(breaks = c("Good", "Bad", "Meh", "Stans only", "Unclassified"),
-                    values = c("#f4bb23", "#a41118", "#0b505c", "#028e34", "grey50"))
-# Add broken y-axis
-# Fix axis titles
-# Format nicely
+                    values = c("#f4bb23", "#a41118", "#0b505c", "#028e34", "grey50")) +
+  labs(x="Predicted probability", y="# of mass features", fill="Manual\nclassification") +
+  ggtitle("Still need to figure out y-axis breaking")
+
+ggsave(plot = pred_prob_class_color_hists_gp, 
+       filename = "pred_prob_class_color_hists.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 3, units = "in", dpi = 300)
 ```
+
+![Figure 3: Histograms showing the estimated likelihood of a given mass feature being categorized as "Good" according to the two-parameter logistic model trained on the combined fully-labeled Falkor and MESOSCOPE datasets.
+Colors indicate the category in which each feature was manually assigned by an expert, with "Stans only" referring to a good peak that was only visible in the standards run alongside the samples.
+Datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (black vertical line) and were otherwise unclassified.](figures/pred_prob_class_color_hists.png){width=70%}
+
+### Model cross-validation with novel datasets
 
@@
 
-lst(all_params, raw_data_params, xcms_params) %>%
+fdr_gpf_cross_train_df <- lst(all_params, raw_data_params, xcms_params) %>%
   imap(function(param_selection, param_name){
@@
   pivot_wider(names_from=error_type, values_from = n) %>%
-  mutate(FDR=round(FP/(FP+TP)*100)) %>%
-  mutate(GPF=round(TP/(FN+TP)*100)) %>%
+  mutate(FDR=FP/(FP+TP)) %>%
+  mutate(GPF=TP/(FN+TP)) %>%
   pivot_longer(c(FDR, GPF)) %>%
@@
   mutate(which_params=factor(which_params, 
-                             labels=c("All\nparams", "Default\nonly", "Raw\ndata params"),
+                             labels=c("All metrics", "XCMS metrics", "Raw data metrics"),
                              levels=c("all_params", "xcms_params", "raw_data_params"))) %>%
-  mutate(cross_type_num=as.numeric(cross_type)) %>%
+  mutate(cross_type_num=as.numeric(cross_type))
+
+train_test_levels <- c("MS3000-MS3000", "FT2040-FT2040", "MS3000-FT2040", "FT2040-MS3000")
+train_test_labels <- c("MESO-MESO", "Falkor-Falkor", "MESO-Falkor", "Falkor-MESO")
+fdr_gpf_cross_train_gp <- fdr_gpf_cross_train_df %>%
+  mutate(train_test=factor(train_test, levels=train_test_levels, labels=train_test_labels)) %>%
+  mutate(name=factor(name, levels=c("%GPF", "%FDR"), labels=c("GPF", "FDR"))) %>%
   ggplot() +
-  geom_point(aes(x=cross_type, y=value, color=train_test, fill=train_test, shape=name)) +
+  geom_point(aes(x=cross_type, y=value, color=train_test, 
+                 fill=train_test, shape=name), size=5) +
   geom_smooth(aes(x=cross_type_num, y=value, group=name), 
@@
                    breaks = c("Same", "Different")) +
-  scale_color_manual(values = c("#0b505c", "#0d8299", "#184d1f", "#028e34"), 
-                     breaks = c("MS3000-MS3000", "MS3000-FT2040", "FT2040-FT2040", "FT2040-MS3000"),
+  scale_color_manual(values = c("#0b505c", "#184d1f", "#0d8299", "#028e34"), 
+                     breaks = train_test_labels,
                      aesthetics = c("color", "fill")) +
-  scale_shape_manual(breaks = c("%FDR", "%GPF"), values = c(25, 17)) +
-  coord_cartesian(ylim = c(0, 100)) +
-  theme_bw()
+  scale_shape_manual(breaks = c("GPF", "FDR"), values = c("\u25B2", "\u25BC")) +
+  coord_cartesian(ylim = c(0, 1)) +
+  scale_y_continuous(labels = scales::label_percent()) +
+  theme_bw() +
+  theme() +
+  labs(x=NULL, y=NULL, color="Train-test", fill="Train-test", shape="Error type") +
+  guides(shape=guide_legend(order=1))
+
+ggsave(plot = fdr_gpf_cross_train_gp, 
+       filename = "fdr_gpf_cross_train.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 3, units = "in", dpi = 300)
 ```
 
+![Figure 4: False discovery rate (%FDR) and fraction of good peaks found (GPF) plotted across different subsets of model parameters.
+Lower %FDR indicates a smaller fraction of false positives among those peaks the model categorized as "Good" using a threshold of 0.5, and higher %GPF indicates a larger fraction of the total good peaks were found using the same threshold.
+Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel.
+Lines of best fit have been estimated and plotted in black on top of the data points.](figures/fdr_gpf_cross_train.png){width=70%}
+
+### Full model probability depends strongly on training data, two-parameter model does not
+
 ```{r red/gold rank dotplot step-down method}
-# Make sure to include overplotting somehow?
 dataset_versions <- c("FT2040_features", "MS3000_features")
 raw_data_params <- c("med_cor", "med_SNR")
+xcms_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", 
+                 "sd_pw", "log_mean_area", "log_sd_area", "sn", 
+                 "f", "scale", "lmin", "feat_npeaks", "n_found", 
+                 "samps_found", "stans_found")
 all_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", "sd_pw",
@@
                 "smp_to_std", "shape_cor", "area_cor", "feat_class")
-dotplot_stepdown_data <- lst(all_params, raw_data_params) %>%
+dotplot_stepdown_data <- lst(all_params, xcms_params, raw_data_params) %>%
   imap(function(param_selection, param_name){
@@
   }) %>% bind_rows()
+
 raw_prob_space_gp <- dotplot_stepdown_data %>%
@@
   filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(which_params=factor(which_params, 
+                             levels=c("all_params", "xcms_params", "raw_data_params"),
+                             labels=c("All metrics", "XCMS metrics", "Raw data metrics"))) %>%
   ggplot() +
   geom_abline(slope = 1, intercept = 0, color="black", lwd=1) +
-  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class)) +
+  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class), alpha=0.2) +
   facet_wrap(~which_params, nrow = 1) +
@@
   scale_y_continuous(breaks = c(0, 0.5, 1)) +
-  labs(x="Falkor-trained model", y="MESO-trained model")
+  coord_fixed() +
+  labs(x="Falkor-trained model (predicted probability)", 
+       y="MESO-trained model\n(predicted probability)") +
+  theme_bw() +
+  theme(legend.position = "none",
+        strip.background = element_rect(fill="white"))
 rank_prob_space_gp <- dotplot_stepdown_data %>%
@@
   pivot_wider(names_from = training_set, values_from = pred_prob) %>%
+  mutate(which_params=factor(which_params, 
+                             levels=c("all_params", "xcms_params", "raw_data_params"),
+                             labels=c("All metrics", "XCMS metrics", "Raw data metrics"))) %>%
   filter(feat_class%in%c("Good", "Bad")) %>%
@@
   geom_abline(slope = 1, intercept = 0, color="black", lwd=1) +
-  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class)) +
+  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class), alpha=0.2) +
   facet_wrap(~which_params, nrow = 1) +
@@
                      labels = c("Min\nrank", "Max\nrank")) +
-  labs(x="Falkor-trained model", y="MESO-trained model") +
-  theme(axis.text.x = element_text(hjust=c(0, 1)))
-plot(gridExtra::arrangeGrob(raw_prob_space_gp, rank_prob_space_gp))
+  labs(x="Falkor-trained model (ranked likelihood)", 
+       y="MESO-trained model\n(ranked likelihood)",
+       color=NULL) +
+  guides(color=guide_legend(override.aes = list(size=4, alpha=0.8))) +
+  coord_fixed() +
+  theme_bw() +
+  theme(axis.text.x = ggtext::element_markdown(hjust=c(0, 1)),
+        # strip.text = element_blank(),
+        strip.background = element_rect(fill="white"),
+        legend.position=c(0.875, 0.25),
+        legend.background = element_blank(),
+        legend.spacing.x = unit(0.003, units = "npc"))
+comb_prob_space_gp <- egg::ggarrange(raw_prob_space_gp, rank_prob_space_gp, draw = FALSE)
+ggsave(plot = comb_prob_space_gp, 
+       filename = "comb_prob_space.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 5, units = "in", dpi = 300)
 ```
+
+![Figure 5: Predicted likelihood of a feature being "Good" according to a model trained on the MESOSCOPE dataset vs a model trained on the Falkor dataset.
+The top row of plots show the exact likelihood predicted by the logistic model across three different subsets of parameters, while the bottom row shows the estimates ranked from least likely to most likely.
+Points are colored by their manually-assigned quality according to an expert.](figures/comb_prob_space.png){width=70%}
+
+Make sure labels are the same as above - probably stick with XCMS for specificity.
+
+Remember to add Pearson and Spearmen correlation coefficients in the corners!
+
+### Error rates on novel data with stricter likelihood threshold
 
@@
   complete(dataset, error_type, fill=list(n=0))
-pred_probs_all %>%
+error_type_all_datasets <- pred_probs_all %>%
   pivot_wider(names_from=error_type, values_from = n) %>%
-  mutate(`%FDR`=`%FP`/(`%FP`+`%TP`)*100) %>%
-  mutate(`%GPF`=`%TP`/(`%FN`+`%TP`)*100) %>%
-  mutate(`%GPF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `%GPF`)) %>%
-  pivot_longer(c(`%FDR`, `%GPF`)) %>%
+  mutate(`FDR`=`%FP`/(`%FP`+`%TP`)) %>%
+  mutate(`GPF`=`%TP`/(`%FN`+`%TP`)) %>%
+  mutate(`GPF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `GPF`)) %>%
+  pivot_longer(c(`FDR`, `GPF`)) %>%
+  drop_na()
+
+fdr_gpf_all_datasets_gp <- error_type_all_datasets %>%
     mutate(desc=case_when(
     is.na(value) ~ NA,
-    name=="%FDR"~ paste0("frac(",`%FP`,",",`%FP`+`%TP`,")"),
-    name=="%GPF"~ paste0("frac(",`%TP`,",",`%FN`+`%TP`,")")
+    name=="FDR"~ paste0("frac(",`%FP`,",",`%FP`+`%TP`,")"),
+    name=="GPF"~ paste0("frac(",`%TP`,",",`%FN`+`%TP`,")")
   )) %>%
@@
     is.na(value) ~ NA,
-    name=="%FDR" & dataset=="Pttime" ~ 1.3,
-    name=="%FDR" ~ -0.3,
-    name=="%GPF" & dataset=="Falkor" ~ 1.3,
-    name=="%GPF" ~ -0.3,
+    name=="FDR" & dataset=="Pttime" ~ 1.3,
+    name=="FDR" ~ -0.3,
+    name=="GPF" ~ 1.3,
   )) %>%
-  drop_na() %>% 
-  mutate(name=factor(name, levels=c("%GPF", "%FDR"))) %>%
-  mutate(dataset=factor(dataset, levels = c("Falkor", "MESOSCOPE", "CultureData", "Pttime"))) %>%
+  mutate(name=factor(name, levels=c("GPF", "FDR"),
+                     labels=c("Fraction of good peaks found (GPF)", 
+                              "False discovery rate (FDR)"))) %>%
+  mutate(dataset=factor(dataset, 
+                        levels = c("Falkor", "MESOSCOPE", "CultureData", "Pttime"),
+                        labels= c("Falkor", "MESO-\nSCOPE", "Culture\nData", "Pttime"))) %>%
   ggplot(aes(x=dataset, y=value)) +
@@
   facet_wrap(~name, scales = "free_y", ncol = 1) +
-  scale_y_continuous(limits = c(0, NA))
+  scale_y_continuous(limits = c(0, NA), labels = scales::label_percent()) +
+  theme_bw() +
+  theme(axis.title = element_blank(),
+        strip.background = element_rect(fill="white"))
+
+ggsave(plot = fdr_gpf_all_datasets_gp, 
+       filename = "fdr_gpf_all_datasets.png", 
+       device = "png", path = "manuscript/figures",
+       width = 3, height = 4, units = "in", dpi = 300)
 
@@
 
-```{r clockplot BMIS}
-if(!file.exists("manuscript/MS3000_norm_areas.csv")){
+![Figure 6: False discovery rate and proportion of total good peaks identified as good by the two-parameter model trained on the combined MESOSCOPE/Falkor dataset and applied to each dataset individually with a threshold of 0.9.
+FDR is calculated by dividing the number of false positives by the total positives produced by the model and GPF is calculated by dividing the number of true positives by the total number of good peaks as identified manually.
+Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fdr_gpf_all_datasets.png){width=30%}
+
+Emphasize the use of a different threshold here!
+
+### Implications for biological conclusions
+
+```{r clockplot BMIS, eval=!file.exists("manuscript/MS3000_norm_areas.csv")}
 source("manuscript/BMIS_peakprep.R")
-}
 ```
 
-```{r clockplot dataset munging}
+```{r clockplot dataset munging, eval=!file.exists("manuscript/MS3000_clockplot_coords.csv")}
 norm_area_data <- read_csv("manuscript/MS3000_norm_areas.csv")
@@
   left_join(file_data %>% select(filename, depth), by = join_by(filename))
-```
 
-```{r clockplot kw_dunn_tests}
 kw_pvals_all <- clockplot_data %>%
@@
                                     "Good peak missed", "Bad peak avoided")))
+
+write_csv(clockplot_complete_df, file = "manuscript/MS3000_clockplot_coords.csv")
 ```
 
-```{r clockplot create circular colormap}
+```{r clockplot}
+resp_class_levels <- c("All equal", 
+    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
+    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
+    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
+    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
+    "Hard to tell")
+clockplot_complete_df <- read_csv(file = "manuscript/MS3000_clockplot_coords.csv") %>%
+  mutate(resp_class=factor(resp_class, levels=resp_class_levels))
+
 cmap_name <- "infinity"
@@
 names(color_scale) <- resp_class_levels
-```
 
-```{r clockplot}
 clock_scale <- 0.66
@@
 axis_y_dodge <- c(2, 2, -2)*clock_scale/25
+
 clockplot_base_gp <- ggplot() +
@@
   coord_fixed(xlim = c(-0.65, 1.3)*clock_scale, 
-              ylim = c(-1.2, 1.2)*clock_scale, 
+              ylim = c(-1, 1)*clock_scale, 
               expand=TRUE) +
@@
   theme_void() +
-  theme(legend.position = "right", legend.title = element_blank()) +
-  guides(colour = guide_legend(override.aes = list(size=4))) +
+  theme(legend.position = "right", 
+        legend.title = element_blank(),
+        plot.background = element_rect(fill="white", color="white")) +
+  guides(colour = guide_legend(override.aes = list(size=4, alpha=1), drop=FALSE)) +
   aes(x=x, y=y, color=resp_class, label=feature)
 
-# clockplot_base_gp +
-#   geom_point(alpha=0.8, data = clockplot_complete_df)
-
-clockplot_base_gp +
-  geom_point(aes(shape=error_type), alpha=0.8, size=2,
+clockplot_complete_gp <- clockplot_base_gp +
+  geom_point(aes(shape=error_type), alpha=0.5, size=2,
              data = clockplot_complete_df) +
@@
                      values = c(19, 3, 4, 1))
+
+ggsave(plot = clockplot_complete_gp, 
+       filename = "clockplot_complete.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 5, units = "in", dpi = 300)
 ```
 
+![Figure 7: Plot of metabolite response to depth shown across three axes using the rank-normalized median value at each depth as the coordinate for that axis.
+Each molecular feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile.
+Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the deep chlorophyll maximum; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}
+
+Make sure I include the threshold used for the above plot (in the legend?)
+
+Make sure I explain the statistical significance and the "Hard to tell" category
+
+See aobut including ONLY the good peaks in a separate plot (inset figure?)
+
 ```{r manual check on clockplot, eval=FALSE}
-library(RaMS)
 # msdata <- readRDS("made_data_MS3000/msdata.rds")
@@
 
+TBD
+
 ## Citations
@@
 https://www.sciencedirect.com/science/article/pii/S0021967314010450?via%3Dihub
+
   - XCMS is the most cited preprocessing tool currently in the metabolomics literature

### 2023-04-19 manuscript/outline.Rmd
Updated methods figures to look nicer(?)

@@
 
-comb_cowplot <- egg::ggarrange(peak_gp, residuals_gp, ncol = 1)
+comb_cowplot <- egg::ggarrange(peak_gp, residuals_gp, ncol = 1, draw = FALSE)
 
@@
   ggtitle("Still need to figure out y-axis breaking")
-
 ggsave(plot = pred_prob_class_color_hists_gp, 
@@
        width = 6.5, height = 3, units = "in", dpi = 300)
+
+# Manual facetting with egg and cowplot
+base_gp <- ggplot() +
+  aes(x=pred_prob, fill=feat_class) +
+  facet_wrap(~dataset, scales="free_y") +
+  scale_fill_manual(breaks = c("Good", "Bad", "Meh", "Stans only", "Unclassified"),
+                    values = c("#f4bb23", "#a41118", "#0b505c", "#028e34", "grey50")) +
+  labs(x="Predicted probability", y="# of mass features", fill="Manual\nclassification") +
+  theme_bw() +
+  theme(axis.text.y.right = element_blank(),
+        axis.ticks.y.right = element_blank(),
+        legend.position = "none")
+axis_breaks <- pred_prob_class_color_hists_df %>%
+  group_by(dataset) %>%
+  mutate(pred_prob=cut(pred_prob, breaks=seq(0, 1, 0.02))) %>%
+  count(pred_prob) %>%
+  slice(1) %>%
+  pull(n, dataset)
+falk_hist_gp <- base_gp + 
+  geom_histogram(breaks=seq(0, 1, 0.02), data = pred_prob_class_color_hists_df %>% 
+                   filter(dataset=="Falkor")) +
+  scale_y_continuous(breaks = c(0, 50, 100, 150, 195),
+                     labels = c(0, 50, 100, 150, axis_breaks["Falkor"])) +
+  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
+  annotate("rect", xmin=0, xmax=1, ymin=180, ymax=185, fill="white", color="black") +
+  annotate("rect", xmin=0, xmax=0.125, ymin=195, ymax=Inf, fill="white", color="grey90") +
+  coord_cartesian(ylim = c(0, 200), expand = FALSE) +
+  theme(axis.title.x = element_blank(), axis.text.x = element_blank())
+meso_hist_gp <- base_gp + 
+  geom_histogram(breaks=seq(0, 1, 0.02), data = pred_prob_class_color_hists_df %>% 
+                   filter(dataset=="MESOSCOPE")) +
+  scale_y_continuous(breaks = c(0, 50, 100, 150, 200, 240),
+                     labels = c(0, 50, 100, 150, 200, axis_breaks["MESOSCOPE"])) +
+  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
+  annotate("rect", xmin=0, xmax=1, ymin=220, ymax=230, fill="white", color="black") +
+  annotate("rect", xmin=0, xmax=0.125, ymin=240, ymax=Inf, fill="white", color="grey90") +
+  coord_cartesian(ylim = c(0, 250), expand = FALSE) +
+  theme(axis.title = element_blank(), axis.text.x = element_blank())
+culture_hist_gp <- base_gp + 
+  geom_histogram(breaks=seq(0, 1, 0.02), data = pred_prob_class_color_hists_df %>% 
+                   filter(dataset=="CultureData")) +
+  scale_y_continuous(breaks = c(0, 50, 100, 116),
+                     labels = c(0, 50, 100, axis_breaks["CultureData"])) +
+  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
+  annotate("rect", xmin=0, xmax=1, ymin=110, ymax=113, fill="white", color="black") +
+  annotate("rect", xmin=0, xmax=0.125, ymin=116, ymax=Inf, fill="white", color="grey90") +
+  coord_cartesian(ylim = c(0, 120), expand = FALSE)
+pttime_hist_gp <- base_gp + 
+  geom_histogram(breaks=seq(0, 1, 0.02), data = pred_prob_class_color_hists_df %>% 
+                   filter(dataset=="Pttime")) +
+  scale_y_continuous(breaks = c(0,100,200,300,400,510),
+                     labels = c(0,100,200,300,400,axis_breaks["Pttime"])) +
+  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
+  annotate("rect", xmin=0, xmax=1, ymin=470, ymax=485, fill="white", color="black") +
+  annotate("rect", xmin=0, xmax=0.125, ymin=510, ymax=Inf, fill="white", color="grey90") +
+  coord_cartesian(ylim = c(0, 520), expand = FALSE) +
+  theme(axis.title.y = element_blank())
+egg_grob <- egg::ggarrange(falk_hist_gp, meso_hist_gp, culture_hist_gp, pttime_hist_gp, draw = FALSE)
+egg_legend <- cowplot::get_legend(culture_hist_gp + theme(legend.position = "right"))
+egg_complete <- cowplot::plot_grid(egg_grob, egg_legend, rel_widths = c(1, 0.25)) +
+  theme(plot.background = element_rect(fill = "white", color=NA))
+ggsave(plot = egg_complete, 
+       filename = "manuscript/figures/pred_prob_class_color_hists.png", 
+       device = "png",
+       width = 6.5, height = 4, units = "in", dpi = 300)
 ```

### 2023-04-20 manuscript/manuscript.Rmd
Starting to write manuscript proper

---
title: 'Picky with peakpicking: assessing chromatographic peak quality with simple
  metrics in metabolomics'
author: "William Kumler"
date: "`r Sys.Date()`"
output:
  word_document:
    reference_docx: template.docx
  pdf_document: default
---

## Abstract

Chromatographic peakpicking continues to represent a significant bottleneck in automated LC-MS workflows. Uncontrolled false discovery rates and the lack of manually-calibrated quality metrics require researchers to visually evaluate individual peaks, requiring large amounts of time and breaking replicability. Here, we manually labeled two oceanographic particulate metabolite datasets to assess the performance of individual peak quality metrics and constructed a predictive model calibrated to the likelihood that visual inspection by an MS expert would include a given mass feature in the downstream analysis. A simple logistic regression model built on two metrics calculated from the raw MS data reduced the fraction of false positives in the analysis from 80-90% down to 1-5% and showed minimal overfitting when applied to novel datasets. We then explored the implications of this quality thresholding on the conclusions obtained by the downstream analysis, concluding that the poor performance of peakpicking algorithms significantly reduces the power of both univariate and multivariate statistical analyses in detecting environmental differences.

## Introduction

Liquid chromatography-mass spectrometry (LC-MS) is a powerful tool for exploring the molecular composition of biological samples. Its rapid sample processing (typically <1 hr run time), low limits of detection (pM-nM range), and ability to characterize novel molecules via fragmentation fingerprints make it a common workhorse for metabolomic research. However, much of the data produced by these instruments goes unused because traditional workflows target only a fraction of the compounds detected based on an *a priori* list of known molecules. Significant efforts in the past two decades have established workflows for untargeted metabolomics [@xcms, @mzmine, @msdial], but these data-driven methods are still handicapped by the poor performance of the core peakpicking algorithms and require manual oversight and curation of the datasets after they're produced by the software. This peakpicking step is typically the very first in the data processing stage, with implications for all downstream analysis and interpretation.

Peakpicking is a well-known problem across many fields of science and especially in LC-MS analysis due to the smooth transition from high-quality peaks into background noise as the limit of detection is approached and crossed, with background signals often sharing many features of high-quality peaks. The imperfection of these algorithms introduces a fundamental tradeoff between false positives and false negatives, where false positives (Type I errors) represent background instrument or chemical noise that has been misclassified as true biological signal and false negatives (Type II errors) indicating biological signals that are misclassified as noise. Existing algorithms tend to be very forgiving of Type I errors and very harsh on Type II, presenting many false positives to the researcher in their attempt to absolutely minimize the number of real molecules missed. This approach makes sense because downstream analyses can always remove additional mass features, but Type II errors cannot be later recovered. However, we would argue that this emphasis on an absolute minimization of the missed peaks does a disservice to scientists for two reasons. First, the abundance of false positives means that peakpicking output cannot be automatically trusted. This mandates manual evaluation of the raw data, a time-consuming and mind-numbing process with a large degree of subjective decision making that cannot be perfectly replicated and scales combinatorially with the number of samples and compounds measured. Second, the presence of false negatives cannot be fully accounted for in the way the false positive rate can. False negatives are introduced in every step of metabolomic analyses, with compounds lost during sample processing, injection, and detection on the instrument. This is a well-known problem in metabolomics and will not be solved by any perfecting of peakpicking algorithms. Instead, emphasis should be placed on allowing the experimenter to set an acceptable threshold for the false discovery rate and accepting that this will inherently add to the number of peaks already lost in the data collection process.

Existing peak-detection softwares do not provide a clear way to do this. Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each molecular feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew which are rarely estimated using a well-documented algorithm. None of these parameters answer the critical question of interest: what is the likelihood that this subset of data corresponds to an actual molecule present in the original sample? This parameter is crucial for downstream analysis because it represents the base rate for error propagation and acceptable thresholds should vary widely by the particular project's goals. In an exploratory analysis, any molecular feature more than 50% likely to be real is perhaps worth considering, while in a confirmatory study this threshold may need to be above 99%. While significant effort has been recently invested into improvements in the peakpicking algorithms, very little has been done to quantify the accuracy and precision of their outputs on the wide variety of datasets from which they are determined.

A single metric of peak quality calibrated to its likelihood of representing an actual molecule has multiple advantages over the current heuristic approaches. First, providing an intuitive metric of peak goodness allows researchers to focus their energy on those features least likely to be noise and most likely to be found in a repeated analysis. This applies both during the manual curation phase in which the very best or most interesting peaks can be visually inspected while those of poor quality can be discarded, as well as in the compound discovery/characterization steps where time and energy spent attempting to determine the structure or molecular formula of a noise signal is wasted. Second, removing low-quality noise signals from the dataset prior to downstream analysis should improve statistical power by reducing the number of hypotheses tested for univariate analyses and increasing the relative power of significant peaks in multivariate analyses. Third, such a metric would allow optimization of peakpicking and chromatographic parameters for untargeted datasets. Rather than measuring the effect of a change on a small subset of known molecules or by counting the total number of peaks found, one could look at the metric's distribution to see whether the change was an improvement in the number and absolute quality of the peaks detected. Fourth, high quality peaks should be independent of software implementations and should be found no matter what algorithms are used, relieving the scientist (and reviewers!) of some of the burden of learning new software or programming languages to check their work. Fifth, this metric would enable more consistent quality control between labs and analyists. Rather than discussing a plethora of parameters or making subjective assessments of peak shape, an idealized metric would provide a simple and intuitive threshold upon which to agree for a given project. This metric would be scripted and provide consistent, reproducible results independent of the particular expert reviewing its performance.

A single comprehensive metric also has significant advantages over multiple individual thresholds. First, a metric calibrated to likelihood is easier and more intuitive to use for classification than relatively arbitrary thresholds of signal-to-noise or peak skewness with ambiguous units and unintuitive interpretations. Second, a single compound metric also has the multivariate advantage over independent thresholds by allowing a good peak to be weak in one area which is compensated for by strong performance in other metrics. Sometimes good peaks do have high SNR or small areas, and thresholding on these metrics independently unnecessarily increases the likelihood of missing signals. Finally, the relative power of these many different metrics is difficult to compute and hard to compare. Is it more important to have an isotope, or a Gaussian peak shape? Is it more likely to be a biological signal if it has a large area, or if it differs between treatment? If so, by what degree? These metrics differ in strength depending on the dataset, but many MS experts have some degree of heuristic knowledge about their relative power for the data they've worked with and analyzed.

An area particularly ripe for LC-MS analysis is that of the open ocean. Low compound and high salt concentrations make this area difficult to study [@BMIS] but its vast size and direct effect on the Earth's carbon cycling make it critical that we understand its dynamics of energy and nutrients on a molecular scale. Metabolites are the currency of chemical exchange both intra- and inter-cellularly, serving as building blocks of larger molecules, regulators of osmotic balance and storage of nutrients, as well as chemical signals on their own. These small molecules serve both as signposts for the complex biological landscape in this highly dynamic region and give a sense of not only who is present but also what ecological roles they're serving and the niches they fill.

In this paper, we use open ocean marine metabolite LC-MS samples to develop and test a variety of chromatographic peak metrics. Here, we construct and validate multiple predictive models of peak quality based on metrics both common in the literature and custom implementations we've found useful in our own analysis. This allows us to connect the physical, chemical, and biological measurements taken regularly around the globe to a molecular-scale perspective of particular organic matter in the ocean by linking the chemical currencies that fuel the planet to the environments in which they're found.

### 2023-04-20 manuscript/outline.Rmd
Updating outline with figure notes from Frank

@@
 DCM = deep chlorophyll maximum, approximately 110 meters.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
+
+Include the descriptor for what "FTXXXX" means.
+
+Note that these were taken from the MESOSCOPE dataset.
+
+Explain the chromatograms (RT/int/ what "goodness" requires?)
+
+Explain that we grabbed the center of the peak +/- 1 (2?) minutes
 
@@
 
+Explain what the different datasets are?
+
 ### Model cross-validation with novel datasets
@@
 Lines of best fit have been estimated and plotted in black on top of the data points.](figures/fdr_gpf_cross_train.png){width=70%}
+
+Dark blue and dark green are too similar, need to separate these better somehow.
+
+Include conclusions in figure legend? Something like "slope of the line is much greater in All metrics and XCMS metrics, indicating overfitting"?
 
@@
 
-Remember to add Pearson and Spearmen correlation coefficients in the corners!
+Remember to add Pearson and Spearman correlation coefficients in the corners!
 
@@
 Emphasize the use of a different threshold here!
+
+Explain why there are no values for GPF in Pttime/CultureData?
 

### 2023-04-25 peakpicking_and_prep.R
Making prefilter explicitly the same in all datasets

@@ -21,7 +21,6 @@ dataset_version <- "Pttime"
 
 output_folder <- paste0("made_data_", dataset_version, "/")
 mzML_files <- list.files(paste0(output_folder, "mzMLs/"), full.names=TRUE)
-prefilter_versioned <- c(3, 1e6)
 
 if(dataset_version%in%c("FT2040", "MS3000")){
   file_data <- data.frame(filename=mzML_files) %>%
@@ -83,7 +82,7 @@ register(BPPARAM = SnowParam(workers = parallel::detectCores()-1,
 cwp <- CentWaveParam(
   ppm = 5, 
   peakwidth = c(20, 80), 
-  prefilter = prefilter_versioned, 
+  prefilter = c(3, 1e6), 
   snthresh = 0, 
   verboseColumns = TRUE, 
   extendLengthMSW = TRUE, 

### 2023-04-25 randomforest_attempts.R
Fixing randomforests after feat_id -> feature rename

@@ -13,7 +13,7 @@ features_extracted <- read_csv(paste0(output_folder, "features_extracted.csv")) 
 
 # Using ALL features up to this point ----
 feature_subset <- features_extracted %>%
-  select(-feat_id, -feat_class) %>%
+  select(-feature, -feat_class) %>%
   data.matrix()
 rfmodel <- randomForest(feature_subset, y=factor(features_extracted$feat_class), 
                         importance=TRUE, proximity = TRUE, keep.forest = TRUE)

### 2023-04-25 manuscript/manuscript.Rmd
Adding methods and the first chunk of results

@@
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_knit$set(root.dir = '..')
+options(readr.show_col_types = FALSE)
+options(tidyverse.quiet = TRUE)
+library(tidyverse)
+library(RaMS)
+```
+
+```{r load data}
+file_data <- read_csv("made_data_MS3000/file_data.csv") %>%
+  mutate(depth=factor(depth, levels=c("15m", "DCM", "175m"))) %>%
+  mutate(filename=basename(filename))
+FT2040_features <- read_csv("made_data_FT2040/features_extracted.csv")
+MS3000_features <- read_csv("made_data_MS3000/features_extracted.csv")
+Pttime_features <- read_csv("made_data_Pttime/features_extracted.csv")
+CultureData_features <- read_csv("made_data_CultureData/features_extracted.csv")
+
+dataset_versions <- c("FT2040_features", "MS3000_features")
+
+raw_data_params <- c("med_cor", "med_SNR")
+xcms_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", 
+                 "sd_pw", "log_mean_area", "log_sd_area", "sn", 
+                 "f", "scale", "lmin", "feat_npeaks", "n_found", 
+                 "samps_found", "stans_found")
+all_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", "sd_pw",
+                "log_mean_area", "log_sd_area", "sn", "f", "scale", 
+                "lmin", "feat_npeaks", "n_found", "samps_found", "stans_found", 
+                "med_cor", "med_SNR", "med_missed_scans", "smp_to_blk", 
+                "smp_to_std", "shape_cor", "area_cor", "feat_class")
+```
 
@@
 This peakpicking step is typically the very first in the data processing stage, with implications for all downstream analysis and interpretation.
+[Mention HILIC chromatography specifically to justify its use here?]
 
@@
 This allows us to connect the physical, chemical, and biological measurements taken regularly around the globe to a molecular-scale perspective of particular organic matter in the ocean by linking the chemical currencies that fuel the planet to the environments in which they're found.
+
+## Methods
+
+### Sample collection
+
+Environmental samples were collected from the North Pacific Subtropical Gyre near Station ALOHA during two research cruises that targeted strong mesoscale eddy features during June/July 2017 and March/April 2018, traversing an area between 28 °N, 156 °W and 23 °N, 161 °W.
+An eddy dipole off the coast of Hawaii was detected using sea-level anomaly (SLA) satellite data and targeted for both a transect across the cyclonic and anticyclonic poles of the eddy dipole.
+The cyclonic pole of the eddy had a maximum negative SLA anomaly of -15 cm in 2017 and -20 cm in 2018, while the anticyclonic center reached +24 cm in 2017 and +21 cm in 2018.
+The 2017 cruise samples were taken along a transect across the eddy dipole while the 208 cruise targeted only the center of each eddy.
+
+Environmental samples were obtained using the onboard CTD rosette to collect water from 15 meters, the deep chlorophyll maximum (DCM), and 175 meters during the 2017 MESOSCOPE cruise and from 25 meters and the DCM during the 2018 Falkor cruise.
+The DCM was determined visually from fluorometer data during the CTD downcast and Niskin bottles were tripped during the return trip to the surface.
+Seawater from each depth was sampled in triplicate by firing one Niskin bottle for each sample.
+Samples were brought to the surface and decanted into prewashed (3x with DI, 3x with sampled seawater) polycarbonate bottles for filtration.
+Samples were filtered by peristaltic pump onto 142mm 0.2 µm Durapore filters held by polycarbonate filter holders on a Masterflex tubing line.
+Pressures were kept as low as possible while still producing a reasonable rate of flow through the filter, approximately 250-500 mL per minute.
+Samples were then removed from the filter holder using solvent-washed tweezers and placed into pre-combusted aluminum foil packets that were then flash-frozen in liquid nitrogen before being stored at -80 °C until extraction.
+A methodological blank was also collected by running filtrate through a new filter and then treated identically to the samples.
+
+Culture samples used as the validation sets for this paper were obtained from existing [blah] and have been previously described by [@Bryn] and [@Pttime paper?].
+
+### Sample processing
+
+Extraction of the environmental samples followed a modified Bligh & Dyer approach as detailed in [@BMIS].
+Briefly, filters were added to PTFE centrifuge tubes with a 1:1 mix of 100 µm and 400 µm silica beads, approximately 2mL -20 °C Optima-grade DCM, and approximately 3mL -20 °C 1:1 methanol/water solution (both also Optima-grade).
+Extraction standards were added during this step.
+The samples were then bead-beaten three times, followed by triplicate washes with fresh methanol/water mixture.
+Samples were then dried down under clean nitrogen gas and warmed using a Fisher-Scientific Reacti-Therm module.
+Dried aqueous fractions were re-dissolved in 380 µL of Optima-grade water and amended with 20 µL isotope-labeled injection standards.
+Additional internal standards were added at this point to measure the variability introduced by chromatography and ionization, and the reconstituted fraction was syringe-filtered to remove any potential clogging material.
+This aqueous fraction was then aliquoted into an HPLC vial for injection on the HILIC column and diluted 1:1 with Optima-grade water.
+A pooled sample was created by combining 20 µL of each sample into the same HPLC vial, and a 1:1 dilution with water half-strength sample was aliquot from that to assess matrix effects and obscuring variation [@BMIS].
+Also run alongside the environmental samples were two mixes of authentic standards in water and in an aliquot of the pooled sample at a variety of concentrations for quality control, annotation, and absolute concentration calculations.
+HPLC vials containing the samples were frozen at -80 °C until thawing shortly before injection.
+
+The CultureData samples were re-run from the frozen aliquots for this paper. The Pttime sample processing is documented [on MW? in their paper?].
+
+### LC conditions
+
+For the MESOSCOPE, Falkor, and CultureData samples a SeQuant ZIC-pHILIC column (5 um particle size, 2.1 mm x 150 mm, from Millipore) was used with 10 mM ammonium carbonate in 85:15 acetonitrile to water (Solvent A) and 10 mM ammonium carbonate in 85:15 water to acetonitrile (Solvent B) at a flow rate of 0.15 mL/min.
+The column was held at 100% A for 2 minutes, ramped to 64% B over 18 minutes, ramped to 100% B over 1 minute, held at 100% B for 5 minutes, and equilibrated at 100% A for 25 minutes (50 minutes total).
+The column was maintained at 30 °C.
+The injection volume was 2 µL for samples and standard mixes.
+When starting a batch, the column was equilibrated at the starting conditions for at least 30 minutes.
+To improve the performance of the HILIC column, we maintained the same injection volume, kept the instrument running water blanks between samples as necessary, and injected standards in a representative matrix (the pooled sample) in addition to standards in water.
+After each batch, the column was flushed with 10 mM ammonium carbonate in 85:15 water to acetonitrile for 20 to 30 minutes.
+LC conditions for the Pttime samples are documented [on MW?
+in their paper?].
+
+### MS conditions
+
+Environmental metabolomic data was collected on a Thermo Q Exactive HF hybrid Orbitrap (QE) mass spectrometer.
+The capillary and auxiliary gas heater temperatures were maintained at 320°C and 100°C, respectively.
+The S-lens RF level was kept at 65, the H-ESI voltage was set to 3.3 kV and sheath gas, auxiliary gas, and sweep gas flow rates were set at 16, 3, and 1, respectively.
+Polarity switching was used with a scan range of 60 to 900 m/z and a resolution of 60,000.
+Calibration was performed every 3-4 days at a target mass of 200 m/z.
+DDA data was collected from the pooled samples for high-confidence annotation of knowns and unknowns.
+All files were then converted to an open-source mzML format and centroided via Proteowizard’s msConvert tool.
+For the Pttime samples, files were pulled directly from Metabolomics Workbench via Project ID [blah] and used in their existing mzXML format.
+
+### Peakpicking, alignment, and grouping with XCMS
+
+The R package XCMS was used to perform peakpicking, retention time correction, and peak correspondence.
+Files were loaded and run separately for each dataset (Falkor, MESOSCOPE, CultureData, and Pttime) using the new "OnDiskMSnExp" infrastructure.
+Default parameters for the CentWave peakpicking algorithm were used except for: ppm, which was set to 5; peakwidth, which was widened to 20-80 seconds; prefilter, for which the intensity threshold was raised to $10^6$; and integrate, which was set to 2 instead of 1.
+snthresh was set to zero because there are known issues with background estimation in this algorithm [@myers2017], and both verboseColumns and the extendLengthMSW parameter were set to TRUE.
+For retention time correction, the Obiwarp method was used except for the CultureData dataset, which was visually inspected and determined not to require correction.
+For the Obiwarp algorithm, the binsize was reduced to 0.1 but all other parameters were left at their defaults or equivalents.
+
+Peak grouping was performed on the two environmental datasets and the Pttime data with a bandwidth of 12, a minFraction of 0.1, binSize of 0.001, and minSamples of 2 but otherwise default arguments.
+CultureData's minFraction was raised to 0.4 but was otherwise identical.
+Sample groups were constructed to consist of the biological replicates for all datasets.
+After peak grouping, peak filling was performed using the fillChromPeaks function with the ppm parameter set to 2.5.
+Finally, mass features with a retention time less than 30 seconds or larger than 20 minutes were removed to avoid interference from the initial and final solvent washes.
+[Check this with Laura]
+
+### Manual inspection and classification
+
+After the full XCMS workflow was completed, the molecular features were visually inspected by a single qualified MS expert.
+For the Falkor and MESOSCOPE datasets, every molecular feature was inspected, while only those features with a predicted probability of 0.9 or higher according to the two-parameter model produced below were for the CultureData and Pttime datasets.
+Inspection consisted of plotting the raw intensity values against the corrected retention-time values for all data points within the *m/z* x RT bounding box determined by the most extreme values for the given feature.
+For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labelling and to more accurately represent what MS experts typically do when assessing the quality of a given peak.
+[Fig 1.] We also decided to ignore missing values and linearly interpolate between known data points rather than filling with zeroes.
+These EICs were then shown to an MS expert for classification into one of 4 categories: Good, Bad, Ambiguous, or Stans only if the peak appeared to only show up in the standards.
+A few randomly-chosen features from the manually-assigned Good and Bad classifications are shown in [Fig.
+1].
+
+```{r grab demo msdata, eval=!file.exists("manuscript/minEIC_df.csv")}
+msdata <- readRDS("made_data_MS3000/msdata.rds")
+
+set.seed(123)
+chosen_feats <- MS3000_features %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  group_by(feat_class) %>%
+  slice_sample(n=5) %>%
+  arrange(feat_class) %>%
+  mutate(feat_id=row_number())
+minEIC_df <- pmap(chosen_feats, function(...){
+  with(list(...), {
+    msdata$EIC[mz%between%pmppm(mean_mz, ppm=5)] %>%
+      filter(rt%between%(mean_rt/60+c(-1, 1))) %>%
+      mutate(feature, feat_class, feat_id)
+  })
+}) %>% 
+  bind_rows() %>%
+  filter(str_detect(filename, "180821")) %>%
+  left_join(file_data, by="filename") %>%
+  filter(samp_type=="Smp")
+write_csv(minEIC_df, "manuscript/minEIC_df.csv")
+```
+
+```{r fig:some good some bad features}
+minEIC_df <- read_csv("manuscript/minEIC_df.csv")
+
+good_chroms_gp <- minEIC_df %>%
+  filter(feat_class=="Good") %>%
+  ggplot() +
+  geom_line(aes(x=rt, y=int, group=filename, color=depth)) +
+  facet_wrap(~feature, nrow = 1, scales = "free") +
+  scale_color_manual(breaks = c("15m", "DCM", "175m"), values = c("#f4bb23", "#028e34", "#0b505c")) +
+  theme_bw() +
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        axis.title = element_blank(),
+        legend.position = "none", 
+        panel.grid = element_blank()) +
+  guides(color=guide_legend(override.aes = list(linewidth=2), title = "Sample depth:  "))
+bad_chroms_gp <- minEIC_df %>%
+  filter(feat_class=="Bad") %>%
+  ggplot() +
+  geom_line(aes(x=rt, y=int, group=filename, color=depth)) +
+  facet_wrap(~feature, nrow = 1, scales = "free") +
+  scale_color_manual(breaks = c("15m", "DCM", "175m"), values = c("#f4bb23", "#028e34", "#0b505c")) +
+  theme_bw() +
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        axis.title = element_blank(),
+        legend.position = "none",
+        panel.grid = element_blank())
+chroms_legend <- cowplot::get_legend(good_chroms_gp + theme(legend.position = "bottom"))
+
+comb_cowplot <- cowplot::plot_grid(good_chroms_gp, bad_chroms_gp, chroms_legend, 
+                                   ncol = 1, rel_heights = c(1, 1, 0.15))
+ggsave(plot = comb_cowplot, 
+       filename = "somegood_somebad_depthcolor_chroms.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 3, units = "in", dpi = 300)
+```
+
+![Figure 1: Randomly selected ion chromatograms from both "Good" (top row) and "Bad" (bottom row) manual classifications, colored by the depth from which the biological sample was taken.
+DCM = deep chlorophyll maximum, approximately 110 meters.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
+
+### Peak feature extraction and metric calculation
+
+Our process of feature engineering was divided into three categories of features: those that could be calculated using a single peak trace, those that required access to the entire file, and those that could only be calculated across multiple files.
+[Rewrite without the "three categories" business ugh] For metrics belonging to the first two categories, summary statistics were used (mean/median, standard deviation) to create a single consensus value for the mass feature as a whole.
+Missing values were ignored during these summary statistic calculations by setting na.rm=TRUE.
+Distributions were visually inspected and highly abnormal metrics were transformed using log-scaling if necessary.
+Using a pairs plot, highly correlated (above a Pearson's r ~ .9) had one of the correlated metrics removed.
+
+The first category consisted largely of information produced directly by XCMS: the very typical *m/z* ratios, retention times, peak widths, and integrated areas as well as the less common and poorly-documented sn, f, scale, and lmin parameters.
+Although we requested additional parameters from XCMS with the verboseColumns = TRUE argument, all of the values returned were NAs and could not be used in the model building so were dropped.
+The total number of peaks found was calculated as the number of peaks divided by the total number of files, and the total number of files in which a peak was initially detected was calculated by subtracting the number of NAs in the peak baseline estimate (which is NA if the peak was filled in) from the total number of files and then dividing by the number of files to normalize across datasets.
+This calculation were performed for the Falkor, MESOSCOPE, and CultureData datasets to estimate the proportion of sample and standard files in which a given peak was found, but could not be supplied for the Pttime dataset because no standard files were available.
+The sn metric contained a large number of zero values to begin with that then became negative infinities when log-scaled, so we replaced those values with zeroes and functionally equated an sn of 0 with an sn of 1.
+
+We also calculated several metrics from the raw mz/rt/int values.
+Using the R package RaMS we extracted the data points falling within each individual peak's mz and retention time bounding box (values between the XCMS-reported min and max) separately for each file.
+The data points were then linearly scaled to fall within the 0-1 range by subtracting the minimum RT and dividing by the maximum RT, then each scaled RT was fit to a beta distribution with α values of 2.5, 3, 4, and 5, and a fixed β value of 5.
+This approach allowed us to approximate a bell curve with increasing degrees of right-skewness and the beta distribution was chosen because it is constrained between 0 and 1 and simple and speedy to generate in R.
+For each α value, Pearson's correlation coefficient was calculated between the beta distribution and the raw data, with the highest value returned as a metric for how peak-shaped the data were [Fig.
+2].
+The beta distribution with the highest *r* was also then used to estimate the noise level within the peak by also scaling both the beta distribution densities and the raw data intensity values as described above, then subtracting the scaled beta distribution from the scaled intensity values, producing the residuals of the fit [Fig.
+2].
+The signal-to-noise ratio (SNR) was calculated by dividing the maximum original peak height by the standard deviation of the residuals multiplied by the maximum height of the original peak.
+This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional molecular signals.
+[@myers?] If there were fewer than 5 data points, a missing value was returned and dropped in subsequent summary calculations.
+Accessing the raw data values also allowed us to calculate the proportion of "missed" scans in a peak for which a retention time exists at other masses in the same sample but for which no data was produced at the selected *m/z* ratio.
+
+```{r how are raw peak shape metrics calculated}
+minEIC_df <- read_csv("manuscript/minEIC_df.csv")
+
+single_EIC <- minEIC_df %>%
+  filter(feature%in%c("FT0588", "FT0280")) %>%
+  filter(filename=="180821_Smp_MS10C215m_A.mzML")
+
+maybe_skews <- c(2.5, 3, 4, 5)
+skew_names <- c("Large skew", "Small skew", "Minimal skew", "Zero skew")
+names(maybe_skews) <- skew_names
+perf_peak_data <- single_EIC %>%
+  group_by(feature) %>%
+  select(feature, rt, int) %>%
+  mutate(rt=(rt-min(rt))/(max(rt)-min(rt))) %>%
+  mutate(int=(int-min(int))/(max(int)-min(int))) %>%
+  cbind(sapply(maybe_skews, function(x){
+    perf_peak <- dbeta(.$rt, shape1=x, shape2=5)
+    perf_peak/max(perf_peak)
+  })) %>%
+  rename(`Raw data`="int") %>%
+  mutate(feature=factor(feature, levels=c("FT0588", "FT0280"), 
+                        labels=c("Good peak", "Bad peak")))
+summary_stats <- perf_peak_data %>%
+  pivot_longer(-c(feature, rt, `Raw data`)) %>%
+  mutate(new_val=value-`Raw data`) %>%
+  group_by(feature, name) %>%
+  summarise(single_cor=round(cor(`Raw data`, value), 2), 
+            single_sd=round(sd(new_val), 2), 
+            .groups = "drop_last") %>%
+  mutate(name=factor(name, levels=skew_names)) %>%
+  mutate(x_coord=c(0.125+(as.numeric(name)-1)/4))
+
+peak_gp <- perf_peak_data %>%
+  pivot_longer(-c(rt, feature)) %>%
+  mutate(name=factor(name, levels=c("Raw data", skew_names))) %>%
+  ggplot() +
+  geom_line(aes(x=rt, y=value, color=name), lwd=1) +
+  geom_rect(xmin=-Inf, xmax=Inf, ymax=-Inf, ymin=-.05, fill="white", color="grey90") +
+  annotate("text", x=0.5, y=-Inf, vjust=-2, label="Pearson's r:") +
+  facet_wrap(~feature) +
+  geom_text(aes(x=x_coord, y=-Inf, label=single_cor, color=name),
+             data = summary_stats, vjust=-.5, show.legend = FALSE) +
+  scale_color_manual(breaks = c("Raw data", skew_names),
+                     values = c("black", viridisLite::plasma(6)[2:5])) +
+  scale_y_continuous(limits = c(-0.3, 1), breaks = c(0, 0.5, 1.0)) +
+  labs(y="Normalized intensity", color=NULL) +
+  theme_bw() +
+  theme(axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        axis.ticks.x = element_blank(),
+        # strip.background = element_rect(fill="white"),
+        legend.position="none",
+        strip.text = element_text(size = 11))
+residuals_gp <- perf_peak_data %>%
+  pivot_longer(-c(rt, feature, `Raw data`)) %>%
+  mutate(new_val=value-`Raw data`) %>%
+  ggplot() +
+  geom_rect(xmin=-Inf, xmax=Inf, ymax=Inf, ymin=0.85, fill="white", color="grey90") +
+  geom_hline(yintercept = 0, lwd=1) +
+  geom_line(aes(x=rt, y=new_val, color=name), lwd=1) +
+  geom_text(aes(x=x_coord, y=Inf, label=single_sd, color=name),
+             data = summary_stats, vjust=3) +
+  annotate("text", x=0.5, y=Inf, vjust=1.5, label="Residual SD:") +
+  facet_grid(~feature) +
+  scale_color_manual(breaks = c("Raw data", skew_names),
+                     values = c("black", viridisLite::plasma(6)[2:5])) +
+  scale_y_continuous(breaks = c(-1, 0, 1), limits = c(-1.1, 1.3)) +
+  labs(x="Normalized retention time", y="Residuals", color=NULL) +
+  theme_bw() +
+  theme(strip.background = element_blank(),
+        strip.text = element_blank(),
+        legend.position = "none")
+
+fig_legend <- cowplot::get_legend(peak_gp + theme(legend.position="right") + 
+                                    guides(color=guide_legend(override.aes = list(lwd=2)), text="none"))
+
+comb_cowplot <- egg::ggarrange(peak_gp, residuals_gp, ncol = 1, draw = FALSE)
+
+comb_peakmetrics_gp <- cowplot::plot_grid(
+  comb_cowplot, fig_legend, rel_widths = c(1, 0.25)
+) + theme(plot.background = element_rect(fill = "white", color=NA))
+
+ggsave(plot = comb_peakmetrics_gp, 
+       filename = "peakmetrics_singlechrom.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 4, units = "in", dpi = 300)
+```
+
+![Figure 2: Method used to calculate the metrics for the two-parameter model from the raw data via comparison to an idealized pseudo-Gaussian peak for both manually identified "Good" and "Bad" peaks.
+Normalization was performed by linearly scaling the raw values into the 0-1 range by subtracting the minimum value and dividing by the maximum.
+Peak shape similarity was measured with Pearson's correlation coefficient and the noise level is estimated as the standard deviation of the residuals after the raw data is subtracted from the idealized peak.](figures/peakmetrics_singlechrom.png){width=70%}
+
+We additionally estimated the existence of a $^{13}$C isotope using a similar method to extract the raw mz/rt/int values within the peak bounding box, then searched the same RT values at an *m/z* delta of +1.003355.
+In places where more than 5 data points existed at both the original mass and the $^{13}$C mass, we again used Pearson's correlation coefficient to estimate the similarity between the two mass traces and used a trapezoidal Riemann sum to estimate the area of the original and isotope peaks.
+The overall feature isotope shape similarity was calculated by taking the median of the correlation coefficients.
+We also calculated the correlation coefficient of the ratio of the $\frac{^{13}C}{^{12}C}$ across multiple files, expecting that a true isotope would have a fixed $\frac{^{13}C}{^{12}C}$ ratio.
+Both the isotope shape similarity and the isotope area correlation were used as metrics in the downstream analysis.
+Peaks for which no isotope signal was detected or had too few scans to calculate the above metrics were imputed with NA values that were again dropped in the calculation of summary statistics.
+
+Finally, we calculated several metrics using a design-of-experiments (DoE) approach for the Falkor and MESOSCOPE datasets which were expected to have a large difference in metabolite composition between those samples collected at the surface (15-25 meters) and those collected deeper in the water column.
+We used ANOVAs to compare the integrated peak area with depth and extracted the p-values of each test, which became an additional potential peak quality metric.
+We also calculated the difference between the mean sample area and the mean blank area and the difference between the mean sample area and the mean area of the standards which also became metrics.
+When there were too many missing values to perform the ANOVA or sample averages, we replaced the calculated statistic with a value a single order of magnitude outside the most extreme value.
+
+### Statistics
+
+All analyses were run in R, version 4.2.2.
+Logistic regressions were performed using base R's `glm` function with elastic net functionality provided by the glmnet package, random forest classifiers were performed using the randomForest package, and initial clustering approaches were explored with the dbscan package.
+
+## Results
+
+```{r summary stats for pred_prob histograms, include=FALSE}
+lst(FT2040_features, MS3000_features, Pttime_features, CultureData_features) %>%
+  bind_rows(.id = "dataset") %>%
+  with(table(dataset, feat_class)) %>%
+  `/`(rowSums(.)) %>%
+  round(2)
+
+train_set <- rbind(FT2040_features, MS3000_features) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=feat_class=="Good")
+
+both_min_model <- paste0("feat_class~", paste(raw_data_params, collapse = "+")) %>%
+  glm(data = train_set, family = binomial)
+both_xcms_model <- paste0("feat_class~", paste(xcms_params, collapse = "+")) %>%
+  glm(data = train_set, family = binomial)
+both_full_model <- paste0("feat_class~", paste(all_params, collapse = "+")) %>%
+  glm(data = train_set, family = binomial)
+
+pred_probs <- bind_rows(lst(
+  Falkor=FT2040_features, MESOSCOPE=MS3000_features,
+  CultureData=CultureData_features, Pttime=Pttime_features), 
+  .id = "dataset") %>%
+  mutate(pred_prob_min=predict(object=both_min_model,newdata = ., type = "response")) %>%
+  mutate(pred_prob_xcms=predict(object=both_xcms_model,newdata = ., type = "response")) %>%
+  mutate(pred_prob_full=predict(object=both_full_model,newdata = ., type = "response"))
+
+pred_probs %>%
+  select(starts_with("pred_prob")) %>%
+  pivot_longer(cols = everything()) %>%
+  mutate(pred_cuts=cut(value, breaks = c(0, 0.01, 0.1, 0.5, 0.9, 0.99, 1))) %>%
+  with(table(name, pred_cuts))
+```
+
+An average of 3,300 molecular features (MFs) were reported by XCMS across the 4 datasets, with the fewest (1495) in the Falkor data and the most in the Pttime samples (7781).
+In the Falkor and MESOSCOPE datasets that were fully labeled by an MS expert, approximately 70% (69% and 73%, respectively) of the features were given a "Bad" designation, corresponding to noise peaks that the expert would not have included in a downstream analysis.
+In both, 5% of the MFs were unable to be assigned confidently to either "Good" or "Bad" classes and 10% were identified as appearing only in the standards, leaving only ~15% of the features classified as "Good" (16% and 12%, respectively).
+
+We used three different multiple logistic regression models to predict the likelihood of each MF being categorized as "Good".
+The first model included all metrics calculated as described above in Methods, the second contained only those parameters immediately available from the XCMS output without revisiting the raw data (the four core peak metrics *m/z*, RT, peak width, area and their standard deviations plus the mysterious lmin, f, and scale values as well as the fraction of peaks, samples, and standards found), and the final model was a simple two-parameter model using only the peak shape and novel SNR metrics.
+In all three cases, the majority of MFs were estimated to have a less than 1% chance of being good.
+The full model and the XCMS model both displayed a strongly bimodal distribution, with a large number of peaks also exceeding a 99% chance of being good, while the two-parameter model had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Fig.
+3).
+
+```{r fig3 pred_prob_class_color_hists}
+base_gp <- ggplot() +
+  aes(x=pred_prob_min, fill=feat_class) +
+  facet_wrap(~dataset, scales="free_y") +
+  scale_fill_manual(breaks = c("Good", "Bad", "Meh", "Stans only", "Unclassified"),
+                    values = c("#f4bb23", "#a41118", "#0b505c", "#028e34", "grey50")) +
+  labs(x="Predicted probability", y="# of mass features", fill="Manual\nclassification") +
+  theme_bw() +
+  theme(axis.text.y.right = element_blank(),
+        axis.ticks.y.right = element_blank(),
+        legend.position = "none")
+axis_breaks <- pred_probs %>%
+  group_by(dataset) %>%
+  mutate(pred_prob_min=cut(pred_prob_min, breaks=seq(0, 1, 0.02))) %>%
+  count(pred_prob_min) %>%
+  slice(1) %>%
+  pull(n, dataset)
+falk_hist_gp <- base_gp + 
+  geom_histogram(breaks=seq(0, 1, 0.02), data = pred_probs %>% 
+                   filter(dataset=="Falkor")) +
+  scale_y_continuous(breaks = c(0, 50, 100, 150, 195),
+                     labels = c(0, 50, 100, 150, axis_breaks["Falkor"])) +
+  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
+  annotate("rect", xmin=0, xmax=1, ymin=180, ymax=185, fill="white", color="black") +
+  annotate("rect", xmin=0, xmax=0.125, ymin=195, ymax=Inf, fill="white", color="grey90") +
+  coord_cartesian(ylim = c(0, 200), expand = FALSE) +
+  theme(axis.title.x = element_blank(), axis.text.x = element_blank())
+meso_hist_gp <- base_gp + 
+  geom_histogram(breaks=seq(0, 1, 0.02), data = pred_probs %>% 
+                   filter(dataset=="MESOSCOPE")) +
+  scale_y_continuous(breaks = c(0, 50, 100, 150, 200, 240),
+                     labels = c(0, 50, 100, 150, 200, axis_breaks["MESOSCOPE"])) +
+  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
+  annotate("rect", xmin=0, xmax=1, ymin=220, ymax=230, fill="white", color="black") +
+  annotate("rect", xmin=0, xmax=0.125, ymin=240, ymax=Inf, fill="white", color="grey90") +
+  coord_cartesian(ylim = c(0, 250), expand = FALSE) +
+  theme(axis.title = element_blank(), axis.text.x = element_blank())
+culture_hist_gp <- base_gp + 
+  geom_histogram(breaks=seq(0, 1, 0.02), data = pred_probs %>% 
+                   filter(dataset=="CultureData")) +
+  scale_y_continuous(breaks = c(0, 50, 100, 116),
+                     labels = c(0, 50, 100, axis_breaks["CultureData"])) +
+  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
+  annotate("rect", xmin=0, xmax=1, ymin=110, ymax=113, fill="white", color="black") +
+  annotate("rect", xmin=0, xmax=0.125, ymin=116, ymax=Inf, fill="white", color="grey90") +
+  coord_cartesian(ylim = c(0, 120), expand = FALSE)
+pttime_hist_gp <- base_gp + 
+  geom_histogram(breaks=seq(0, 1, 0.02), data = pred_probs %>% 
+                   filter(dataset=="Pttime")) +
+  scale_y_continuous(breaks = c(0,100,200,300,400,510),
+                     labels = c(0,100,200,300,400,axis_breaks["Pttime"])) +
+  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
+  annotate("rect", xmin=0, xmax=1, ymin=470, ymax=485, fill="white", color="black") +
+  annotate("rect", xmin=0, xmax=0.125, ymin=510, ymax=Inf, fill="white", color="grey90") +
+  coord_cartesian(ylim = c(0, 520), expand = FALSE) +
+  theme(axis.title.y = element_blank())
+egg_grob <- egg::ggarrange(falk_hist_gp, meso_hist_gp, culture_hist_gp, pttime_hist_gp, draw = FALSE)
+egg_legend <- cowplot::get_legend(culture_hist_gp + theme(legend.position = "right"))
+egg_complete <- cowplot::plot_grid(egg_grob, egg_legend, rel_widths = c(1, 0.25)) +
+  theme(plot.background = element_rect(fill = "white", color=NA))
+ggsave(plot = egg_complete, 
+       filename = "manuscript/figures/pred_prob_class_color_hists.png", 
+       device = "png",
+       width = 6.5, height = 4, units = "in", dpi = 300)
+```
+
+![Figure 3: Histograms showing the estimated likelihood of a given mass feature being categorized as "Good" according to the two-parameter logistic model trained on the combined fully-labeled Falkor and MESOSCOPE datasets.
+Colors indicate the category in which each feature was manually assigned by an expert, with "Stans only" referring to a good peak that was only visible in the standards run alongside the samples.
+Datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (black vertical line) and were otherwise unclassified.](figures/pred_prob_class_color_hists.png){width=90%}
+
+In each case, we determined each molecular feature as a true positive (TP) if it was predicted to be Good and was manually classified as Good, a true negative if both predicted and classifed as Bad, a false positive if predicted to be Good but manually classified as Bad, and a false negative if predicted to be Bad but was in fact manually classified as Good.
+This allowed us to additionally define two useful measures of success, the traditionally-defined false discovery rate (FDR, defined as 1-precision or the number of false positives divided by the total number of predicted positives) and the percentage of good peaks found (GPF, also known as the recall or sensitivity and defined as the number of true positives divided by the total number of positive predictions).
+These two measures allowed us to monitor two key questions typically asked of a predictive model in this context: first, what fraction of the data predicted to be good is actually noise; and second, what fraction of the good data am I actually recovering in the downstream analysis?
+
+The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when [predicting] using the full dataset, with FDR values in the 5-10% range and 80-90% GPF values implying that a large majority of the good peaks passed the threshold.
+The XCMS metrics performed slightly worse, with FDR values in the 10-15% range and GPF values closer to 75%.
+The two-parameter model performed worst when tested internally, with an FDR of about 20% and %GPF also around 75% (Figure 4).
+However, when the models were trained on a different dataset than the one they were used to predict classifications for, nearly all models had similar performance with %FDR around 10-25 and %GPF around 60-80.
+This dramatic drop in performance on out-of-training data strongly implied that the full and XCMS-only models were highly overfit, while the two-parameter model had probably reached stability.
+Also notably, the model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more peaks recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE [and it's unclear why this should be the case] (Figure 4).
+
+```{r FDR vs GPF cross-train triangle figure feature selection}
+fdr_gpf_cross_train_df <- lst(all_params, raw_data_params, xcms_params) %>%
+  imap(function(param_selection, param_name){
+    lapply(dataset_versions, function(train_set){
+      full_model <- get(train_set) %>% 
+        filter(feat_class%in%c("Good", "Bad")) %>%
+        select(feat_class, all_of(param_selection)) %>%
+        mutate(feat_class=feat_class=="Good") %>%
+        glm(formula=feat_class~., family = binomial)
+      lapply(dataset_versions, function(test_set){
+        get(test_set) %>%
+          mutate(pred_prob=predict(object=full_model,newdata = ., type = "response")) %>%
+          mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
+          select(feat_class, starts_with("pred_class")) %>%
+          mutate(train_test=paste(train_set, test_set, sep = "-")) %>%
+          mutate(train_test=str_remove_all(train_test, "_features")) %>%
+          mutate(cross_type=ifelse(train_set==test_set, "Same", "Different"))
+      }) %>% bind_rows()
+    }) %>% bind_rows() %>% mutate(which_params=param_name)
+  }) %>% bind_rows() %>%
+  mutate(error_type=case_when(
+    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
+    feat_class=="Bad" & pred_class=="Good" ~ "FP",
+    feat_class=="Good" & pred_class=="Bad" ~ "FN",
+    feat_class=="Good" & pred_class=="Good" ~ "TP"
+  )) %>%
+  group_by(train_test, which_params, cross_type) %>%
+  count(error_type) %>%
+  ungroup() %>%
+  mutate(cross_type=factor(cross_type, levels=c("Same", "Different"))) %>%
+  pivot_wider(names_from=error_type, values_from = n) %>%
+  mutate(FDR=FP/(FP+TP)) %>%
+  mutate(GPF=TP/(FN+TP)) %>%
+  pivot_longer(c(FDR, GPF)) %>%
+  mutate(name=paste0("%", name)) %>%
+  mutate(which_params=factor(which_params, 
+                             labels=c("All metrics", "XCMS metrics", "Raw data metrics"),
+                             levels=c("all_params", "xcms_params", "raw_data_params"))) %>%
+  mutate(cross_type_num=as.numeric(cross_type))
+
+train_test_levels <- c("MS3000-MS3000", "FT2040-FT2040", "MS3000-FT2040", "FT2040-MS3000")
+train_test_labels <- c("MESO-MESO", "Falkor-Falkor", "MESO-Falkor", "Falkor-MESO")
+fdr_gpf_cross_train_gp <- fdr_gpf_cross_train_df %>%
+  mutate(train_test=factor(train_test, levels=train_test_levels, labels=train_test_labels)) %>%
+  mutate(name=factor(name, levels=c("%GPF", "%FDR"), labels=c("GPF", "FDR"))) %>%
+  ggplot() +
+  geom_point(aes(x=cross_type, y=value, color=train_test, 
+                 fill=train_test, shape=name), size=5) +
+  geom_smooth(aes(x=cross_type_num, y=value, group=name), 
+              method = "lm", formula="y~x", color="black", se=FALSE) +
+  facet_grid(~which_params) +
+  scale_x_discrete(labels = c("Same\ndataset\n", "Different\ndataset\n"),
+                   breaks = c("Same", "Different")) +
+  scale_color_manual(values = c("#0b505c", "#184d1f", "#0d8299", "#028e34"), 
+                     breaks = train_test_labels,
+                     aesthetics = c("color", "fill")) +
+  scale_shape_manual(breaks = c("GPF", "FDR"), values = c("\u25B2", "\u25BC")) +
+  coord_cartesian(ylim = c(0, 1)) +
+  scale_y_continuous(labels = scales::label_percent()) +
+  theme_bw() +
+  theme() +
+  labs(x=NULL, y=NULL, color="Train-test", fill="Train-test", shape="Error type") +
+  guides(shape=guide_legend(order=1))
+
+ggsave(plot = fdr_gpf_cross_train_gp, 
+       filename = "fdr_gpf_cross_train.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 3, units = "in", dpi = 300)
+```
+
+![Figure 4: False discovery rate (%FDR) and fraction of good peaks found (GPF) plotted across different subsets of model parameters.
+Lower %FDR indicates a smaller fraction of false positives among those peaks the model categorized as "Good" using a threshold of 0.5, and higher %GPF indicates a larger fraction of the total good peaks were found using the same threshold.
+Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel.
+Lines of best fit have been estimated and plotted in black on top of the data points.](figures/fdr_gpf_cross_train.png){width=90%}
+
+To further explore questions of model stability and the potential for overfitting, we compared the predictions from a Falkor-trained model to a MESOSCOPE-trained model.
+In the ideal situation, these predictions are the same because the values are independent of a given training set's quirks, but this was not what we found for the full and XCMS-only models (Figure 5).
+We compared the absolute likelihood values using Pearson's correlation coefficient and found that the two-parameter models had the highest $r$ value of 0.996, while the full models and the XCMS-trained models had $r$ values of 0.856 and 0.896, respectively (Figure 5).
+Finally, we compared the model likelihoods in ranked space with Spearman's correlation coefficient to test the question of whether or not the *ranking* was similar even if the absolute values differed.
+Here, we found an intensification of the effect described before, with an even higher $r$ for the two-parameter model of 0.998 but lower $r$ values for the full and XCMS-trained model of 0.818 and 0.849, respectively (Figure 5).
+
+```{r red/gold rank dotplot step-down method}
+dotplot_stepdown_data <- lst(all_params, xcms_params, raw_data_params) %>%
+  imap(function(param_selection, param_name){
+    lapply(dataset_versions, function(train_set){
+      model_i <- get(train_set) %>% 
+        filter(feat_class%in%c("Good", "Bad")) %>%
+        select(feat_class, all_of(param_selection)) %>%
+        mutate(feat_class=feat_class=="Good") %>%
+        glm(formula=feat_class~., family = binomial)
+      MS3000_features %>%
+        mutate(pred_prob=predict(model_i, ., type="response")) %>%
+        select(feature, feat_class, pred_prob) %>%
+        mutate(which_params=param_name) %>%
+        mutate(training_set=train_set)
+    }) %>% bind_rows()
+  }) %>% bind_rows()
+
+stepdown_cors <- dotplot_stepdown_data %>%
+  group_by(training_set, which_params) %>%
+  pivot_wider(names_from = training_set, values_from = c(pred_prob)) %>%
+  summarize(pcor=cor(FT2040_features, MS3000_features, method = "pearson"),
+            scor=cor(FT2040_features, MS3000_features, method = "spearman"))
+
+raw_prob_space_gp <- dotplot_stepdown_data %>%
+  pivot_wider(names_from = training_set, values_from = pred_prob) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(which_params=factor(
+    which_params, levels=c("all_params", "xcms_params", "raw_data_params"),
+    labels=c("All metrics", "XCMS metrics", "Raw data metrics"))) %>%
+  ggplot() +
+  geom_abline(slope = 1, intercept = 0, color="black", lwd=1) +
+  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class), alpha=0.2) +
+  facet_wrap(~which_params, nrow = 1) +
+  scale_fill_manual(values = c("#a41118", "#f4bb23"), breaks = c("Bad", "Good"), 
+                    aesthetics = c("color", "fill")) +
+  scale_x_continuous(breaks = c(0, 0.5, 1)) +
+  scale_y_continuous(breaks = c(0, 0.5, 1)) +
+  coord_fixed() +
+  labs(x="Falkor-trained model (predicted probability)", 
+       y="MESO-trained model\n(predicted probability)") +
+  theme_bw() +
+  theme(legend.position = "none",
+        strip.background = element_rect(fill="white"))
+rank_prob_space_gp <- dotplot_stepdown_data %>%
+  group_by(training_set, which_params) %>%
+  mutate(pred_prob=rank(pred_prob)) %>%
+  pivot_wider(names_from = training_set, values_from = pred_prob) %>%
+  mutate(which_params=factor(
+    which_params, 
+    levels=c("all_params", "xcms_params", "raw_data_params"),
+    labels=c("All metrics", "XCMS metrics", "Raw data metrics"))) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  ggplot() +
+  geom_abline(slope = 1, intercept = 0, color="black", lwd=1) +
+  geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class), alpha=0.2) +
+  facet_wrap(~which_params, nrow = 1) +
+  scale_fill_manual(values = c("#a41118", "#f4bb23"), breaks = c("Bad", "Good"), 
+                    aesthetics = c("color", "fill")) +
+  scale_x_continuous(breaks = c(0, nrow(MS3000_features)), 
+                     labels = c("Min rank", "Max rank")) +
+  scale_y_continuous(breaks = c(0, nrow(MS3000_features)),
+                     labels = c("Min\nrank", "Max\nrank")) +
+  labs(x="Falkor-trained model (ranked likelihood)", 
+       y="MESO-trained model\n(ranked likelihood)",
+       color=NULL) +
+  guides(color=guide_legend(override.aes = list(size=4, alpha=0.8))) +
+  coord_fixed() +
+  theme_bw() +
+  theme(axis.text.x = ggtext::element_markdown(hjust=c(0, 1)),
+        # strip.text = element_blank(),
+        strip.background = element_rect(fill="white"),
+        legend.position=c(0.875, 0.25),
+        legend.background = element_blank(),
+        legend.spacing.x = unit(0.003, units = "npc"))
+comb_prob_space_gp <- egg::ggarrange(raw_prob_space_gp, rank_prob_space_gp, draw = FALSE)
+ggsave(plot = comb_prob_space_gp, 
+       filename = "comb_prob_space.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 5, units = "in", dpi = 300)
+```
+
+![Figure 5: Predicted likelihood of a feature being "Good" according to a model trained on the MESOSCOPE dataset vs a model trained on the Falkor dataset.
+The top row of plots show the exact likelihood predicted by the logistic model across three different subsets of parameters, while the bottom row shows the estimates ranked from least likely to most likely.
+Points are colored by their manually-assigned quality according to an expert.](figures/comb_prob_space.png){width=90%}
+
+We also looked at the estimates produced by the two models and compared them with the combined model trained on the combined datasets to assess the model stability directly.
+A majority of the time, the estimates from the two models disagreed by more than 2x the standard error of the estimate and frequently disagreed by that amount depending on the other parameters included in the model.
+Some parameters disagreed not only in magnitude but also in sign, with the Falkor-trained full model increasing peak goodness likelihood with larger ppm variation and a wider peak width, while the MESOSCOPE-trained full model had negative estimates for each of these parameters.
+Notably, the peak shape and novel SNR parameters in the two-parameter model were among the most robust to training model variation, potentially explaining the consistency described above (Supp.
+figure X).
+
+```{r model stability dotplot errorbars supp figure}
+both_features <- rbind(FT2040_features, MS3000_features)
+dataset_versions_both <- c(dataset_versions, "both_features")
+model_param_estimates <- lst(all_params, xcms_params, raw_data_params) %>%
+  imap(function(param_selection, param_name){
+    lapply(dataset_versions_both, function(train_set){
+      model_params <- get(train_set) %>% 
+        filter(feat_class%in%c("Good", "Bad")) %>%
+        select(feat_class, all_of(param_selection)) %>%
+        mutate(feat_class=feat_class=="Good") %>%
+        glm(formula=feat_class~., family = binomial) %>%
+        broom::tidy() %>%
+        mutate(train_set)
+    }) %>% bind_rows() %>% mutate(param_name)
+  }) %>% bind_rows()
+suppfig_model_param_estimates_gp <- model_param_estimates %>%
+  mutate(param_name=str_remove(param_name, "_params")) %>%
+  ggplot(aes(x=param_name, color=train_set)) +
+  geom_hline(yintercept = 0) +
+  geom_point(aes(y=estimate), position = position_dodge(width=0.2)) +
+  geom_errorbar(aes(ymin=estimate-2*std.error, ymax=estimate+2*std.error),
+                position = position_dodge(width=0.2), width=0.3) +
+  facet_wrap(~term, scales="free_y", ncol = 4) +
+  theme_bw() +
+  theme(legend.position = "top", axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
+ggsave(plot = suppfig_model_param_estimates_gp, 
+       filename = "suppfig_model_param_estimates.png", 
+       device = "png", path = "manuscript/supplement",
+       width = 6.5, height = 9, units = "in", dpi = 300)
+```
+
+We also tested the robustness of the model under a smaller training set, emulating a situation in which only a fraction of the data was available or only a portion of the mass features had been labeled.
+This allowed us to test the required sample size for the different models, with a larger sample size presumably required for the models with more parameters.
+Because no parameter was present in all 3 models, we looked at the top 2 most significant parameters from each model: average *m/z* and peak shape for the full model, average peakwidth and the standard deviation in retention time for the XCMS model, and peak shape and SNR for the two-parameter model.
+For the XCMS and two-parameter model, we found reasonably good convergence in a dataset containing half the mass features with most values falling within two standard errors of the estimate, while the full model required closer to 80% of the mass features to produce estimates consistent with the original model (Supp.
+figure X).
+
+```{r subsampling model stability supp figure}
+set.seed(123)
+both_data <- rbind(MS3000_features, FT2040_features) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=feat_class=="Good")
+
+best_estimates <- model_param_estimates %>%
+  filter(train_set=="both_features") %>%
+  filter(term!="(Intercept)") %>%
+  group_by(param_name) %>%
+  arrange(p.value) %>%
+  slice(1:2)
+
+frac_data <- map(c(0.2, 0.5, 0.8), function(sample_fraction){
+  replicate(10, {
+    train_set <- slice_sample(both_data, prop = sample_fraction)
+    lst(all_params, xcms_params, raw_data_params) %>%
+      imap(function(param_selection, param_name){
+        train_set %>%
+          select(feat_class, all_of(param_selection)) %>%
+          glm(formula=feat_class~., family = binomial) %>%
+          broom::tidy() %>%
+          mutate(param_name)
+    }) %>% bind_rows()
+  }, simplify = FALSE) %>% bind_rows() %>%
+    mutate(sample_fraction)
+}) %>% bind_rows()
+
+sample_model_estimates <- best_estimates %>%
+  select(term, train_set, param_name) %>%
+  left_join(frac_data, multiple = "all")
+
+gplist <- c("all_params", "xcms_params", "raw_data_params") %>%
+  lapply(function(which_params){
+    ggplot() +
+      geom_hline(yintercept = 0) +
+      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error*2, 
+                    ymax=estimate+std.error*2),
+                fill="grey80", data=best_estimates %>% filter(param_name==which_params)) +
+      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error, 
+                    ymax=estimate+std.error),
+                fill="grey50", data=best_estimates %>% filter(param_name==which_params)) +
+      geom_point(aes(x=sample_fraction, y=estimate), 
+                 data=sample_model_estimates %>% filter(param_name==which_params),
+                 alpha=0.5) +
+      ggh4x::facet_grid2(param_name~term, scales = "free_y", independent = "y") +
+      scale_x_continuous(breaks = c(0, 0.2, 0.5, 0.8, 1), limits = c(0, 1)) +
+      theme_bw()
+  })
+sampled_model_est_gp <- do.call(what = egg::ggarrange, c(gplist, ncol=1))
+ggsave(plot = sampled_model_est_gp, 
+       filename = "sampled_model_est.png", 
+       device = "png", path = "manuscript/supplement",
+       width = 6.5, height = 5, units = "in", dpi = 300)
+```
+
+Given the reasonably compelling evidence above favoring the two-parameter model on novel datasets, we applied this trained model on two additional datasets that differed significantly from the training data.

### 2023-04-25 manuscript/outline.Rmd
Adding results outline and figure notes

@@
 
+Include the molecular mass in the feature label? If known, include name?
+
+Put the actual good/bad label in the figure somewhere
+
+What's the importance of the different colors?
+
 Note that these were taken from the MESOSCOPE dataset.
 
-Explain the chromatograms (RT/int/ what "goodness" requires?)
+Explain the chromatograms (RT/int/ what "goodness" requires?) explain x/y axis in figure legend, maybe do in bulk across all at once?
 
@@
 
+Good figure, spacing between text and labels and data
+
 ## Results
+
+Data description
+
+  - Counts of MFs in each manually-assigned category
+  - Proportion of good/bad in each dataset
+  - Supplemental figure: class colored histograms with each metric?
+  - Supplemental figure: class colored pairs plot?
+
+Logistic regressions
+
+  - Predictions are strongly binomial for all and xcms, less so for two-parameter model
+    - Discuss why we dropped the p-value metric?
+  - Performance within a given dataset vs across datasets and model selection
+    - Introduce FDR and GPF metrics
+      - Motivated by actual questions
+      - FDR = 1-precision (what fraction are likely false?)
+      - GPF = recall/sensitivity (what percent did I recover?)
+    - Very low FDR (5-10%) and high GPF (75-90%) *within* datasets for all metrics and xcms metrics, worse for raw data (~20% FDR, ~75% GPF)
+    - Higher FDR and GPF *across* datasets for all and xcms, about the same for raw data, leaves them all performing about the same(!)
+    - Figure 4
+    - Overfitting concerns (line slope) on cross-dataset
+    - Step-down methodology and independence from training set (Figure 5)
+  - Model stability
+    - How similar are the model parameters to each other when trained on Falkor/MESO/both?
+      - Supplemental figure: 3x errorbar plots for Falkor/MESO/both model estimates +/- SEs
+    - What fraction of the dataset did we actually need to label? (Answer: basically the whole thing, possibly even more)
+      - Supplemental figure: subset-based training with individual dots for min (maybe full?) model estimates added to full model estimates +/- SE bars
+  - Threshold selection and FDR in CultureData/Pttime (Figure 6)
+    - Higher threshold more representative of real research/applications in situations where 50% chance of being a peak is too high
+    - Additional external validation for FDR that still shows increase, albeit small (0-1% -> 2-5%)
+    - Unable to check GPF without fully labelling the datasets
+
+Other ML strategies
+
+  - Comparison to RF things
+  - Comparison to glmnet things
+
+Biological implications
+
+  - Clockplot and univariate statistics
+    - Report proportions of MFs in each category?
+    - Many "bad" still have strong biology trend, not a sufficient measure on its own
+    - Missing out on some good peaks (far left clockplot)
+    - Do the FDR corrected p-value comparisons with more/less peaks by threshold
+
+  - PERMANOVA table
+    - Better performance somehow with 90% threshold than manually labeled "good" peaks?
+    - Emphasize the lack of power when all peaks are included in the analysis with 1% and 0% thresholds
+  - Actual biology things
+    - Most metabolites track with biomass
+    - Surface-specific largely MAAs, some new?
+    - Degradation products or bacterial things accumulating at depth?
 
@@
 
+Make sure the break gap is consistent
+
+Maybe turn into two rows, one zoomed and one unzoomed?
+
+Differentiate the 0.9 line from the axis break line - don't like the vertical black line, looks like data
+
+Convert "meh" to "ambiguous"
+
 ### Model cross-validation with novel datasets
@@
 Include conclusions in figure legend? Something like "slope of the line is much greater in All metrics and XCMS metrics, indicating overfitting"?
+
+Is averaging the two points for the best fit line best? Lines sometimes have different slopes
+
+Is GPF worth using instead of more common jargon like 1-precision
+
+Maybe simpler metric
+
+Line feels a little heavy and too dark, put behind data, maybe make line grey instead of dark black
 
@@
 
-Make sure labels are the same as above - probably stick with XCMS for specificity.
+Make sure labels are the same as above - probably stick with XCMS for specificity. (Could just go in tables?)
 
 Remember to add Pearson and Spearman correlation coefficients in the corners!
+
+Maybe avoid "rank" phrasing? Instead of best/worst?
 
@@
 Explain why there are no values for GPF in Pttime/CultureData?
+
+Render as a table and see how that looks instead?
+
+Include Pttime definition in the figure caption?
+
+Maybe include the facet captions as the y-axis label instead? Anitra doesn't love how it looks as-is
 
@@
 
+Switch the shapes to make empty circles "good peaks missed"
+
+Add some internal circle to help render information about the overplotting in the center? Give a bar graph breakdown for the  things within the circle?
+
 ```{r manual check on clockplot, eval=FALSE}
@@
 
+Biological circumstances to test:
+1. Using ALL peaks produced by XCMS
+2. Using only the manually-categorized "Good" peaks
+3. Using the model-categorized "Good" peaks at threshold 0.99
+4. Using the model-categorized "Good" peaks at threshold 0.5
+5. Using the model-categorized "Good" peaks at threshold 0.01
+
+```{r}
+library(vegan)
+MS3000_metadata <- file_data %>%
+  filter(samp_type=="Smp") %>%
+  filter(str_detect(filename, "180821")) %>%
+  select(filename, depth) %>%
+  column_to_rownames("filename")
+norm_area_data <- read_csv("manuscript/MS3000_norm_areas.csv")
+
+both_min_model <- rbind(FT2040_features, MS3000_features) %>%
+  select(feat_class, med_cor, med_SNR) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=feat_class=="Good") %>%
+  glm(formula=feat_class~., family = binomial)
+pred_probs <- MS3000_features %>%
+  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
+  select(feature, feat_class, pred_prob)
+
+wide_data <- norm_area_data %>%
+  group_by(feature) %>%
+  filter(var(norm_area)>0) %>%
+  # mutate(norm_area=rank(norm_area)) %>%
+  mutate(norm_area=scale(norm_area)[,1]) %>%
+  left_join(pred_probs) %>% 
+  pivot_wider(names_from = "filename", values_from = norm_area) %>%
+  column_to_rownames("feature")
+
+mat_data <- wide_data %>%
+  # filter(feat_class=="Good") %>%
+  # filter(pred_prob>0.9) %>% 
+  # filter(pred_prob>0.5) %>%
+  # filter(pred_prob>0.1) %>%
+  select(-feat_class, -pred_prob) %>%
+  data.matrix() %>%
+  t()
+
+adonis2(mat_data~depth, data = MS3000_metadata, method = "manhattan")
+nmds_output <- metaMDS(mat_data, k = 2, distance = "manhattan", trace = 0, 
+                       autotransform = FALSE, noshare = FALSE, wascores = FALSE)
+nmds_output$points %>%
+  as.data.frame() %>%
+  cbind(MS3000_metadata) %>%
+  rownames_to_column("filename") %>%
+  select(filename, starts_with("MDS"), depth) %>%
+  mutate(basename=str_extract(filename, "MS\\d+C\\d+.*(?=_[A-C])")) %>%
+  ggplot(aes(x=MDS1, y=MDS2)) +
+  geom_polygon(aes(fill=depth, color=depth, group=basename), alpha=0.5) +
+  annotate(geom = "label", x=Inf, y=Inf, 
+           label=paste("Stress =", round(nmds_output$stress, 2)),
+           hjust=1, vjust=1, label.r=unit(0, "in")) +
+  theme_minimal()
+```
+
+```{r}
+mat_all <- wide_data
+mat_good <- wide_data %>% filter(feat_class=="Good")
+mat_90 <- wide_data %>% filter(pred_prob>0.9)
+mat_50 <- wide_data %>% filter(pred_prob>0.5)
+mat_10 <- wide_data %>% filter(pred_prob>0.1)
+mat_01 <- wide_data %>% filter(pred_prob>0.01)
+
+mat_list <- lst(mat_good, mat_90, mat_50, mat_10, mat_01, mat_all)
+thresh_tests <- lapply(mat_list, function(mat_given){
+  mat_data <- mat_given %>%
+    select(-feat_class, -pred_prob) %>%
+    data.matrix() %>%
+    t()
+  perm_output <- adonis2(mat_data~depth, data = MS3000_metadata, method = "manhattan")
+  nmds_output <- metaMDS(mat_data, k = 2, distance = "manhattan", trace = 0,
+                         autotransform = FALSE, noshare = FALSE, wascores = FALSE)
+  nmds_gp <- nmds_output$points %>%
+    as.data.frame() %>%
+    cbind(MS3000_metadata) %>%
+    rownames_to_column("filename") %>%
+    select(filename, starts_with("MDS"), depth) %>%
+    mutate(basename=str_extract(filename, "MS\\d+C\\d+.*(?=_[A-C])")) %>%
+    ggplot(aes(x=MDS1, y=MDS2)) +
+    geom_polygon(aes(fill=depth, color=depth, group=basename), alpha=0.5) +
+    annotate(geom = "label", x=Inf, y=Inf, 
+             label=paste("Stress =", round(nmds_output$stress, 2)),
+             hjust=1, vjust=1, label.r=unit(0, "in")) +
+    theme_minimal()
+  list(perm_output, nmds_output, nmds_gp)
+}) %>% set_names(names(mat_list))
+
+imap(thresh_tests, function(thresh_i, thresh_name){
+  as.data.frame(thresh_i[[1]][1,]) %>% mutate(thresh_name)
+}) %>% bind_rows() %>%
+  select(thresh_name, R2, `F`, `Pr(>F)`) %>%
+  `rownames<-`(NULL) %>%
+  column_to_rownames("thresh_name") %>%
+  knitr::kable()
+
+lapply(thresh_tests, `[[`, 3) %>%
+  do.call(what=gridExtra::grid.arrange)
+```
+
+See which/why peaks differ between predicted and actual quality to start explaining why 90% threshold performs better than the "Good" cutoff
+
 ## Discussion
 
-TBD
+  - Fundamental tradeoff between GPF and FDR should be determined by the scientific question
+  - FDR and GPF rates are very rarely reported but critically important to include
+  - Full model and XCMS model are definitely overfit
+    - Steep performance drop on new data
+    - Disagreement based on training set
+  - Only tested on HILIC data, expectations for RP columns
+  - Using medians/means as summary statistics ignores differences between treatments (i.e.
+lots of blanks blows up the median): maybe better to use max/top3?
+Maybe better to implement independently for each sample class?
+  - XCMS alone does not provide enough information to distinguish good peaks from bad with much confidence
 
@@
 ## Unstructured thoughts
+
+Journal target: BMC Bioinformatics
+Alternative: PLoS Comp Bio (Too high-impact? Not as many MS papers)
+Alternative: MDPI? That's where 255,000 was published
+Alternative II: Bioinformatics (very bio-focused)
+
 Pleas:

### 2023-04-29 manuscript/manuscript.Rmd
Added Fig 6 and some initial results description

@@
 Given the reasonably compelling evidence above favoring the two-parameter model on novel datasets, we applied this trained model on two additional datasets that differed significantly from the training data.
+The CultureData samples were produced in the Ingalls lab like the MESOSCOPE and Falkor samples, but represent data from a variety of phytoplankton and bacterial cultures in fresh and salt water rather than environmental samples.
+The Pttime dataset was discovered on Metabolomics Workbench [@MW id] and consists of Phaeodactylum triconutum cultures collected at a variety of timepoints from both pelleted cells and the released exudate.
+This dataset was chosen because of the similar LC-MS setup used as a benchmark comparison for the performance other labs with similar setups may expect to achieve.
+Each of these datasets were only fractionally labeled, with those peaks above the 0.9 likelihood threshold according to the two-parameter model reviewed manually and categorized.
+
+This stricter threshold was used
+
+```{r FDR vs GPF with additional datasets}
+# Currently a fairly simple figure, may need additional data?
+both_min_model <- rbind(FT2040_features, MS3000_features) %>%
+  select(feat_class, med_cor, med_SNR) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=feat_class=="Good") %>%
+  glm(formula=feat_class~., family = binomial)
+pred_probs_all <- bind_rows(lst(
+  Falkor=FT2040_features, MESOSCOPE=MS3000_features,
+  CultureData=CultureData_features, Pttime=Pttime_features), 
+  .id = "dataset") %>%
+  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
+  mutate(pred_class=ifelse(pred_prob>0.9, "Good", "Bad")) %>%
+  mutate(error_type=case_when(
+    feat_class=="Good" & pred_class=="Good" ~ "%TP",
+    feat_class=="Good" & pred_class=="Bad" ~ "%FN",
+    feat_class=="Bad" & pred_class=="Good" ~ "%FP",
+    feat_class=="Bad" & pred_class=="Bad" ~ "%TN",
+    TRUE ~ "Incomparable"
+  )) %>%
+  mutate(dataset=factor(dataset, levels=c("MESOSCOPE", "Falkor", "CultureData", "Pttime"))) %>%
+  filter(error_type!="Incomparable") %>%
+  group_by(dataset, error_type) %>%
+  summarise(n=n(), .groups = 'drop') %>%
+  complete(dataset, error_type, fill=list(n=0))
+error_type_all_datasets <- pred_probs_all %>%
+  pivot_wider(names_from=error_type, values_from = n) %>%
+  mutate(`FDR`=`%FP`/(`%FP`+`%TP`)) %>%
+  mutate(`GPF`=`%TP`/(`%FN`+`%TP`)) %>%
+  mutate(`GPF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `GPF`)) %>%
+  pivot_longer(c(`FDR`, `GPF`)) %>%
+  drop_na()
+
+fdr_gpf_all_datasets_gp <- error_type_all_datasets %>%
+    mutate(desc=case_when(
+    is.na(value) ~ NA,
+    name=="FDR"~ paste0("frac(",`%FP`,",",`%FP`+`%TP`,")"),
+    name=="GPF"~ paste0("frac(",`%TP`,",",`%FN`+`%TP`,")")
+  )) %>%
+  mutate(vjust_custom=case_when(
+    is.na(value) ~ NA,
+    name=="FDR" & dataset=="Pttime" ~ 1.3,
+    name=="FDR" ~ -0.3,
+    name=="GPF" ~ 1.3,
+  )) %>%
+  mutate(name=factor(name, levels=c("GPF", "FDR"),
+                     labels=c("Fraction of good peaks found (GPF)", 
+                              "False discovery rate (FDR)"))) %>%
+  mutate(dataset=factor(dataset, 
+                        levels = c("Falkor", "MESOSCOPE", "CultureData", "Pttime"),
+                        labels= c("Falkor", "MESO-\nSCOPE", "Culture\nData", "Pttime"))) %>%
+  ggplot(aes(x=dataset, y=value)) +
+  geom_point(size=3) +
+  geom_text(aes(label=desc, vjust=vjust_custom), parse=TRUE) +
+  facet_wrap(~name, scales = "free_y", ncol = 1) +
+  scale_y_continuous(limits = c(0, NA), labels = scales::label_percent()) +
+  theme_bw() +
+  theme(axis.title = element_blank(),
+        strip.background = element_rect(fill="white"))
+
+ggsave(plot = fdr_gpf_all_datasets_gp, 
+       filename = "fdr_gpf_all_datasets.png", 
+       device = "png", path = "manuscript/figures",
+       width = 3, height = 4, units = "in", dpi = 300)
+
+# Confusion matrices for confirmation
+# pred_probs_all
+# pred_probs_all %>%
+#   ggplot(aes(x=error_type, y=n)) +
+#   geom_col(color="#0b505c", fill="#0b505c") +
+#   geom_text(aes(label=n), vjust=-0.2, color="#0b505c") +
+#   facet_wrap(~dataset, nrow=1) +
+#   theme_bw() +
+#   theme(axis.title = element_blank())
+```
+
+![Figure 6: False discovery rate and proportion of total good peaks identified as good by the two-parameter model trained on the combined MESOSCOPE/Falkor dataset and applied to each dataset individually with a threshold of 0.9.
+FDR is calculated by dividing the number of false positives by the total positives produced by the model and GPF is calculated by dividing the number of true positives by the total number of good peaks as identified manually.
+Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fdr_gpf_all_datasets.png){width=30%}

### 2023-05-02 manuscript/manuscript.Rmd
Adding regularized regression discussion and clockplot code

@@
 
-All analyses were run in R, version 4.2.2.
-Logistic regressions were performed using base R's `glm` function with elastic net functionality provided by the glmnet package, random forest classifiers were performed using the randomForest package, and initial clustering approaches were explored with the dbscan package.
+We also tested whether the performance could be improved with elastic net regression or random forest models.
+These models handle correlated variables better than ordinary least squares regression, so we also included several additional implementations of the peak shape and novel SNR parameters when summarizing across multiple files, using a max and a median of the top-three best values rather than just the overall median as well as a log-transformed version of the median peak shape calculated as $median(log_{10}(1-r))$ where $r$ is Pearson's correlation coefficient, as described above (Figure 2).
+Cross-validation was used to select the optimal tuning parameter $\lambda$ with `glmnet` package's `cv.glmnet` for an elastic net penalty (α) of 0, 0.5, and 1.
+Random forests were implemented using the randomForest package with default settings and a factor-type response vector to ensure classification was used rather than regression.
+
+We explored the implications of applying this model to the MESOSCOPE dataset at a variety of thresholds.
+In univariate space, we used nonparametric Kruskal-Wallis analyses of variance to measure the difference between the surface (15m), DCM (~110m), and 175m samples because the metabolite peak areas could not be assumed to be normally distributed.
+These univariate tests were then controlled for multiply hypothesis testing using R's `p.adjust` function with method `fdr`.[@ref] We also performed post-hoc Dunn tests provided by the `rstatix` package to categorize the response to depth for those mass features for which the KW test was significant, with responses falling into one of the 14 classes possible when permuting the sign and significance of the Dunn test.[@ref] p-values obtained from the Dunn tests were not FDR controlled because it was used as a categorization tool rather than a null hypothesis test.
+In multivariate space, we used a permutational MANOVA (PERMANOVA) [@ref] provided by the vegan package's `adonis2` function to test for multivariate differences in structure of the metabolome with depth.[@ref] We ran multiple PERMANOVAs with a different subset of mass features included each time, corresponding to using the output from XCMS directly, likelihood thresholds of 0.01, 0.1, 0.5, 0.9, and finally only those peaks manually annotated as good.
+
+All analyses were run in R, version 4.2.2, and code is available on the Github at https://github.com/wkumler/MS_metrics.[@ref]
 
@@
 
+### Dataset characterization
+
 An average of 3,300 molecular features (MFs) were reported by XCMS across the 4 datasets, with the fewest (1495) in the Falkor data and the most in the Pttime samples (7781).
@@
 In both, 5% of the MFs were unable to be assigned confidently to either "Good" or "Bad" classes and 10% were identified as appearing only in the standards, leaving only ~15% of the features classified as "Good" (16% and 12%, respectively).
+
+### Logistic regression performance
 
@@
 
+### Model stability under different training sets and subset of labels
+
 To further explore questions of model stability and the potential for overfitting, we compared the predictions from a Falkor-trained model to a MESOSCOPE-trained model.
@@
   })
-sampled_model_est_gp <- do.call(what = egg::ggarrange, c(gplist, ncol=1))
+sampled_model_est_gp <- do.call(what = egg::ggarrange, c(gplist, ncol=1, draw=FALSE))
 ggsave(plot = sampled_model_est_gp, 
@@
 ```
+
+### Performance of a stricter threshold on novel datasets
 
@@
 
-This stricter threshold was used
+This stricter threshold was chosen because we felt less comfortable interpreting results based on mass features that were only 50% likely to be real, but did not feel the need to be so strict with this exploratory analysis that we wanted to limit it to 99+% likelihood peaks.
+This threshold also struck a nice balance between the number of peaks we estimated to be necessary for robust testing while still remaining reasonable to manually label.
+For the CultureData dataset, we obtained 1,790 total mass features, 192 of which had predicted likelihoods above 0.9.
+Of these, 151 were identified manually as "Good", 21 were given "Ambiguous" designations, and only 3 were flagged as "Bad", with the remaining 17 appearing only in the standards.
+For the Pttime dataset, 7,781 were obtained with 400 flagged by the model as "Good".
+348 were truly good peaks, 35 were ambiguous, and 17 were "Bad".
+No standards were run during this analysis, so there were no features in that category.
+[There exists, however, some uncontrolled confirmation bias here in the manual labelling because the MS expert knew that these peaks had already passed some quality threshold and were more likely to be good than not.]
+
+With the stricter threshold, we obtained FDR rates consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for the Falkor, MESOSCOPE, CultureData, and Pttime respectively [Figure 6].
+Of course, this low error rate required that we miss out on additional valuable features, with only a fraction of the total good peaks making it past this threshold.
+In both the Falkor and MESOSCOPE datasets, the fraction of good peaks found was below 40% with actual values of 39.4% and 26.5%, respectively.
+Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the %GPF but expect it to be in a similar range (Figure 6).
 
@@
   mutate(pred_class=ifelse(pred_prob>0.9, "Good", "Bad")) %>%
+  # count(pred_class, dataset)
   mutate(error_type=case_when(
@@
   )) %>%
+  # count(feat_class, pred_class, dataset) %>% filter(dataset=="Pttime")
   mutate(dataset=factor(dataset, levels=c("MESOSCOPE", "Falkor", "CultureData", "Pttime"))) %>%
@@
 Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fdr_gpf_all_datasets.png){width=30%}
+
+### Regularized regression and random forests perform about the same
+
+None of the advanced ML models significantly improved performance when cross-validated between the MESOSCOPE and Falkor datasets when measured by both initial performance and the performance drop when applied across datasets.
+All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GPF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent Type II error acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp.
+figure X).
+The random forest model had perfect predictive capacity when tested internally on the training data (FDR=0%, GPF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp.
+figure X).
+In each case, the performance drop when applied to a novel dataset was more extreme than the simple two-parameter model described above.
+
+```{r elastic net and random forest}
+library(glmnet)
+net_cross_train_df <- lst(ridge=0, elastic=0.5, lasso=1) %>%
+  imap(function(alpha_selection, alpha_name){
+    lapply(dataset_versions, function(train_set){
+      net_x <- get(train_set) %>% 
+        filter(feat_class%in%c("Good", "Bad")) %>%
+        select(-feat_class, -feature) %>%
+        data.matrix()
+      net_y <- get(train_set) %>% 
+        filter(feat_class%in%c("Good", "Bad")) %>%
+        mutate(feat_class=feat_class=="Good") %>%
+        pull(feat_class)
+      net_model <- cv.glmnet(x=net_x, y=net_y, family="binomial", alpha=alpha_selection)
+
+      lapply(dataset_versions, function(test_set){
+        get(test_set) %>%
+          select(-feat_class, -feature) %>%
+          mutate(pred_prob=predict(object=net_model, newx = data.matrix(.), 
+                                   type = "response")[,"lambda.1se"]) %>%
+          mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
+          mutate(feat_class=get(test_set)$feat_class) %>%
+          select(feat_class, pred_class) %>%
+          mutate(train_test=paste(train_set, test_set, sep = "-")) %>%
+          mutate(train_test=str_remove_all(train_test, "_features")) %>%
+          mutate(cross_type=ifelse(train_set==test_set, "Same", "Different"))
+      }) %>% bind_rows()
+    }) %>% bind_rows() %>% mutate(which_alpha=alpha_name)
+  }) %>% bind_rows()
+library(randomForest)
+rf_cross_train_df <- lapply(dataset_versions, function(train_set){
+  rf_x <- get(train_set) %>% 
+    select(-medtop3_cor, -max_cor, -log_med_cor, -max_SNR) %>%
+    select(med_SNR, med_cor, feat_class, feature) %>%
+    filter(feat_class%in%c("Good", "Bad")) %>%
+    select(-feat_class, -feature)
+  rf_y <- get(train_set) %>% 
+    filter(feat_class%in%c("Good", "Bad")) %>%
+    pull(feat_class) %>%
+    factor()
+  rfmodel <- randomForest(rf_x, rf_y)
+
+  lapply(dataset_versions, function(test_set){
+    get(test_set) %>%
+      select(-feat_class, -feature) %>%
+      mutate(pred_class=predict(object=rfmodel, newdata = ., type = "response")) %>%
+      mutate(feat_class=get(test_set)$feat_class) %>%
+      select(feat_class, pred_class) %>%
+      mutate(train_test=paste(train_set, test_set, sep = "-")) %>%
+      mutate(train_test=str_remove_all(train_test, "_features")) %>%
+      mutate(cross_type=ifelse(train_set==test_set, "Same", "Different"))
+      }) %>% bind_rows()
+    }) %>% bind_rows() %>%
+  mutate(which_alpha="rf")
+
+fdr_gpf_cross_train_df <- net_cross_train_df %>%
+  rbind(rf_cross_train_df) %>%
+    mutate(error_type=case_when(
+    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
+    feat_class=="Bad" & pred_class=="Good" ~ "FP",
+    feat_class=="Good" & pred_class=="Bad" ~ "FN",
+    feat_class=="Good" & pred_class=="Good" ~ "TP"
+  )) %>%
+  drop_na() %>%
+  group_by(train_test, which_alpha, cross_type) %>%
+  count(error_type) %>%
+  ungroup() %>%
+  mutate(cross_type=factor(cross_type, levels=c("Same", "Different"))) %>%
+  pivot_wider(names_from=error_type, values_from = n, values_fill = 0) %>%
+  mutate(FDR=FP/(FP+TP)) %>%
+  mutate(GPF=TP/(FN+TP)) %>%
+  mutate(cross_type_num=as.numeric(cross_type))
+train_test_levels <- c("MS3000-MS3000", "FT2040-FT2040", "MS3000-FT2040", "FT2040-MS3000")
+train_test_labels <- c("MESO-MESO", "Falkor-Falkor", "MESO-Falkor", "Falkor-MESO")
+fdr_gpf_cross_train_gp <- fdr_gpf_cross_train_df %>%
+  mutate(train_test=factor(train_test, levels=train_test_levels, labels=train_test_labels)) %>%
+  mutate(which_alpha=factor(which_alpha, levels=c("ridge", "elastic", "lasso", "rf"),
+                            labels=c("Ridge: α=0", "Elastic: α=0.5", "Lasso: α=1", "Random forest"))) %>%
+  pivot_longer(c(FDR, GPF)) %>%
+  ggplot() +
+  geom_smooth(aes(x=cross_type_num, y=value, group=name), 
+              method = "lm", formula="y~x", color="black", se=FALSE) +
+  geom_point(aes(x=cross_type, y=value, color=train_test, 
+                 fill=train_test, shape=name), size=5, alpha=0.8) +
+  facet_grid(~which_alpha) +
+  scale_x_discrete(labels = c("Same\ndataset\n", "Different\ndataset\n"),
+                   breaks = c("Same", "Different")) +
+  scale_color_manual(values = c("#0b505c", "#184d1f", "#0d8299", "#028e34"), 
+                     breaks = train_test_labels,
+                     aesthetics = c("color", "fill")) +
+  scale_shape_manual(breaks = c("GPF", "FDR"), values = c("\u25B2", "\u25BC")) +
+  coord_cartesian(ylim = c(0, 1)) +
+  scale_y_continuous(labels = scales::label_percent()) +
+  theme_bw() +
+  theme() +
+  labs(x=NULL, y=NULL, color="Train-test", fill="Train-test", shape="Error type") +
+  guides(shape=guide_legend(order=1))
+
+ggsave(plot = fdr_gpf_cross_train_gp, 
+       filename = "regregrf_fdrgpf_cross_train.png", 
+       device = "png", path = "manuscript/supplement",
+       width = 6.5, height = 3, units = "in", dpi = 300)
+```
+
+### Implications for biological conclusions
+
+[Figure 7] shows the distribution of all mass features identified by XCMS and their categorization according to both the biochemical response to depth as well as the classifications assigned by the MS expert and the two-parameter model at a 50% likelihood threshold.
+
+The abundance of noise peaks located at the center of the plot and therefore lacking much biological trend with depth validates the conventional heuristic of relying on downstream statistical tests to separate signal from noise (with the assumption that noise will be fixed across all samples and unaffected by environmental structuring of samples).
+However, it is also clear that there are both some good peaks that do not have trends with depth (and are therefore interesting for bucking the biomass signal) as well as a good number of bad peaks that do change across sample depths.
+These bad peaks that have biological signals appear to be frequently due to the tail of a peak being integrated as a separate molecular feature, which then retains the biological signal of the actual feature.
+
+Most metabolites track with biomass and are consistently abundant through the surface and DCM but drop off at depth, as prior measurements of particulate carbon and beam attenuation have shown [@benedetto].
+However, several groups of compounds diverge intriguingly from this bulk biomass trend, with some molecules found almost exclusively at the surface and others found to increase with depth rather than decrease.
+Several mass features that were found most exclusively at the surface were putatively identified as mycosporine-like amino acids (MAAs), which have been shown to have UV photoprotective properties.[@ref] Although we were unable to calculate their exact concentrations due to a lack of authentic standards, peak areas were approximately 100 to 1000 times higher at 15 meters than their values averaged between the DCM and 175 meter samples.
+The abundance of MAAs in this particular category indicates that there may be other MAA molecules with this depth pattern that have not yet been characterized also among these unknowns.
+
+```{r clockplot BMIS, eval=!file.exists("manuscript/MS3000_norm_areas.csv")}
+library(xcms)
+file_data <- read_csv("made_data_MS3000/file_data.csv") %>%
+  mutate(filename=basename(filename))
+msnexp_filled <- readRDS("made_data_MS3000/msnexp_filled.rds")
+peak_data_long <- msnexp_filled %>%
+  chromPeaks() %>%
+  as.data.frame() %>%
+  rownames_to_column() %>%
+  mutate(peakidx=row_number()) %>%
+  select(sample, peakidx, into)
+peak_data <- msnexp_filled %>%
+  featureDefinitions() %>%
+  as.data.frame() %>%
+  filter(rtmed>30 & rtmed<1200) %>%
+  select(peakidx) %>%
+  rownames_to_column("feature") %>%
+  unnest_longer(peakidx) %>%
+  left_join(peak_data_long, by = join_by(peakidx)) %>%
+  mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
+  complete(feature, filename, fill = list(into=0)) %>%
+  left_join(file_data, by = join_by(filename)) %>%
+  filter(str_detect(filename, "180821"))
+
+source("manuscript/IS_integrations/is_combine.R")
+IS_areas <- read_csv("manuscript/IS_integrations/all_IS.csv") %>%
+  mutate(filename=paste0(filename, ".mzML")) %>%
+  rename(area="IS_area") %>%
+  filter(!str_detect(filename, "QC")) %>%
+  pivot_wider(names_from=IS_name, values_from = "IS_area") %>%
+  mutate("Inj_vol"=ifelse(str_detect(filename, "Poo.*Half"), 0.5, 1)) %>%
+  pivot_longer(-filename, names_to = "IS_name", values_to = "IS_area")
+pooled_IS_areas <- IS_areas %>%
+  filter(str_detect(filename, "_Poo_"))
+
+BMIS <- peak_data %>%
+  filter(str_detect(filename, "_Poo_")) %>%
+  full_join(pooled_IS_areas, multiple = "all", by = join_by(filename)) %>%
+  group_by(feature, IS_name) %>%
+  mutate(plot_area=into/IS_area*mean(IS_area)) %>%
+  summarise(cv=sd(plot_area)/mean(plot_area)) %>%
+  # left_join(init_cvs) %>%
+  arrange(cv) %>%
+  slice(1) %>%
+  select(feature, IS_name)
+
+BMISed_data <- peak_data %>%
+  filter(samp_type=="Smp") %>%
+  select(feature, filename, into) %>%
+  left_join(BMIS, by = join_by(feature)) %>%
+  left_join(IS_areas, by = join_by(filename, IS_name)) %>%
+  group_by(feature) %>%
+  mutate(norm_area=into/IS_area*mean(IS_area)) %>%
+  select(feature, filename, norm_area)
+
+norm_area_data <- BMISed_data %>%
+  arrange(norm_area) %>%
+  group_by(feature, filename) %>%
+  slice(1) %>%
+  ungroup()
+
+if(any(is.na(norm_area_data$norm_area)))stop("Found NAs in norm_area!")
+if(!all(table(norm_area_data$feature)==99))stop("Found not 99 samps in plotready!")
+write_csv(norm_area_data, "manuscript/MS3000_norm_areas.csv")
+```
+
+```{r clockplot dataset munging, eval=!file.exists("manuscript/MS3000_clockplot_coords.csv")}
+norm_area_data <- read_csv("manuscript/MS3000_norm_areas.csv")
+
+both_min_model <- rbind(FT2040_features, MS3000_features) %>%
+  select(feat_class, med_cor, med_SNR) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=feat_class=="Good") %>%
+  glm(formula=feat_class~., family = binomial)
+best_feats <- MS3000_features %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
+  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
+  select(feature, feat_class, pred_class)
+clockplot_data <- norm_area_data %>%
+  right_join(best_feats, by = join_by(feature)) %>%
+  left_join(file_data %>% select(filename, depth), by = join_by(filename))
+
+kw_pvals_all <- clockplot_data %>%
+  nest(data=-c(feature, pred_class)) %>%
+  mutate(tidyaov=map(
+    data, ~ broom::tidy(kruskal.test(.x$norm_area~.x$depth))), 
+  ) %>%
+  unnest(tidyaov) %>%
+  select(-parameter, -method) %>%
+  mutate(cor_pval_all=p.adjust(p.value, method="fdr")) %>%
+  group_by(pred_class) %>%
+  mutate(fdr_pval_byclass=p.adjust(p.value, method="fdr")) %>%
+  ungroup()
+
+resp_class_levels <- c("All equal", 
+    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
+    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
+    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
+    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
+    "Hard to tell")
+kw_classes_all <- clockplot_data %>%
+  group_by(feature) %>%
+  mutate(norm_area=rank(norm_area)) %>%
+  filter(var(norm_area)>0) %>%
+  rstatix::dunn_test(norm_area~depth, p.adjust.method = "none") %>%
+  mutate(contrast=paste(group2, group1, sep = "_")) %>%
+  select(feature, contrast, est=statistic, pval=p.adj) %>%
+  pivot_wider(names_from = contrast, values_from = c(pval, est)) %>%
+  mutate(resp_class=case_when(
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m<0 & est_175m_DCM<0 ~ "15m > DCM > 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m<0 & est_175m_DCM<0 ~ "DCM > 15m > 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m>0 & est_175m_DCM>0 ~ "DCM < 15m < 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m>0 & est_175m_DCM>0 ~ "15m < DCM < 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m>0 & est_175m_DCM<0 ~ "DCM > 175m > 15m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m<0 & est_175m_DCM>0 ~ "DCM < 175m < 15m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM>0.05 & est_DCM_15m<0 & est_175m_15m<0 ~ "15m > DCM = 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM>0.05 & est_DCM_15m>0 & est_175m_15m>0 ~ "15m < DCM = 175m",
+    pval_DCM_15m>0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_175m_15m<0 & est_175m_DCM<0 ~ "15m = DCM > 175m",
+    pval_DCM_15m>0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_175m_15m>0 & est_175m_DCM>0 ~ "15m = DCM < 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m>0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_DCM<0 ~ "DCM > 15m = 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m>0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_DCM>0 ~ "DCM < 15m = 175m",
+    pval_DCM_15m>0.05 & pval_175m_15m>0.05 & pval_175m_DCM>0.05 ~ "All equal",
+    TRUE ~ "Hard to tell"
+  )) %>%
+  select(feature, resp_class) %>%
+  mutate(resp_class=factor(resp_class, levels=resp_class_levels))
+
+clockplot_xy_spots <- clockplot_data %>%
+  group_by(feature) %>%
+  mutate(norm_area=rank(norm_area)) %>%
+  group_by(feature, depth) %>%
+  summarise(mean_fold_change=mean(norm_area/50), .groups = "drop") %>% 
+  # div by 50 bc 99 files and median 1:99 is 50
+  pivot_wider(names_from=depth, values_from=mean_fold_change) %>%
+  mutate(x=DCM-1) %>%           # cursed but works??
+  mutate(y=`15m`+DCM/2-1.5) %>% # super fucking cursed
+  select(feature, x, y)
+
+clockplot_complete_df <- clockplot_data %>%
+  distinct(feature, feat_class, pred_class) %>%
+  left_join(clockplot_xy_spots, by = join_by(feature)) %>%
+  left_join(kw_pvals_all %>% select(feature, fdr_pval_byclass), by = join_by(feature)) %>%
+  left_join(kw_classes_all, by = join_by(feature)) %>%
+  mutate(error_type=case_when(
+    pred_class=="Good" & feat_class=="Good" ~ "TP",
+    pred_class=="Good" & feat_class=="Bad" ~ "FP",
+    pred_class=="Bad" & feat_class=="Good" ~ "FN",
+    pred_class=="Bad" & feat_class=="Bad" ~ "TN"
+  )) %>%
+  mutate(error_type=factor(error_type, levels=c("TP", "FP", "FN", "TN"),
+                           labels=c("Good peak found", "Bad peak included",
+                                    "Good peak missed", "Bad peak avoided")))
+
+write_csv(clockplot_complete_df, file = "manuscript/MS3000_clockplot_coords.csv")
+```
+
+```{r clockplot}
+resp_class_levels <- c("All equal", 
+    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
+    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
+    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
+    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
+    "Hard to tell")
+clockplot_complete_df <- read_csv(file = "manuscript/MS3000_clockplot_coords.csv") %>%
+  mutate(resp_class=factor(resp_class, levels=resp_class_levels))
+
+cmap_name <- "infinity"
+cmap_url <- paste0("https://raw.githubusercontent.com/1313e/CMasher/master/",
+                   "cmasher/colormaps/", cmap_name, "/", cmap_name, "_norm.txt")
+cmtable <- read.table(cmap_url)
+cmscale <- rgb(cmtable[,1], cmtable[,2], cmtable[,3])
+color_scale <- c("grey50", head(cmscale[seq(1, length(cmscale), length.out=13)], 12), "grey80")
+names(color_scale) <- resp_class_levels
+
+clock_scale <- 0.66
+axis_xends <- c(-1/2, 1.1, -1/2)*clock_scale
+axis_yends <- c(sqrt(3)/2, 0, -sqrt(3)/2)*clock_scale
+axis_y_dodge <- c(2, 2, -2)*clock_scale/25
+
+clockplot_base_gp <- ggplot() +
+  annotate("segment", x=c(0,0,0), y=c(0,0,0), 
+           xend=axis_xends, yend=axis_yends,
+           arrow=arrow(length=unit(0.02, "npc"), type = "closed")) +
+  annotate("segment", x=c(0,0,0), y=c(0,0,0),
+           xend=-axis_xends, yend=-axis_yends,
+           lty = 2) +
+  annotate("label", x=axis_xends, y=axis_yends+axis_y_dodge,
+           label=c("15m axis", "DCM axis", "175m axis"), 
+           fill="#FFFFFFAA", label.size=0) +
+  coord_fixed(xlim = c(-0.65, 1.3)*clock_scale, 
+              ylim = c(-1, 1)*clock_scale, 
+              expand=TRUE) +
+  scale_color_manual(values = color_scale, breaks = names(color_scale), drop=FALSE) +
+  theme_void() +
+  theme(legend.position = "right", 
+        legend.title = element_blank(),
+        plot.background = element_rect(fill="white", color="white")) +
+  guides(colour = guide_legend(override.aes = list(size=4, alpha=1), drop=FALSE)) +
+  aes(x=x, y=y, color=resp_class, label=feature)
+
+clockplot_complete_gp <- clockplot_base_gp +
+  geom_point(aes(shape=error_type), alpha=0.5, size=2,
+             data = clockplot_complete_df) +
+  scale_shape_manual(breaks = c("Good peak found", "Bad peak included",
+                                "Good peak missed", "Bad peak avoided"), 
+                     values = c(19, 3, 4, 1))
+
+ggsave(plot = clockplot_complete_gp, 
+       filename = "clockplot_complete.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 5, units = "in", dpi = 300)
+```
+
+![Figure 7: Plot of metabolite response to depth shown across three axes using the rank-normalized median value at each depth as the coordinate for that axis.
+Each molecular feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile.
+Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the deep chlorophyll maximum; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}

### 2023-05-02 manuscript/outline.Rmd
Adding regularized regression discussion and clockplot code

@@
     - Report proportions of MFs in each category?
+    - Big central thing where all the noise ended up
     - Many "bad" still have strong biology trend, not a sufficient measure on its own

### 2023-05-02 manuscript/BMIS_peakprep.R
Moving BMIS into manuscript directly

@@ -1,90 +1,0 @@
-
-
-# Setup ----
-library(tidyverse)
-library(xcms)
-library(data.table)
-file_data <- read_csv("made_data_MS3000/file_data.csv") %>%
-  mutate(filename=basename(filename))
-msnexp_filled <- readRDS("made_data_MS3000/msnexp_filled.rds")
-
-# Reorganize peak data from XCMS into clean format ----
-peak_data_long <- msnexp_filled %>%
-  chromPeaks() %>%
-  as.data.frame() %>%
-  rownames_to_column() %>%
-  mutate(peakidx=row_number()) %>%
-  select(sample, peakidx, into)
-peak_data <- msnexp_filled %>%
-  featureDefinitions() %>%
-  as.data.frame() %>%
-  filter(rtmed%between%c(30, 1200)) %>%
-  # filter(mzmed%between%pmppm(116.071154, 5))
-  select(peakidx) %>%
-  rownames_to_column("feature") %>%
-  unnest_longer(peakidx) %>%
-  left_join(peak_data_long, by = join_by(peakidx)) %>%
-  mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
-  complete(feature, filename, fill = list(into=0)) %>%
-  left_join(file_data, by = join_by(filename)) %>%
-  filter(str_detect(filename, "180821"))
-
-# Perform B-MIS ----
-source("manuscript/IS_integrations/is_combine.R")
-IS_areas <- read_csv("manuscript/IS_integrations/all_IS.csv") %>%
-  mutate(filename=paste0(filename, ".mzML")) %>%
-  rename(area="IS_area") %>%
-  filter(!str_detect(filename, "QC")) %>%
-  pivot_wider(names_from=IS_name, values_from = "IS_area") %>%
-  mutate("Inj_vol"=ifelse(str_detect(filename, "Poo.*Half"), 0.5, 1)) %>%
-  pivot_longer(-filename, names_to = "IS_name", values_to = "IS_area")
-pooled_IS_areas <- IS_areas %>%
-  filter(str_detect(filename, "_Poo_"))
-
-# Check using Proline
-peak_data %>%
-  select(feature, filename, into, depth) %>%
-  filter(str_detect(filename, "_Poo_")) %>%
-  filter(feature=="FT0532") %>%
[…]

### 2023-05-04 manuscript/manuscript.Rmd
Draft 0.5 done, sending to Anitra for discussion of results

@@
 We then explored the implications of this quality thresholding on the conclusions obtained by the downstream analysis, concluding that the poor performance of peakpicking algorithms significantly reduces the power of both univariate and multivariate statistical analyses in detecting environmental differences.
+We finally show that depth is a major driver of variability in the metabolome in high-quality environmental metabolomics datasets and identify several interesting metabolite groups for future investigation.
 
@@
 Notably, the peak shape and novel SNR parameters in the two-parameter model were among the most robust to training model variation, potentially explaining the consistency described above (Supp.
-figure X).
-
-```{r model stability dotplot errorbars supp figure}
-both_features <- rbind(FT2040_features, MS3000_features)
-dataset_versions_both <- c(dataset_versions, "both_features")
-model_param_estimates <- lst(all_params, xcms_params, raw_data_params) %>%
-  imap(function(param_selection, param_name){
-    lapply(dataset_versions_both, function(train_set){
-      model_params <- get(train_set) %>% 
-        filter(feat_class%in%c("Good", "Bad")) %>%
-        select(feat_class, all_of(param_selection)) %>%
-        mutate(feat_class=feat_class=="Good") %>%
-        glm(formula=feat_class~., family = binomial) %>%
-        broom::tidy() %>%
-        mutate(train_set)
-    }) %>% bind_rows() %>% mutate(param_name)
-  }) %>% bind_rows()
-suppfig_model_param_estimates_gp <- model_param_estimates %>%
-  mutate(param_name=str_remove(param_name, "_params")) %>%
-  ggplot(aes(x=param_name, color=train_set)) +
-  geom_hline(yintercept = 0) +
-  geom_point(aes(y=estimate), position = position_dodge(width=0.2)) +
-  geom_errorbar(aes(ymin=estimate-2*std.error, ymax=estimate+2*std.error),
-                position = position_dodge(width=0.2), width=0.3) +
-  facet_wrap(~term, scales="free_y", ncol = 4) +
-  theme_bw() +
-  theme(legend.position = "top", axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
-ggsave(plot = suppfig_model_param_estimates_gp, 
-       filename = "suppfig_model_param_estimates.png", 
-       device = "png", path = "manuscript/supplement",
-       width = 6.5, height = 9, units = "in", dpi = 300)
-```
+figure 1).
 
@@
 For the XCMS and two-parameter model, we found reasonably good convergence in a dataset containing half the mass features with most values falling within two standard errors of the estimate, while the full model required closer to 80% of the mass features to produce estimates consistent with the original model (Supp.
-figure X).
-
-```{r subsampling model stability supp figure}
-set.seed(123)
-both_data <- rbind(MS3000_features, FT2040_features) %>%
-  filter(feat_class%in%c("Good", "Bad")) %>%
-  mutate(feat_class=feat_class=="Good")
-
-best_estimates <- model_param_estimates %>%
-  filter(train_set=="both_features") %>%
-  filter(term!="(Intercept)") %>%
-  group_by(param_name) %>%
-  arrange(p.value) %>%
-  slice(1:2)
-
-frac_data <- map(c(0.2, 0.5, 0.8), function(sample_fraction){
-  replicate(10, {
-    train_set <- slice_sample(both_data, prop = sample_fraction)
-    lst(all_params, xcms_params, raw_data_params) %>%
-      imap(function(param_selection, param_name){
-        train_set %>%
-          select(feat_class, all_of(param_selection)) %>%
-          glm(formula=feat_class~., family = binomial) %>%
-          broom::tidy() %>%
-          mutate(param_name)
-    }) %>% bind_rows()
-  }, simplify = FALSE) %>% bind_rows() %>%
-    mutate(sample_fraction)
-}) %>% bind_rows()
-
-sample_model_estimates <- best_estimates %>%
-  select(term, train_set, param_name) %>%
-  left_join(frac_data, multiple = "all")
-
-gplist <- c("all_params", "xcms_params", "raw_data_params") %>%
-  lapply(function(which_params){
-    ggplot() +
-      geom_hline(yintercept = 0) +
-      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error*2, 
-                    ymax=estimate+std.error*2),
-                fill="grey80", data=best_estimates %>% filter(param_name==which_params)) +
-      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error, 
-                    ymax=estimate+std.error),
-                fill="grey50", data=best_estimates %>% filter(param_name==which_params)) +
-      geom_point(aes(x=sample_fraction, y=estimate), 
-                 data=sample_model_estimates %>% filter(param_name==which_params),
-                 alpha=0.5) +
-      ggh4x::facet_grid2(param_name~term, scales = "free_y", independent = "y") +
-      scale_x_continuous(breaks = c(0, 0.2, 0.5, 0.8, 1), limits = c(0, 1)) +
-      theme_bw()
-  })
-sampled_model_est_gp <- do.call(what = egg::ggarrange, c(gplist, ncol=1, draw=FALSE))
-ggsave(plot = sampled_model_est_gp, 
-       filename = "sampled_model_est.png", 
-       device = "png", path = "manuscript/supplement",
-       width = 6.5, height = 5, units = "in", dpi = 300)
-```
+figure 2).
 
@@
 All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GPF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent Type II error acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp.
-figure X).
+figure 3).
 The random forest model had perfect predictive capacity when tested internally on the training data (FDR=0%, GPF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp.
-figure X).
+figure 3).
 In each case, the performance drop when applied to a novel dataset was more extreme than the simple two-parameter model described above.
 
+### Implications for biological conclusions
+
+#### Univariate techniques
+
+[Figure 7] shows the distribution of all mass features identified by XCMS and their categorization according to both the biochemical response to depth as well as the classifications assigned by the MS expert and the two-parameter model at a 50% likelihood threshold.
+
+The abundance of noise peaks located at the center of the plot and therefore lacking much biological trend with depth validates the conventional heuristic of relying on downstream statistical tests to separate signal from noise (with the assumption that noise will be fixed across all samples and unaffected by environmental structuring of samples).
+However, it is also clear that there are both some good peaks that do not have trends with depth (and are therefore interesting for bucking the biomass signal) as well as a good number of bad peaks that do change across sample depths.
+These bad peaks that have biological signals appear to be frequently due to the tail of a peak being integrated as a separate molecular feature, which then retains the biological signal of the actual feature.
+
+Although the identification of bad peaks predominantly improves univariate analysis by removing those features from consideration in individual molecular narratives, filtering out the bad peaks should also improve the power of the individual analyses by reducing the number of hypotheses that must then be corrected for later.
+Multiple hypothesis testing corrections rely on the number and shape of the p-value distribution to make estimates of the "true" p-values for each compound, and removing a large number of noise peaks will result in lower p-values for those peaks that are flagged as good.
+
+Most metabolites track with biomass and are consistently abundant through the surface and DCM but drop off at depth, as prior measurements of particulate carbon and beam attenuation have shown [@benedetto].
+However, several groups of compounds diverge intriguingly from this bulk biomass trend, with some molecules found almost exclusively at the surface and others found to increase with depth rather than decrease.
+Several mass features that were found dominantly at the surface were putatively identified as mycosporine-like amino acids (MAAs), which have been shown to have UV photoprotective properties.[@ref] Although we were unable to calculate their exact concentrations due to a lack of authentic standards, peak areas were approximately 100 to 1000 times higher at 15 meters than their values averaged between the DCM and 175 meter samples.
+The abundance of MAAs in this particular category indicates that there may be other MAA molecules with this depth pattern that have not yet been characterized also among these unknowns.
+
+```{r clockplot BMIS, eval=!file.exists("manuscript/MS3000_norm_areas.csv")}
+library(xcms)
+file_data <- read_csv("made_data_MS3000/file_data.csv") %>%
+  mutate(filename=basename(filename))
+msnexp_filled <- readRDS("made_data_MS3000/msnexp_filled.rds")
+peak_data_long <- msnexp_filled %>%
+  chromPeaks() %>%
+  as.data.frame() %>%
+  rownames_to_column() %>%
+  mutate(peakidx=row_number()) %>%
+  select(sample, peakidx, into)
+peak_data <- msnexp_filled %>%
+  featureDefinitions() %>%
+  as.data.frame() %>%
+  filter(rtmed>30 & rtmed<1200) %>%
+  select(peakidx) %>%
+  rownames_to_column("feature") %>%
+  unnest_longer(peakidx) %>%
+  left_join(peak_data_long, by = join_by(peakidx)) %>%
+  mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
+  complete(feature, filename, fill = list(into=0)) %>%
+  left_join(file_data, by = join_by(filename)) %>%
+  filter(str_detect(filename, "180821"))
+
+source("manuscript/IS_integrations/is_combine.R")
+IS_areas <- read_csv("manuscript/IS_integrations/all_IS.csv") %>%
+  mutate(filename=paste0(filename, ".mzML")) %>%
+  rename(area="IS_area") %>%
+  filter(!str_detect(filename, "QC")) %>%
+  pivot_wider(names_from=IS_name, values_from = "IS_area") %>%
+  mutate("Inj_vol"=ifelse(str_detect(filename, "Poo.*Half"), 0.5, 1)) %>%
+  pivot_longer(-filename, names_to = "IS_name", values_to = "IS_area")
+pooled_IS_areas <- IS_areas %>%
+  filter(str_detect(filename, "_Poo_"))
+
+BMIS <- peak_data %>%
+  filter(str_detect(filename, "_Poo_")) %>%
+  full_join(pooled_IS_areas, multiple = "all", by = join_by(filename)) %>%
+  group_by(feature, IS_name) %>%
+  mutate(plot_area=into/IS_area*mean(IS_area)) %>%
+  summarise(cv=sd(plot_area)/mean(plot_area)) %>%
+  # left_join(init_cvs) %>%
+  arrange(cv) %>%
+  slice(1) %>%
+  select(feature, IS_name)
+
+BMISed_data <- peak_data %>%
+  filter(samp_type=="Smp") %>%
+  select(feature, filename, into) %>%
+  left_join(BMIS, by = join_by(feature)) %>%
+  left_join(IS_areas, by = join_by(filename, IS_name)) %>%
+  group_by(feature) %>%
+  mutate(norm_area=into/IS_area*mean(IS_area)) %>%
+  select(feature, filename, norm_area)
+
+norm_area_data <- BMISed_data %>%
+  arrange(norm_area) %>%
+  group_by(feature, filename) %>%
+  slice(1) %>%
+  ungroup()
+
+if(any(is.na(norm_area_data$norm_area)))stop("Found NAs in norm_area!")
+if(!all(table(norm_area_data$feature)==99))stop("Found not 99 samps in plotready!")
+write_csv(norm_area_data, "manuscript/MS3000_norm_areas.csv")
+```
+
+```{r clockplot dataset munging, eval=!file.exists("manuscript/MS3000_clockplot_coords.csv")}
+norm_area_data <- read_csv("manuscript/MS3000_norm_areas.csv")
+
+both_min_model <- rbind(FT2040_features, MS3000_features) %>%
+  select(feat_class, med_cor, med_SNR) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=feat_class=="Good") %>%
+  glm(formula=feat_class~., family = binomial)
+best_feats <- MS3000_features %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
+  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
+  select(feature, feat_class, pred_class)
+clockplot_data <- norm_area_data %>%
+  right_join(best_feats, by = join_by(feature)) %>%
+  left_join(file_data %>% select(filename, depth), by = join_by(filename))
+
+kw_pvals_all <- clockplot_data %>%
+  nest(data=-c(feature, pred_class)) %>%
+  mutate(tidyaov=map(
+    data, ~ broom::tidy(kruskal.test(.x$norm_area~.x$depth))), 
+  ) %>%
+  unnest(tidyaov) %>%
+  select(-parameter, -method) %>%
+  mutate(cor_pval_all=p.adjust(p.value, method="fdr")) %>%
+  group_by(pred_class) %>%
+  mutate(fdr_pval_byclass=p.adjust(p.value, method="fdr")) %>%
+  ungroup()
+
+resp_class_levels <- c("All equal", 
+    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
+    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
+    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
+    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
+    "Hard to tell")
+kw_classes_all <- clockplot_data %>%
+  group_by(feature) %>%
+  mutate(norm_area=rank(norm_area)) %>%
+  filter(var(norm_area)>0) %>%
+  rstatix::dunn_test(norm_area~depth, p.adjust.method = "none") %>%
+  mutate(contrast=paste(group2, group1, sep = "_")) %>%
+  select(feature, contrast, est=statistic, pval=p.adj) %>%
+  pivot_wider(names_from = contrast, values_from = c(pval, est)) %>%
+  mutate(resp_class=case_when(
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m<0 & est_175m_DCM<0 ~ "15m > DCM > 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m<0 & est_175m_DCM<0 ~ "DCM > 15m > 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m>0 & est_175m_DCM>0 ~ "DCM < 15m < 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m>0 & est_175m_DCM>0 ~ "15m < DCM < 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m>0 & est_175m_DCM<0 ~ "DCM > 175m > 15m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m<0 & est_175m_DCM>0 ~ "DCM < 175m < 15m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM>0.05 & est_DCM_15m<0 & est_175m_15m<0 ~ "15m > DCM = 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM>0.05 & est_DCM_15m>0 & est_175m_15m>0 ~ "15m < DCM = 175m",
+    pval_DCM_15m>0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_175m_15m<0 & est_175m_DCM<0 ~ "15m = DCM > 175m",
+    pval_DCM_15m>0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_175m_15m>0 & est_175m_DCM>0 ~ "15m = DCM < 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m>0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_DCM<0 ~ "DCM > 15m = 175m",
+    pval_DCM_15m<0.05 & pval_175m_15m>0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_DCM>0 ~ "DCM < 15m = 175m",
+    pval_DCM_15m>0.05 & pval_175m_15m>0.05 & pval_175m_DCM>0.05 ~ "All equal",
+    TRUE ~ "Hard to tell"
+  )) %>%
+  select(feature, resp_class) %>%
+  mutate(resp_class=factor(resp_class, levels=resp_class_levels))
+
+clockplot_xy_spots <- clockplot_data %>%
+  group_by(feature) %>%
+  mutate(norm_area=rank(norm_area)) %>%
+  group_by(feature, depth) %>%
+  summarise(mean_fold_change=mean(norm_area/50), .groups = "drop") %>% 
+  # div by 50 bc 99 files and median 1:99 is 50
+  pivot_wider(names_from=depth, values_from=mean_fold_change) %>%
+  mutate(x=DCM-1) %>%           # cursed but works??
+  mutate(y=`15m`+DCM/2-1.5) %>% # super fucking cursed
+  select(feature, x, y)
+
+clockplot_complete_df <- clockplot_data %>%
+  distinct(feature, feat_class, pred_class) %>%
+  left_join(clockplot_xy_spots, by = join_by(feature)) %>%
+  left_join(kw_pvals_all %>% select(feature, fdr_pval_byclass), by = join_by(feature)) %>%
+  left_join(kw_classes_all, by = join_by(feature)) %>%
+  mutate(error_type=case_when(
+    pred_class=="Good" & feat_class=="Good" ~ "TP",
+    pred_class=="Good" & feat_class=="Bad" ~ "FP",
+    pred_class=="Bad" & feat_class=="Good" ~ "FN",
+    pred_class=="Bad" & feat_class=="Bad" ~ "TN"
+  )) %>%
+  mutate(error_type=factor(error_type, levels=c("TP", "FP", "FN", "TN"),
+                           labels=c("Good peak found", "Bad peak included",
+                                    "Good peak missed", "Bad peak avoided")))
+
+write_csv(clockplot_complete_df, file = "manuscript/MS3000_clockplot_coords.csv")
+```
+
+```{r clockplot}
+resp_class_levels <- c("All equal", 
+    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
+    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
+    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
+    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
+    "Hard to tell")
+clockplot_complete_df <- read_csv(file = "manuscript/MS3000_clockplot_coords.csv") %>%
+  mutate(resp_class=factor(resp_class, levels=resp_class_levels))
+
+cmap_name <- "infinity"
+cmap_url <- paste0("https://raw.githubusercontent.com/1313e/CMasher/master/",
+                   "cmasher/colormaps/", cmap_name, "/", cmap_name, "_norm.txt")
+cmtable <- read.table(cmap_url)
+cmscale <- rgb(cmtable[,1], cmtable[,2], cmtable[,3])
+color_scale <- c("grey50", head(cmscale[seq(1, length(cmscale), length.out=13)], 12), "grey80")
+names(color_scale) <- resp_class_levels
+
+clock_scale <- 0.66
+axis_xends <- c(-1/2, 1.1, -1/2)*clock_scale
+axis_yends <- c(sqrt(3)/2, 0, -sqrt(3)/2)*clock_scale
+axis_y_dodge <- c(2, 2, -2)*clock_scale/25
+
+clockplot_base_gp <- ggplot() +
+  annotate("segment", x=c(0,0,0), y=c(0,0,0), 
+           xend=axis_xends, yend=axis_yends,
+           arrow=arrow(length=unit(0.02, "npc"), type = "closed")) +
+  annotate("segment", x=c(0,0,0), y=c(0,0,0),
+           xend=-axis_xends, yend=-axis_yends,
+           lty = 2) +
+  annotate("label", x=axis_xends, y=axis_yends+axis_y_dodge,
+           label=c("15m axis", "DCM axis", "175m axis"), 
+           fill="#FFFFFFAA", label.size=0) +
+  coord_fixed(xlim = c(-0.65, 1.3)*clock_scale, 
+              ylim = c(-1, 1)*clock_scale, 
+              expand=TRUE) +
+  scale_color_manual(values = color_scale, breaks = names(color_scale), drop=FALSE) +
+  theme_void() +
+  theme(legend.position = "right", 
+        legend.title = element_blank(),
+        plot.background = element_rect(fill="white", color="white")) +
+  guides(colour = guide_legend(override.aes = list(size=4, alpha=1), drop=FALSE)) +
+  aes(x=x, y=y, color=resp_class, label=feature)
+
+clockplot_complete_gp <- clockplot_base_gp +
+  geom_point(aes(shape=error_type), alpha=0.5, size=2,
+             data = clockplot_complete_df) +
+  scale_shape_manual(breaks = c("Good peak found", "Bad peak included",
+                                "Good peak missed", "Bad peak avoided"), 
+                     values = c(19, 3, 4, 1))
+
+ggsave(plot = clockplot_complete_gp, 
+       filename = "clockplot_complete.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 5, units = "in", dpi = 300)
+```
+
+![Figure 7: Plot of metabolite response to depth shown across three axes using the rank-normalized median value at each depth as the coordinate for that axis.
+Each molecular feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile.
+Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the deep chlorophyll maximum; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}
+
+#### Multivariate techniques
+
+Multivariate statistics also benefit from the reduced FDR when applying the two-parameter model.
+For the PERMANOVAs, we found that the proportion of variance explained (R$^2$) and the pseudo-F statistic increased monotonically with the likelihood threshold used to subset the data (Table 1).
+
+```{r multivariate setup}
+set.seed(123)
+MS3000_metadata <- file_data %>%
+  filter(samp_type=="Smp") %>%
+  filter(str_detect(filename, "180821")) %>%
+  select(filename, depth) %>%
+  column_to_rownames("filename")
+norm_area_data <- read_csv("manuscript/MS3000_norm_areas.csv")
+both_min_model <- rbind(FT2040_features, MS3000_features) %>%
+  select(feat_class, med_cor, med_SNR) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=feat_class=="Good") %>%
+  glm(formula=feat_class~., family = binomial)
+pred_probs <- MS3000_features %>%
+  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
+  select(feature, feat_class, pred_prob)
+wide_data <- norm_area_data %>%
+  group_by(feature) %>%
+  filter(var(norm_area)>0) %>%
+  # mutate(norm_area=rank(norm_area)) %>%
+  mutate(norm_area=scale(norm_area)[,1]) %>%
+  left_join(pred_probs, by="feature") %>% 
+  pivot_wider(names_from = "filename", values_from = norm_area) %>%
+  column_to_rownames("feature")
+
+mat_all <- wide_data
+mat_good <- wide_data %>% filter(feat_class=="Good")
+mat_90 <- wide_data %>% filter(pred_prob>0.9)
+mat_50 <- wide_data %>% filter(pred_prob>0.5)
+mat_10 <- wide_data %>% filter(pred_prob>0.1)
+mat_01 <- wide_data %>% filter(pred_prob>0.01)
+mat_list <- lst(mat_good, mat_90, mat_50, mat_10, mat_01, mat_all)
+```
+
+```{r PERMANOVA results}
+thresh_tests <- imap(mat_list, function(mat_given, mat_name){
+  mat_data <- mat_given %>%
+    select(-feat_class, -pred_prob) %>%
+    data.matrix() %>%
+    t()
+  perm_out <- vegan::adonis2(mat_data~depth, data = MS3000_metadata, method = "manhattan")
+  as.data.frame(perm_out[1,]) %>% 
+    mutate(thresh_name=mat_name) %>%
+    mutate(n=ncol(mat_data))
+}) %>% 
+  set_names(names(mat_list)) %>%
+  bind_rows() %>%
+  select(thresh_name, n, R2, `F`, `Pr(>F)`) %>%
+  `rownames<-`(NULL) %>%
+  column_to_rownames("thresh_name") %>%
+  knitr::kable()
+```
+
+In each test, the permutational p-value obtained was less than 0.001, indicating that the differences between samples due to depth were unlikely to be due to chance in all cases.
+However, the pseudo-F was much larger with higher thresholds, scaling from around 8.5 when thresholding at a 1% likelihood to 42 when thresholded at a 90% likelihood.
+
+|         |    n|   R2|    F| Pr(>F)|
+|:--------|----:|----:|----:|------:|
+|mat_good |  249| 0.44| 38.0|  0.001|
+|mat_90   |   75| 0.46| 42.0|  0.001|
+|mat_50   |  287| 0.34| 25.3|  0.001|
+|mat_10   |  516| 0.25| 16.0|  0.001|
+|mat_01   | 1129| 0.15|  8.5|  0.001|
+|mat_all  | 2086| 0.10|  5.5|  0.001|
+
+*Table 1: Number of mass features (n), variance explained (R2), pseudo-F statistic (F), and permutational p-value (Pr(>F)) estimates obtained from performing a permutational MANOVA on subsets of the full mass feature selection according to variable likelihood thresholds.
+[Needs cleaning, will fix]*
+
+We also tested the inclusion of all the features identified with XCMS (corresponding to a 0% threshold) and the results when only the manually-identified "Good" features were included.
+The default XCMS output continued the trend observed above, as expected, with the least variance explained and the lowest R$^2$ value.
+Subsetting for the "Good" peaks only, however, did not actually return the highest F-ratio or R$^2$, instead falling between the 50% and 90% thresholds for these two metrics.
+In large part this is due to the much smaller number of features: 249 features were manually labeled as Good, while only 75 exceeded the 90% likelihood threshold.
+
+The relative power of identifying only the very best peaks was also illustrated visually with non-metric multidimensional scaling (NMDS) plots (Supp.
+figure 4).
+In these common exploratory plots, the features with likelihoods above 50% strongly separated by depth while lower thresholds disguised the true signal and had higher stress values.
+Performing an NMDS on the manually-identified Good peaks resulted in output nearly indistinguishable from those of the 90% and 50% thresholds, while the 0.1 and 0.01 thresholds looked almost identical to the same analysis performed on all the XCMS features.
+
+## Discussion
+
+### Simple logistic regression models show good performance and minimal overfitting
+
+  - Recap performance metrics
+    - Discuss the relative performance and compare to assumptions in lit for
+      - Large peak area
+      - Has an isotope
+      - p-values
+    - Discuss use of summary statistics and the potential need for a blocked model by sample type
+  - Emphasize the best performance of the two-parameter model
+    - Full model and XCMS model definitely overfit, same for regularized/random forest
+    - XCMS alone does not provide enough information to accurately distinguish good from noise
+      - Likely true for all peakpickers
+      - Plea: report mz/rt bounding box for individual peaks so additional metrics can be calculated
+    - These params very simple (and fast!) to estimate
+      - Easy to do better with fancier models (e.g. loess smoothing for SNR)
+    - Why do all the models seem to approach ~25% FDR and ~75% GPF on the novel datasets?
+  - Limited to HILIC data at the moment but does seem to perform well across labs
+
+### Unable to create strict control of the FDR but a single threshold still makes life easier
+
+  - Initial goal: set an FDR, choose a likelihood threshold based on it
+    - Not doable here but maybe in future work?
+    - Pivoted to recommending step-down manual labelling
+  - Plea: report your FDR! Especially if you're manually labelling!
+    - Necessary for error propagation anyway
+  - Future work: alter chromatography/peakpicking params to maximize # of peaks above 0.9 threshold
+
+### Biological results do depend on the threshold used
+
+  - Large amounts of noise reduce the power of both univariate and multivariate stats
+    - Trade-off with loss of interesting signals at higher thresholds (cmpds on clockplot left)
+    - Possible to get wildly different results with different thresholds (% variance explained)
+    - Manual evaluation for interesting signals still strongly recommended
+  - Depth is a major driver of metabolome composition in the gyre
+  - Majority of metabolites track with biomass metrics
+  - Some interesting metabolites diverge from dominant biomass trend
+    - Some new MAAs in clockplot top left? (Still need to check this w MSMS)
+    - Degradation products / bacterial specialties at bottom?
+
+## Conclusions
+
+TBD
+
+## Acknowledgements
+
+## References
+
+## Supplement
+
+```{r model stability dotplot errorbars supp figure}
+both_features <- rbind(FT2040_features, MS3000_features)
+dataset_versions_both <- c(dataset_versions, "both_features")
+model_param_estimates <- lst(all_params, xcms_params, raw_data_params) %>%
+  imap(function(param_selection, param_name){
+    lapply(dataset_versions_both, function(train_set){
+      model_params <- get(train_set) %>% 
+        filter(feat_class%in%c("Good", "Bad")) %>%
+        select(feat_class, all_of(param_selection)) %>%
+        mutate(feat_class=feat_class=="Good") %>%
+        glm(formula=feat_class~., family = binomial) %>%
+        broom::tidy() %>%
+        mutate(train_set)
+    }) %>% bind_rows() %>% mutate(param_name)
+  }) %>% bind_rows()
+suppfig_model_param_estimates_gp <- model_param_estimates %>%
+  mutate(param_name=str_remove(param_name, "_params")) %>%
+  ggplot(aes(x=param_name, color=train_set)) +
+  geom_hline(yintercept = 0) +
+  geom_point(aes(y=estimate), position = position_dodge(width=0.2)) +
+  geom_errorbar(aes(ymin=estimate-2*std.error, ymax=estimate+2*std.error),
+                position = position_dodge(width=0.2), width=0.3) +
+  facet_wrap(~term, scales="free_y", ncol = 4) +
+  theme_bw() +
+  theme(legend.position = "top", axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
+ggsave(plot = suppfig_model_param_estimates_gp, 
+       filename = "suppfig_model_param_estimates.png", 
+       device = "png", path = "manuscript/supplement",
+       width = 6.5, height = 9, units = "in", dpi = 300)
+```
+
+![Supplemental figure 1: Model parameter estimates for each of the metrics in the full model, additionally broken down by their inclusion in the two-parameter (raw_data) and XCMS-exclusively models.
+Colors correspond to the dataset used to train the logistic regression model, with "both" indicating a combined model using all manually-labeled features across both datasets.
+<Explanations of the actual facet names forthcoming.>](supplement/suppfig_model_param_estimates.png){width=90%}
+
+```{r subsampling model stability supp figure}
+set.seed(123)
+both_data <- rbind(MS3000_features, FT2040_features) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=feat_class=="Good")
+
+best_estimates <- model_param_estimates %>%
+  filter(train_set=="both_features") %>%
+  filter(term!="(Intercept)") %>%
+  group_by(param_name) %>%
+  arrange(p.value) %>%
+  slice(1:2)
+
+frac_data <- map(c(0.2, 0.5, 0.8), function(sample_fraction){
+  replicate(10, {
+    train_set <- slice_sample(both_data, prop = sample_fraction)
+    lst(all_params, xcms_params, raw_data_params) %>%
+      imap(function(param_selection, param_name){
+        train_set %>%
+          select(feat_class, all_of(param_selection)) %>%
+          glm(formula=feat_class~., family = binomial) %>%
+          broom::tidy() %>%
+          mutate(param_name)
+    }) %>% bind_rows()
+  }, simplify = FALSE) %>% bind_rows() %>%
+    mutate(sample_fraction)
+}) %>% bind_rows()
+
+sample_model_estimates <- best_estimates %>%
+  select(term, train_set, param_name) %>%
+  left_join(frac_data, multiple = "all", by=join_by(term, param_name))
+
+gplist <- c("all_params", "xcms_params", "raw_data_params") %>%
+  lapply(function(which_params){
+    ggplot() +
+      geom_hline(yintercept = 0) +
+      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error*2, 
+                    ymax=estimate+std.error*2),
+                fill="grey80", data=best_estimates %>% filter(param_name==which_params)) +
+      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error, 
+                    ymax=estimate+std.error),
+                fill="grey50", data=best_estimates %>% filter(param_name==which_params)) +
+      geom_point(aes(x=sample_fraction, y=estimate), 
+                 data=sample_model_estimates %>% filter(param_name==which_params),
+                 alpha=0.5) +
+      ggh4x::facet_grid2(param_name~term, scales = "free_y", independent = "y") +
+      scale_x_continuous(breaks = c(0, 0.2, 0.5, 0.8, 1), limits = c(0, 1)) +
+      theme_bw()
+  })
+sampled_model_est_gp <- do.call(what = egg::ggarrange, c(gplist, ncol=1, draw=FALSE))
+ggsave(plot = sampled_model_est_gp, 
+       filename = "sampled_model_est.png", 
+       device = "png", path = "manuscript/supplement",
+       width = 6.5, height = 5, units = "in", dpi = 300)
+```
+
+![Supplemental figure 2: Robustness of the two most significant metrics across the full (all_params), XCMS-only (xcms_params), and two-parameter (raw_data_params) models.
+The x-axis corresponds to the fraction of the data used to train the model and the y-coordinate shows the estimated value for the specified term in the subset across 10-fold replicated subsampling.
+The grey bar in the background corresponds to the estimate of the full model +/- 1SE (thinner dark grey bar) and 2SE (thicker light grey bar).](supplement/sampled_model_est.png){width=90%}
+
 ```{r elastic net and random forest}
-library(glmnet)
+set.seed(123)
 net_cross_train_df <- lst(ridge=0, elastic=0.5, lasso=1) %>%
@@
         pull(feat_class)
-      net_model <- cv.glmnet(x=net_x, y=net_y, family="binomial", alpha=alpha_selection)
+      net_model <- glmnet::cv.glmnet(x=net_x, y=net_y, family="binomial", alpha=alpha_selection)
 
@@
   }) %>% bind_rows()
-library(randomForest)
 rf_cross_train_df <- lapply(dataset_versions, function(train_set){
@@
     factor()
-  rfmodel <- randomForest(rf_x, rf_y)
+  rfmodel <- randomForest::randomForest(rf_x, rf_y)
 
@@
 
-### Implications for biological conclusions
-
-[Figure 7] shows the distribution of all mass features identified by XCMS and their categorization according to both the biochemical response to depth as well as the classifications assigned by the MS expert and the two-parameter model at a 50% likelihood threshold.
-
-The abundance of noise peaks located at the center of the plot and therefore lacking much biological trend with depth validates the conventional heuristic of relying on downstream statistical tests to separate signal from noise (with the assumption that noise will be fixed across all samples and unaffected by environmental structuring of samples).
-However, it is also clear that there are both some good peaks that do not have trends with depth (and are therefore interesting for bucking the biomass signal) as well as a good number of bad peaks that do change across sample depths.
-These bad peaks that have biological signals appear to be frequently due to the tail of a peak being integrated as a separate molecular feature, which then retains the biological signal of the actual feature.
-
-Most metabolites track with biomass and are consistently abundant through the surface and DCM but drop off at depth, as prior measurements of particulate carbon and beam attenuation have shown [@benedetto].
-However, several groups of compounds diverge intriguingly from this bulk biomass trend, with some molecules found almost exclusively at the surface and others found to increase with depth rather than decrease.
-Several mass features that were found most exclusively at the surface were putatively identified as mycosporine-like amino acids (MAAs), which have been shown to have UV photoprotective properties.[@ref] Although we were unable to calculate their exact concentrations due to a lack of authentic standards, peak areas were approximately 100 to 1000 times higher at 15 meters than their values averaged between the DCM and 175 meter samples.
-The abundance of MAAs in this particular category indicates that there may be other MAA molecules with this depth pattern that have not yet been characterized also among these unknowns.
-
-```{r clockplot BMIS, eval=!file.exists("manuscript/MS3000_norm_areas.csv")}
-library(xcms)
-file_data <- read_csv("made_data_MS3000/file_data.csv") %>%
-  mutate(filename=basename(filename))
-msnexp_filled <- readRDS("made_data_MS3000/msnexp_filled.rds")
-peak_data_long <- msnexp_filled %>%
-  chromPeaks() %>%
+![Supplemental figure 3: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.](supplement/regregrf_fdrgpf_cross_train.png){width=90%}
+
+```{r NMDS results}
+thresh_NMDS_df <- imap(mat_list, function(mat_given, mat_name){
+  mat_data <- mat_given %>%
+    select(-feat_class, -pred_prob) %>%
+    data.matrix() %>%
+    t()
+  nmds_output <- vegan::metaMDS(mat_data, k = 2, distance = "manhattan", trace = 0,
+                         autotransform = FALSE, noshare = FALSE, wascores = FALSE)
+  nmds_gp <- nmds_output$points %>%
     as.data.frame() %>%
-  rownames_to_column() %>%
-  mutate(peakidx=row_number()) %>%
-  select(sample, peakidx, into)
-peak_data <- msnexp_filled %>%
-  featureDefinitions() %>%
-  as.data.frame() %>%
-  filter(rtmed>30 & rtmed<1200) %>%
-  select(peakidx) %>%
-  rownames_to_column("feature") %>%
-  unnest_longer(peakidx) %>%
-  left_join(peak_data_long, by = join_by(peakidx)) %>%
-  mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
-  complete(feature, filename, fill = list(into=0)) %>%
-  left_join(file_data, by = join_by(filename)) %>%
-  filter(str_detect(filename, "180821"))
-
-source("manuscript/IS_integrations/is_combine.R")
-IS_areas <- read_csv("manuscript/IS_integrations/all_IS.csv") %>%
-  mutate(filename=paste0(filename, ".mzML")) %>%
-  rename(area="IS_area") %>%
-  filter(!str_detect(filename, "QC")) %>%
-  pivot_wider(names_from=IS_name, values_from = "IS_area") %>%
-  mutate("Inj_vol"=ifelse(str_detect(filename, "Poo.*Half"), 0.5, 1)) %>%
-  pivot_longer(-filename, names_to = "IS_name", values_to = "IS_area")
-pooled_IS_areas <- IS_areas %>%
-  filter(str_detect(filename, "_Poo_"))
-
-BMIS <- peak_data %>%
-  filter(str_detect(filename, "_Poo_")) %>%
-  full_join(pooled_IS_areas, multiple = "all", by = join_by(filename)) %>%
-  group_by(feature, IS_name) %>%
-  mutate(plot_area=into/IS_area*mean(IS_area)) %>%
-  summarise(cv=sd(plot_area)/mean(plot_area)) %>%
-  # left_join(init_cvs) %>%
-  arrange(cv) %>%
-  slice(1) %>%
-  select(feature, IS_name)
-
-BMISed_data <- peak_data %>%
-  filter(samp_type=="Smp") %>%
-  select(feature, filename, into) %>%
-  left_join(BMIS, by = join_by(feature)) %>%
-  left_join(IS_areas, by = join_by(filename, IS_name)) %>%
-  group_by(feature) %>%
-  mutate(norm_area=into/IS_area*mean(IS_area)) %>%
-  select(feature, filename, norm_area)
-
-norm_area_data <- BMISed_data %>%
-  arrange(norm_area) %>%
-  group_by(feature, filename) %>%
-  slice(1) %>%
-  ungroup()
-
-if(any(is.na(norm_area_data$norm_area)))stop("Found NAs in norm_area!")
-if(!all(table(norm_area_data$feature)==99))stop("Found not 99 samps in plotready!")
-write_csv(norm_area_data, "manuscript/MS3000_norm_areas.csv")
-```
-
-```{r clockplot dataset munging, eval=!file.exists("manuscript/MS3000_clockplot_coords.csv")}
-norm_area_data <- read_csv("manuscript/MS3000_norm_areas.csv")
-
-both_min_model <- rbind(FT2040_features, MS3000_features) %>%
-  select(feat_class, med_cor, med_SNR) %>%
-  filter(feat_class%in%c("Good", "Bad")) %>%
-  mutate(feat_class=feat_class=="Good") %>%
-  glm(formula=feat_class~., family = binomial)
-best_feats <- MS3000_features %>%
-  filter(feat_class%in%c("Good", "Bad")) %>%
-  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
-  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
-  select(feature, feat_class, pred_class)
-clockplot_data <- norm_area_data %>%
-  right_join(best_feats, by = join_by(feature)) %>%
-  left_join(file_data %>% select(filename, depth), by = join_by(filename))
-
-kw_pvals_all <- clockplot_data %>%
-  nest(data=-c(feature, pred_class)) %>%
-  mutate(tidyaov=map(
-    data, ~ broom::tidy(kruskal.test(.x$norm_area~.x$depth))), 
-  ) %>%
-  unnest(tidyaov) %>%
-  select(-parameter, -method) %>%
-  mutate(cor_pval_all=p.adjust(p.value, method="fdr")) %>%
-  group_by(pred_class) %>%
-  mutate(fdr_pval_byclass=p.adjust(p.value, method="fdr")) %>%
-  ungroup()
-
-resp_class_levels <- c("All equal", 
-    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
-    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
-    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
-    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
-    "Hard to tell")
-kw_classes_all <- clockplot_data %>%
-  group_by(feature) %>%
-  mutate(norm_area=rank(norm_area)) %>%
-  filter(var(norm_area)>0) %>%
-  rstatix::dunn_test(norm_area~depth, p.adjust.method = "none") %>%
-  mutate(contrast=paste(group2, group1, sep = "_")) %>%
-  select(feature, contrast, est=statistic, pval=p.adj) %>%
-  pivot_wider(names_from = contrast, values_from = c(pval, est)) %>%
-  mutate(resp_class=case_when(
-    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m<0 & est_175m_DCM<0 ~ "15m > DCM > 175m",
-    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m<0 & est_175m_DCM<0 ~ "DCM > 15m > 175m",
-    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m>0 & est_175m_DCM>0 ~ "DCM < 15m < 175m",
-    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m>0 & est_175m_DCM>0 ~ "15m < DCM < 175m",
-    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m>0 & est_175m_DCM<0 ~ "DCM > 175m > 15m",
-    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m<0 & est_175m_DCM>0 ~ "DCM < 175m < 15m",
-    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM>0.05 & est_DCM_15m<0 & est_175m_15m<0 ~ "15m > DCM = 175m",
-    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM>0.05 & est_DCM_15m>0 & est_175m_15m>0 ~ "15m < DCM = 175m",
-    pval_DCM_15m>0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_175m_15m<0 & est_175m_DCM<0 ~ "15m = DCM > 175m",
-    pval_DCM_15m>0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_175m_15m>0 & est_175m_DCM>0 ~ "15m = DCM < 175m",
-    pval_DCM_15m<0.05 & pval_175m_15m>0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_DCM<0 ~ "DCM > 15m = 175m",
-    pval_DCM_15m<0.05 & pval_175m_15m>0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_DCM>0 ~ "DCM < 15m = 175m",
-    pval_DCM_15m>0.05 & pval_175m_15m>0.05 & pval_175m_DCM>0.05 ~ "All equal",
-    TRUE ~ "Hard to tell"
+    cbind(MS3000_metadata) %>%
+    rownames_to_column("filename") %>%
+    mutate(mat_name) %>%
+    mutate(stress=nmds_output$stress)
+}) %>% 
+  bind_rows() %>%
+  select(mat_name, filename, starts_with("MDS"), depth, stress) %>%
+  mutate(basename=str_extract(filename, "MS\\d+C\\d+.*(?=_[A-C])")) %>%
+  mutate(mat_name=factor(
+    mat_name, levels=c("mat_good", "mat_90", "mat_50", "mat_10", "mat_01", "mat_all"),
+    labels=c("Only good", paste("Threshold", c(0.9, 0.5, 0.1, 0.01)), "All features")
   )) %>%
-  select(feature, resp_class) %>%
-  mutate(resp_class=factor(resp_class, levels=resp_class_levels))
-
-clockplot_xy_spots <- clockplot_data %>%
-  group_by(feature) %>%
-  mutate(norm_area=rank(norm_area)) %>%
-  group_by(feature, depth) %>%
-  summarise(mean_fold_change=mean(norm_area/50), .groups = "drop") %>% 
-  # div by 50 bc 99 files and median 1:99 is 50
-  pivot_wider(names_from=depth, values_from=mean_fold_change) %>%
-  mutate(x=DCM-1) %>%           # cursed but works??
-  mutate(y=`15m`+DCM/2-1.5) %>% # super fucking cursed
-  select(feature, x, y)
-
-clockplot_complete_df <- clockplot_data %>%
-  distinct(feature, feat_class, pred_class) %>%
-  left_join(clockplot_xy_spots, by = join_by(feature)) %>%
-  left_join(kw_pvals_all %>% select(feature, fdr_pval_byclass), by = join_by(feature)) %>%
-  left_join(kw_classes_all, by = join_by(feature)) %>%
-  mutate(error_type=case_when(
-    pred_class=="Good" & feat_class=="Good" ~ "TP",
-    pred_class=="Good" & feat_class=="Bad" ~ "FP",
-    pred_class=="Bad" & feat_class=="Good" ~ "FN",
-    pred_class=="Bad" & feat_class=="Bad" ~ "TN"
-  )) %>%
-  mutate(error_type=factor(error_type, levels=c("TP", "FP", "FN", "TN"),
-                           labels=c("Good peak found", "Bad peak included",
-                                    "Good peak missed", "Bad peak avoided")))
-
-write_csv(clockplot_complete_df, file = "manuscript/MS3000_clockplot_coords.csv")
-```
-
-```{r clockplot}
-resp_class_levels <- c("All equal", 
-    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
-    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
-    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
-    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
-    "Hard to tell")
-clockplot_complete_df <- read_csv(file = "manuscript/MS3000_clockplot_coords.csv") %>%
-  mutate(resp_class=factor(resp_class, levels=resp_class_levels))
-
-cmap_name <- "infinity"
-cmap_url <- paste0("https://raw.githubusercontent.com/1313e/CMasher/master/",
-                   "cmasher/colormaps/", cmap_name, "/", cmap_name, "_norm.txt")
-cmtable <- read.table(cmap_url)
-cmscale <- rgb(cmtable[,1], cmtable[,2], cmtable[,3])
-color_scale <- c("grey50", head(cmscale[seq(1, length(cmscale), length.out=13)], 12), "grey80")
-names(color_scale) <- resp_class_levels
-
-clock_scale <- 0.66
-axis_xends <- c(-1/2, 1.1, -1/2)*clock_scale
-axis_yends <- c(sqrt(3)/2, 0, -sqrt(3)/2)*clock_scale
-axis_y_dodge <- c(2, 2, -2)*clock_scale/25
-
-clockplot_base_gp <- ggplot() +
-  annotate("segment", x=c(0,0,0), y=c(0,0,0), 
-           xend=axis_xends, yend=axis_yends,
-           arrow=arrow(length=unit(0.02, "npc"), type = "closed")) +
-  annotate("segment", x=c(0,0,0), y=c(0,0,0),
-           xend=-axis_xends, yend=-axis_yends,
-           lty = 2) +
-  annotate("label", x=axis_xends, y=axis_yends+axis_y_dodge,
-           label=c("15m axis", "DCM axis", "175m axis"), 
-           fill="#FFFFFFAA", label.size=0) +
-  coord_fixed(xlim = c(-0.65, 1.3)*clock_scale, 
-              ylim = c(-1, 1)*clock_scale, 
-              expand=TRUE) +
-  scale_color_manual(values = color_scale, breaks = names(color_scale), drop=FALSE) +
-  theme_void() +
-  theme(legend.position = "right", 
-        legend.title = element_blank(),
-        plot.background = element_rect(fill="white", color="white")) +
-  guides(colour = guide_legend(override.aes = list(size=4, alpha=1), drop=FALSE)) +
-  aes(x=x, y=y, color=resp_class, label=feature)
-
-clockplot_complete_gp <- clockplot_base_gp +
-  geom_point(aes(shape=error_type), alpha=0.5, size=2,
-             data = clockplot_complete_df) +
-  scale_shape_manual(breaks = c("Good peak found", "Bad peak included",
-                                "Good peak missed", "Bad peak avoided"), 
-                     values = c(19, 3, 4, 1))
-
-ggsave(plot = clockplot_complete_gp, 
-       filename = "clockplot_complete.png", 
-       device = "png", path = "manuscript/figures",
-       width = 6.5, height = 5, units = "in", dpi = 300)
-```
-
-![Figure 7: Plot of metabolite response to depth shown across three axes using the rank-normalized median value at each depth as the coordinate for that axis.
-Each molecular feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile.
-Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the deep chlorophyll maximum; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}
+  mutate(stress=format(stress, digits=2))
+thresh_NMDS_gp <- thresh_NMDS_df %>%
+  ggplot() +
+  geom_vline(xintercept = 0) +
+  geom_hline(yintercept = 0) +
+  geom_polygon(aes(x=MDS1, y=MDS2, fill=depth, color=depth, group=basename), alpha=0.5) +
+  geom_label(aes(label=stress), x=Inf, y=Inf, 
+             label.r = unit(0, "in"), hjust=1, vjust=1,
+             data=thresh_NMDS_df %>% distinct(mat_name, stress)) +
+  facet_wrap(~mat_name, scales = "free", ncol=3) +
+  scale_color_manual(values=viridis::viridis(9, direction = -1)[c(2,5,8)], 
+                     breaks = c("15m", "DCM", "175m"),
+                     aesthetics = c("color", "fill")) +
+  theme_bw() +
+  theme(aspect.ratio = 1)
+ggsave(plot = thresh_NMDS_gp, filename = "thresh_NMDS_plots.png", 
+       device = "png", path = "manuscript/supplement",
+       width = 6.5, height = 4, units = "in", dpi = 300)
+```
+
+![Supplemental figure 4: NMDS plots of metabolite similarity according to sample depth across multiple likelihood thresholds.
+Triplicate samples are represented by the vertices of the triangles and colored by the depth from which they were sampled (DCM = deep chlorophyll maximum, ~110m).
+"Only good" refers to those features manually labeled as "Good".
+NMDS stress values are reported to 3 significant figures in the upper right of the plot.](supplement/thresh_NMDS_plots.png){width=90%}

### 2023-05-27 manuscript/manuscript.Rmd
Splitting abstract into BMC sections
Restructured to follow BMC guidelines (methods at end)
Massively reduced intro length

@@
 
+### Background
+
 Chromatographic peakpicking continues to represent a significant bottleneck in automated LC-MS workflows.
@@
 Here, we manually labeled two oceanographic particulate metabolite datasets to assess the performance of individual peak quality metrics and constructed a predictive model calibrated to the likelihood that visual inspection by an MS expert would include a given mass feature in the downstream analysis.
+
+### Results
+
 A simple logistic regression model built on two metrics calculated from the raw MS data reduced the fraction of false positives in the analysis from 80-90% down to 1-5% and showed minimal overfitting when applied to novel datasets.
@@
 
-## Introduction
+### Conclusions
+
+TBD
+
+## Background
 
@@
 Its rapid sample processing (typically <1 hr run time), low limits of detection (pM-nM range), and ability to characterize novel molecules via fragmentation fingerprints make it a common workhorse for metabolomic research.
-However, much of the data produced by these instruments goes unused because traditional workflows target only a fraction of the compounds detected based on an *a priori* list of known molecules.
-Significant efforts in the past two decades have established workflows for untargeted metabolomics [@xcms, @mzmine, @msdial], but these data-driven methods are still handicapped by the poor performance of the core peakpicking algorithms and require manual oversight and curation of the datasets after they're produced by the software.
-This peakpicking step is typically the very first in the data processing stage, with implications for all downstream analysis and interpretation.
-[Mention HILIC chromatography specifically to justify its use here?]
-
-Peakpicking is a well-known problem across many fields of science and especially in LC-MS analysis due to the smooth transition from high-quality peaks into background noise as the limit of detection is approached and crossed, with background signals often sharing many features of high-quality peaks.
-The imperfection of these algorithms introduces a fundamental tradeoff between false positives and false negatives, where false positives (Type I errors) represent background instrument or chemical noise that has been misclassified as true biological signal and false negatives (Type II errors) indicating biological signals that are misclassified as noise.
-Existing algorithms tend to be very forgiving of Type I errors and very harsh on Type II, presenting many false positives to the researcher in their attempt to absolutely minimize the number of real molecules missed.
-This approach makes sense because downstream analyses can always remove additional mass features, but Type II errors cannot be later recovered.
-However, we would argue that this emphasis on an absolute minimization of the missed peaks does a disservice to scientists for two reasons.
-First, the abundance of false positives means that peakpicking output cannot be automatically trusted.
-This mandates manual evaluation of the raw data, a time-consuming and mind-numbing process with a large degree of subjective decision making that cannot be perfectly replicated and scales combinatorially with the number of samples and compounds measured.
-Second, the presence of false negatives cannot be fully accounted for in the way the false positive rate can.
-False negatives are introduced in every step of metabolomic analyses, with compounds lost during sample processing, injection, and detection on the instrument.
-This is a well-known problem in metabolomics and will not be solved by any perfecting of peakpicking algorithms.
-Instead, emphasis should be placed on allowing the experimenter to set an acceptable threshold for the false discovery rate and accepting that this will inherently add to the number of peaks already lost in the data collection process.
+In the past two decades, data-driven methods have established workflows for untargeted metabolomics but the imperfect performance of the core peakpicking algorithms continue to require manual oversight and curation.
+This problem has been exacerbated by the increased use of non-traditional chromatography such as hydrophilic interaction which tends to produce noisier peaks.
+
+Noisy data and imperfect detection algorithms introduce a tradeoff between false positives (where background instrument or chemical noise is misclassified as biological signal) and false negatives (where real signals are undetected).
+Existing algorithms favor false positives because downstream analyses can always remove erroneous mass features, but false negatives cannot be later recovered.
+However, this approach creates two additional downstream problems.
+First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of peaks, a task that scales combinatorially with the number of samples and compounds measured.
+Second, false negatives cannot be fully accounted for in the way false positives can because false negatives are introduced in every step of metabolomic analyses from sample collection through injection.
+Instead of minimizing false negatives, we believe that emphasis should be placed on allowing the experimenter to set a threshold for the false discovery rate and accept that this will inherently add to the number of peaks already lost in the data collection process.
 
 Existing peak-detection softwares do not provide a clear way to do this.
-Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each molecular feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew which are rarely estimated using a well-documented algorithm.
-None of these parameters answer the critical question of interest: what is the likelihood that this subset of data corresponds to an actual molecule present in the original sample?
+Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each molecular feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew.
+None of these parameters answer the critical question about the likelihood that a given feature corresponds to a molecule present in the original sample.
 This parameter is crucial for downstream analysis because it represents the base rate for error propagation and acceptable thresholds should vary widely by the particular project's goals.
 In an exploratory analysis, any molecular feature more than 50% likely to be real is perhaps worth considering, while in a confirmatory study this threshold may need to be above 99%.
-While significant effort has been recently invested into improvements in the peakpicking algorithms, very little has been done to quantify the accuracy and precision of their outputs on the wide variety of datasets from which they are determined.
-
-A single metric of peak quality calibrated to its likelihood of representing an actual molecule has multiple advantages over the current heuristic approaches.
-First, providing an intuitive metric of peak goodness allows researchers to focus their energy on those features least likely to be noise and most likely to be found in a repeated analysis.
-This applies both during the manual curation phase in which the very best or most interesting peaks can be visually inspected while those of poor quality can be discarded, as well as in the compound discovery/characterization steps where time and energy spent attempting to determine the structure or molecular formula of a noise signal is wasted.
-Second, removing low-quality noise signals from the dataset prior to downstream analysis should improve statistical power by reducing the number of hypotheses tested for univariate analyses and increasing the relative power of significant peaks in multivariate analyses.
-Third, such a metric would allow optimization of peakpicking and chromatographic parameters for untargeted datasets.
-Rather than measuring the effect of a change on a small subset of known molecules or by counting the total number of peaks found, one could look at the metric's distribution to see whether the change was an improvement in the number and absolute quality of the peaks detected.
-Fourth, high quality peaks should be independent of software implementations and should be found no matter what algorithms are used, relieving the scientist (and reviewers!) of some of the burden of learning new software or programming languages to check their work.
-Fifth, this metric would enable more consistent quality control between labs and analyists.
-Rather than discussing a plethora of parameters or making subjective assessments of peak shape, an idealized metric would provide a simple and intuitive threshold upon which to agree for a given project.
-This metric would be scripted and provide consistent, reproducible results independent of the particular expert reviewing its performance.
-
-A single comprehensive metric also has significant advantages over multiple individual thresholds.
-First, a metric calibrated to likelihood is easier and more intuitive to use for classification than relatively arbitrary thresholds of signal-to-noise or peak skewness with ambiguous units and unintuitive interpretations.
-Second, a single compound metric also has the multivariate advantage over independent thresholds by allowing a good peak to be weak in one area which is compensated for by strong performance in other metrics.
-Sometimes good peaks do have high SNR or small areas, and thresholding on these metrics independently unnecessarily increases the likelihood of missing signals.
-Finally, the relative power of these many different metrics is difficult to compute and hard to compare.
-Is it more important to have an isotope, or a Gaussian peak shape?
-Is it more likely to be a biological signal if it has a large area, or if it differs between treatment?
-If so, by what degree?
-These metrics differ in strength depending on the dataset, but many MS experts have some degree of heuristic knowledge about their relative power for the data they've worked with and analyzed.
+While significant effort has been invested in improving the peakpicking algorithms, very little has been done to quantify the accuracy and precision of their outputs across the wide variety of datasets to which they are applied.
+
+A single parameter of peak quality also facilitates downstream analyses in multiple ways.
+This metric would improve statistical power by reducing the number of effective hypotheses tested and allow researchers to focus effort on features least likely to be noise.
+Additionally, this parameter could be optimized to improve peakpicking and chromatographic settings independently of the software used or inter-lab variability when scripted to provide consistent, reproducible results independent of the particular expert reviewing its performance.
+Constructing such a single comprehensive metric calibrated to likelihood is also more effective than multiple independent thresholds because it has meaningful units, does not require estimating the relative power of individual metrics, and allows a good peak to compensate for weak performance in one area with strong performance in other metrics.
+
+[Where to talk about the peakpicking vs integration? Here we just looked at peakpicking]
 
@@
 Here, we construct and validate multiple predictive models of peak quality based on metrics both common in the literature and custom implementations we've found useful in our own analysis.
-This allows us to connect the physical, chemical, and biological measurements taken regularly around the globe to a molecular-scale perspective of particular organic matter in the ocean by linking the chemical currencies that fuel the planet to the environments in which they're found.
-
-## Methods
-
-### Sample collection
-
-Environmental samples were collected from the North Pacific Subtropical Gyre near Station ALOHA during two research cruises that targeted strong mesoscale eddy features during June/July 2017 and March/April 2018, traversing an area between 28 °N, 156 °W and 23 °N, 161 °W.
-An eddy dipole off the coast of Hawaii was detected using sea-level anomaly (SLA) satellite data and targeted for both a transect across the cyclonic and anticyclonic poles of the eddy dipole.
-The cyclonic pole of the eddy had a maximum negative SLA anomaly of -15 cm in 2017 and -20 cm in 2018, while the anticyclonic center reached +24 cm in 2017 and +21 cm in 2018.
-The 2017 cruise samples were taken along a transect across the eddy dipole while the 208 cruise targeted only the center of each eddy.
-
-Environmental samples were obtained using the onboard CTD rosette to collect water from 15 meters, the deep chlorophyll maximum (DCM), and 175 meters during the 2017 MESOSCOPE cruise and from 25 meters and the DCM during the 2018 Falkor cruise.
-The DCM was determined visually from fluorometer data during the CTD downcast and Niskin bottles were tripped during the return trip to the surface.
-Seawater from each depth was sampled in triplicate by firing one Niskin bottle for each sample.
-Samples were brought to the surface and decanted into prewashed (3x with DI, 3x with sampled seawater) polycarbonate bottles for filtration.
-Samples were filtered by peristaltic pump onto 142mm 0.2 µm Durapore filters held by polycarbonate filter holders on a Masterflex tubing line.
-Pressures were kept as low as possible while still producing a reasonable rate of flow through the filter, approximately 250-500 mL per minute.
-Samples were then removed from the filter holder using solvent-washed tweezers and placed into pre-combusted aluminum foil packets that were then flash-frozen in liquid nitrogen before being stored at -80 °C until extraction.
-A methodological blank was also collected by running filtrate through a new filter and then treated identically to the samples.
-
-Culture samples used as the validation sets for this paper were obtained from existing [blah] and have been previously described by [@Bryn] and [@Pttime paper?].
-
-### Sample processing
-
-Extraction of the environmental samples followed a modified Bligh & Dyer approach as detailed in [@BMIS].
-Briefly, filters were added to PTFE centrifuge tubes with a 1:1 mix of 100 µm and 400 µm silica beads, approximately 2mL -20 °C Optima-grade DCM, and approximately 3mL -20 °C 1:1 methanol/water solution (both also Optima-grade).
-Extraction standards were added during this step.
-The samples were then bead-beaten three times, followed by triplicate washes with fresh methanol/water mixture.
-Samples were then dried down under clean nitrogen gas and warmed using a Fisher-Scientific Reacti-Therm module.
-Dried aqueous fractions were re-dissolved in 380 µL of Optima-grade water and amended with 20 µL isotope-labeled injection standards.
-Additional internal standards were added at this point to measure the variability introduced by chromatography and ionization, and the reconstituted fraction was syringe-filtered to remove any potential clogging material.
-This aqueous fraction was then aliquoted into an HPLC vial for injection on the HILIC column and diluted 1:1 with Optima-grade water.
-A pooled sample was created by combining 20 µL of each sample into the same HPLC vial, and a 1:1 dilution with water half-strength sample was aliquot from that to assess matrix effects and obscuring variation [@BMIS].
-Also run alongside the environmental samples were two mixes of authentic standards in water and in an aliquot of the pooled sample at a variety of concentrations for quality control, annotation, and absolute concentration calculations.
-HPLC vials containing the samples were frozen at -80 °C until thawing shortly before injection.
-
-The CultureData samples were re-run from the frozen aliquots for this paper. The Pttime sample processing is documented [on MW? in their paper?].
-
-### LC conditions
-
-For the MESOSCOPE, Falkor, and CultureData samples a SeQuant ZIC-pHILIC column (5 um particle size, 2.1 mm x 150 mm, from Millipore) was used with 10 mM ammonium carbonate in 85:15 acetonitrile to water (Solvent A) and 10 mM ammonium carbonate in 85:15 water to acetonitrile (Solvent B) at a flow rate of 0.15 mL/min.
-The column was held at 100% A for 2 minutes, ramped to 64% B over 18 minutes, ramped to 100% B over 1 minute, held at 100% B for 5 minutes, and equilibrated at 100% A for 25 minutes (50 minutes total).
-The column was maintained at 30 °C.
-The injection volume was 2 µL for samples and standard mixes.
-When starting a batch, the column was equilibrated at the starting conditions for at least 30 minutes.
-To improve the performance of the HILIC column, we maintained the same injection volume, kept the instrument running water blanks between samples as necessary, and injected standards in a representative matrix (the pooled sample) in addition to standards in water.
-After each batch, the column was flushed with 10 mM ammonium carbonate in 85:15 water to acetonitrile for 20 to 30 minutes.
-LC conditions for the Pttime samples are documented [on MW?
-in their paper?].
-
-### MS conditions
-
-Environmental metabolomic data was collected on a Thermo Q Exactive HF hybrid Orbitrap (QE) mass spectrometer.
-The capillary and auxiliary gas heater temperatures were maintained at 320°C and 100°C, respectively.
-The S-lens RF level was kept at 65, the H-ESI voltage was set to 3.3 kV and sheath gas, auxiliary gas, and sweep gas flow rates were set at 16, 3, and 1, respectively.
-Polarity switching was used with a scan range of 60 to 900 m/z and a resolution of 60,000.
-Calibration was performed every 3-4 days at a target mass of 200 m/z.
-DDA data was collected from the pooled samples for high-confidence annotation of knowns and unknowns.
-All files were then converted to an open-source mzML format and centroided via Proteowizard’s msConvert tool.
-For the Pttime samples, files were pulled directly from Metabolomics Workbench via Project ID [blah] and used in their existing mzXML format.
-
-### Peakpicking, alignment, and grouping with XCMS
-
-The R package XCMS was used to perform peakpicking, retention time correction, and peak correspondence.
-Files were loaded and run separately for each dataset (Falkor, MESOSCOPE, CultureData, and Pttime) using the new "OnDiskMSnExp" infrastructure.
-Default parameters for the CentWave peakpicking algorithm were used except for: ppm, which was set to 5; peakwidth, which was widened to 20-80 seconds; prefilter, for which the intensity threshold was raised to $10^6$; and integrate, which was set to 2 instead of 1.
-snthresh was set to zero because there are known issues with background estimation in this algorithm [@myers2017], and both verboseColumns and the extendLengthMSW parameter were set to TRUE.
-For retention time correction, the Obiwarp method was used except for the CultureData dataset, which was visually inspected and determined not to require correction.
-For the Obiwarp algorithm, the binsize was reduced to 0.1 but all other parameters were left at their defaults or equivalents.
-
-Peak grouping was performed on the two environmental datasets and the Pttime data with a bandwidth of 12, a minFraction of 0.1, binSize of 0.001, and minSamples of 2 but otherwise default arguments.
-CultureData's minFraction was raised to 0.4 but was otherwise identical.
-Sample groups were constructed to consist of the biological replicates for all datasets.
-After peak grouping, peak filling was performed using the fillChromPeaks function with the ppm parameter set to 2.5.
-Finally, mass features with a retention time less than 30 seconds or larger than 20 minutes were removed to avoid interference from the initial and final solvent washes.
-[Check this with Laura]
-
-### Manual inspection and classification
-
-After the full XCMS workflow was completed, the molecular features were visually inspected by a single qualified MS expert.
-For the Falkor and MESOSCOPE datasets, every molecular feature was inspected, while only those features with a predicted probability of 0.9 or higher according to the two-parameter model produced below were for the CultureData and Pttime datasets.
-Inspection consisted of plotting the raw intensity values against the corrected retention-time values for all data points within the *m/z* x RT bounding box determined by the most extreme values for the given feature.
-For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labelling and to more accurately represent what MS experts typically do when assessing the quality of a given peak.
-[Fig 1.] We also decided to ignore missing values and linearly interpolate between known data points rather than filling with zeroes.
-These EICs were then shown to an MS expert for classification into one of 4 categories: Good, Bad, Ambiguous, or Stans only if the peak appeared to only show up in the standards.
-A few randomly-chosen features from the manually-assigned Good and Bad classifications are shown in [Fig.
-1].
-
-```{r grab demo msdata, eval=!file.exists("manuscript/minEIC_df.csv")}
-msdata <- readRDS("made_data_MS3000/msdata.rds")
-
-set.seed(123)
-chosen_feats <- MS3000_features %>%
-  filter(feat_class%in%c("Good", "Bad")) %>%
-  group_by(feat_class) %>%
-  slice_sample(n=5) %>%
-  arrange(feat_class) %>%
-  mutate(feat_id=row_number())
-minEIC_df <- pmap(chosen_feats, function(...){
-  with(list(...), {
-    msdata$EIC[mz%between%pmppm(mean_mz, ppm=5)] %>%
-      filter(rt%between%(mean_rt/60+c(-1, 1))) %>%
-      mutate(feature, feat_class, feat_id)
-  })
-}) %>% 
-  bind_rows() %>%
-  filter(str_detect(filename, "180821")) %>%
-  left_join(file_data, by="filename") %>%
-  filter(samp_type=="Smp")
-write_csv(minEIC_df, "manuscript/minEIC_df.csv")
-```
-
-```{r fig:some good some bad features}
-minEIC_df <- read_csv("manuscript/minEIC_df.csv")
-
-good_chroms_gp <- minEIC_df %>%
-  filter(feat_class=="Good") %>%
-  ggplot() +
-  geom_line(aes(x=rt, y=int, group=filename, color=depth)) +
-  facet_wrap(~feature, nrow = 1, scales = "free") +
-  scale_color_manual(breaks = c("15m", "DCM", "175m"), values = c("#f4bb23", "#028e34", "#0b505c")) +
-  theme_bw() +
-  theme(axis.text = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title = element_blank(),
-        legend.position = "none", 
-        panel.grid = element_blank()) +
-  guides(color=guide_legend(override.aes = list(linewidth=2), title = "Sample depth:  "))
-bad_chroms_gp <- minEIC_df %>%
-  filter(feat_class=="Bad") %>%
-  ggplot() +
-  geom_line(aes(x=rt, y=int, group=filename, color=depth)) +
-  facet_wrap(~feature, nrow = 1, scales = "free") +
-  scale_color_manual(breaks = c("15m", "DCM", "175m"), values = c("#f4bb23", "#028e34", "#0b505c")) +
-  theme_bw() +
-  theme(axis.text = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title = element_blank(),
-        legend.position = "none",
-        panel.grid = element_blank())
-chroms_legend <- cowplot::get_legend(good_chroms_gp + theme(legend.position = "bottom"))
-
-comb_cowplot <- cowplot::plot_grid(good_chroms_gp, bad_chroms_gp, chroms_legend, 
-                                   ncol = 1, rel_heights = c(1, 1, 0.15))
-ggsave(plot = comb_cowplot, 
-       filename = "somegood_somebad_depthcolor_chroms.png", 
-       device = "png", path = "manuscript/figures",
-       width = 6.5, height = 3, units = "in", dpi = 300)
-```
-
-![Figure 1: Randomly selected ion chromatograms from both "Good" (top row) and "Bad" (bottom row) manual classifications, colored by the depth from which the biological sample was taken.
-DCM = deep chlorophyll maximum, approximately 110 meters.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
-
-### Peak feature extraction and metric calculation
-
-Our process of feature engineering was divided into three categories of features: those that could be calculated using a single peak trace, those that required access to the entire file, and those that could only be calculated across multiple files.
-[Rewrite without the "three categories" business ugh] For metrics belonging to the first two categories, summary statistics were used (mean/median, standard deviation) to create a single consensus value for the mass feature as a whole.
-Missing values were ignored during these summary statistic calculations by setting na.rm=TRUE.
-Distributions were visually inspected and highly abnormal metrics were transformed using log-scaling if necessary.
-Using a pairs plot, highly correlated (above a Pearson's r ~ .9) had one of the correlated metrics removed.
-
-The first category consisted largely of information produced directly by XCMS: the very typical *m/z* ratios, retention times, peak widths, and integrated areas as well as the less common and poorly-documented sn, f, scale, and lmin parameters.
-Although we requested additional parameters from XCMS with the verboseColumns = TRUE argument, all of the values returned were NAs and could not be used in the model building so were dropped.
-The total number of peaks found was calculated as the number of peaks divided by the total number of files, and the total number of files in which a peak was initially detected was calculated by subtracting the number of NAs in the peak baseline estimate (which is NA if the peak was filled in) from the total number of files and then dividing by the number of files to normalize across datasets.
-This calculation were performed for the Falkor, MESOSCOPE, and CultureData datasets to estimate the proportion of sample and standard files in which a given peak was found, but could not be supplied for the Pttime dataset because no standard files were available.
-The sn metric contained a large number of zero values to begin with that then became negative infinities when log-scaled, so we replaced those values with zeroes and functionally equated an sn of 0 with an sn of 1.
-
-We also calculated several metrics from the raw mz/rt/int values.
-Using the R package RaMS we extracted the data points falling within each individual peak's mz and retention time bounding box (values between the XCMS-reported min and max) separately for each file.
-The data points were then linearly scaled to fall within the 0-1 range by subtracting the minimum RT and dividing by the maximum RT, then each scaled RT was fit to a beta distribution with α values of 2.5, 3, 4, and 5, and a fixed β value of 5.
-This approach allowed us to approximate a bell curve with increasing degrees of right-skewness and the beta distribution was chosen because it is constrained between 0 and 1 and simple and speedy to generate in R.
-For each α value, Pearson's correlation coefficient was calculated between the beta distribution and the raw data, with the highest value returned as a metric for how peak-shaped the data were [Fig.
-2].
-The beta distribution with the highest *r* was also then used to estimate the noise level within the peak by also scaling both the beta distribution densities and the raw data intensity values as described above, then subtracting the scaled beta distribution from the scaled intensity values, producing the residuals of the fit [Fig.
-2].
-The signal-to-noise ratio (SNR) was calculated by dividing the maximum original peak height by the standard deviation of the residuals multiplied by the maximum height of the original peak.
-This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional molecular signals.
-[@myers?] If there were fewer than 5 data points, a missing value was returned and dropped in subsequent summary calculations.
-Accessing the raw data values also allowed us to calculate the proportion of "missed" scans in a peak for which a retention time exists at other masses in the same sample but for which no data was produced at the selected *m/z* ratio.
-
-```{r how are raw peak shape metrics calculated}
-minEIC_df <- read_csv("manuscript/minEIC_df.csv")
-
-single_EIC <- minEIC_df %>%
-  filter(feature%in%c("FT0588", "FT0280")) %>%
-  filter(filename=="180821_Smp_MS10C215m_A.mzML")
-
-maybe_skews <- c(2.5, 3, 4, 5)
-skew_names <- c("Large skew", "Small skew", "Minimal skew", "Zero skew")
-names(maybe_skews) <- skew_names
-perf_peak_data <- single_EIC %>%
-  group_by(feature) %>%
-  select(feature, rt, int) %>%
-  mutate(rt=(rt-min(rt))/(max(rt)-min(rt))) %>%
-  mutate(int=(int-min(int))/(max(int)-min(int))) %>%
-  cbind(sapply(maybe_skews, function(x){
-    perf_peak <- dbeta(.$rt, shape1=x, shape2=5)
-    perf_peak/max(perf_peak)
-  })) %>%
-  rename(`Raw data`="int") %>%
-  mutate(feature=factor(feature, levels=c("FT0588", "FT0280"), 
-                        labels=c("Good peak", "Bad peak")))
-summary_stats <- perf_peak_data %>%
-  pivot_longer(-c(feature, rt, `Raw data`)) %>%
-  mutate(new_val=value-`Raw data`) %>%
-  group_by(feature, name) %>%
-  summarise(single_cor=round(cor(`Raw data`, value), 2), 
-            single_sd=round(sd(new_val), 2), 
-            .groups = "drop_last") %>%
-  mutate(name=factor(name, levels=skew_names)) %>%
-  mutate(x_coord=c(0.125+(as.numeric(name)-1)/4))
-
-peak_gp <- perf_peak_data %>%
-  pivot_longer(-c(rt, feature)) %>%
-  mutate(name=factor(name, levels=c("Raw data", skew_names))) %>%
-  ggplot() +
-  geom_line(aes(x=rt, y=value, color=name), lwd=1) +
-  geom_rect(xmin=-Inf, xmax=Inf, ymax=-Inf, ymin=-.05, fill="white", color="grey90") +
-  annotate("text", x=0.5, y=-Inf, vjust=-2, label="Pearson's r:") +
-  facet_wrap(~feature) +
-  geom_text(aes(x=x_coord, y=-Inf, label=single_cor, color=name),
-             data = summary_stats, vjust=-.5, show.legend = FALSE) +
-  scale_color_manual(breaks = c("Raw data", skew_names),
-                     values = c("black", viridisLite::plasma(6)[2:5])) +
-  scale_y_continuous(limits = c(-0.3, 1), breaks = c(0, 0.5, 1.0)) +
-  labs(y="Normalized intensity", color=NULL) +
-  theme_bw() +
-  theme(axis.text.x = element_blank(),
-        axis.title.x = element_blank(),
-        axis.ticks.x = element_blank(),
-        # strip.background = element_rect(fill="white"),
-        legend.position="none",
-        strip.text = element_text(size = 11))
-residuals_gp <- perf_peak_data %>%
-  pivot_longer(-c(rt, feature, `Raw data`)) %>%
-  mutate(new_val=value-`Raw data`) %>%
-  ggplot() +
-  geom_rect(xmin=-Inf, xmax=Inf, ymax=Inf, ymin=0.85, fill="white", color="grey90") +
-  geom_hline(yintercept = 0, lwd=1) +
-  geom_line(aes(x=rt, y=new_val, color=name), lwd=1) +
-  geom_text(aes(x=x_coord, y=Inf, label=single_sd, color=name),
-             data = summary_stats, vjust=3) +
-  annotate("text", x=0.5, y=Inf, vjust=1.5, label="Residual SD:") +
-  facet_grid(~feature) +
-  scale_color_manual(breaks = c("Raw data", skew_names),
-                     values = c("black", viridisLite::plasma(6)[2:5])) +
-  scale_y_continuous(breaks = c(-1, 0, 1), limits = c(-1.1, 1.3)) +
-  labs(x="Normalized retention time", y="Residuals", color=NULL) +
-  theme_bw() +
-  theme(strip.background = element_blank(),
-        strip.text = element_blank(),
-        legend.position = "none")
-
-fig_legend <- cowplot::get_legend(peak_gp + theme(legend.position="right") + 
-                                    guides(color=guide_legend(override.aes = list(lwd=2)), text="none"))
-
-comb_cowplot <- egg::ggarrange(peak_gp, residuals_gp, ncol = 1, draw = FALSE)
-
-comb_peakmetrics_gp <- cowplot::plot_grid(
-  comb_cowplot, fig_legend, rel_widths = c(1, 0.25)
-) + theme(plot.background = element_rect(fill = "white", color=NA))
-
-ggsave(plot = comb_peakmetrics_gp, 
-       filename = "peakmetrics_singlechrom.png", 
-       device = "png", path = "manuscript/figures",
-       width = 6.5, height = 4, units = "in", dpi = 300)
-```
-
-![Figure 2: Method used to calculate the metrics for the two-parameter model from the raw data via comparison to an idealized pseudo-Gaussian peak for both manually identified "Good" and "Bad" peaks.
-Normalization was performed by linearly scaling the raw values into the 0-1 range by subtracting the minimum value and dividing by the maximum.
-Peak shape similarity was measured with Pearson's correlation coefficient and the noise level is estimated as the standard deviation of the residuals after the raw data is subtracted from the idealized peak.](figures/peakmetrics_singlechrom.png){width=70%}
-
-We additionally estimated the existence of a $^{13}$C isotope using a similar method to extract the raw mz/rt/int values within the peak bounding box, then searched the same RT values at an *m/z* delta of +1.003355.
-In places where more than 5 data points existed at both the original mass and the $^{13}$C mass, we again used Pearson's correlation coefficient to estimate the similarity between the two mass traces and used a trapezoidal Riemann sum to estimate the area of the original and isotope peaks.
-The overall feature isotope shape similarity was calculated by taking the median of the correlation coefficients.
-We also calculated the correlation coefficient of the ratio of the $\frac{^{13}C}{^{12}C}$ across multiple files, expecting that a true isotope would have a fixed $\frac{^{13}C}{^{12}C}$ ratio.
-Both the isotope shape similarity and the isotope area correlation were used as metrics in the downstream analysis.
-Peaks for which no isotope signal was detected or had too few scans to calculate the above metrics were imputed with NA values that were again dropped in the calculation of summary statistics.
-
-Finally, we calculated several metrics using a design-of-experiments (DoE) approach for the Falkor and MESOSCOPE datasets which were expected to have a large difference in metabolite composition between those samples collected at the surface (15-25 meters) and those collected deeper in the water column.
-We used ANOVAs to compare the integrated peak area with depth and extracted the p-values of each test, which became an additional potential peak quality metric.
-We also calculated the difference between the mean sample area and the mean blank area and the difference between the mean sample area and the mean area of the standards which also became metrics.
-When there were too many missing values to perform the ANOVA or sample averages, we replaced the calculated statistic with a value a single order of magnitude outside the most extreme value.
-
-### Statistics
-
-We also tested whether the performance could be improved with elastic net regression or random forest models.
-These models handle correlated variables better than ordinary least squares regression, so we also included several additional implementations of the peak shape and novel SNR parameters when summarizing across multiple files, using a max and a median of the top-three best values rather than just the overall median as well as a log-transformed version of the median peak shape calculated as $median(log_{10}(1-r))$ where $r$ is Pearson's correlation coefficient, as described above (Figure 2).
-Cross-validation was used to select the optimal tuning parameter $\lambda$ with `glmnet` package's `cv.glmnet` for an elastic net penalty (α) of 0, 0.5, and 1.
-Random forests were implemented using the randomForest package with default settings and a factor-type response vector to ensure classification was used rather than regression.
-
-We explored the implications of applying this model to the MESOSCOPE dataset at a variety of thresholds.
-In univariate space, we used nonparametric Kruskal-Wallis analyses of variance to measure the difference between the surface (15m), DCM (~110m), and 175m samples because the metabolite peak areas could not be assumed to be normally distributed.
-These univariate tests were then controlled for multiply hypothesis testing using R's `p.adjust` function with method `fdr`.[@ref] We also performed post-hoc Dunn tests provided by the `rstatix` package to categorize the response to depth for those mass features for which the KW test was significant, with responses falling into one of the 14 classes possible when permuting the sign and significance of the Dunn test.[@ref] p-values obtained from the Dunn tests were not FDR controlled because it was used as a categorization tool rather than a null hypothesis test.
-In multivariate space, we used a permutational MANOVA (PERMANOVA) [@ref] provided by the vegan package's `adonis2` function to test for multivariate differences in structure of the metabolome with depth.[@ref] We ran multiple PERMANOVAs with a different subset of mass features included each time, corresponding to using the output from XCMS directly, likelihood thresholds of 0.01, 0.1, 0.5, 0.9, and finally only those peaks manually annotated as good.
-
-All analyses were run in R, version 4.2.2, and code is available on the Github at https://github.com/wkumler/MS_metrics.[@ref]
+This allows us to connect the physical, chemical, and biological measurements taken regularly around the globe to a molecular-scale perspective of particulate organic matter in the ocean by linking the chemical currencies that fuel the planet to the environments in which they're found.
 
@@
 
+## Methods
+
+### Sample collection
+
+Environmental samples were collected from the North Pacific Subtropical Gyre near Station ALOHA during two research cruises that targeted strong mesoscale eddy features during June/July 2017 and March/April 2018, traversing an area between 28 °N, 156 °W and 23 °N, 161 °W.
+An eddy dipole off the coast of Hawaii was detected using sea-level anomaly (SLA) satellite data and targeted for both a transect across the cyclonic and anticyclonic poles of the eddy dipole.
+The cyclonic pole of the eddy had a maximum negative SLA anomaly of -15 cm in 2017 and -20 cm in 2018, while the anticyclonic center reached +24 cm in 2017 and +21 cm in 2018.
+The 2017 cruise samples were taken along a transect across the eddy dipole while the 208 cruise targeted only the center of each eddy.
+
+Environmental samples were obtained using the onboard CTD rosette to collect water from 15 meters, the deep chlorophyll maximum (DCM), and 175 meters during the 2017 MESOSCOPE cruise and from 25 meters and the DCM during the 2018 Falkor cruise.
+The DCM was determined visually from fluorometer data during the CTD downcast and Niskin bottles were tripped during the return trip to the surface.
+Seawater from each depth was sampled in triplicate by firing one Niskin bottle for each sample.
+Samples were brought to the surface and decanted into prewashed (3x with DI, 3x with sampled seawater) polycarbonate bottles for filtration.
+Samples were filtered by peristaltic pump onto 142mm 0.2 µm Durapore filters held by polycarbonate filter holders on a Masterflex tubing line.
+Pressures were kept as low as possible while still producing a reasonable rate of flow through the filter, approximately 250-500 mL per minute.
+Samples were then removed from the filter holder using solvent-washed tweezers and placed into pre-combusted aluminum foil packets that were then flash-frozen in liquid nitrogen before being stored at -80 °C until extraction.
+A methodological blank was also collected by running filtrate through a new filter and then treated identically to the samples.
+
+Culture samples used as the validation sets for this paper were obtained from existing [blah] and have been previously described by [@Bryn] and [@Pttime paper?].
+
+### Sample processing
+
+Extraction of the environmental samples followed a modified Bligh & Dyer approach as detailed in [@BMIS].
+Briefly, filters were added to PTFE centrifuge tubes with a 1:1 mix of 100 µm and 400 µm silica beads, approximately 2mL -20 °C Optima-grade DCM, and approximately 3mL -20 °C 1:1 methanol/water solution (both also Optima-grade).
+Extraction standards were added during this step.
+The samples were then bead-beaten three times, followed by triplicate washes with fresh methanol/water mixture.
+Samples were then dried down under clean nitrogen gas and warmed using a Fisher-Scientific Reacti-Therm module.
+Dried aqueous fractions were re-dissolved in 380 µL of Optima-grade water and amended with 20 µL isotope-labeled injection standards.
+Additional internal standards were added at this point to measure the variability introduced by chromatography and ionization, and the reconstituted fraction was syringe-filtered to remove any potential clogging material.
+This aqueous fraction was then aliquoted into an HPLC vial for injection on the HILIC column and diluted 1:1 with Optima-grade water.
+A pooled sample was created by combining 20 µL of each sample into the same HPLC vial, and a 1:1 dilution with water half-strength sample was aliquot from that to assess matrix effects and obscuring variation [@BMIS].
+Also run alongside the environmental samples were two mixes of authentic standards in water and in an aliquot of the pooled sample at a variety of concentrations for quality control, annotation, and absolute concentration calculations.
+HPLC vials containing the samples were frozen at -80 °C until thawing shortly before injection.
+
+The CultureData samples were re-run from the frozen aliquots for this paper. The Pttime sample processing is documented [on MW? in their paper?].
+
+### LC conditions
+
+For the MESOSCOPE, Falkor, and CultureData samples a SeQuant ZIC-pHILIC column (5 um particle size, 2.1 mm x 150 mm, from Millipore) was used with 10 mM ammonium carbonate in 85:15 acetonitrile to water (Solvent A) and 10 mM ammonium carbonate in 85:15 water to acetonitrile (Solvent B) at a flow rate of 0.15 mL/min.
+The column was held at 100% A for 2 minutes, ramped to 64% B over 18 minutes, ramped to 100% B over 1 minute, held at 100% B for 5 minutes, and equilibrated at 100% A for 25 minutes (50 minutes total).
+The column was maintained at 30 °C.
+The injection volume was 2 µL for samples and standard mixes.
+When starting a batch, the column was equilibrated at the starting conditions for at least 30 minutes.
+To improve the performance of the HILIC column, we maintained the same injection volume, kept the instrument running water blanks between samples as necessary, and injected standards in a representative matrix (the pooled sample) in addition to standards in water.
+After each batch, the column was flushed with 10 mM ammonium carbonate in 85:15 water to acetonitrile for 20 to 30 minutes.
+LC conditions for the Pttime samples are documented [on MW?
+in their paper?].
+
+### MS conditions
+
+Environmental metabolomic data was collected on a Thermo Q Exactive HF hybrid Orbitrap (QE) mass spectrometer.
+The capillary and auxiliary gas heater temperatures were maintained at 320°C and 100°C, respectively.
+The S-lens RF level was kept at 65, the H-ESI voltage was set to 3.3 kV and sheath gas, auxiliary gas, and sweep gas flow rates were set at 16, 3, and 1, respectively.
+Polarity switching was used with a scan range of 60 to 900 m/z and a resolution of 60,000.
+Calibration was performed every 3-4 days at a target mass of 200 m/z.
+DDA data was collected from the pooled samples for high-confidence annotation of knowns and unknowns.
+All files were then converted to an open-source mzML format and centroided via Proteowizard’s msConvert tool.
+For the Pttime samples, files were pulled directly from Metabolomics Workbench via Project ID [blah] and used in their existing mzXML format.
+
+### Peakpicking, alignment, and grouping with XCMS
+
+The R package XCMS was used to perform peakpicking, retention time correction, and peak correspondence.
+Files were loaded and run separately for each dataset (Falkor, MESOSCOPE, CultureData, and Pttime) using the new "OnDiskMSnExp" infrastructure.
+Default parameters for the CentWave peakpicking algorithm were used except for: ppm, which was set to 5; peakwidth, which was widened to 20-80 seconds; prefilter, for which the intensity threshold was raised to $10^6$; and integrate, which was set to 2 instead of 1.
+snthresh was set to zero because there are known issues with background estimation in this algorithm [@myers2017], and both verboseColumns and the extendLengthMSW parameter were set to TRUE.
+For retention time correction, the Obiwarp method was used except for the CultureData dataset, which was visually inspected and determined not to require correction.
+For the Obiwarp algorithm, the binsize was reduced to 0.1 but all other parameters were left at their defaults or equivalents.
+
+Peak grouping was performed on the two environmental datasets and the Pttime data with a bandwidth of 12, a minFraction of 0.1, binSize of 0.001, and minSamples of 2 but otherwise default arguments.
+CultureData's minFraction was raised to 0.4 but was otherwise identical.
+Sample groups were constructed to consist of the biological replicates for all datasets.
+After peak grouping, peak filling was performed using the fillChromPeaks function with the ppm parameter set to 2.5.
+Finally, mass features with a retention time less than 30 seconds or larger than 20 minutes were removed to avoid interference from the initial and final solvent washes.
+[Check this with Laura]
+
+### Manual inspection and classification
+
+After the full XCMS workflow was completed, the molecular features were visually inspected by a single qualified MS expert.
+For the Falkor and MESOSCOPE datasets, every molecular feature was inspected, while only those features with a predicted probability of 0.9 or higher according to the two-parameter model produced below were for the CultureData and Pttime datasets.
+Inspection consisted of plotting the raw intensity values against the corrected retention-time values for all data points within the *m/z* x RT bounding box determined by the most extreme values for the given feature.
+For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labelling and to more accurately represent what MS experts typically do when assessing the quality of a given peak.
+[Fig 1.] We also decided to ignore missing values and linearly interpolate between known data points rather than filling with zeroes.
+These EICs were then shown to an MS expert for classification into one of 4 categories: Good, Bad, Ambiguous, or Stans only if the peak appeared to only show up in the standards.
+A few randomly-chosen features from the manually-assigned Good and Bad classifications are shown in [Fig.
+1].
+
+```{r grab demo msdata, eval=!file.exists("manuscript/minEIC_df.csv")}
+msdata <- readRDS("made_data_MS3000/msdata.rds")
+
+set.seed(123)
+chosen_feats <- MS3000_features %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  group_by(feat_class) %>%
+  slice_sample(n=5) %>%
+  arrange(feat_class) %>%
+  mutate(feat_id=row_number())
+minEIC_df <- pmap(chosen_feats, function(...){
+  with(list(...), {
+    msdata$EIC[mz%between%pmppm(mean_mz, ppm=5)] %>%
+      filter(rt%between%(mean_rt/60+c(-1, 1))) %>%
+      mutate(feature, feat_class, feat_id)
+  })
+}) %>% 
+  bind_rows() %>%
+  filter(str_detect(filename, "180821")) %>%
+  left_join(file_data, by="filename") %>%
+  filter(samp_type=="Smp")
+write_csv(minEIC_df, "manuscript/minEIC_df.csv")
+```
+
+```{r fig:some good some bad features}
+minEIC_df <- read_csv("manuscript/minEIC_df.csv")
+
+good_chroms_gp <- minEIC_df %>%
+  filter(feat_class=="Good") %>%
+  ggplot() +
+  geom_line(aes(x=rt, y=int, group=filename, color=depth)) +
+  facet_wrap(~feature, nrow = 1, scales = "free") +
+  scale_color_manual(breaks = c("15m", "DCM", "175m"), values = c("#f4bb23", "#028e34", "#0b505c")) +
+  theme_bw() +
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        axis.title = element_blank(),
+        legend.position = "none", 
+        panel.grid = element_blank()) +
+  guides(color=guide_legend(override.aes = list(linewidth=2), title = "Sample depth:  "))
+bad_chroms_gp <- minEIC_df %>%
+  filter(feat_class=="Bad") %>%
+  ggplot() +
+  geom_line(aes(x=rt, y=int, group=filename, color=depth)) +
+  facet_wrap(~feature, nrow = 1, scales = "free") +
+  scale_color_manual(breaks = c("15m", "DCM", "175m"), values = c("#f4bb23", "#028e34", "#0b505c")) +
+  theme_bw() +
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        axis.title = element_blank(),
+        legend.position = "none",
+        panel.grid = element_blank())
+chroms_legend <- cowplot::get_legend(good_chroms_gp + theme(legend.position = "bottom"))
+
+comb_cowplot <- cowplot::plot_grid(good_chroms_gp, bad_chroms_gp, chroms_legend, 
+                                   ncol = 1, rel_heights = c(1, 1, 0.15))
+ggsave(plot = comb_cowplot, 
+       filename = "somegood_somebad_depthcolor_chroms.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 3, units = "in", dpi = 300)
+```
+
+![Figure 1: Randomly selected ion chromatograms from both "Good" (top row) and "Bad" (bottom row) manual classifications, colored by the depth from which the biological sample was taken.
+DCM = deep chlorophyll maximum, approximately 110 meters.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
+
+### Peak feature extraction and metric calculation
+
+Our process of feature engineering was divided into three categories of features: those that could be calculated using a single peak trace, those that required access to the entire file, and those that could only be calculated across multiple files.
+[Rewrite without the "three categories" business ugh] For metrics belonging to the first two categories, summary statistics were used (mean/median, standard deviation) to create a single consensus value for the mass feature as a whole.
+Missing values were ignored during these summary statistic calculations by setting na.rm=TRUE.
+Distributions were visually inspected and highly abnormal metrics were transformed using log-scaling if necessary.
+Using a pairs plot, highly correlated (above a Pearson's r ~ .9) had one of the correlated metrics removed.
+
+The first category consisted largely of information produced directly by XCMS: the very typical *m/z* ratios, retention times, peak widths, and integrated areas as well as the less common and poorly-documented sn, f, scale, and lmin parameters.
+Although we requested additional parameters from XCMS with the verboseColumns = TRUE argument, all of the values returned were NAs and could not be used in the model building so were dropped.
+The total number of peaks found was calculated as the number of peaks divided by the total number of files, and the total number of files in which a peak was initially detected was calculated by subtracting the number of NAs in the peak baseline estimate (which is NA if the peak was filled in) from the total number of files and then dividing by the number of files to normalize across datasets.
+This calculation were performed for the Falkor, MESOSCOPE, and CultureData datasets to estimate the proportion of sample and standard files in which a given peak was found, but could not be supplied for the Pttime dataset because no standard files were available.
+The sn metric contained a large number of zero values to begin with that then became negative infinities when log-scaled, so we replaced those values with zeroes and functionally equated an sn of 0 with an sn of 1.
+
+We also calculated several metrics from the raw mz/rt/int values.
+Using the R package RaMS we extracted the data points falling within each individual peak's mz and retention time bounding box (values between the XCMS-reported min and max) separately for each file.
+The data points were then linearly scaled to fall within the 0-1 range by subtracting the minimum RT and dividing by the maximum RT, then each scaled RT was fit to a beta distribution with α values of 2.5, 3, 4, and 5, and a fixed β value of 5.
+This approach allowed us to approximate a bell curve with increasing degrees of right-skewness and the beta distribution was chosen because it is constrained between 0 and 1 and simple and speedy to generate in R.
+For each α value, Pearson's correlation coefficient was calculated between the beta distribution and the raw data, with the highest value returned as a metric for how peak-shaped the data were [Fig.
+2].
+The beta distribution with the highest *r* was also then used to estimate the noise level within the peak by also scaling both the beta distribution densities and the raw data intensity values as described above, then subtracting the scaled beta distribution from the scaled intensity values, producing the residuals of the fit [Fig.
+2].
+The signal-to-noise ratio (SNR) was calculated by dividing the maximum original peak height by the standard deviation of the residuals multiplied by the maximum height of the original peak.
+This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional molecular signals.
+[@myers?] If there were fewer than 5 data points, a missing value was returned and dropped in subsequent summary calculations.
+Accessing the raw data values also allowed us to calculate the proportion of "missed" scans in a peak for which a retention time exists at other masses in the same sample but for which no data was produced at the selected *m/z* ratio.
+
+```{r how are raw peak shape metrics calculated}
+minEIC_df <- read_csv("manuscript/minEIC_df.csv")
+
+single_EIC <- minEIC_df %>%
+  filter(feature%in%c("FT0588", "FT0280")) %>%
+  filter(filename=="180821_Smp_MS10C215m_A.mzML")
+
+maybe_skews <- c(2.5, 3, 4, 5)
+skew_names <- c("Large skew", "Small skew", "Minimal skew", "Zero skew")
+names(maybe_skews) <- skew_names
+perf_peak_data <- single_EIC %>%
+  group_by(feature) %>%
+  select(feature, rt, int) %>%
+  mutate(rt=(rt-min(rt))/(max(rt)-min(rt))) %>%
+  mutate(int=(int-min(int))/(max(int)-min(int))) %>%
+  cbind(sapply(maybe_skews, function(x){
+    perf_peak <- dbeta(.$rt, shape1=x, shape2=5)
+    perf_peak/max(perf_peak)
+  })) %>%
+  rename(`Raw data`="int") %>%
+  mutate(feature=factor(feature, levels=c("FT0588", "FT0280"), 
+                        labels=c("Good peak", "Bad peak")))
+summary_stats <- perf_peak_data %>%
+  pivot_longer(-c(feature, rt, `Raw data`)) %>%
+  mutate(new_val=value-`Raw data`) %>%
+  group_by(feature, name) %>%
+  summarise(single_cor=round(cor(`Raw data`, value), 2), 
+            single_sd=round(sd(new_val), 2), 
+            .groups = "drop_last") %>%
+  mutate(name=factor(name, levels=skew_names)) %>%
+  mutate(x_coord=c(0.125+(as.numeric(name)-1)/4))
+
+peak_gp <- perf_peak_data %>%
+  pivot_longer(-c(rt, feature)) %>%
+  mutate(name=factor(name, levels=c("Raw data", skew_names))) %>%
+  ggplot() +
+  geom_line(aes(x=rt, y=value, color=name), lwd=1) +
+  geom_rect(xmin=-Inf, xmax=Inf, ymax=-Inf, ymin=-.05, fill="white", color="grey90") +
+  annotate("text", x=0.5, y=-Inf, vjust=-2, label="Pearson's r:") +
+  facet_wrap(~feature) +
+  geom_text(aes(x=x_coord, y=-Inf, label=single_cor, color=name),
+             data = summary_stats, vjust=-.5, show.legend = FALSE) +
+  scale_color_manual(breaks = c("Raw data", skew_names),
+                     values = c("black", viridisLite::plasma(6)[2:5])) +
+  scale_y_continuous(limits = c(-0.3, 1), breaks = c(0, 0.5, 1.0)) +
+  labs(y="Normalized intensity", color=NULL) +
+  theme_bw() +
+  theme(axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        axis.ticks.x = element_blank(),
+        # strip.background = element_rect(fill="white"),
+        legend.position="none",
+        strip.text = element_text(size = 11))
+residuals_gp <- perf_peak_data %>%
+  pivot_longer(-c(rt, feature, `Raw data`)) %>%
+  mutate(new_val=value-`Raw data`) %>%
+  ggplot() +
+  geom_rect(xmin=-Inf, xmax=Inf, ymax=Inf, ymin=0.85, fill="white", color="grey90") +
+  geom_hline(yintercept = 0, lwd=1) +
+  geom_line(aes(x=rt, y=new_val, color=name), lwd=1) +
+  geom_text(aes(x=x_coord, y=Inf, label=single_sd, color=name),
+             data = summary_stats, vjust=3) +
+  annotate("text", x=0.5, y=Inf, vjust=1.5, label="Residual SD:") +
+  facet_grid(~feature) +
+  scale_color_manual(breaks = c("Raw data", skew_names),
+                     values = c("black", viridisLite::plasma(6)[2:5])) +
+  scale_y_continuous(breaks = c(-1, 0, 1), limits = c(-1.1, 1.3)) +
+  labs(x="Normalized retention time", y="Residuals", color=NULL) +
+  theme_bw() +
+  theme(strip.background = element_blank(),
+        strip.text = element_blank(),
+        legend.position = "none")
+
+fig_legend <- cowplot::get_legend(peak_gp + theme(legend.position="right") + 
+                                    guides(color=guide_legend(override.aes = list(lwd=2)), text="none"))
+
+comb_cowplot <- egg::ggarrange(peak_gp, residuals_gp, ncol = 1, draw = FALSE)
+
+comb_peakmetrics_gp <- cowplot::plot_grid(
+  comb_cowplot, fig_legend, rel_widths = c(1, 0.25)
+) + theme(plot.background = element_rect(fill = "white", color=NA))
+
+ggsave(plot = comb_peakmetrics_gp, 
+       filename = "peakmetrics_singlechrom.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 4, units = "in", dpi = 300)
+```
+
+![Figure 2: Method used to calculate the metrics for the two-parameter model from the raw data via comparison to an idealized pseudo-Gaussian peak for both manually identified "Good" and "Bad" peaks.
+Normalization was performed by linearly scaling the raw values into the 0-1 range by subtracting the minimum value and dividing by the maximum.
+Peak shape similarity was measured with Pearson's correlation coefficient and the noise level is estimated as the standard deviation of the residuals after the raw data is subtracted from the idealized peak.](figures/peakmetrics_singlechrom.png){width=70%}
+
+We additionally estimated the existence of a $^{13}$C isotope using a similar method to extract the raw mz/rt/int values within the peak bounding box, then searched the same RT values at an *m/z* delta of +1.003355.
+In places where more than 5 data points existed at both the original mass and the $^{13}$C mass, we again used Pearson's correlation coefficient to estimate the similarity between the two mass traces and used a trapezoidal Riemann sum to estimate the area of the original and isotope peaks.
+The overall feature isotope shape similarity was calculated by taking the median of the correlation coefficients.
+We also calculated the correlation coefficient of the ratio of the $\frac{^{13}C}{^{12}C}$ across multiple files, expecting that a true isotope would have a fixed $\frac{^{13}C}{^{12}C}$ ratio.
+Both the isotope shape similarity and the isotope area correlation were used as metrics in the downstream analysis.
+Peaks for which no isotope signal was detected or had too few scans to calculate the above metrics were imputed with NA values that were again dropped in the calculation of summary statistics.
+
+Finally, we calculated several metrics using a design-of-experiments (DoE) approach for the Falkor and MESOSCOPE datasets which were expected to have a large difference in metabolite composition between those samples collected at the surface (15-25 meters) and those collected deeper in the water column.
+We used ANOVAs to compare the integrated peak area with depth and extracted the p-values of each test, which became an additional potential peak quality metric.
+We also calculated the difference between the mean sample area and the mean blank area and the difference between the mean sample area and the mean area of the standards which also became metrics.
+When there were too many missing values to perform the ANOVA or sample averages, we replaced the calculated statistic with a value a single order of magnitude outside the most extreme value.
+
+### Statistics
+
+We also tested whether the performance could be improved with elastic net regression or random forest models.
+These models handle correlated variables better than ordinary least squares regression, so we also included several additional implementations of the peak shape and novel SNR parameters when summarizing across multiple files, using a max and a median of the top-three best values rather than just the overall median as well as a log-transformed version of the median peak shape calculated as $median(log_{10}(1-r))$ where $r$ is Pearson's correlation coefficient, as described above (Figure 2).
+Cross-validation was used to select the optimal tuning parameter $\lambda$ with `glmnet` package's `cv.glmnet` for an elastic net penalty (α) of 0, 0.5, and 1.
+Random forests were implemented using the randomForest package with default settings and a factor-type response vector to ensure classification was used rather than regression.
+
+We explored the implications of applying this model to the MESOSCOPE dataset at a variety of thresholds.
+In univariate space, we used nonparametric Kruskal-Wallis analyses of variance to measure the difference between the surface (15m), DCM (~110m), and 175m samples because the metabolite peak areas could not be assumed to be normally distributed.
+These univariate tests were then controlled for multiply hypothesis testing using R's `p.adjust` function with method `fdr`.[@ref] We also performed post-hoc Dunn tests provided by the `rstatix` package to categorize the response to depth for those mass features for which the KW test was significant, with responses falling into one of the 14 classes possible when permuting the sign and significance of the Dunn test.[@ref] p-values obtained from the Dunn tests were not FDR controlled because it was used as a categorization tool rather than a null hypothesis test.
+In multivariate space, we used a permutational MANOVA (PERMANOVA) [@ref] provided by the vegan package's `adonis2` function to test for multivariate differences in structure of the metabolome with depth.[@ref] We ran multiple PERMANOVAs with a different subset of mass features included each time, corresponding to using the output from XCMS directly, likelihood thresholds of 0.01, 0.1, 0.5, 0.9, and finally only those peaks manually annotated as good.
+
+All analyses were run in R, version 4.2.2, and code is available on the Github at https://github.com/wkumler/MS_metrics.[@ref]
+
 ## Acknowledgements

### 2023-05-27 manuscript/manuscript.Rmd
Updating abstract to implement Anitra's comments on it

@@
 Uncontrolled false discovery rates and the lack of manually-calibrated quality metrics require researchers to visually evaluate individual peaks, requiring large amounts of time and breaking replicability.
-Here, we manually labeled two oceanographic particulate metabolite datasets to assess the performance of individual peak quality metrics and constructed a predictive model calibrated to the likelihood that visual inspection by an MS expert would include a given mass feature in the downstream analysis.
+This problem is exacerbated in noisy environmental datasets and for novel separation methods such as hydrophilic interaction columns in metabolomics, creating a demand for a simple, intuitive, and robust metric of peak quality.
 
@@
 
-A simple logistic regression model built on two metrics calculated from the raw MS data reduced the fraction of false positives in the analysis from 80-90% down to 1-5% and showed minimal overfitting when applied to novel datasets.
+Here, we manually labeled four HILIC oceanographic particulate metabolite datasets to assess the performance of individual peak quality metrics and constructed a predictive model calibrated to the likelihood that visual inspection by an MS expert would include a given mass feature in the downstream analysis.
+We found that a simple logistic regression model built on two metrics calculated from the raw MS data reduced the fraction of false positives in the analysis from 80-90% down to 1-5% and showed minimal overfitting when applied to novel datasets.
 We then explored the implications of this quality thresholding on the conclusions obtained by the downstream analysis, concluding that the poor performance of peakpicking algorithms significantly reduces the power of both univariate and multivariate statistical analyses in detecting environmental differences.
-We finally show that depth is a major driver of variability in the metabolome in high-quality environmental metabolomics datasets and identify several interesting metabolite groups for future investigation.
+We finally show that depth is a major driver of variability in the metabolome and identify several interesting metabolite groups for future investigation.
 

### 2023-05-29 manuscript/manuscript.Rmd
Refactoring results to be results only, part 1

@@
 
-An area particularly ripe for LC-MS analysis is that of the open ocean.
+An area particularly ripe for metabolomic analysis is that of the open ocean.
 Low compound and high salt concentrations make this area difficult to study [@BMIS] but its vast size and direct effect on the Earth's carbon cycling make it critical that we understand its dynamics of energy and nutrients on a molecular scale.
-Metabolites are the currency of chemical exchange both intra- and inter-cellularly, serving as building blocks of larger molecules, regulators of osmotic balance and storage of nutrients, as well as chemical signals on their own.
+Metabolites are the currency of chemical exchange both intra- and inter-cellularly, serving as building blocks of larger molecules, regulators of osmotic balance and storage of nutrients, as well as important chemical signals on their own.
 These small molecules serve both as signposts for the complex biological landscape in this highly dynamic region and give a sense of not only who is present but also what ecological roles they're serving and the niches they fill.
@@
 
-We used three different multiple logistic regression models to predict the likelihood of each MF being categorized as "Good".
-The first model included all metrics calculated as described above in Methods, the second contained only those parameters immediately available from the XCMS output without revisiting the raw data (the four core peak metrics *m/z*, RT, peak width, area and their standard deviations plus the mysterious lmin, f, and scale values as well as the fraction of peaks, samples, and standards found), and the final model was a simple two-parameter model using only the peak shape and novel SNR metrics.
-In all three cases, the majority of MFs were estimated to have a less than 1% chance of being good.
-The full model and the XCMS model both displayed a strongly bimodal distribution, with a large number of peaks also exceeding a 99% chance of being good, while the two-parameter model had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Fig.
+According to all three logistic regressions (see Methods), the majority of MFs were estimated to have a less than 1% chance of being good.
+The full model (containing all possible peak metrics) and the XCMS model (built on only those metrics calculated from the XCMS output) both displayed a strongly bimodal distribution, with a large number of peaks also exceeding a 99% chance of being good, while the two-parameter model (consisting of the novel SNR metric and the peak shape correlation metric) had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Fig.
 3).
@@
 
-In each case, we determined each molecular feature as a true positive (TP) if it was predicted to be Good and was manually classified as Good, a true negative if both predicted and classifed as Bad, a false positive if predicted to be Good but manually classified as Bad, and a false negative if predicted to be Bad but was in fact manually classified as Good.
-This allowed us to additionally define two useful measures of success, the traditionally-defined false discovery rate (FDR, defined as 1-precision or the number of false positives divided by the total number of predicted positives) and the percentage of good peaks found (GPF, also known as the recall or sensitivity and defined as the number of true positives divided by the total number of positive predictions).
-These two measures allowed us to monitor two key questions typically asked of a predictive model in this context: first, what fraction of the data predicted to be good is actually noise; and second, what fraction of the good data am I actually recovering in the downstream analysis?
-
 The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when [predicting] using the full dataset, with FDR values in the 5-10% range and 80-90% GPF values implying that a large majority of the good peaks passed the threshold.
@@
 However, when the models were trained on a different dataset than the one they were used to predict classifications for, nearly all models had similar performance with %FDR around 10-25 and %GPF around 60-80.
-This dramatic drop in performance on out-of-training data strongly implied that the full and XCMS-only models were highly overfit, while the two-parameter model had probably reached stability.
-Also notably, the model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more peaks recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE [and it's unclear why this should be the case] (Figure 4).
+The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more peaks recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE.
+(Figure 4)
 
@@
 
-To further explore questions of model stability and the potential for overfitting, we compared the predictions from a Falkor-trained model to a MESOSCOPE-trained model.
-In the ideal situation, these predictions are the same because the values are independent of a given training set's quirks, but this was not what we found for the full and XCMS-only models (Figure 5).
-We compared the absolute likelihood values using Pearson's correlation coefficient and found that the two-parameter models had the highest $r$ value of 0.996, while the full models and the XCMS-trained models had $r$ values of 0.856 and 0.896, respectively (Figure 5).
-Finally, we compared the model likelihoods in ranked space with Spearman's correlation coefficient to test the question of whether or not the *ranking* was similar even if the absolute values differed.
-Here, we found an intensification of the effect described before, with an even higher $r$ for the two-parameter model of 0.998 but lower $r$ values for the full and XCMS-trained model of 0.818 and 0.849, respectively (Figure 5).
+We found that the predictions made from a Falkor-trained dataset consistently differed from a MESOSCOPE-trained dataset for the full and XCMS-only models.
+In the raw probability space, the two-parameter models had the highest $r$ value of 0.996, while the full models and the XCMS-trained models had $r$ values of 0.856 and 0.896, respectively.
+When compared in ranked space, we found an intensification of the effect described previously, with an even higher $r$ for the two-parameter model of 0.998 but lower $r$ values for the full and XCMS-trained model of 0.818 and 0.849, respectively (Figure 5).
 
@@
 
-We also looked at the estimates produced by the two models and compared them with the combined model trained on the combined datasets to assess the model stability directly.
-A majority of the time, the estimates from the two models disagreed by more than 2x the standard error of the estimate and frequently disagreed by that amount depending on the other parameters included in the model.
+A majority of the time, the estimates from the two models disagreed by more than 2x the standard error of the estimate.
 Some parameters disagreed not only in magnitude but also in sign, with the Falkor-trained full model increasing peak goodness likelihood with larger ppm variation and a wider peak width, while the MESOSCOPE-trained full model had negative estimates for each of these parameters.
-Notably, the peak shape and novel SNR parameters in the two-parameter model were among the most robust to training model variation, potentially explaining the consistency described above (Supp.
+Notably, the peak shape and novel SNR parameters used in the two-parameter model were among the most robust to training model variation, potentially explaining the consistency described above (Supp.
 figure 1).
 
-We also tested the robustness of the model under a smaller training set, emulating a situation in which only a fraction of the data was available or only a portion of the mass features had been labeled.
-This allowed us to test the required sample size for the different models, with a larger sample size presumably required for the models with more parameters.
-Because no parameter was present in all 3 models, we looked at the top 2 most significant parameters from each model: average *m/z* and peak shape for the full model, average peakwidth and the standard deviation in retention time for the XCMS model, and peak shape and SNR for the two-parameter model.
-For the XCMS and two-parameter model, we found reasonably good convergence in a dataset containing half the mass features with most values falling within two standard errors of the estimate, while the full model required closer to 80% of the mass features to produce estimates consistent with the original model (Supp.
+When testing model stability under a smaller sample size, we found reasonably good convergence in a dataset containing half the mass features with most values falling within two standard errors of the estimate for the XCMS and two-parameter model, while the full model required closer to 80% of the mass features to produce estimates consistent with the original model (Supp.
 figure 2).
 
+### Regularized regression and random forests perform about the same
+
+None of the penalized regression models significantly improved cross-validated performance between the MESOSCOPE and Falkor datasets when measured by both initial performance and the performance drop when applied across datasets.
+All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GPF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent false negative acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp.
+figure 3).
+The random forest model, interestingly, had perfect predictive capacity when tested internally on the training data (FDR=0%, GPF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp.
+figure 3).
+In each case, the performance drop when applied to a novel dataset was more extreme than the simple two-parameter model described above.
+
 ### Performance of a stricter threshold on novel datasets
 
-Given the reasonably compelling evidence above favoring the two-parameter model on novel datasets, we applied this trained model on two additional datasets that differed significantly from the training data.
-The CultureData samples were produced in the Ingalls lab like the MESOSCOPE and Falkor samples, but represent data from a variety of phytoplankton and bacterial cultures in fresh and salt water rather than environmental samples.
-The Pttime dataset was discovered on Metabolomics Workbench [@MW id] and consists of Phaeodactylum triconutum cultures collected at a variety of timepoints from both pelleted cells and the released exudate.
-This dataset was chosen because of the similar LC-MS setup used as a benchmark comparison for the performance other labs with similar setups may expect to achieve.
-Each of these datasets were only fractionally labeled, with those peaks above the 0.9 likelihood threshold according to the two-parameter model reviewed manually and categorized.
-
-This stricter threshold was chosen because we felt less comfortable interpreting results based on mass features that were only 50% likely to be real, but did not feel the need to be so strict with this exploratory analysis that we wanted to limit it to 99+% likelihood peaks.
-This threshold also struck a nice balance between the number of peaks we estimated to be necessary for robust testing while still remaining reasonable to manually label.
+We settled on a 90% likelihood threshold for application to novel datasets because it struck a nice balance between the number of peaks we estimated to be necessary for robust testing while still remaining reasonable to manually label.
 For the CultureData dataset, we obtained 1,790 total mass features, 192 of which had predicted likelihoods above 0.9.
@@
 No standards were run during this analysis, so there were no features in that category.
-[There exists, however, some uncontrolled confirmation bias here in the manual labelling because the MS expert knew that these peaks had already passed some quality threshold and were more likely to be good than not.]
 
 With the stricter threshold, we obtained FDR rates consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for the Falkor, MESOSCOPE, CultureData, and Pttime respectively [Figure 6].
-Of course, this low error rate required that we miss out on additional valuable features, with only a fraction of the total good peaks making it past this threshold.
+Of course, this low error rate required that we miss out on additional potentially valuable features, with only a fraction of the total good peaks making it past this threshold.
 In both the Falkor and MESOSCOPE datasets, the fraction of good peaks found was below 40% with actual values of 39.4% and 26.5%, respectively.
@@
 
-### Regularized regression and random forests perform about the same
-
-None of the advanced ML models significantly improved performance when cross-validated between the MESOSCOPE and Falkor datasets when measured by both initial performance and the performance drop when applied across datasets.
-All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GPF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent Type II error acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp.
-figure 3).
-The random forest model had perfect predictive capacity when tested internally on the training data (FDR=0%, GPF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp.
-figure 3).
-In each case, the performance drop when applied to a novel dataset was more extreme than the simple two-parameter model described above.
-
 ### Implications for biological conclusions
@@
 #### Univariate techniques
-
-[Figure 7] shows the distribution of all mass features identified by XCMS and their categorization according to both the biochemical response to depth as well as the classifications assigned by the MS expert and the two-parameter model at a 50% likelihood threshold.
-
-The abundance of noise peaks located at the center of the plot and therefore lacking much biological trend with depth validates the conventional heuristic of relying on downstream statistical tests to separate signal from noise (with the assumption that noise will be fixed across all samples and unaffected by environmental structuring of samples).
-However, it is also clear that there are both some good peaks that do not have trends with depth (and are therefore interesting for bucking the biomass signal) as well as a good number of bad peaks that do change across sample depths.
-These bad peaks that have biological signals appear to be frequently due to the tail of a peak being integrated as a separate molecular feature, which then retains the biological signal of the actual feature.
-
-Although the identification of bad peaks predominantly improves univariate analysis by removing those features from consideration in individual molecular narratives, filtering out the bad peaks should also improve the power of the individual analyses by reducing the number of hypotheses that must then be corrected for later.
-Multiple hypothesis testing corrections rely on the number and shape of the p-value distribution to make estimates of the "true" p-values for each compound, and removing a large number of noise peaks will result in lower p-values for those peaks that are flagged as good.
-
-Most metabolites track with biomass and are consistently abundant through the surface and DCM but drop off at depth, as prior measurements of particulate carbon and beam attenuation have shown [@benedetto].
-However, several groups of compounds diverge intriguingly from this bulk biomass trend, with some molecules found almost exclusively at the surface and others found to increase with depth rather than decrease.
-Several mass features that were found dominantly at the surface were putatively identified as mycosporine-like amino acids (MAAs), which have been shown to have UV photoprotective properties.[@ref] Although we were unable to calculate their exact concentrations due to a lack of authentic standards, peak areas were approximately 100 to 1000 times higher at 15 meters than their values averaged between the DCM and 175 meter samples.
-The abundance of MAAs in this particular category indicates that there may be other MAA molecules with this depth pattern that have not yet been characterized also among these unknowns.
 
@@
 
+A majority of the features (1,323 of 1,832 total) in the MESOSCOPE dataset had no significant trend with depth, with FDR-controlled Kruskal-Wallis p-values exceeding 0.05 (Figure 7).
+The largest category that did have a trend with depth was the 15m = DCM > 175m category, containing 118 molecular features, with largest peak areas distributed evenly between the 15 meter and deep chlorophyll maximum (DCM) samples, while the 175 meter samples had significantly smaller areas.
+The similar but statistically distinct categories of 15m > DCM > 175m and DCM > 15m > 175m had 68 and 35 features, respectively, and together indicate that many molecules are highly abundant throughout the surface ocean, down to the DCM layer and decrease in concentration thereafter.
+A surprising number of features were also found to have DCM minima (DCM < 15m = 175m, 26 features) or linear increases with depth (15m < DCM < 175m, 12 features) given the few environmental parameters that have these trends (Figure 7).
+
+A different story emerged, however, when the bad peaks were removed from this analysis.
+Good features were most commonly found to have their highest concentrations at the DCM or the surface, rather than being fixed with respect to depth.
+Of the 182 good peaks, less than a fifth had no trend with depth (44/249) and a majority had unequivocally lowest values in the 175 meter samples (those with 15m/DCM > 175m, 145 features).
+The two-parameter model, when applied with a 50% likelihood threshold, also recovered this general feature distribution and classified many of the features with no significant depth signal as likely to be bad (Figure 7B, currently in Slack?).
+
+Additionally, a large number of features manually identified as bad nonetheless had significant differences with depth.
+This was surprising because we had assumed that bad peaks corresponded to instrument noise, which we did not expect to have any biological trend.
+Further investigation of a few randomly selected bad features with a biological difference revealed the reason behind this: most of those investigated were actually tails of other peaks.
+Integrating just the tail of a peak retains the biological signal of the full peak while still looking visually like instrument noise, thereby introducing pseudoreplication in the feature space.
+
+The model did fail to recover some interesting biological variation, however.
+Two features of particular interest were those good peaks with a DCM minimum (DCM < 15m = 175m), both of which were missed by the two-parameter model.
+These features possess an unexpected biological signal that does not track with depth or other common oceanographic parameters, thereby potentially representing an interesting biomarker or pathway that's downregulated in excess of the biomass effect.
+[Maybe better for discussion?]
+
+[MAA paragraph?] Several mass features that were found dominantly at the surface were putatively identified as mycosporine-like amino acids (MAAs), which have been shown to have UV photoprotective properties.[@ref] Although we were unable to calculate their exact concentrations due to a lack of authentic standards, peak areas were approximately 100 to 1000 times higher at 15 meters than their values averaged between the DCM and 175 meter samples.
+The abundance of MAAs in this particular category indicates that there may be other MAA molecules with this depth pattern that have not yet been characterized also among these unknowns.
+
 ```{r clockplot}
@@
 
-Multivariate statistics also benefit from the reduced FDR when applying the two-parameter model.
+Multivariate statistics also benefitted from the reduced FDR when applying the two-parameter model.
 For the PERMANOVAs, we found that the proportion of variance explained (R$^2$) and the pseudo-F statistic increased monotonically with the likelihood threshold used to subset the data (Table 1).
@@
 
-The relative power of identifying only the very best peaks was also illustrated visually with non-metric multidimensional scaling (NMDS) plots (Supp.
-figure 4).
+The relative power of identifying only the very best peaks was also illustrated visually with non-metric multidimensional scaling (NMDS) plots (Figure XX).
 In these common exploratory plots, the features with likelihoods above 50% strongly separated by depth while lower thresholds disguised the true signal and had higher stress values.
 Performing an NMDS on the manually-identified Good peaks resulted in output nearly indistinguishable from those of the 90% and 50% thresholds, while the 0.1 and 0.01 thresholds looked almost identical to the same analysis performed on all the XCMS features.
+
+```{r NMDS results}
+thresh_NMDS_df <- imap(mat_list, function(mat_given, mat_name){
+  mat_data <- mat_given %>%
+    select(-feat_class, -pred_prob) %>%
+    data.matrix() %>%
+    t()
+  nmds_output <- vegan::metaMDS(mat_data, k = 2, distance = "manhattan", trace = 0,
+                         autotransform = FALSE, noshare = FALSE, wascores = FALSE)
+  nmds_gp <- nmds_output$points %>%
+    as.data.frame() %>%
+    cbind(MS3000_metadata) %>%
+    rownames_to_column("filename") %>%
+    mutate(mat_name) %>%
+    mutate(stress=nmds_output$stress)
+}) %>% 
+  bind_rows() %>%
+  select(mat_name, filename, starts_with("MDS"), depth, stress) %>%
+  mutate(basename=str_extract(filename, "MS\\d+C\\d+.*(?=_[A-C])")) %>%
+  mutate(mat_name=factor(
+    mat_name, levels=c("mat_good", "mat_90", "mat_50", "mat_10", "mat_01", "mat_all"),
+    labels=c("Only good", paste("Threshold", c(0.9, 0.5, 0.1, 0.01)), "All features")
+  )) %>%
+  mutate(stress=format(stress, digits=2))
+thresh_NMDS_gp <- thresh_NMDS_df %>%
+  ggplot() +
+  geom_vline(xintercept = 0) +
+  geom_hline(yintercept = 0) +
+  geom_polygon(aes(x=MDS1, y=MDS2, fill=depth, color=depth, group=basename), alpha=0.5) +
+  geom_label(aes(label=stress), x=Inf, y=Inf, 
+             label.r = unit(0, "in"), hjust=1, vjust=1,
+             data=thresh_NMDS_df %>% distinct(mat_name, stress)) +
+  facet_wrap(~mat_name, scales = "free", ncol=3) +
+  scale_color_manual(values=viridis::viridis(9, direction = -1)[c(2,5,8)], 
+                     breaks = c("15m", "DCM", "175m"),
+                     aesthetics = c("color", "fill")) +
+  theme_bw() +
+  theme(aspect.ratio = 1)
+ggsave(plot = thresh_NMDS_gp, filename = "thresh_NMDS_plots.png", 
+       device = "png", path = "manuscript/figures",
+       width = 6.5, height = 4, units = "in", dpi = 300)
+```
+
+![Figure XX: NMDS plots of metabolite similarity according to sample depth across multiple likelihood thresholds.
+Triplicate samples are represented by the vertices of the triangles and colored by the depth from which they were sampled (DCM = deep chlorophyll maximum, ~110m).
+"Only good" refers to those features manually labeled as "Good".
+NMDS stress values are reported to 3 significant figures in the upper right of the plot.](figures/thresh_NMDS_plots.png){width=90%}
 
@@
 
-### Statistics
+### Regressions and model development
+
+We used three different multiple logistic regression models to predict the likelihood of each MF being categorized as "Good".
+The first model included all metrics calculated as described above in Methods, the second contained only those parameters immediately available from the XCMS output without revisiting the raw data (the four core peak metrics *m/z*, RT, peak width, area and their standard deviations plus the mysterious lmin, f, and scale values as well as the fraction of peaks, samples, and standards found), and the final model was a simple two-parameter model using only the peak shape and novel SNR metrics.
+
+In each case, we categorized each molecular feature as a true positive (TP) if it was predicted to be Good and was manually classified as Good, a true negative if both predicted and classifed as Bad, a false positive if predicted to be Good but manually classified as Bad, and a false negative if predicted to be Bad but was in fact manually classified as Good.
+This allowed us to additionally define two useful measures of success, the traditionally-defined false discovery rate (FDR, defined as 1-precision or the number of false positives divided by the total number of predicted positives) and the percentage of good peaks found (GPF, also known as the recall or sensitivity and defined as the number of true positives divided by the total number of positive predictions).
+
+To further explore questions of model stability and the potential for overfitting, we compared the predictions from a Falkor-trained model to a MESOSCOPE-trained model.
+This comparison was done in both the raw probability space as well as a rank-ordered space to test whether the most extreme likelihood (i.e.
+very best and very worst) peaks were consistently found to be most extreme independently of the actual likelihood predicted.
+For the raw probability space we compared the predictions using Pearson's correlation coefficient, while Spearman's rank-ordered coefficient was used for the ranked space.
+We also looked at the estimates produced by these two models and compared them with the combined model trained on both datasets combined to assess the model stability directly.
+
+We also tested the robustness of the model under a smaller training set, emulating a situation in which only a fraction of the data was available or only a portion of the mass features had been labeled.
+This allowed us to test the required sample size for the different models, with a larger sample size presumably required for the models with more parameters.
+Because no parameter was present in all 3 models, we looked at the top 2 most significant parameters from each model: average *m/z* and peak shape for the full model, average peakwidth and the standard deviation in retention time for the XCMS model, and peak shape and SNR for the two-parameter model.
 
@@
 Random forests were implemented using the randomForest package with default settings and a factor-type response vector to ensure classification was used rather than regression.
+
+### Application of the model to novel datasets
+
+After exploring the different models described above and determining that the two-parameter model would likely perform most consistently on novel datasets, we applied this trained model on two additional datasets that differed significantly from the training data.
+The CultureData dataset was produced in the Ingalls lab like MESOSCOPE and Falkor, but represent data from a variety of phytoplankton and bacterial cultures in fresh and salt water rather than environmental samples.
+
+The Pttime dataset was discovered on Metabolomics Workbench [@MW id] and consists of *Phaeodactylum tricornutum* cultures collected at a variety of timepoints from both pelleted cells and the released exudate.
+This dataset was chosen because of the similar LC-MS setup used as a benchmark comparison for the performance other labs with similar setups may expect to achieve using the trained model directly.
+Each of these datasets were only fractionally labeled, with those peaks above the 0.9 likelihood threshold according to the two-parameter model reviewed manually and categorized.
+
+This stricter threshold was chosen because we felt less comfortable interpreting results based on mass features that were only 50% likely to be real, but did not feel the need to be so strict with this exploratory analysis that we wanted to limit it to 99+% likelihood peaks.
+
+### Using variable thresholds to determine effects on biological conclusions
 
@@
 ![Supplemental figure 3: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.](supplement/regregrf_fdrgpf_cross_train.png){width=90%}
-
-```{r NMDS results}
-thresh_NMDS_df <- imap(mat_list, function(mat_given, mat_name){
-  mat_data <- mat_given %>%
-    select(-feat_class, -pred_prob) %>%
-    data.matrix() %>%
-    t()
-  nmds_output <- vegan::metaMDS(mat_data, k = 2, distance = "manhattan", trace = 0,
-                         autotransform = FALSE, noshare = FALSE, wascores = FALSE)
-  nmds_gp <- nmds_output$points %>%
-    as.data.frame() %>%
-    cbind(MS3000_metadata) %>%
-    rownames_to_column("filename") %>%
-    mutate(mat_name) %>%
-    mutate(stress=nmds_output$stress)
-}) %>% 
-  bind_rows() %>%
-  select(mat_name, filename, starts_with("MDS"), depth, stress) %>%
-  mutate(basename=str_extract(filename, "MS\\d+C\\d+.*(?=_[A-C])")) %>%
-  mutate(mat_name=factor(
-    mat_name, levels=c("mat_good", "mat_90", "mat_50", "mat_10", "mat_01", "mat_all"),
-    labels=c("Only good", paste("Threshold", c(0.9, 0.5, 0.1, 0.01)), "All features")
-  )) %>%
-  mutate(stress=format(stress, digits=2))
-thresh_NMDS_gp <- thresh_NMDS_df %>%
-  ggplot() +
-  geom_vline(xintercept = 0) +
-  geom_hline(yintercept = 0) +
-  geom_polygon(aes(x=MDS1, y=MDS2, fill=depth, color=depth, group=basename), alpha=0.5) +
-  geom_label(aes(label=stress), x=Inf, y=Inf, 
-             label.r = unit(0, "in"), hjust=1, vjust=1,
-             data=thresh_NMDS_df %>% distinct(mat_name, stress)) +
-  facet_wrap(~mat_name, scales = "free", ncol=3) +
-  scale_color_manual(values=viridis::viridis(9, direction = -1)[c(2,5,8)], 
-                     breaks = c("15m", "DCM", "175m"),
-                     aesthetics = c("color", "fill")) +
-  theme_bw() +
-  theme(aspect.ratio = 1)
-ggsave(plot = thresh_NMDS_gp, filename = "thresh_NMDS_plots.png", 
-       device = "png", path = "manuscript/supplement",
-       width = 6.5, height = 4, units = "in", dpi = 300)
-```
-
-![Supplemental figure 4: NMDS plots of metabolite similarity according to sample depth across multiple likelihood thresholds.
-Triplicate samples are represented by the vertices of the triangles and colored by the depth from which they were sampled (DCM = deep chlorophyll maximum, ~110m).
-"Only good" refers to those features manually labeled as "Good".
-NMDS stress values are reported to 3 significant figures in the upper right of the plot.](supplement/thresh_NMDS_plots.png){width=90%}

### 2023-06-01 manuscript/manuscript.Rmd
Adding latest edits, references, most of new discussion

@@
   pdf_document: default
+bibliography: refs.bib
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, eval=FALSE)
 knitr::opts_knit$set(root.dir = '..')
@@
 
-Here, we manually labeled four HILIC oceanographic particulate metabolite datasets to assess the performance of individual peak quality metrics and constructed a predictive model calibrated to the likelihood that visual inspection by an MS expert would include a given mass feature in the downstream analysis.
-We found that a simple logistic regression model built on two metrics calculated from the raw MS data reduced the fraction of false positives in the analysis from 80-90% down to 1-5% and showed minimal overfitting when applied to novel datasets.
-We then explored the implications of this quality thresholding on the conclusions obtained by the downstream analysis, concluding that the poor performance of peakpicking algorithms significantly reduces the power of both univariate and multivariate statistical analyses in detecting environmental differences.
-We finally show that depth is a major driver of variability in the metabolome and identify several interesting metabolite groups for future investigation.
+Here, we manually labeled four HILIC oceanographic particulate metabolite datasets to assess the performance of individual peak quality metrics.
+We used these datasets to construct a predictive model calibrated to the likelihood that visual inspection by an MS expert would include a given mass feature in the downstream analysis.
+We found that a simple logistic regression model built on two metrics calculated from the raw MS data reduced the fraction of false positives in the analysis from 70-80% down to 1-5% and showed minimal overfitting when applied to novel datasets.
+We then explored the implications of this quality thresholding on the conclusions obtained by the downstream analysis and found that while only 10% of the variance in the dataset could be explained by depth in the default output from the peakpicker, approximately 40% of the variance was explained when restricted to high-quality peaks alone.
 
@@
 
-TBD
+We conclude that the poor performance of peakpicking algorithms significantly reduces the power of both univariate and multivariate statistical analyses to detect environmental differences.
+We demonstrate that simple models built on intuitive metrics and derived from the raw data are more robust and can outperform more complex models when applied to new data.
+Finally, we show that in properly curated datasets, depth is a major driver of variability in the marine microbial metabolome and identify several interesting metabolite trends for future investigation.
 
@@
 In the past two decades, data-driven methods have established workflows for untargeted metabolomics but the imperfect performance of the core peakpicking algorithms continue to require manual oversight and curation.
-This problem has been exacerbated by the increased use of non-traditional chromatography such as hydrophilic interaction which tends to produce noisier peaks.
+This problem has been exacerbated by the increased use of non-traditional chromatography such as hydrophilic interaction which tends to produce noisier peaks.[@Bajad2006; @Myers2017; @Gika2019]
 
 Noisy data and imperfect detection algorithms introduce a tradeoff between false positives (where background instrument or chemical noise is misclassified as biological signal) and false negatives (where real signals are undetected).
-Existing algorithms favor false positives because downstream analyses can always remove erroneous mass features, but false negatives cannot be later recovered.
-However, this approach creates two additional downstream problems.
+Existing algorithms tend to favor false positives because downstream analyses can always remove erroneous mass features, but false negatives cannot be later recovered.[@Pirttila2022; @Gloaguen2022] However, this approach creates two additional downstream problems.
 First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of peaks, a task that scales combinatorially with the number of samples and compounds measured.
@@
 Existing peak-detection softwares do not provide a clear way to do this.
-Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each molecular feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew.
-None of these parameters answer the critical question about the likelihood that a given feature corresponds to a molecule present in the original sample.
+Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each molecular feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew.[@Pirttila2022] None of these parameters answer the critical question about the likelihood that a given feature corresponds to a molecule present in the original sample.
 This parameter is crucial for downstream analysis because it represents the base rate for error propagation and acceptable thresholds should vary widely by the particular project's goals.
 In an exploratory analysis, any molecular feature more than 50% likely to be real is perhaps worth considering, while in a confirmatory study this threshold may need to be above 99%.
-While significant effort has been invested in improving the peakpicking algorithms, very little has been done to quantify the accuracy and precision of their outputs across the wide variety of datasets to which they are applied.
+Despite significant effort invested in improving the peakpicking algorithms, very little has been done to quantify the accuracy and precision of their outputs across the wide variety of datasets to which they are applied.
 
@@
 This metric would improve statistical power by reducing the number of effective hypotheses tested and allow researchers to focus effort on features least likely to be noise.
-Additionally, this parameter could be optimized to improve peakpicking and chromatographic settings independently of the software used or inter-lab variability when scripted to provide consistent, reproducible results independent of the particular expert reviewing its performance.
-Constructing such a single comprehensive metric calibrated to likelihood is also more effective than multiple independent thresholds because it has meaningful units, does not require estimating the relative power of individual metrics, and allows a good peak to compensate for weak performance in one area with strong performance in other metrics.
+Additionally, this parameter could be optimized to improve peakpicking and chromatographic settings independently of the software used and minimize inter-lab variability when scripted to provide consistent, reproducible results independent of the particular expert reviewing its performance.
+Constructing such a single comprehensive metric calibrated to likelihood is also more effective than multiple independent thresholds because it has meaningful units, does not require estimating the relative power of individual metrics, and allows a good peak to compensate for weak performance in one area with strong performance in other metrics, e.g.
+as implemented in @Pirttila2022.
 
@@
 
-An area particularly ripe for metabolomic analysis is that of the open ocean.
-Low compound and high salt concentrations make this area difficult to study [@BMIS] but its vast size and direct effect on the Earth's carbon cycling make it critical that we understand its dynamics of energy and nutrients on a molecular scale.
-Metabolites are the currency of chemical exchange both intra- and inter-cellularly, serving as building blocks of larger molecules, regulators of osmotic balance and storage of nutrients, as well as important chemical signals on their own.
-These small molecules serve both as signposts for the complex biological landscape in this highly dynamic region and give a sense of not only who is present but also what ecological roles they're serving and the niches they fill.
+An area particularly ripe for metabolomic analysis is that of the open ocean.[@KidoSoule2015] Low compound and high salt concentrations make this area difficult to study but its vast size and direct effect on the Earth's biogeochemistry make it critical that we understand its dynamics of energy and nutrients on a molecular scale.[@Boysen2018] Metabolites are the currency of chemical exchange both intra- and inter-cellularly, serving as building blocks of larger molecules, regulators of osmotic balance and storage of nutrients, as well as important chemical signals on their own.
+These small molecules serve both as signposts for the complex biological landscape in this highly dynamic region and give a sense of not only who is present but also what ecological roles they're serving and the niches they fill.[@KidoSoule2015, @Boysen2021, @Heal2021]
 
@@
 
-An average of 3,300 molecular features (MFs) were reported by XCMS across the 4 datasets, with the fewest (1495) in the Falkor data and the most in the Pttime samples (7781).
+An average of 3,300 molecular features (MFs) were reported by XCMS across the 4 datasets, with the fewest (1,495) in the Falkor data and the most in the Pttime samples (7,781).
 In the Falkor and MESOSCOPE datasets that were fully labeled by an MS expert, approximately 70% (69% and 73%, respectively) of the features were given a "Bad" designation, corresponding to noise peaks that the expert would not have included in a downstream analysis.
@@
 
-According to all three logistic regressions (see Methods), the majority of MFs were estimated to have a less than 1% chance of being good.
+According to all three logistic regression models (see Methods), the majority of MFs were estimated to have a less than 1% chance of being good.
 The full model (containing all possible peak metrics) and the XCMS model (built on only those metrics calculated from the XCMS output) both displayed a strongly bimodal distribution, with a large number of peaks also exceeding a 99% chance of being good, while the two-parameter model (consisting of the novel SNR metric and the peak shape correlation metric) had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Fig.
@@
 
-The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when [predicting] using the full dataset, with FDR values in the 5-10% range and 80-90% GPF values implying that a large majority of the good peaks passed the threshold.
+The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR values in the 5-10% range and 80-90% GPF values implying that a large majority of the good peaks passed the threshold with very little noise included.
 The XCMS metrics performed slightly worse, with FDR values in the 10-15% range and GPF values closer to 75%.
 The two-parameter model performed worst when tested internally, with an FDR of about 20% and %GPF also around 75% (Figure 4).
-However, when the models were trained on a different dataset than the one they were used to predict classifications for, nearly all models had similar performance with %FDR around 10-25 and %GPF around 60-80.
-The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more peaks recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE.
-(Figure 4)
+However, when the models were trained on a different dataset than the one they were used to predict classifications for, all models had similar performance with %FDR around 10-25 and %GPF around 60-80.
+The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more peaks recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE (Figure 4).
 
@@
 
-### Model stability under different training sets and subset of labels
+### Model stability under different training sets
 
@@
 
-We settled on a 90% likelihood threshold for application to novel datasets because it struck a nice balance between the number of peaks we estimated to be necessary for robust testing while still remaining reasonable to manually label.
+We settled on a 90% likelihood threshold for application to novel datasets because it struck a balance between the number of peaks we estimated to be necessary for robust testing while still remaining reasonable to manually label.
 For the CultureData dataset, we obtained 1,790 total mass features, 192 of which had predicted likelihoods above 0.9.
@@
 
-With the stricter threshold, we obtained FDR rates consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for the Falkor, MESOSCOPE, CultureData, and Pttime respectively [Figure 6].
+With the stricter threshold, we obtained FDR values consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for Falkor, MESOSCOPE, CultureData, and Pttime respectively [Figure 6].
 Of course, this low error rate required that we miss out on additional potentially valuable features, with only a fraction of the total good peaks making it past this threshold.
-In both the Falkor and MESOSCOPE datasets, the fraction of good peaks found was below 40% with actual values of 39.4% and 26.5%, respectively.
+In both the Falkor and MESOSCOPE datasets, less than half of the good peaks were labeled as such, with actual values of 39.4% and 26.5%, respectively.
 Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the %GPF but expect it to be in a similar range (Figure 6).
@@
 
-A majority of the features (1,323 of 1,832 total) in the MESOSCOPE dataset had no significant trend with depth, with FDR-controlled Kruskal-Wallis p-values exceeding 0.05 (Figure 7).
+A majority of the features (1,323 of 1,832 total) in the original, non-thresholded MESOSCOPE dataset had no significant trend with depth, with FDR-controlled Kruskal-Wallis p-values exceeding 0.05 (Figure 7).
 The largest category that did have a trend with depth was the 15m = DCM > 175m category, containing 118 molecular features, with largest peak areas distributed evenly between the 15 meter and deep chlorophyll maximum (DCM) samples, while the 175 meter samples had significantly smaller areas.
-The similar but statistically distinct categories of 15m > DCM > 175m and DCM > 15m > 175m had 68 and 35 features, respectively, and together indicate that many molecules are highly abundant throughout the surface ocean, down to the DCM layer and decrease in concentration thereafter.
+The similar but statistically distinct categories of 15m > DCM > 175m and DCM > 15m > 175m had 68 and 35 features, respectively, and together indicate that many molecules are highly abundant throughout the surface ocean down to the DCM layer and decrease in concentration thereafter.
 A surprising number of features were also found to have DCM minima (DCM < 15m = 175m, 26 features) or linear increases with depth (15m < DCM < 175m, 12 features) given the few environmental parameters that have these trends (Figure 7).
@@
 
-[MAA paragraph?] Several mass features that were found dominantly at the surface were putatively identified as mycosporine-like amino acids (MAAs), which have been shown to have UV photoprotective properties.[@ref] Although we were unable to calculate their exact concentrations due to a lack of authentic standards, peak areas were approximately 100 to 1000 times higher at 15 meters than their values averaged between the DCM and 175 meter samples.
+[MAA paragraph?] Several mass features that were found dominantly at the surface were putatively identified as mycosporine-like amino acids (MAAs), which have been shown to have UV photoprotective properties.[@ref] Although we were unable to calculate their exact concentrations due to a lack of authentic standards, peak areas were approximately 100 to 1000 times higher in the 15 meter samples than the values averaged between the DCM and 175 meter samples.
 The abundance of MAAs in this particular category indicates that there may be other MAA molecules with this depth pattern that have not yet been characterized also among these unknowns.
@@
 
-In each test, the permutational p-value obtained was less than 0.001, indicating that the differences between samples due to depth were unlikely to be due to chance in all cases.
+In each test, the permutational p-value obtained was less than 0.001, indicating that the differences between samples due to depth were never likely to be due entirely to chance.
 However, the pseudo-F was much larger with higher thresholds, scaling from around 8.5 when thresholding at a 1% likelihood to 42 when thresholded at a 90% likelihood.
@@
 
-### Simple logistic regression models show good performance and minimal overfitting
-
-  - Recap performance metrics
-    - Discuss the relative performance and compare to assumptions in lit for
-      - Large peak area
-      - Has an isotope
-      - p-values
-    - Discuss use of summary statistics and the potential need for a blocked model by sample type
-  - Emphasize the best performance of the two-parameter model
-    - Full model and XCMS model definitely overfit, same for regularized/random forest
-    - XCMS alone does not provide enough information to accurately distinguish good from noise
-      - Likely true for all peakpickers
-      - Plea: report mz/rt bounding box for individual peaks so additional metrics can be calculated
-    - These params very simple (and fast!) to estimate
-      - Easy to do better with fancier models (e.g. loess smoothing for SNR)
-    - Why do all the models seem to approach ~25% FDR and ~75% GPF on the novel datasets?
-  - Limited to HILIC data at the moment but does seem to perform well across labs
-
-### Unable to create strict control of the FDR but a single threshold still makes life easier
-
-  - Initial goal: set an FDR, choose a likelihood threshold based on it
-    - Not doable here but maybe in future work?
-    - Pivoted to recommending step-down manual labelling
-  - Plea: report your FDR! Especially if you're manually labelling!
-    - Necessary for error propagation anyway
-  - Future work: alter chromatography/peakpicking params to maximize # of peaks above 0.9 threshold
-
-### Biological results do depend on the threshold used
-
-  - Large amounts of noise reduce the power of both univariate and multivariate stats
-    - Trade-off with loss of interesting signals at higher thresholds (cmpds on clockplot left)
-    - Possible to get wildly different results with different thresholds (% variance explained)
-    - Manual evaluation for interesting signals still strongly recommended
-  - Depth is a major driver of metabolome composition in the gyre
-  - Majority of metabolites track with biomass metrics
-  - Some interesting metabolites diverge from dominant biomass trend
-    - Some new MAAs in clockplot top left? (Still need to check this w MSMS)
-    - Degradation products / bacterial specialties at bottom?
+We used two fully-labeled and two partially-labeled HILIC LC-MS datasets to assess the performance of the XCMS algorithm and construct a robust model of peak quality.
+To measure performance, we used two measurements of success closely tied to intuitive questions about a dataset: the percentage of good peaks found (GPF, also known as the recall or sensitivity) and the percentage of bad peaks included or false discovery rate (FDR).
+We also chose these metrics because they are independent of the number of true negatives detected.
+This was a concern because we did not want to create a situation in which the algorithm is rewarded for artificially increasing the number of overall bad peaks, only to later remove them.
+Similarly, we decided against using the F$_1$ score as an overall summary statistic because false negatives and false positives have very different implications in this context and should be treated separately.
+
+One of the major ways in which this manuscript differs from prior work is its focus on summary statistics calculated across multiple files.[@Pirttila2022; @Guo2021; @Muller2020] Most existing peakpicking literature uses the single-file EIC peak as the core bit of training data, but that approach ignores critical information obtained elsewhere in the MS run that can change the judgement made on a single chromatogram.
+Features that are high quality are typically represented in multiple files, especially in QC pooled samples, and a feature that only has a peak in a single file is typically regarded as highly suspect, if not removed entirely.
+Classifying an entire feature at once not only has the advantage of reducing the amount of manual labor by a factor equal to the number of files in the run (typically 10s-100s smaller) but is also a better representation of the judgement made by an MS expert.
+Of course, this approach inherits all the difficulties of summary statistic calculations.
+The authors spent many hours discussing the implications of differences between mean and median values, but in general this approach produced more confidence than any single-file EIC.
+
+### Two-parameter logistic regression model with raw data metrics showed the most reliable performance
+
+These two measures of success were then used to explore different types of classification models for separating good molecular features (MFs) from bad, with a particular focus on quantifying the likelihood of each class rather than just returning the label.
+We found that a simple two-parameter logistic regression model trained on two novel metrics of peak quality had reasonably good performance on the training set and was highly robust when applied to novel datasets.
+
+The logistic regression in particular was favored over the random forest and regularized regression we tried due to their similar performance and increased interpretability (Supp.
+figure XX).
+In the literature, more complex models have also shown similar or better performance but typically require larger training sets and are fundamentally uninterpretable.[@Pirttila2022 and references therein]
+
+```{r, eval=FALSE}
+FT2040_features %>%
+  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
+  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  # with(table(feat_class, pred_class))
+  with(caret::confusionMatrix(as.factor(pred_class), as.factor(feat_class)))
+# Falkor if everything was just classified as "Bad"
+(980+56)/(980+56+45+196) # 0.824
+
+MS3000_features %>%
+  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
+  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  # with(table(feat_class, pred_class))
+  with(caret::confusionMatrix(as.factor(pred_class), as.factor(feat_class)))
+# MESOSCOPE if everything was just classified as "Bad"
+(1547+36)/(1547+36+67+182) # 0.864
+```
+
+@Guo2021 (EVA): Accuracy of 90-95% but did not distinguish between false positives and false negatives.
+Likely driven by the large number of bad peaks - here, our accuracy with a threshold of 0.5 was 92.1% for Falkor and 94.4% for MESOSCOPE.
+
+Accuracy is a problematic metric because if just predicting everything as "Bad" was 82.4% and 86.4%, respectively.
+Might be slightly facetious, but this was essentially what happened for a good number of the individual predictors - none of the smoothing curves exceeded 50% likelihood (Supp figure XX).
+
+This is why we chose to measure GPF and FDR instead of total accuracy. Not directly comparable datasets though and we've definitely got the advantage of using a half-internal training set.
+
+@Melnikov2020 (peakonly): Accuracy of 89%, precision of 97% and a recall of 89%
+Versus our precision (PPV) of 95.6% and recall (sensitivity) of 94.6% for Falkor
+Versus our precision (PPV) of 95.9% and recall (sensitivity) of 97.7% for MESOSCOPE
+@Gloaguen2022 (NeatMS): Claims universally superior performance to peakonly
+
+These deep learning methods in particular are challenging because they do not produce overall quality scores and the tradeoff between false positives and false negatives is fixed regardless of experimental priorities.
+
+The two metrics in the final model were derived from the raw EIC data rather than any summary statistic calculated by XCMS automaticallyand were implemented because they matched our intuition about what makes an MF look good to an MS expert.
+It is worth noting that these are very simple metrics and therefore fast to calculate, but we expect that more complicated metrics could perform even better.
+For example, using the data within the peak boundaries for SNR calculation rather than data outside of them is not known to the authors to be implemented elsewhere but could be further improved by more advanced smoothing methods rather than using the residuals directly.
+Additionally, the calculation of peak shape using a Pearson's correlation to an idealized curve was not expected to be especially powerful given prior research and that the centWave algorithm essentially uses this information already during the wavelet fitting, but still proved to be a highly effective check on noise inclusion.[@Ipsen2010] This metric could be improved with more careful summary statistics that account for the differences between samples.
+Currently, the use of the overall median value does a reasonable job at identifying MFs that appear in many samples but performs poorly when detecting MFs that appear in only a few.
+Also worth noting is that the calculation of any new metrics such as these that rely on access to the raw data require exact specification of the maximum and minimum *m/z* and retention time for a peak, values that are not always returned by peakpicking algorithms.
+
+We were surprised at the poor performance of several other metrics.
+The isotope information in particular was expected to be a very strong predictor of peak quality given previous work that uses this metric extensively.[@Libiseller2015; @Treutler2016; @ElAbiead2021] We learned that many noise peaks still have reliable isotopes (perhaps unsurprising, given that the noise is in fact often caused by solvents or contaminants that are still chemical in nature) and that many real peaks are simply too low-intensity to have them in this kind of dilute environmental sample.
+We also showed that the automatically calculated SNR parameter from XCMS is not especially useful in distinguishing signal from noise.
+After inspecting a selection of peaks that had anomalous values for this metric, we are inclined to agree with [@Myers2017] and conclude that this is often due to insufficient data outside of the peak for a robust calculation of noise level.
+Finally, we were surprised to find essentially no predictive power offered by peak area or intensity, with good peaks distributed almost identically to the bad peaks in this space.
+This cautions strongly against an arbitrarily-decided intensity threshold for winnowing down the number of MFs, in agreement with previous work.[@Houriet2022; @Barupal2021]
+
+We settled on the highly reduced model of just two parameters because we found that additional parameters often improved performance on the training set but did not do so significantly for the novel datasets where the application of such a model is actually useful (Figure XX).
+One important caveat to note is that for the partially-labelled CultureData and Pttime datasets, there exists an uncontrolled degree of experimenter bias because the MS expert responsible for labeling did know that these peaks were all expected to be good.
+However, given that we do still see poor-quality peaks in this set indicates that this was not an overwhelming bias.
+The drastic drop in performance on out-of-sample data was particularly concerning because it creates overconfidence in the true level of noise actually ending up in the final dataset.
+
+We also found that this reduced two-parameter model was largely independent of the particular training set used, unlike in the more complex models (Figure XX).
+This was true in both absolute likelihood as well as rank-ordered space, a particularly important distinction when one imagines manually labeling "down" the dataset where the researcher desires to start viewing the chromatograms associated with the very best features and eventually reaching a point where enough MFs have been reviewed or bad MFs are consistent enough that they decide to forgo additional labeling.
+
+A final benefit to the reduced model is the smaller training set required to reach stability (Supp figure XX).
+This reduced size means that a useful model could be trained using only a fraction of the MFs identified in a sample set and then used to predict the quality for the remainder of the features.
+This reduction in training set size was not as significant as we expected, however, with several hundred features requiring manual labeling before even the two most stable parameters reached a consensus.
+
+### Biological conclusions vary significantly by feature quality
+
+We found that the conclusions obtained from the metabolomic datasets differed in significant ways depending on the quality threshold used to remove bad MFs from the downstream analysis.
+In the multivariate case, we ran the same analysis of PERMANOVAs and NMDS plots on various subsets of the original XCMS output and found that the effect size of depth was strongly influenced by by the threshold chosen.
+This is unsurprising given that most noise peaks should not have a biological signal to begin with, but is troubling for interpreting analyses where the FDR is not reported or the dataset not manually reviewed because the absence of a notable effect could simply be due to the overwhelming degree of noise in the default output.
+
+In the univariate case, we showed that while noise MFs are predominantly absent of a large biological signal, there are many that still have a significant biological trend.
+While some of these are inherently due to the likelihood of getting a small p-value with enough attempts despite FDR correction, a larger number of these poor-quality MFs were due to partial integration in which only the tail of a feature was integrated.
+This essentially duplicates the signal of the original peak in later analyses and should be removed.
+The real features showed a strong biological trend of high concentration throughout the surface ocean and down through the deep chlorophyll maximum (DCM), with most features equally abundant at 15 meters and this ~110 meter depth feature before dropping off at depth.
+This pattern tracks well with previous reports of biomass from the same sample site as well as earlier literature.[@Barone2022; @Heal2021] Critically, this also highlights the danger of noise MFs when additional normalizations are later applied.
+Scaling metabolomic data to biomass measurements is a common technique, and yet here it would have caused an enormous number of false positives that would have appeared to be intriguingly enriched below the DCM.
 
@@
 
-TBD
+The large number of noise peaks present in metabolomics datasets can be controlled using a simple logistic classification model.
+We trained such a model on two full-labeled open ocean HILIC datasets and found that the best performing parameters in the model for were a custom signal-to-noise metric and a test of similarity to a bell curve.
+This model showed robustness to overfitting, independence from the training set, and a reduced degree of manual labeling required.
+With this model, we showed how the distribution of metabolites in the open ocean is strongly affected by depth and categorized molecules according to their depth response.
+This distribution reproduces measures of bulk biomass but highlights several molecules of interest that diverge from the overall trend.
 
@@
 
-Culture samples used as the validation sets for this paper were obtained from existing [blah] and have been previously described by [@Bryn] and [@Pttime paper?].
+Culture samples used as the validation sets for this paper have been previously described by [@Durham2022] and on Metabolomics Workbench (Project ID PR001317).
 
@@
 
-Extraction of the environmental samples followed a modified Bligh & Dyer approach as detailed in [@BMIS].
+Extraction of the environmental samples followed a modified Bligh & Dyer approach as detailed in [@Boysen2018].
 Briefly, filters were added to PTFE centrifuge tubes with a 1:1 mix of 100 µm and 400 µm silica beads, approximately 2mL -20 °C Optima-grade DCM, and approximately 3mL -20 °C 1:1 methanol/water solution (both also Optima-grade).
@@
 This aqueous fraction was then aliquoted into an HPLC vial for injection on the HILIC column and diluted 1:1 with Optima-grade water.
-A pooled sample was created by combining 20 µL of each sample into the same HPLC vial, and a 1:1 dilution with water half-strength sample was aliquot from that to assess matrix effects and obscuring variation [@BMIS].
+A pooled sample was created by combining 20 µL of each sample into the same HPLC vial, and a 1:1 dilution with water half-strength sample was aliquot from that to assess matrix effects and obscuring variation [@Boysen2018].
 Also run alongside the environmental samples were two mixes of authentic standards in water and in an aliquot of the pooled sample at a variety of concentrations for quality control, annotation, and absolute concentration calculations.
@@
 
-The CultureData samples were re-run from the frozen aliquots for this paper. The Pttime sample processing is documented [on MW? in their paper?].
+The CultureData samples were re-run from the frozen aliquots for this paper. The Pttime sample processing is documented on Metabolomics Workbench where it has been assigned Project ID PR001317.
 
@@
 After each batch, the column was flushed with 10 mM ammonium carbonate in 85:15 water to acetonitrile for 20 to 30 minutes.
-LC conditions for the Pttime samples are documented [on MW?
-in their paper?].
+LC conditions for the Pttime samples are documented on Metabolomics Workbench where it has been assigned Project ID PR001317.
 
@@
 All files were then converted to an open-source mzML format and centroided via Proteowizard’s msConvert tool.
-For the Pttime samples, files were pulled directly from Metabolomics Workbench via Project ID [blah] and used in their existing mzXML format.
+For the Pttime samples, files were pulled directly from Metabolomics Workbench via Project ID PR001317 and used in their existing mzXML format.
 
@@
 
-The R package XCMS was used to perform peakpicking, retention time correction, and peak correspondence.
-Files were loaded and run separately for each dataset (Falkor, MESOSCOPE, CultureData, and Pttime) using the new "OnDiskMSnExp" infrastructure.
+The R package XCMS was used to perform peakpicking, retention time correction, and peak correspondence.[@Smith2006, @Tautenhahn2008] Files were loaded and run separately for each dataset (Falkor, MESOSCOPE, CultureData, and Pttime) using the new "OnDiskMSnExp" infrastructure.
 Default parameters for the CentWave peakpicking algorithm were used except for: ppm, which was set to 5; peakwidth, which was widened to 20-80 seconds; prefilter, for which the intensity threshold was raised to $10^6$; and integrate, which was set to 2 instead of 1.
-snthresh was set to zero because there are known issues with background estimation in this algorithm [@myers2017], and both verboseColumns and the extendLengthMSW parameter were set to TRUE.
-For retention time correction, the Obiwarp method was used except for the CultureData dataset, which was visually inspected and determined not to require correction.
-For the Obiwarp algorithm, the binsize was reduced to 0.1 but all other parameters were left at their defaults or equivalents.
+snthresh was set to zero because there are known issues with background estimation in this algorithm [@Myers2017], and both verboseColumns and the extendLengthMSW parameter were set to TRUE.
+For retention time correction, the Obiwarp method was used except for the CultureData dataset, which was visually inspected and determined not to require correction.[@Benton2010] For the Obiwarp algorithm, the binsize was reduced to 0.1 but all other parameters were left at their defaults or equivalents.
 
@@
 Finally, mass features with a retention time less than 30 seconds or larger than 20 minutes were removed to avoid interference from the initial and final solvent washes.
-[Check this with Laura]
 
@@
 After the full XCMS workflow was completed, the molecular features were visually inspected by a single qualified MS expert.
-For the Falkor and MESOSCOPE datasets, every molecular feature was inspected, while only those features with a predicted probability of 0.9 or higher according to the two-parameter model produced below were for the CultureData and Pttime datasets.
+For the Falkor and MESOSCOPE datasets, every molecular feature was inspected, while only those features with a predicted probability of 0.9 or higher according to the two-parameter model produced below were inspected for the CultureData and Pttime datasets.
 Inspection consisted of plotting the raw intensity values against the corrected retention-time values for all data points within the *m/z* x RT bounding box determined by the most extreme values for the given feature.
-For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labelling and to more accurately represent what MS experts typically do when assessing the quality of a given peak.
+For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labeling and to more accurately represent what MS experts typically do when assessing the quality of a given peak.
 [Fig 1.] We also decided to ignore missing values and linearly interpolate between known data points rather than filling with zeroes.
@@
 The signal-to-noise ratio (SNR) was calculated by dividing the maximum original peak height by the standard deviation of the residuals multiplied by the maximum height of the original peak.
-This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional molecular signals.
-[@myers?] If there were fewer than 5 data points, a missing value was returned and dropped in subsequent summary calculations.
+This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional molecular signals.[@Myers2017] If there were fewer than 5 data points, a missing value was returned and dropped in subsequent summary calculations.
 Accessing the raw data values also allowed us to calculate the proportion of "missed" scans in a peak for which a retention time exists at other masses in the same sample but for which no data was produced at the selected *m/z* ratio.
@@
 In each case, we categorized each molecular feature as a true positive (TP) if it was predicted to be Good and was manually classified as Good, a true negative if both predicted and classifed as Bad, a false positive if predicted to be Good but manually classified as Bad, and a false negative if predicted to be Bad but was in fact manually classified as Good.
-This allowed us to additionally define two useful measures of success, the traditionally-defined false discovery rate (FDR, defined as 1-precision or the number of false positives divided by the total number of predicted positives) and the percentage of good peaks found (GPF, also known as the recall or sensitivity and defined as the number of true positives divided by the total number of positive predictions).
+This allowed us to additionally define two useful measures of success, the traditionally-defined false discovery rate (FDR, defined as 1-precision or the number of false positives divided by the total number of predicted positives) and the percentage of good peaks found (GPF, also known as the recall or sensitivity and defined as the number of true positives divided by the total number of actual positives).
 
@@
 
-The Pttime dataset was discovered on Metabolomics Workbench [@MW id] and consists of *Phaeodactylum tricornutum* cultures collected at a variety of timepoints from both pelleted cells and the released exudate.
+The Pttime dataset was discovered on Metabolomics Workbench where it has been assigned Project ID PR001317.
+The data can be accessed directly via it's Project DOI: 10.21228/M8GH6P.
+This project dataset consists of *Phaeodactylum tricornutum* cultures collected at a variety of timepoints from both pelleted cells and the released exudate.
 This dataset was chosen because of the similar LC-MS setup used as a benchmark comparison for the performance other labs with similar setups may expect to achieve using the trained model directly.
+
 Each of these datasets were only fractionally labeled, with those peaks above the 0.9 likelihood threshold according to the two-parameter model reviewed manually and categorized.
-
 This stricter threshold was chosen because we felt less comfortable interpreting results based on mass features that were only 50% likely to be real, but did not feel the need to be so strict with this exploratory analysis that we wanted to limit it to 99+% likelihood peaks.
@@
 In univariate space, we used nonparametric Kruskal-Wallis analyses of variance to measure the difference between the surface (15m), DCM (~110m), and 175m samples because the metabolite peak areas could not be assumed to be normally distributed.
-These univariate tests were then controlled for multiply hypothesis testing using R's `p.adjust` function with method `fdr`.[@ref] We also performed post-hoc Dunn tests provided by the `rstatix` package to categorize the response to depth for those mass features for which the KW test was significant, with responses falling into one of the 14 classes possible when permuting the sign and significance of the Dunn test.[@ref] p-values obtained from the Dunn tests were not FDR controlled because it was used as a categorization tool rather than a null hypothesis test.
-In multivariate space, we used a permutational MANOVA (PERMANOVA) [@ref] provided by the vegan package's `adonis2` function to test for multivariate differences in structure of the metabolome with depth.[@ref] We ran multiple PERMANOVAs with a different subset of mass features included each time, corresponding to using the output from XCMS directly, likelihood thresholds of 0.01, 0.1, 0.5, 0.9, and finally only those peaks manually annotated as good.
-
-All analyses were run in R, version 4.2.2, and code is available on the Github at https://github.com/wkumler/MS_metrics.[@ref]
+These univariate tests were then controlled for multiply hypothesis testing using R's `p.adjust` function with method `fdr`.[@Benjamini1995] We also performed post-hoc Dunn tests provided by the `rstatix` package to categorize the response to depth for those mass features for which the KW test was significant, with responses falling into one of the 14 classes possible when permuting the sign and significance of the Dunn test.[@Dunn1964] p-values obtained from the Dunn tests were not FDR controlled because it was used as a categorization tool rather than a null hypothesis test.
+In multivariate space, we used a permutational MANOVA (PERMANOVA) [@Anderson2014] provided by the vegan package's `adonis2` function to test for multivariate differences in structure of the metabolome with depth.[@veganpkg] We ran multiple PERMANOVAs with a different subset of mass features included each time, corresponding to using the output from XCMS directly, likelihood thresholds of 0.01, 0.1, 0.5, 0.9, and finally only those peaks manually annotated as good.
+
+All analyses were run in R, version 4.2.2, and code is available on the Github at https://github.com/wkumler/MS_metrics.[@R]
+
+## Abbreviations
+
+DCM: Deep Chlorophyll Maximum
+
+SNR: Signal to Noise Ratio
+
+MF: Molecular Feature
+
+GPF: Good Peaks Found
+
+FDR: False Discovery Rate
+
+HILIC: Hydrophilic Interaction Liquid Chromatography
+
+EIC: Extracted Ion Chromatogram
+
+LC: Liquid Chromatography
+
+MS: Mass Spectrometry
 
@@
 ## References
+
+<div id="refs"></div>
 

### 2023-06-01 manuscript/manuscript.Rmd
Added second Myers ref, finished spicy literature comparison

@@
 In the past two decades, data-driven methods have established workflows for untargeted metabolomics but the imperfect performance of the core peakpicking algorithms continue to require manual oversight and curation.
-This problem has been exacerbated by the increased use of non-traditional chromatography such as hydrophilic interaction which tends to produce noisier peaks.[@Bajad2006; @Myers2017; @Gika2019]
+This problem has been exacerbated by the increased use of non-traditional chromatography such as hydrophilic interaction which tends to produce noisier peaks.[@Bajad2006; @Myers2017b; @Gika2019]
 
@@
 Existing algorithms tend to favor false positives because downstream analyses can always remove erroneous mass features, but false negatives cannot be later recovered.[@Pirttila2022; @Gloaguen2022] However, this approach creates two additional downstream problems.
-First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of peaks, a task that scales combinatorially with the number of samples and compounds measured.
-Second, false negatives cannot be fully accounted for in the way false positives can because false negatives are introduced in every step of metabolomic analyses from sample collection through injection.
+First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of peaks, a task that scales combinatorially with the number of samples and compounds measured.[@Myers2017a] Second, false negatives cannot be fully accounted for in the way false positives can because false negatives are introduced in every step of metabolomic analyses from sample collection through injection.
 Instead of minimizing false negatives, we believe that emphasis should be placed on allowing the experimenter to set a threshold for the false discovery rate and accept that this will inherently add to the number of peaks already lost in the data collection process.
@@
 We used two fully-labeled and two partially-labeled HILIC LC-MS datasets to assess the performance of the XCMS algorithm and construct a robust model of peak quality.
-To measure performance, we used two measurements of success closely tied to intuitive questions about a dataset: the percentage of good peaks found (GPF, also known as the recall or sensitivity) and the percentage of bad peaks included or false discovery rate (FDR).
-We also chose these metrics because they are independent of the number of true negatives detected.
-This was a concern because we did not want to create a situation in which the algorithm is rewarded for artificially increasing the number of overall bad peaks, only to later remove them.
-Similarly, we decided against using the F$_1$ score as an overall summary statistic because false negatives and false positives have very different implications in this context and should be treated separately.
-
-One of the major ways in which this manuscript differs from prior work is its focus on summary statistics calculated across multiple files.[@Pirttila2022; @Guo2021; @Muller2020] Most existing peakpicking literature uses the single-file EIC peak as the core bit of training data, but that approach ignores critical information obtained elsewhere in the MS run that can change the judgement made on a single chromatogram.
-Features that are high quality are typically represented in multiple files, especially in QC pooled samples, and a feature that only has a peak in a single file is typically regarded as highly suspect, if not removed entirely.
+To measure performance, we used two measurements of success closely tied to intuitive questions about a dataset: the percentage of total good peaks found (GPF, also known as recall or sensitivity) and the percentage of bad peaks included or false discovery rate (FDR).
+We decided against using the F$_1$ score as an overall summary statistic because false negatives and false positives have very different implications in this context and should be treated separately.
+
+One of the major ways in which this manuscript differs from prior work is its focus on summary statistics calculated across multiple files.
+Most existing peakpicking literature uses the single-file EIC peak as the core bit of training data, but that approach ignores critical information obtained elsewhere in the MS run that can change the judgement made on a single chromatogram.[@Pirttila2022; @Guo2021; @Muller2020] Features that are high quality are typically represented in multiple files, especially in quality control pooled samples, and a feature that only has a peak in a single file is typically regarded as highly suspect, if not removed entirely.
 Classifying an entire feature at once not only has the advantage of reducing the amount of manual labor by a factor equal to the number of files in the run (typically 10s-100s smaller) but is also a better representation of the judgement made by an MS expert.
@@
 We found that a simple two-parameter logistic regression model trained on two novel metrics of peak quality had reasonably good performance on the training set and was highly robust when applied to novel datasets.
-
 The logistic regression in particular was favored over the random forest and regularized regression we tried due to their similar performance and increased interpretability (Supp.
 figure XX).
-In the literature, more complex models have also shown similar or better performance but typically require larger training sets and are fundamentally uninterpretable.[@Pirttila2022 and references therein]
 
@@
 
-@Guo2021 (EVA): Accuracy of 90-95% but did not distinguish between false positives and false negatives.
-Likely driven by the large number of bad peaks - here, our accuracy with a threshold of 0.5 was 92.1% for Falkor and 94.4% for MESOSCOPE.
-
-Accuracy is a problematic metric because if just predicting everything as "Bad" was 82.4% and 86.4%, respectively.
-Might be slightly facetious, but this was essentially what happened for a good number of the individual predictors - none of the smoothing curves exceeded 50% likelihood (Supp figure XX).
-
-This is why we chose to measure GPF and FDR instead of total accuracy. Not directly comparable datasets though and we've definitely got the advantage of using a half-internal training set.
-
-@Melnikov2020 (peakonly): Accuracy of 89%, precision of 97% and a recall of 89%
-Versus our precision (PPV) of 95.6% and recall (sensitivity) of 94.6% for Falkor
-Versus our precision (PPV) of 95.9% and recall (sensitivity) of 97.7% for MESOSCOPE
-@Gloaguen2022 (NeatMS): Claims universally superior performance to peakonly
-
-These deep learning methods in particular are challenging because they do not produce overall quality scores and the tradeoff between false positives and false negatives is fixed regardless of experimental priorities.
-
-The two metrics in the final model were derived from the raw EIC data rather than any summary statistic calculated by XCMS automaticallyand were implemented because they matched our intuition about what makes an MF look good to an MS expert.
+The model's performance is similar to the highly complex convolutional neural networks currently implemented in the literature.
+@Guo2021 presented EVA and reported an accuracy of 90-95%, a range inclusive of our accuracy on both the Falkor (92.1%) and MESOSCOPE (94.4%) when using the two-parameter model with a likelihood threshold of 0.5.
+However, we note that accuracy alone can be a highly misleading statistic to report when working with datasets as unbalanced as these because very high accuracy can be obtained by simply classifying everything as bad, with a strong incentive to actually *increase* the number of bad peaks detected while doing so.
+This strategy, when applied to our data, returned accuracies in the 80-90% range despite being a useless classifier for downstream analysis.
+This is perhaps a slightly facetious argument, but many of the single-parameter logistic models we tested never exceeded a 50% likelihood and essentially functioned as this type of classifier (Supp figure XX).
+
+The class imbalance is partially why we chose to measure GPF and FDR instead of total accuracy.
+When reported, our model performs similarly on both metrics, albeit with a more limited application to HILIC-Orbitrap data.
+Compared to the peakonly algorithm of @Melnikov2020, we obtained higher accuracy scores (they report an accuracy of 89%) and higher recall (peakonly's recall was also 89%, while recall (%GPF) on our data was 94.6% and 97.7% for the Falkor and MESOSCOPE datasets, respectively).
+However, peakonly outperforms our model in precision with an FDR of 3% while our final FDR was 4.4% for the Falkor dataset and 4.1% for MESOSCOPE.
+
+@Gloaguen2022 later introduced NeatMS, another CNN, and compared it directly to peakonly to claim equivalent or superior performance across a range of dilution factors.
+However, they do not report total precision or recall metrics in a comprehensive untargeted way but instead focus only on assessing the model's performance on known chemical standards.
+Additionally, they report a percentage of standards found for the peakonly model applied to their data and find that its performance is significantly lower (79.4%) than the recall reported in @Melnikov2020, perhaps indicating that the peakonly model is still overfit.
+
+We would argue that the deep learning models, while promising, do not currently show significant enough improvement over a simple logistic model to justify their black-box nature.
+In particular, they provide no way to control the tradeoff between false positives and false negatives and no relative ranking of individual MF quality beyond the broad bins into which they are placed.
+[This is spicy, might be better to soften it a bit?]
+
+The two metrics in the final model were derived from the raw EIC data rather than any summary statistic calculated by XCMS automatically and were implemented because they matched our intuition about what makes an MF look good to an MS expert.
 It is worth noting that these are very simple metrics and therefore fast to calculate, but we expect that more complicated metrics could perform even better.
-For example, using the data within the peak boundaries for SNR calculation rather than data outside of them is not known to the authors to be implemented elsewhere but could be further improved by more advanced smoothing methods rather than using the residuals directly.
-Additionally, the calculation of peak shape using a Pearson's correlation to an idealized curve was not expected to be especially powerful given prior research and that the centWave algorithm essentially uses this information already during the wavelet fitting, but still proved to be a highly effective check on noise inclusion.[@Ipsen2010] This metric could be improved with more careful summary statistics that account for the differences between samples.
+For example, the method of using the data within the peak boundaries for SNR calculation rather than data outside of them is not known to the authors to be implemented elsewhere but could be further improved by more advanced smoothing methods rather than using the residuals directly.
+Additionally, the calculation of peak shape using a Pearson's correlation to an idealized curve was not expected to be especially powerful given prior research (e.g.
+@Ipsen2010) and that the centWave algorithm essentially uses this information already during the wavelet fitting, but still proved to be a highly effective parameter.
+This metric could be improved with more careful summary statistics that account for the differences between samples.
 Currently, the use of the overall median value does a reasonable job at identifying MFs that appear in many samples but performs poorly when detecting MFs that appear in only a few.
@@
 We also showed that the automatically calculated SNR parameter from XCMS is not especially useful in distinguishing signal from noise.
-After inspecting a selection of peaks that had anomalous values for this metric, we are inclined to agree with [@Myers2017] and conclude that this is often due to insufficient data outside of the peak for a robust calculation of noise level.
+After inspecting a selection of peaks that had anomalous values for this metric, we are inclined to agree with @Myers2017a and conclude that this is often due to insufficient data outside of the peak for a robust calculation of noise level.
 Finally, we were surprised to find essentially no predictive power offered by peak area or intensity, with good peaks distributed almost identically to the bad peaks in this space.
 This cautions strongly against an arbitrarily-decided intensity threshold for winnowing down the number of MFs, in agreement with previous work.[@Houriet2022; @Barupal2021]
+
+RSD/CV problems also noted by @Gloaguen2022!
 
@@
 Default parameters for the CentWave peakpicking algorithm were used except for: ppm, which was set to 5; peakwidth, which was widened to 20-80 seconds; prefilter, for which the intensity threshold was raised to $10^6$; and integrate, which was set to 2 instead of 1.
-snthresh was set to zero because there are known issues with background estimation in this algorithm [@Myers2017], and both verboseColumns and the extendLengthMSW parameter were set to TRUE.
+snthresh was set to zero because there are known issues with background estimation in this algorithm [@Myers2017a], and both verboseColumns and the extendLengthMSW parameter were set to TRUE.
 For retention time correction, the Obiwarp method was used except for the CultureData dataset, which was visually inspected and determined not to require correction.[@Benton2010] For the Obiwarp algorithm, the binsize was reduced to 0.1 but all other parameters were left at their defaults or equivalents.
@@
 The signal-to-noise ratio (SNR) was calculated by dividing the maximum original peak height by the standard deviation of the residuals multiplied by the maximum height of the original peak.
-This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional molecular signals.[@Myers2017] If there were fewer than 5 data points, a missing value was returned and dropped in subsequent summary calculations.
+This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional molecular signals.[@Myers2017b] If there were fewer than 5 data points, a missing value was returned and dropped in subsequent summary calculations.
 Accessing the raw data values also allowed us to calculate the proportion of "missed" scans in a peak for which a retention time exists at other masses in the same sample but for which no data was produced at the selected *m/z* ratio.

### 2023-06-02 manuscript/manuscript.Rmd
Updated figure numbers (probably got them all but who knows)

@@
 According to all three logistic regression models (see Methods), the majority of MFs were estimated to have a less than 1% chance of being good.
-The full model (containing all possible peak metrics) and the XCMS model (built on only those metrics calculated from the XCMS output) both displayed a strongly bimodal distribution, with a large number of peaks also exceeding a 99% chance of being good, while the two-parameter model (consisting of the novel SNR metric and the peak shape correlation metric) had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Fig.
-3).
-
-```{r fig3 pred_prob_class_color_hists}
+The full model (containing all possible peak metrics) and the XCMS model (built on only those metrics calculated from the XCMS output) both displayed a strongly bimodal distribution, with a large number of peaks also exceeding a 99% chance of being good, while the two-parameter model (consisting of the novel SNR metric and the peak shape correlation metric) had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Figure 1).
+
+```{r Fig1 pred_prob_class_color_hists}
 base_gp <- ggplot() +
@@
 
-![Figure 3: Histograms showing the estimated likelihood of a given mass feature being categorized as "Good" according to the two-parameter logistic model trained on the combined fully-labeled Falkor and MESOSCOPE datasets.
+![Figure 1: Histograms showing the estimated likelihood of a given mass feature being categorized as "Good" according to the two-parameter logistic model trained on the combined fully-labeled Falkor and MESOSCOPE datasets.
 Colors indicate the category in which each feature was manually assigned by an expert, with "Stans only" referring to a good peak that was only visible in the standards run alongside the samples.
@@
 The XCMS metrics performed slightly worse, with FDR values in the 10-15% range and GPF values closer to 75%.
-The two-parameter model performed worst when tested internally, with an FDR of about 20% and %GPF also around 75% (Figure 4).
+The two-parameter model performed worst when tested internally, with an FDR of about 20% and %GPF also around 75% (Figure 2).
 However, when the models were trained on a different dataset than the one they were used to predict classifications for, all models had similar performance with %FDR around 10-25 and %GPF around 60-80.
-The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more peaks recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE (Figure 4).
-
-```{r FDR vs GPF cross-train triangle figure feature selection}
+The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more peaks recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE (Figure 2).
+
+```{r Fig2 FDR vs GPF cross-train triangle figure feature selection}
 fdr_gpf_cross_train_df <- lst(all_params, raw_data_params, xcms_params) %>%
@@
 
-![Figure 4: False discovery rate (%FDR) and fraction of good peaks found (GPF) plotted across different subsets of model parameters.
+![Figure 2: False discovery rate (%FDR) and fraction of good peaks found (GPF) plotted across different subsets of model parameters.
 Lower %FDR indicates a smaller fraction of false positives among those peaks the model categorized as "Good" using a threshold of 0.5, and higher %GPF indicates a larger fraction of the total good peaks were found using the same threshold.
@@
 In the raw probability space, the two-parameter models had the highest $r$ value of 0.996, while the full models and the XCMS-trained models had $r$ values of 0.856 and 0.896, respectively.
-When compared in ranked space, we found an intensification of the effect described previously, with an even higher $r$ for the two-parameter model of 0.998 but lower $r$ values for the full and XCMS-trained model of 0.818 and 0.849, respectively (Figure 5).
-
-```{r red/gold rank dotplot step-down method}
+When compared in ranked space, we found an intensification of the effect described previously, with an even higher $r$ for the two-parameter model of 0.998 but lower $r$ values for the full and XCMS-trained model of 0.818 and 0.849, respectively (Figure 3).
+
+```{r Fig3 red/gold rank dotplot step-down method}
 dotplot_stepdown_data <- lst(all_params, xcms_params, raw_data_params) %>%
@@
 
-![Figure 5: Predicted likelihood of a feature being "Good" according to a model trained on the MESOSCOPE dataset vs a model trained on the Falkor dataset.
+![Figure 3: Predicted likelihood of a feature being "Good" according to a model trained on the MESOSCOPE dataset vs a model trained on the Falkor dataset.
 The top row of plots show the exact likelihood predicted by the logistic model across three different subsets of parameters, while the bottom row shows the estimates ranked from least likely to most likely.
@@
 
-With the stricter threshold, we obtained FDR values consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for Falkor, MESOSCOPE, CultureData, and Pttime respectively [Figure 6].
+With the stricter threshold, we obtained FDR values consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for Falkor, MESOSCOPE, CultureData, and Pttime respectively (Figure 4).
 Of course, this low error rate required that we miss out on additional potentially valuable features, with only a fraction of the total good peaks making it past this threshold.
 In both the Falkor and MESOSCOPE datasets, less than half of the good peaks were labeled as such, with actual values of 39.4% and 26.5%, respectively.
-Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the %GPF but expect it to be in a similar range (Figure 6).
+Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the %GPF but expect it to be in a similar range (Figure 4).
 
@@
 
-![Figure 6: False discovery rate and proportion of total good peaks identified as good by the two-parameter model trained on the combined MESOSCOPE/Falkor dataset and applied to each dataset individually with a threshold of 0.9.
+![Figure 4: False discovery rate and proportion of total good peaks identified as good by the two-parameter model trained on the combined MESOSCOPE/Falkor dataset and applied to each dataset individually with a threshold of 0.9.
 FDR is calculated by dividing the number of false positives by the total positives produced by the model and GPF is calculated by dividing the number of true positives by the total number of good peaks as identified manually.
@@
 
-A majority of the features (1,323 of 1,832 total) in the original, non-thresholded MESOSCOPE dataset had no significant trend with depth, with FDR-controlled Kruskal-Wallis p-values exceeding 0.05 (Figure 7).
+A majority of the features (1,323 of 1,832 total) in the original, non-thresholded MESOSCOPE dataset had no significant trend with depth, with FDR-controlled Kruskal-Wallis p-values exceeding 0.05 (Figure 5).
 The largest category that did have a trend with depth was the 15m = DCM > 175m category, containing 118 molecular features, with largest peak areas distributed evenly between the 15 meter and deep chlorophyll maximum (DCM) samples, while the 175 meter samples had significantly smaller areas.
@@
 Of the 182 good peaks, less than a fifth had no trend with depth (44/249) and a majority had unequivocally lowest values in the 175 meter samples (those with 15m/DCM > 175m, 145 features).
-The two-parameter model, when applied with a 50% likelihood threshold, also recovered this general feature distribution and classified many of the features with no significant depth signal as likely to be bad (Figure 7B, currently in Slack?).
+The two-parameter model, when applied with a 50% likelihood threshold, also recovered this general feature distribution and classified many of the features with no significant depth signal as likely to be bad [Figure 7B, currently in Slack?].
 
@@
 
-```{r clockplot}
+```{r Fig5 clockplot}
 resp_class_levels <- c("All equal", 
@@
 
-![Figure 7: Plot of metabolite response to depth shown across three axes using the rank-normalized median value at each depth as the coordinate for that axis.
+![Figure 5: Plot of metabolite response to depth shown across three axes using the rank-normalized median value at each depth as the coordinate for that axis.
 Each molecular feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile.
@@
 
-```{r NMDS results}
+```{r Fig6 NMDS results}
 thresh_NMDS_df <- imap(mat_list, function(mat_given, mat_name){
@@
 
-![Figure XX: NMDS plots of metabolite similarity according to sample depth across multiple likelihood thresholds.
+![Figure 6: NMDS plots of metabolite similarity according to sample depth across multiple likelihood thresholds.
 Triplicate samples are represented by the vertices of the triangles and colored by the depth from which they were sampled (DCM = deep chlorophyll maximum, ~110m).
@@
 Inspection consisted of plotting the raw intensity values against the corrected retention-time values for all data points within the *m/z* x RT bounding box determined by the most extreme values for the given feature.
-For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labeling and to more accurately represent what MS experts typically do when assessing the quality of a given peak.
-[Fig 1.] We also decided to ignore missing values and linearly interpolate between known data points rather than filling with zeroes.
+For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labeling and to more accurately represent what MS experts typically do when assessing the quality of a given peak (Figure 7).
+We also decided to ignore missing values and linearly interpolate between known data points rather than filling with zeroes.
 These EICs were then shown to an MS expert for classification into one of 4 categories: Good, Bad, Ambiguous, or Stans only if the peak appeared to only show up in the standards.
-A few randomly-chosen features from the manually-assigned Good and Bad classifications are shown in [Fig.
-1].
+A few randomly-chosen features from the manually-assigned Good and Bad classifications are shown in Figure 7.
 
@@
 
-```{r fig:some good some bad features}
+```{r Fig7 some good some bad features}
 minEIC_df <- read_csv("manuscript/minEIC_df.csv")
@@
 
-![Figure 1: Randomly selected ion chromatograms from both "Good" (top row) and "Bad" (bottom row) manual classifications, colored by the depth from which the biological sample was taken.
+![Figure 7: Randomly selected ion chromatograms from both "Good" (top row) and "Bad" (bottom row) manual classifications, colored by the depth from which the biological sample was taken.
 DCM = deep chlorophyll maximum, approximately 110 meters.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
@@
 Distributions were visually inspected and highly abnormal metrics were transformed using log-scaling if necessary.
-Using a pairs plot, highly correlated (above a Pearson's r ~ .9) had one of the correlated metrics removed.
+Using a pairs plot, highly correlated (above a Pearson's r ~ 0.9) had one of the correlated metrics removed.
 
@@
 This approach allowed us to approximate a bell curve with increasing degrees of right-skewness and the beta distribution was chosen because it is constrained between 0 and 1 and simple and speedy to generate in R.
-For each α value, Pearson's correlation coefficient was calculated between the beta distribution and the raw data, with the highest value returned as a metric for how peak-shaped the data were [Fig.
-2].
-The beta distribution with the highest *r* was also then used to estimate the noise level within the peak by also scaling both the beta distribution densities and the raw data intensity values as described above, then subtracting the scaled beta distribution from the scaled intensity values, producing the residuals of the fit [Fig.
-2].
+For each α value, Pearson's correlation coefficient was calculated between the beta distribution and the raw data, with the highest value returned as a metric for how peak-shaped the data were (Figure 8).
+The beta distribution with the highest *r* was also then used to estimate the noise level within the peak by also scaling both the beta distribution densities and the raw data intensity values as described above, then subtracting the scaled beta distribution from the scaled intensity values, producing the residuals of the fit (Figure 8).
 The signal-to-noise ratio (SNR) was calculated by dividing the maximum original peak height by the standard deviation of the residuals multiplied by the maximum height of the original peak.
@@
 
-```{r how are raw peak shape metrics calculated}
+```{r Fig8 how are raw peak shape metrics calculated}
 minEIC_df <- read_csv("manuscript/minEIC_df.csv")
@@
 
-![Figure 2: Method used to calculate the metrics for the two-parameter model from the raw data via comparison to an idealized pseudo-Gaussian peak for both manually identified "Good" and "Bad" peaks.
+![Figure 8: Method used to calculate the metrics for the two-parameter model from the raw data via comparison to an idealized pseudo-Gaussian peak for both manually identified "Good" and "Bad" peaks.
 Normalization was performed by linearly scaling the raw values into the 0-1 range by subtracting the minimum value and dividing by the maximum.
@@
 
-## Acknowledgements
+## Declarations
+### Ethics approval and consent to participate
+
+Not applicable
+
+### Consent for publication
+
+Not applicable
+
+### Availability of data and materials
+
+The raw mzML files are all available on Metabolomics Workbench.
+Falkor, MESOSCOPE, and CultureData can be found under Project ID [blah] and accessed directly using the Project ID http://dx.doi.org/[blah], while Pttime is located under Project ID PR001317 and can be accessed directly using its Project DOI: http://dx.doi.org/10.21228/M8GH6P.
+Code and other raw data is available on the GitHub repository at https://github.com/wkumler/MS_metrics.
+The manuscript has been rendered as a single R Markdown document with analyses contained within for reproducibility.
+
+### Competing interests
+
+The authors declare that they have no competing interests
+
+### Funding
+
+[TBD, get from Anitra. eScience?]
+
+### Authors' contributions
+### Acknowledgements
 

### 2023-06-02 feature_extraction.R
Adding CV as additional parameter in model

@@ -34,7 +34,7 @@ trapz <- function(x, y) {
 # dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
 # dataset_version <- "CultureData"
-dataset_version <- "Pttime"
+# dataset_version <- "Pttime"
 output_folder <- paste0("made_data_", dataset_version, "/")
 
 file_data <- read_csv(paste0(output_folder, "file_data.csv")) %>%
@@ -62,6 +62,14 @@ peak_data <- msnexp_filled %>%
 
 
 # Simple feature extraction - those provided by XCMS directly ----
+
+cv <- function(vec, robust=FALSE){
+  if(robust){
+    mad(vec, na.rm=TRUE)/median(vec, na.rm = TRUE)
+  } else {
+    sd(vec, na.rm = TRUE)/mean(vec, na.rm=TRUE)
+  }
+}
 n_files <- nrow(file_data)
 n_samps <- sum(file_data$samp_type=="Smp")
 n_stans <- sum(file_data$samp_type=="Std")
@@ -82,7 +90,9 @@ simple_feats <- peak_data %>%
             feat_npeaks=n()/n_files,
             n_found=(n_files-sum(is.na(intb)))/n_files,
             samps_found=1-sum(is.na(intb) & samp_type=="Smp")/n_samps,
-            stans_found=1-sum(is.na(intb) & samp_type=="Std")/n_stans
+            stans_found=1-sum(is.na(intb) & samp_type=="Std")/n_stans,
+            pooled_cv=cv(into[samp_type=="Poo"]),
+            pooled_cv_rob=cv(into[samp_type=="Poo"], robust=TRUE)
             ) %>%
   mutate(sn=ifelse(is.infinite(sn), 0, sn))
 write.csv(simple_feats, paste0(output_folder, "simple_feats.csv"), row.names = FALSE)
@@ -355,7 +365,8 @@ features_extracted <- simple_feats %>%
   left_join(classified_feats) %>%
   filter(mean_rt%between%c(30, 1200))
 if(dataset_version=="Pttime"){
-  features_extracted <- select(-c(samps_found, stans_found, smp_to_blk, smp_to_std),
+  features_extracted <- select(-c(samps_found, stans_found, smp_to_blk, smp_to_std,
+                                  pooled_cv, pooled_cv_rob),
                                .data = features_extracted)
 }
 write.csv(features_extracted, paste0(output_folder, "features_extracted.csv"), 

### 2023-06-02 peakpicking_and_prep.R

@@ -17,7 +17,7 @@ options(pillar.sigfig=7)
 # dataset_version <- "FT2040"
 # dataset_version <- "MS3000"
 # dataset_version <- "CultureData"
-dataset_version <- "Pttime"
+# dataset_version <- "Pttime"
 
 output_folder <- paste0("made_data_", dataset_version, "/")
 mzML_files <- list.files(paste0(output_folder, "mzMLs/"), full.names=TRUE)

### 2023-06-02 manuscript/manuscript.Rmd
Adding CV as additional parameter in model
Reran models with the CV parameter

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, eval=FALSE)
+knitr::opts_chunk$set(echo = FALSE)
 knitr::opts_knit$set(root.dir = '..')
@@
 library(RaMS)
+```
+
+```{r run analysis, eval=FALSE}
+for(dataset_version in c("FT2040", "MS3000", "CultureData", "Pttime")){
+  print(paste("Processing", dataset_version))
+  # source("peakpicking_and_prep.R")
+  source("feature_extraction.R")
+}
 ```
@@
                  "f", "scale", "lmin", "feat_npeaks", "n_found", 
-                 "samps_found", "stans_found")
+                 "samps_found", "stans_found", "pooled_cv", "pooled_cv_rob")
 all_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", "sd_pw",
@@
                 "med_cor", "med_SNR", "med_missed_scans", "smp_to_blk", 
-                "smp_to_std", "shape_cor", "area_cor", "feat_class")
+                "smp_to_std", "pooled_cv", "pooled_cv_rob", "shape_cor", 
+                "area_cor")
 ```

### 2023-06-02 feature_extraction.R

@@ -55,7 +55,7 @@ peak_data <- msnexp_filled %>%
   unnest_longer(peakidx) %>%
   rename_with(~paste0("feat_", .x), .cols = -peakidx) %>%
   dplyr::rename(feature="feat_id") %>%
-  left_join(peak_data_long) %>%
+  left_join(peak_data_long, by = join_by(peakidx)) %>%
   mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
   complete(feature, filename)
 
@@ -74,7 +74,7 @@ n_files <- nrow(file_data)
 n_samps <- sum(file_data$samp_type=="Smp")
 n_stans <- sum(file_data$samp_type=="Std")
 simple_feats <- peak_data %>%
-  left_join(file_data) %>%
+  left_join(file_data, by="filename") %>%
   group_by(feature) %>%
   fill(starts_with("feat"), .direction = "downup") %>%
   summarise(mean_mz=unique(feat_mzmed), 
@@ -94,7 +94,9 @@ simple_feats <- peak_data %>%
             pooled_cv=cv(into[samp_type=="Poo"]),
             pooled_cv_rob=cv(into[samp_type=="Poo"], robust=TRUE)
             ) %>%
-  mutate(sn=ifelse(is.infinite(sn), 0, sn))
+  mutate(sn=ifelse(is.infinite(sn), 0, sn)) %>%
+  mutate(pooled_cv=ifelse(is.na(pooled_cv), 1, pooled_cv)) %>%
+  mutate(pooled_cv_rob=ifelse(is.na(pooled_cv_rob), 1, pooled_cv_rob))
 write.csv(simple_feats, paste0(output_folder, "simple_feats.csv"), row.names = FALSE)
 
 # Calculate peak shape metrics from EICs ----
@@ -351,18 +353,18 @@ stan_diffs <- read_csv(paste0(output_folder, "stan_diffs.csv"))
 feat_isodata <- read_csv(paste0(output_folder, "feat_isodata.csv"))
 classified_feats <- read_csv(paste0(output_folder, "classified_feats.csv")) %>%
   select(feature, feat_class) %>%
-  right_join(data.frame(feature=simple_feats$feature)) %>%
+  right_join(data.frame(feature=simple_feats$feature), by="feature") %>%
   mutate(feat_class=ifelse(is.na(feat_class), "Unclassified", feat_class)) %>%
   arrange(feature)
 
 features_extracted <- simple_feats %>%
[…]

### 2023-06-02 feature_extraction.R
Replacing NAs with medians instead of 1s

@@ -95,8 +95,8 @@ simple_feats <- peak_data %>%
             pooled_cv_rob=cv(into[samp_type=="Poo"], robust=TRUE)
             ) %>%
   mutate(sn=ifelse(is.infinite(sn), 0, sn)) %>%
-  mutate(pooled_cv=ifelse(is.na(pooled_cv), 1, pooled_cv)) %>%
-  mutate(pooled_cv_rob=ifelse(is.na(pooled_cv_rob), 1, pooled_cv_rob))
+  mutate(pooled_cv=ifelse(is.na(pooled_cv), median(pooled_cv, na.rm=TRUE), pooled_cv)) %>%
+  mutate(pooled_cv_rob=ifelse(is.na(pooled_cv_rob), median(pooled_cv, na.rm=TRUE), pooled_cv_rob))
 write.csv(simple_feats, paste0(output_folder, "simple_feats.csv"), row.names = FALSE)
 
 # Calculate peak shape metrics from EICs ----

### 2023-06-05 manuscript/manuscript.Rmd
Fixing comparison to CNN models, adding RSD discussion, subheader labels

@@
 
+#### Performance relative to other peak picking models
+
 ```{r, eval=FALSE}
@@
   filter(feat_class%in%c("Good", "Bad")) %>%
-  # with(table(feat_class, pred_class))
-  with(caret::confusionMatrix(as.factor(pred_class), as.factor(feat_class)))
+  with(table(feat_class, pred_class)) %>%
+  caret::confusionMatrix(positive="Good", mode="everything")
 # Falkor if everything was just classified as "Bad"
@@
   filter(feat_class%in%c("Good", "Bad")) %>%
-  # with(table(feat_class, pred_class))
-  with(caret::confusionMatrix(as.factor(pred_class), as.factor(feat_class)))
+  with(table(feat_class, pred_class)) %>%
+  caret::confusionMatrix(positive="Good", mode="everything")
 # MESOSCOPE if everything was just classified as "Bad"
 (1547+36)/(1547+36+67+182) # 0.864
-```
-
-The model's performance is similar to the highly complex convolutional neural networks currently implemented in the literature.
-@Guo2021 presented EVA and reported an accuracy of 90-95%, a range inclusive of our accuracy on both the Falkor (92.1%) and MESOSCOPE (94.4%) when using the two-parameter model with a likelihood threshold of 0.5.
+
+bind_rows(MS3000_features, FT2040_features) %>%
+  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
+  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  with(table(feat_class, pred_class)) %>%
+  caret::confusionMatrix(positive="Good", mode="everything")
+```
+
+This model performs less well than previous reports in the literature, but is highly simplified.
+@Guo2021 presented EVA and reported an accuracy of 90-95%, a range inclusive of our accuracy on both the Falkor (92.1%) and MESOSCOPE (94.4%) datasets when using a likelihood threshold of 0.5.
 However, we note that accuracy alone can be a highly misleading statistic to report when working with datasets as unbalanced as these because very high accuracy can be obtained by simply classifying everything as bad, with a strong incentive to actually *increase* the number of bad peaks detected while doing so.
@@
 The class imbalance is partially why we chose to measure GPF and FDR instead of total accuracy.
-When reported, our model performs similarly on both metrics, albeit with a more limited application to HILIC-Orbitrap data.
-Compared to the peakonly algorithm of @Melnikov2020, we obtained higher accuracy scores (they report an accuracy of 89%) and higher recall (peakonly's recall was also 89%, while recall (%GPF) on our data was 94.6% and 97.7% for the Falkor and MESOSCOPE datasets, respectively).
-However, peakonly outperforms our model in precision with an FDR of 3% while our final FDR was 4.4% for the Falkor dataset and 4.1% for MESOSCOPE.
+@Melnikov2020 reported these values as well in their presentation of peakonly, relative to which we obtained higher accuracy (they report 89% accuracy) but worse GPF and FDR than their model (89% and 3%, respectively, relative to our 80% GPF and 23% FDR overall).
 
@@
 
-We would argue that the deep learning models, while promising, do not currently show significant enough improvement over a simple logistic model to justify their black-box nature.
-In particular, they provide no way to control the tradeoff between false positives and false negatives and no relative ranking of individual MF quality beyond the broad bins into which they are placed.
-[This is spicy, might be better to soften it a bit?]
+While the model we present here has reduced performance relative to the CNNs, we would argue that its utility is not in maximizing performance but instead in maximizing interpretability.
+In particular, the CNNs provide no way to control the tradeoff between false positives and false negatives and no relative ranking of individual MF quality beyond the broad bins into which they are placed or explanation of relative metrics strengths for later analyses.
+
+#### Assessing the relative power of individual metrics
 
@@
 
+```{r}
+bind_rows(MS3000_features, FT2040_features) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=feat_class=="Good") %>%
+  select(feat_class, all_of(all_params)) %>%
+  glm(formula=feat_class~., family = binomial) %>%
+  broom::tidy() %>%
+  arrange(p.value) %>%
+  print(n=Inf)
+
+bind_rows(MS3000_features, FT2040_features) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  ggplot() +
+  geom_histogram(aes(x=mean_mz, fill=feat_class)) +
+  facet_wrap(~feat_class, ncol = 1, scales = "free_y")
+bind_rows(MS3000_features, FT2040_features) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=ifelse(feat_class=="Good", 1, 0)) %>%
+  ggplot(aes(x=mean_mz, y=feat_class)) +
+  geom_jitter(height=0.1, alpha=0.2) +
+  geom_smooth(method="glm", method.args=list(family=binomial), formula = "y~x")
+```
+
 We were surprised at the poor performance of several other metrics.
 The isotope information in particular was expected to be a very strong predictor of peak quality given previous work that uses this metric extensively.[@Libiseller2015; @Treutler2016; @ElAbiead2021] We learned that many noise peaks still have reliable isotopes (perhaps unsurprising, given that the noise is in fact often caused by solvents or contaminants that are still chemical in nature) and that many real peaks are simply too low-intensity to have them in this kind of dilute environmental sample.
+
+The relative standard deviation (RSD), also called the coefficient of variance, among pooled samples is another parameter that performed poorly given its general acceptance as a quality scoring metric.
+In the full model, neither the traditional calculation of RSD (standard deviation divided by the mean) nor the robust implementation (median absolute deviation divided by the median) were significant parameters.
+This result was also reported by @Gloaguen2022 who noted that while the RSD was typically lower for high-quality features there were many noise peaks with low RSDs as well.
+We also note that though they do not report the statistical significance of the RSD shift across feature quality, it doesn't appear to be hugely significant without a very large sample size.
+
 We also showed that the automatically calculated SNR parameter from XCMS is not especially useful in distinguishing signal from noise.
 After inspecting a selection of peaks that had anomalous values for this metric, we are inclined to agree with @Myers2017a and conclude that this is often due to insufficient data outside of the peak for a robust calculation of noise level.
+
 Finally, we were surprised to find essentially no predictive power offered by peak area or intensity, with good peaks distributed almost identically to the bad peaks in this space.
@@
 
-RSD/CV problems also noted by @Gloaguen2022!
+#### Model selection and simplification
 

### 2023-06-05 manuscript/manuscript.Rmd
Updating authorship

@@
   metrics in metabolomics'
-author: "William Kumler"
+author: "William Kumler, Bryna Hazelton, and Anitra Ingalls"
 date: "`r Sys.Date()`"

### 2023-06-05 manuscript/manuscript.Rmd
Adding to CNN discussion after talking with Bryna

@@
   with(table(feat_class, pred_class)) %>%
-  caret::confusionMatrix(positive="Good", mode="everything")
+  caret::confusionMatrix(positive="Bad", mode="everything")
 ```
@@
 @Guo2021 presented EVA and reported an accuracy of 90-95%, a range inclusive of our accuracy on both the Falkor (92.1%) and MESOSCOPE (94.4%) datasets when using a likelihood threshold of 0.5.
-However, we note that accuracy alone can be a highly misleading statistic to report when working with datasets as unbalanced as these because very high accuracy can be obtained by simply classifying everything as bad, with a strong incentive to actually *increase* the number of bad peaks detected while doing so.
+However, we note that accuracy alone can be a highly misleading statistic to report when working with unbalanced datasets because very high accuracy can be obtained by simply classifying everything as bad, with a strong incentive to actually *increase* the number of bad peaks detected while doing so.
 This strategy, when applied to our data, returned accuracies in the 80-90% range despite being a useless classifier for downstream analysis.
-This is perhaps a slightly facetious argument, but many of the single-parameter logistic models we tested never exceeded a 50% likelihood and essentially functioned as this type of classifier (Supp figure XX).
-
-The class imbalance is partially why we chose to measure GPF and FDR instead of total accuracy.
-@Melnikov2020 reported these values as well in their presentation of peakonly, relative to which we obtained higher accuracy (they report 89% accuracy) but worse GPF and FDR than their model (89% and 3%, respectively, relative to our 80% GPF and 23% FDR overall).
-
-@Gloaguen2022 later introduced NeatMS, another CNN, and compared it directly to peakonly to claim equivalent or superior performance across a range of dilution factors.
+This is perhaps a facetious argument, but many of the single-parameter logistic models we tested never exceeded a 50% likelihood and essentially functioned as this type of classifier.
+
+The class imbalance, with mostly poor quality MFs, is partially why we chose to measure precision and recall instead of total accuracy.
+However, precision and recall can also be ambiguous when the positive class is not specified and the raw confusion matrices are unavailable, thus our very precise use of the FDR and GPF metrics as well as providing the raw confusion matrices in Supplemental Table XX.
+@Melnikov2020 reported precision and recall in their presentation of `peakonly`, relative to which we obtained higher accuracy (they report 89% accuracy) but worse GPF and FDR (89% and 3%, respectively, relative to our 77.1% GPF and 19.6% FDR overall).
+However, if we report precision and recall with the positive class set to "Bad", essentially trying to predict poor-quality MFs instead of good ones, our precision becomes 96.5% and our recall 95.7% due to the strong prior information about most MFs being bad.
+
+@Gloaguen2022 later introduced NeatMS, another CNN, and compared it directly to `peakonly` to claim equivalent or superior performance across a range of dilution factors.
 However, they do not report total precision or recall metrics in a comprehensive untargeted way but instead focus only on assessing the model's performance on known chemical standards.
-Additionally, they report a percentage of standards found for the peakonly model applied to their data and find that its performance is significantly lower (79.4%) than the recall reported in @Melnikov2020, perhaps indicating that the peakonly model is still overfit.
-
-While the model we present here has reduced performance relative to the CNNs, we would argue that its utility is not in maximizing performance but instead in maximizing interpretability.
-In particular, the CNNs provide no way to control the tradeoff between false positives and false negatives and no relative ranking of individual MF quality beyond the broad bins into which they are placed or explanation of relative metrics strengths for later analyses.
+They report a percentage of standards found for the `peakonly` model applied to their data and find that its performance is significantly lower (79.4%) than the recall reported in @Melnikov2020, perhaps indicating that the `peakonly` model is still overfit.
+
+While the model we present here likely has reduced performance relative to the CNNs, we would argue that its utility is not in maximizing performance but instead in maximizing interpretability.
+In particular, the CNNs provide no way to control the tradeoff between false positives and false negatives and no relative ranking of individual MF quality beyond the broad bins into which they are placed or explanation of relative metric strength for later analyses.
 
@@
 
-```{r}
-bind_rows(MS3000_features, FT2040_features) %>%
+```{r, eval=FALSE}
+bind_rows(MS3000_features, FT2040_features)
+FT2040_features %>%
   filter(feat_class%in%c("Good", "Bad")) %>%

### 2023-06-05 manuscript/manuscript.Rmd
Found the Kantz paper which does many of the same things

@@
 Classifying an entire feature at once not only has the advantage of reducing the amount of manual labor by a factor equal to the number of files in the run (typically 10s-100s smaller) but is also a better representation of the judgement made by an MS expert.
-Of course, this approach inherits all the difficulties of summary statistic calculations.
-The authors spent many hours discussing the implications of differences between mean and median values, but in general this approach produced more confidence than any single-file EIC.
+An exemplary implementation of this multi-file approach in prior work is reported in @Kantz2019, who compared the multi-file summary statistic model to a deep neural network and came to many similar conclusions.
 
@@
 
-These two measures of success were then used to explore different types of classification models for separating good molecular features (MFs) from bad, with a particular focus on quantifying the likelihood of each class rather than just returning the label.
-We found that a simple two-parameter logistic regression model trained on two novel metrics of peak quality had reasonably good performance on the training set and was highly robust when applied to novel datasets.
-The logistic regression in particular was favored over the random forest and regularized regression we tried due to their similar performance and increased interpretability (Supp.
-figure XX).
-
-#### Performance relative to other peak picking models
-
 ```{r, eval=FALSE}
+both_min_model <- rbind(FT2040_features, MS3000_features) %>%
+  select(feat_class, med_cor, med_SNR) %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_class=feat_class=="Good") %>%
+  glm(formula=feat_class~., family = binomial)
+
 FT2040_features %>%
@@
   with(table(feat_class, pred_class)) %>%
-  caret::confusionMatrix(positive="Bad", mode="everything")
-```
-
-This model performs less well than previous reports in the literature, but is highly simplified.
+  caret::confusionMatrix(positive="Good", mode="prec_recall")
+```
+
+We explored several different types of classification models for separating good molecular features (MFs) from bad, with a particular focus on quantifying the likelihood of each class rather than just returning the label.
+We found that a simple two-parameter logistic regression model trained on two novel metrics of peak quality had reasonably good performance on the training set and was highly robust when applied to novel datasets.
+The logistic regression in particular was favored over the random forest and regularized regression we tried due to their similar performance and increased interpretability (Supp.
+figure XX).
+
+This model outperforms the previously reported logistic regression model in @Kantz2019 and is highly simplified.
+There, they used a nineteen-parameter multiple logistic regression model and found a maximum performance of 80% GPF and an FDR of 34% on a cross-validated second cohort, similar to our cross-dataset testing.
+Our final two-parameter model also had an 80% GPF at a 0.5 likelihood threshold, but a significantly lower FDR of ~22.9%.
+This increased performance is likely due to our use of the metrics recalculated from the raw data, as their metrics were only calculated from the default MzMine2 peak parameters reported: similar to our XCMS-only model.
+
+#### Performance relative to recent deep learning methods
+
 @Guo2021 presented EVA and reported an accuracy of 90-95%, a range inclusive of our accuracy on both the Falkor (92.1%) and MESOSCOPE (94.4%) datasets when using a likelihood threshold of 0.5.
@@
 
-While the model we present here likely has reduced performance relative to the CNNs, we would argue that its utility is not in maximizing performance but instead in maximizing interpretability.
+While the model we present here likely has reduced performance relative to the CNNs, we would argue that its utility is not in maximizing performance but instead in maximizing interpretability, as previously noted by @Kantz2019.
 In particular, the CNNs provide no way to control the tradeoff between false positives and false negatives and no relative ranking of individual MF quality beyond the broad bins into which they are placed or explanation of relative metric strength for later analyses.
@@
 Currently, the use of the overall median value does a reasonable job at identifying MFs that appear in many samples but performs poorly when detecting MFs that appear in only a few.
-Also worth noting is that the calculation of any new metrics such as these that rely on access to the raw data require exact specification of the maximum and minimum *m/z* and retention time for a peak, values that are not always returned by peakpicking algorithms.
+Also worth noting is that the calculation of any new metrics such as these that rely on access to the raw data require exact specification of the maximum and minimum *m/z* and retention time for a peak, values that are not always returned by peakpicking algorithms and must be recalculated, as in @Kantz2019.
 
@@
   facet_wrap(~feat_class, ncol = 1, scales = "free_y")
-bind_rows(MS3000_features, FT2040_features) %>%
+bind_rows(, FT2040_features)
+MS3000_features %>%
   filter(feat_class%in%c("Good", "Bad")) %>%

### 2023-06-06 manuscript/manuscript.Rmd
Fixing clockplot bug depending on factor levels

@@
   right_join(best_feats, by = join_by(feature)) %>%
-  left_join(file_data %>% select(filename, depth), by = join_by(filename))
+  left_join(file_data %>% select(filename, depth), by = join_by(filename)) %>%
+  mutate(depth=factor(depth, levels=c("15m", "DCM", "175m")))
 

### 2023-06-06 manuscript/manuscript.Rmd
Adding first set of confusion matrices

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, eval=FALSE)
+knitr::opts_chunk$set(echo = FALSE, eval=TRUE)
 knitr::opts_knit$set(root.dir = '..')
@@
   # source("peakpicking_and_prep.R")
-  source("feature_extraction.R")
+  # source("feature_extraction.R")
 }
@@
   mutate(FDR=FP/(FP+TP)) %>%
-  mutate(GPF=TP/(FN+TP)) %>%
+  mutate(GPF=TP/(FN+TP))
+
+train_test_levels <- c("MS3000-MS3000", "FT2040-FT2040", "MS3000-FT2040", "FT2040-MS3000")
+train_test_labels <- c("MESO-MESO", "Falkor-Falkor", "MESO-Falkor", "Falkor-MESO")
+fdr_gpf_cross_train_gp <- fdr_gpf_cross_train_df %>%
   pivot_longer(c(FDR, GPF)) %>%
@@
                              levels=c("all_params", "xcms_params", "raw_data_params"))) %>%
-  mutate(cross_type_num=as.numeric(cross_type))
-
-train_test_levels <- c("MS3000-MS3000", "FT2040-FT2040", "MS3000-FT2040", "FT2040-MS3000")
-train_test_labels <- c("MESO-MESO", "Falkor-Falkor", "MESO-Falkor", "Falkor-MESO")
-fdr_gpf_cross_train_gp <- fdr_gpf_cross_train_df %>%
+  mutate(cross_type_num=as.numeric(cross_type)) %>%
   mutate(train_test=factor(train_test, levels=train_test_levels, labels=train_test_labels)) %>%
@@
   mutate(filename=paste0(filename, ".mzML")) %>%
-  rename(area="IS_area") %>%
+  dplyr::rename(area="IS_area") %>%
   filter(!str_detect(filename, "QC")) %>%
@@
 
-The two metrics in the final model were derived from the raw EIC data rather than any summary statistic calculated by XCMS automatically and were implemented because they matched our intuition about what makes an MF look good to an MS expert.
+Although the deep learning models show promise for peak quality recalibration, many mass-spectrometrists are reluctant to jump fully to their black-box nature.
+For this reason, we also reported here the relative power of individual parameters in our full model and use the results to dispel several myths about which parameters are useful in distinguishing signal from noise.
+
+The two metrics in the final model were rederived from the raw EIC data because they matched our intuition about what makes an MF look good to an MS expert.
 It is worth noting that these are very simple metrics and therefore fast to calculate, but we expect that more complicated metrics could perform even better.
@@
 
-The relative standard deviation (RSD), also called the coefficient of variance, among pooled samples is another parameter that performed poorly given its general acceptance as a quality scoring metric.
+The relative standard deviation (RSD), also called the coefficient of variance, among pooled samples is another parameter that performed surprisingly poorly given its general acceptance as a quality scoring metric.
 In the full model, neither the traditional calculation of RSD (standard deviation divided by the mean) nor the robust implementation (median absolute deviation divided by the median) were significant parameters.
@@
 
-We settled on the highly reduced model of just two parameters because we found that additional parameters often improved performance on the training set but did not do so significantly for the novel datasets where the application of such a model is actually useful (Figure XX).
+We settled on the highly reduced model of just two parameters because we found that additional parameters often improved performance on the training set but did not do so significantly for the novel datasets where the application of such a model is actually useful (Figure 2).
 One important caveat to note is that for the partially-labelled CultureData and Pttime datasets, there exists an uncontrolled degree of experimenter bias because the MS expert responsible for labeling did know that these peaks were all expected to be good.
@@
 
-We also found that this reduced two-parameter model was largely independent of the particular training set used, unlike in the more complex models (Figure XX).
+We also found that this reduced two-parameter model was largely independent of the particular training set used, unlike in the more complex models (Figure 3).
 This was true in both absolute likelihood as well as rank-ordered space, a particularly important distinction when one imagines manually labeling "down" the dataset where the researcher desires to start viewing the chromatograms associated with the very best features and eventually reaching a point where enough MFs have been reviewed or bad MFs are consistent enough that they decide to forgo additional labeling.
@@
   })) %>%
-  rename(`Raw data`="int") %>%
+  dplyr::rename(`Raw data`="int") %>%
   mutate(feature=factor(feature, levels=c("FT0588", "FT0280"), 
@@
 
-fdr_gpf_cross_train_df <- net_cross_train_df %>%
+fdr_gpf_cross_train_df_fancy <- net_cross_train_df %>%
   rbind(rf_cross_train_df) %>%
@@
 train_test_labels <- c("MESO-MESO", "Falkor-Falkor", "MESO-Falkor", "Falkor-MESO")
-fdr_gpf_cross_train_gp <- fdr_gpf_cross_train_df %>%
+fdr_gpf_cross_train_gp <- fdr_gpf_cross_train_df_fancy %>%
   mutate(train_test=factor(train_test, levels=train_test_levels, labels=train_test_labels)) %>%
@@
 ![Supplemental figure 3: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.](supplement/regregrf_fdrgpf_cross_train.png){width=90%}
+
+```{r confusion matrices for all the models for supplement}
+model_levels <- c("all_params", "xcms_params", "raw_data_params", 
+                  "ridge", "elastic", "lasso", "rf")
+model_labels <- c("Full model", "XCMS model", "Two-parameter model",
+                  "Ridge regression", "Elastic net", "Lasso regression", 
+                  "Random forest")
+fdr_gpf_individual <- fdr_gpf_cross_train_df %>%
+  bind_rows(fdr_gpf_cross_train_df_fancy %>% dplyr::rename(which_params=which_alpha)) %>%
+  separate("train_test", into = c("train", "test"), sep = "-") %>%
+  select(-`NA`) %>%
+  select(model_type=which_params, train, test, TP, TN, FP, FN, FDR, GPF) %>%
+  mutate(model_type=factor(model_type, levels=model_levels, labels=model_labels)) %>%
+  arrange(model_type)
+fdr_gpf_full_model <- bind_rows(lst(FT2040_features, MS3000_features), .id = "dataset") %>%
+  mutate(pred_prob=predict(object=both_min_model, newdata = ., type = "response")) %>%
+  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
+  mutate(error_type=case_when(
+    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
+    feat_class=="Bad" & pred_class=="Good" ~ "FP",
+    feat_class=="Good" & pred_class=="Bad" ~ "FN",
+    feat_class=="Good" & pred_class=="Good" ~ "TP"
+  )) %>%
+  drop_na() %>%
+  bind_rows(combined=., each=., .id = "calc_type") %>% 
+  mutate(calc_type=ifelse(calc_type=="each", dataset, calc_type)) %>%
+  select(-dataset) %>%
+  group_by(calc_type) %>%
+  count(error_type) %>%
+  ungroup() %>%
+  pivot_wider(names_from=error_type, values_from = n, values_fill = 0) %>%
+  mutate(FDR=FP/(FP+TP)) %>%
+  mutate(GPF=TP/(FN+TP)) %>%
+  mutate(model_type="Two-parameter model") %>%
+  mutate(train="Both") %>%
+  mutate(test=case_when(
+    calc_type=="FT2040_features"~"FT2040",
+    calc_type=="MS3000_features"~"MS3000",
+    calc_type=="combined"~"Both"
+  )) %>%
+  select(model_type, train, test, TP, TN, FP, FN, FDR, GPF)
+
+bind_rows(fdr_gpf_individual, fdr_gpf_full_model) %>%
+  write_csv("manuscript/supplement/model_confusion_matrices.csv")
+```

### 2023-06-06 manuscript/manuscript.Rmd
Adding stricter threshold confusion matrices

@@
   mutate(`GPF`=`%TP`/(`%FN`+`%TP`)) %>%
-  mutate(`GPF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `GPF`)) %>%
+  mutate(`GPF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `GPF`))
+
+fdr_gpf_all_datasets_gp <- error_type_all_datasets %>%
   pivot_longer(c(`FDR`, `GPF`)) %>%
-  drop_na()
-
-fdr_gpf_all_datasets_gp <- error_type_all_datasets %>%
+  drop_na() %>%
   mutate(desc=case_when(
@@
 
-```{r subsampling model stability supp figure}
+```{r subsampling model stability supp figure, warning=FALSE}
 set.seed(123)
@@
 
-bind_rows(fdr_gpf_individual, fdr_gpf_full_model) %>%
+fdr_gpf_strict_all <- bind_rows(lst(FT2040_features, MS3000_features), .id = "dataset") %>%
+  mutate(pred_prob=predict(object=both_min_model, newdata = ., type = "response")) %>%
+  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
+  mutate(error_type=case_when(
+    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
+    feat_class=="Bad" & pred_class=="Good" ~ "FP",
+    feat_class=="Good" & pred_class=="Bad" ~ "FN",
+    feat_class=="Good" & pred_class=="Good" ~ "TP"
+  )) %>%
+  drop_na() %>%
+  bind_rows(combined=., each=., .id = "calc_type") %>% 
+  mutate(calc_type=ifelse(calc_type=="each", dataset, calc_type)) %>%
+  select(-dataset) %>%
+  group_by(calc_type) %>%
+  count(error_type) %>%
+  ungroup() %>%
+  pivot_wider(names_from=error_type, values_from = n, values_fill = 0) %>%
+  mutate(FDR=FP/(FP+TP)) %>%
+  mutate(GPF=TP/(FN+TP)) %>%
+  mutate(model_type="Two-parameter model") %>%
+  mutate(train="Both") %>%
+  mutate(test=case_when(
+    calc_type=="FT2040_features"~"FT2040",
+    calc_type=="MS3000_features"~"MS3000",
+    calc_type=="combined"~"Both"
+  )) %>%
+  select(model_type, train, test, TP, TN, FP, FN, FDR, GPF)
+
+fdr_gpf_stricter_thresh <- error_type_all_datasets %>%
+  set_names(c("test", "FN", "FP", "TN", "TP", "FDR", "GPF")) %>%
+  mutate(model_type="Both (threshold 0.9)") %>%
+  mutate(train="Both") %>%
+  select(model_type, train, test, TP, TN, FP, FN, FDR, GPF)
+
+bind_rows(fdr_gpf_individual, fdr_gpf_full_model, fdr_gpf_stricter_thresh) %>%
   write_csv("manuscript/supplement/model_confusion_matrices.csv")

### 2023-06-06 manuscript/manuscript.Rmd
Switched from "peak" to "MF" terminology

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, eval=TRUE)
+knitr::opts_chunk$set(echo = FALSE, eval=FALSE)
 knitr::opts_knit$set(root.dir = '..')
@@
 Existing algorithms tend to favor false positives because downstream analyses can always remove erroneous mass features, but false negatives cannot be later recovered.[@Pirttila2022; @Gloaguen2022] However, this approach creates two additional downstream problems.
-First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of peaks, a task that scales combinatorially with the number of samples and compounds measured.[@Myers2017a] Second, false negatives cannot be fully accounted for in the way false positives can because false negatives are introduced in every step of metabolomic analyses from sample collection through injection.
-Instead of minimizing false negatives, we believe that emphasis should be placed on allowing the experimenter to set a threshold for the false discovery rate and accept that this will inherently add to the number of peaks already lost in the data collection process.
+First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of molecular features (MFs), a task that scales combinatorially with the number of samples and compounds measured.[@Myers2017a] Second, false negatives cannot be fully accounted for in the way false positives can because false negatives are introduced in every step of metabolomic analyses from sample collection through injection.
+Instead of minimizing false negatives, we believe that emphasis should be placed on allowing the experimenter to set a threshold for the false discovery rate and accept that this will inherently add to the number of MFs already lost in the data collection process.
 
@@
 
-A single parameter of peak quality also facilitates downstream analyses in multiple ways.
+A single parameter of MF quality also facilitates downstream analyses in multiple ways.
 This metric would improve statistical power by reducing the number of effective hypotheses tested and allow researchers to focus effort on features least likely to be noise.
 Additionally, this parameter could be optimized to improve peakpicking and chromatographic settings independently of the software used and minimize inter-lab variability when scripted to provide consistent, reproducible results independent of the particular expert reviewing its performance.
-Constructing such a single comprehensive metric calibrated to likelihood is also more effective than multiple independent thresholds because it has meaningful units, does not require estimating the relative power of individual metrics, and allows a good peak to compensate for weak performance in one area with strong performance in other metrics, e.g.
+Constructing such a single comprehensive metric calibrated to likelihood is also more effective than multiple independent thresholds because it has meaningful units, does not require estimating the relative power of individual metrics, and allows a good MF to compensate for weak performance in one area with strong performance in other metrics, e.g.
 as implemented in @Pirttila2022.
@@
 In this paper, we use open ocean marine metabolite LC-MS samples to develop and test a variety of chromatographic peak metrics.
-Here, we construct and validate multiple predictive models of peak quality based on metrics both common in the literature and custom implementations we've found useful in our own analysis.
+Here, we construct and validate multiple predictive models of MF quality based on metrics both common in the literature and custom implementations we've found useful in our own analysis.
 This allows us to connect the physical, chemical, and biological measurements taken regularly around the globe to a molecular-scale perspective of particulate organic matter in the ocean by linking the chemical currencies that fuel the planet to the environments in which they're found.
@@
 An average of 3,300 molecular features (MFs) were reported by XCMS across the 4 datasets, with the fewest (1,495) in the Falkor data and the most in the Pttime samples (7,781).
-In the Falkor and MESOSCOPE datasets that were fully labeled by an MS expert, approximately 70% (69% and 73%, respectively) of the features were given a "Bad" designation, corresponding to noise peaks that the expert would not have included in a downstream analysis.
+In the Falkor and MESOSCOPE datasets that were fully labeled by an MS expert, approximately 70% (69% and 73%, respectively) of the features were given a "Bad" designation, corresponding to noise MFs that the expert would not have included in a downstream analysis.
 In both, 5% of the MFs were unable to be assigned confidently to either "Good" or "Bad" classes and 10% were identified as appearing only in the standards, leaving only ~15% of the features classified as "Good" (16% and 12%, respectively).
@@
 According to all three logistic regression models (see Methods), the majority of MFs were estimated to have a less than 1% chance of being good.
-The full model (containing all possible peak metrics) and the XCMS model (built on only those metrics calculated from the XCMS output) both displayed a strongly bimodal distribution, with a large number of peaks also exceeding a 99% chance of being good, while the two-parameter model (consisting of the novel SNR metric and the peak shape correlation metric) had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Figure 1).
+The full model (containing all possible peak metrics) and the XCMS model (built on only those metrics calculated from the XCMS output) both displayed a strongly bimodal distribution, with a large number of MFs also exceeding a 99% chance of being good, while the two-parameter model (consisting of the novel SNR metric and the peak shape correlation metric) had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Figure 1).
 
@@
 ![Figure 1: Histograms showing the estimated likelihood of a given mass feature being categorized as "Good" according to the two-parameter logistic model trained on the combined fully-labeled Falkor and MESOSCOPE datasets.
-Colors indicate the category in which each feature was manually assigned by an expert, with "Stans only" referring to a good peak that was only visible in the standards run alongside the samples.
+Colors indicate the category in which each feature was manually assigned by an expert, with "Stans only" referring to a good molecular feature that was only visible in the standards run alongside the samples.
 Datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (black vertical line) and were otherwise unclassified.](figures/pred_prob_class_color_hists.png){width=90%}
 
-The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR values in the 5-10% range and 80-90% GPF values implying that a large majority of the good peaks passed the threshold with very little noise included.
+The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR values in the 5-10% range and 80-90% GPF values implying that a large majority of the good MFs passed the threshold with very little noise included.
 The XCMS metrics performed slightly worse, with FDR values in the 10-15% range and GPF values closer to 75%.
@@
 However, when the models were trained on a different dataset than the one they were used to predict classifications for, all models had similar performance with %FDR around 10-25 and %GPF around 60-80.
-The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more peaks recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE (Figure 2).
+The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more MFs recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE (Figure 2).
 
@@
 ![Figure 2: False discovery rate (%FDR) and fraction of good peaks found (GPF) plotted across different subsets of model parameters.
-Lower %FDR indicates a smaller fraction of false positives among those peaks the model categorized as "Good" using a threshold of 0.5, and higher %GPF indicates a larger fraction of the total good peaks were found using the same threshold.
+Lower %FDR indicates a smaller fraction of false positives among those molecular features the model categorized as "Good" using a threshold of 0.5, and higher %GPF indicates a larger fraction of the total good features were found using the same threshold.
 Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel.
@@
 A majority of the time, the estimates from the two models disagreed by more than 2x the standard error of the estimate.
-Some parameters disagreed not only in magnitude but also in sign, with the Falkor-trained full model increasing peak goodness likelihood with larger ppm variation and a wider peak width, while the MESOSCOPE-trained full model had negative estimates for each of these parameters.
+Some parameters disagreed not only in magnitude but also in sign, with the Falkor-trained full model increasing MF goodness likelihood with larger ppm variation and a wider peak width, while the MESOSCOPE-trained full model had negative estimates for each of these parameters.
 Notably, the peak shape and novel SNR parameters used in the two-parameter model were among the most robust to training model variation, potentially explaining the consistency described above (Supp.
@@
 
-We settled on a 90% likelihood threshold for application to novel datasets because it struck a balance between the number of peaks we estimated to be necessary for robust testing while still remaining reasonable to manually label.
+We settled on a 90% likelihood threshold for application to novel datasets because it struck a balance between the number of MFs we estimated to be necessary for robust testing while still remaining reasonable to manually label.
 For the CultureData dataset, we obtained 1,790 total mass features, 192 of which had predicted likelihoods above 0.9.
@@
 For the Pttime dataset, 7,781 were obtained with 400 flagged by the model as "Good".
-348 were truly good peaks, 35 were ambiguous, and 17 were "Bad".
+348 were truly good MFs, 35 were ambiguous, and 17 were "Bad".
 No standards were run during this analysis, so there were no features in that category.
@@
 With the stricter threshold, we obtained FDR values consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for Falkor, MESOSCOPE, CultureData, and Pttime respectively (Figure 4).
-Of course, this low error rate required that we miss out on additional potentially valuable features, with only a fraction of the total good peaks making it past this threshold.
-In both the Falkor and MESOSCOPE datasets, less than half of the good peaks were labeled as such, with actual values of 39.4% and 26.5%, respectively.
+Of course, this low error rate required that we miss out on additional potentially valuable features, with only a fraction of the total good MFs making it past this threshold.
+In both the Falkor and MESOSCOPE datasets, less than half of the good MFs were labeled as such, with actual values of 39.4% and 26.5%, respectively.
 Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the %GPF but expect it to be in a similar range (Figure 4).
@@
 
-![Figure 4: False discovery rate and proportion of total good peaks identified as good by the two-parameter model trained on the combined MESOSCOPE/Falkor dataset and applied to each dataset individually with a threshold of 0.9.
-FDR is calculated by dividing the number of false positives by the total positives produced by the model and GPF is calculated by dividing the number of true positives by the total number of good peaks as identified manually.
+![Figure 4: False discovery rate and proportion of total good features identified as good by the two-parameter model trained on the combined MESOSCOPE/Falkor dataset and applied to each dataset individually with a threshold of 0.9.
+FDR is calculated by dividing the number of false positives by the total positives produced by the model and GPF is calculated by dividing the number of true positives by the total number of good features as identified manually.
 Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fdr_gpf_all_datasets.png){width=30%}
@@
 
-A different story emerged, however, when the bad peaks were removed from this analysis.
+A different story emerged, however, when the bad MFs were removed from this analysis.
 Good features were most commonly found to have their highest concentrations at the DCM or the surface, rather than being fixed with respect to depth.
-Of the 182 good peaks, less than a fifth had no trend with depth (44/249) and a majority had unequivocally lowest values in the 175 meter samples (those with 15m/DCM > 175m, 145 features).
+Of the 182 good MFs, less than a fifth had no trend with depth (44/249) and a majority had unequivocally lowest values in the 175 meter samples (those with 15m/DCM > 175m, 145 features).
 The two-parameter model, when applied with a 50% likelihood threshold, also recovered this general feature distribution and classified many of the features with no significant depth signal as likely to be bad [Figure 7B, currently in Slack?].
@@
 Additionally, a large number of features manually identified as bad nonetheless had significant differences with depth.
-This was surprising because we had assumed that bad peaks corresponded to instrument noise, which we did not expect to have any biological trend.
-Further investigation of a few randomly selected bad features with a biological difference revealed the reason behind this: most of those investigated were actually tails of other peaks.
+This was surprising because we had assumed that bad MFs corresponded to instrument noise, which we did not expect to have any biological trend.
+Further investigation of a few randomly selected bad features with a biological difference revealed the reason behind this: most of those investigated were actually tails of other MFs.
 Integrating just the tail of a peak retains the biological signal of the full peak while still looking visually like instrument noise, thereby introducing pseudoreplication in the feature space.
@@
 The model did fail to recover some interesting biological variation, however.
-Two features of particular interest were those good peaks with a DCM minimum (DCM < 15m = 175m), both of which were missed by the two-parameter model.
+Two features of particular interest were those good MFs with a DCM minimum (DCM < 15m = 175m), both of which were missed by the two-parameter model.
 These features possess an unexpected biological signal that does not track with depth or other common oceanographic parameters, thereby potentially representing an interesting biomarker or pathway that's downregulated in excess of the biomass effect.
@@
 The default XCMS output continued the trend observed above, as expected, with the least variance explained and the lowest R$^2$ value.
-Subsetting for the "Good" peaks only, however, did not actually return the highest F-ratio or R$^2$, instead falling between the 50% and 90% thresholds for these two metrics.
+Subsetting for the "Good" MFs only, however, did not actually return the highest F-ratio or R$^2$, instead falling between the 50% and 90% thresholds for these two metrics.
 In large part this is due to the much smaller number of features: 249 features were manually labeled as Good, while only 75 exceeded the 90% likelihood threshold.
 
-The relative power of identifying only the very best peaks was also illustrated visually with non-metric multidimensional scaling (NMDS) plots (Figure XX).
-In these common exploratory plots, the features with likelihoods above 50% strongly separated by depth while lower thresholds disguised the true signal and had higher stress values.
-Performing an NMDS on the manually-identified Good peaks resulted in output nearly indistinguishable from those of the 90% and 50% thresholds, while the 0.1 and 0.01 thresholds looked almost identical to the same analysis performed on all the XCMS features.
+The relative power of identifying only the very best MFs was also illustrated visually with non-metric multidimensional scaling (NMDS) plots (Figure XX).
+In these common exploratory plots, the MFs with likelihoods above 50% strongly separated by depth while lower thresholds disguised the true signal and had higher stress values.
+Performing an NMDS on the manually-identified Good MFs resulted in output nearly indistinguishable from those of the 90% and 50% thresholds, while the 0.1 and 0.01 thresholds looked almost identical to the same analysis performed on all the XCMS features.
 
@@
 We used two fully-labeled and two partially-labeled HILIC LC-MS datasets to assess the performance of the XCMS algorithm and construct a robust model of peak quality.
-To measure performance, we used two measurements of success closely tied to intuitive questions about a dataset: the percentage of total good peaks found (GPF, also known as recall or sensitivity) and the percentage of bad peaks included or false discovery rate (FDR).
+To measure performance, we used two measurements of success closely tied to intuitive questions about a dataset: the percentage of total good peaks found (GPF, also known as recall or sensitivity) and the percentage of bad molecular features (MFs) included, also known as the false discovery rate (FDR).
 We decided against using the F$_1$ score as an overall summary statistic because false negatives and false positives have very different implications in this context and should be treated separately.
@@
 @Guo2021 presented EVA and reported an accuracy of 90-95%, a range inclusive of our accuracy on both the Falkor (92.1%) and MESOSCOPE (94.4%) datasets when using a likelihood threshold of 0.5.
-However, we note that accuracy alone can be a highly misleading statistic to report when working with unbalanced datasets because very high accuracy can be obtained by simply classifying everything as bad, with a strong incentive to actually *increase* the number of bad peaks detected while doing so.
+However, we note that accuracy alone can be a highly misleading statistic to report when working with unbalanced datasets because very high accuracy can be obtained by simply classifying everything as bad, with a strong incentive to actually *increase* the number of bad MFs detected while doing so.
 This strategy, when applied to our data, returned accuracies in the 80-90% range despite being a useless classifier for downstream analysis.
@@
 We were surprised at the poor performance of several other metrics.
-The isotope information in particular was expected to be a very strong predictor of peak quality given previous work that uses this metric extensively.[@Libiseller2015; @Treutler2016; @ElAbiead2021] We learned that many noise peaks still have reliable isotopes (perhaps unsurprising, given that the noise is in fact often caused by solvents or contaminants that are still chemical in nature) and that many real peaks are simply too low-intensity to have them in this kind of dilute environmental sample.
+The isotope information in particular was expected to be a very strong predictor of MF quality given previous work that uses this metric extensively.[@Libiseller2015; @Treutler2016; @ElAbiead2021] We learned that many noise MFs still have reliable isotopes (perhaps unsurprising, given that the noise is in fact often caused by solvents or contaminants that are still chemical in nature) and that many real MFs are simply too low-intensity to have them in this kind of dilute environmental sample.
 
@@
 In the full model, neither the traditional calculation of RSD (standard deviation divided by the mean) nor the robust implementation (median absolute deviation divided by the median) were significant parameters.
-This result was also reported by @Gloaguen2022 who noted that while the RSD was typically lower for high-quality features there were many noise peaks with low RSDs as well.
+This result was also reported by @Gloaguen2022 who noted that while the RSD was typically lower for high-quality features there were many noise MFs with low RSDs as well.
 We also note that though they do not report the statistical significance of the RSD shift across feature quality, it doesn't appear to be hugely significant without a very large sample size.
@@
 We also showed that the automatically calculated SNR parameter from XCMS is not especially useful in distinguishing signal from noise.
-After inspecting a selection of peaks that had anomalous values for this metric, we are inclined to agree with @Myers2017a and conclude that this is often due to insufficient data outside of the peak for a robust calculation of noise level.
-
-Finally, we were surprised to find essentially no predictive power offered by peak area or intensity, with good peaks distributed almost identically to the bad peaks in this space.
+After inspecting a selection of MFs that had anomalous values for this metric, we are inclined to agree with @Myers2017a and conclude that this is often due to insufficient data outside of the peak for a robust calculation of noise level.
+
+Finally, we were surprised to find essentially no predictive power offered by peak area or intensity, with good MFs distributed almost identically to the bad MFs in this space.
 This cautions strongly against an arbitrarily-decided intensity threshold for winnowing down the number of MFs, in agreement with previous work.[@Houriet2022; @Barupal2021]
@@
 We settled on the highly reduced model of just two parameters because we found that additional parameters often improved performance on the training set but did not do so significantly for the novel datasets where the application of such a model is actually useful (Figure 2).
-One important caveat to note is that for the partially-labelled CultureData and Pttime datasets, there exists an uncontrolled degree of experimenter bias because the MS expert responsible for labeling did know that these peaks were all expected to be good.
-However, given that we do still see poor-quality peaks in this set indicates that this was not an overwhelming bias.
+One important caveat to note is that for the partially-labelled CultureData and Pttime datasets, there exists an uncontrolled degree of experimenter bias because the MS expert responsible for labeling did know that these MFs were all expected to be good.
+However, given that we do still see poor-quality MFs in this set indicates that this was not an overwhelming bias.
 The drastic drop in performance on out-of-sample data was particularly concerning because it creates overconfidence in the true level of noise actually ending up in the final dataset.
@@
 In the multivariate case, we ran the same analysis of PERMANOVAs and NMDS plots on various subsets of the original XCMS output and found that the effect size of depth was strongly influenced by by the threshold chosen.
-This is unsurprising given that most noise peaks should not have a biological signal to begin with, but is troubling for interpreting analyses where the FDR is not reported or the dataset not manually reviewed because the absence of a notable effect could simply be due to the overwhelming degree of noise in the default output.
+This is unsurprising given that most noise MFs should not have a biological signal to begin with, but is troubling for interpreting analyses where the FDR is not reported or the dataset not manually reviewed because the absence of a notable effect could simply be due to the overwhelming degree of noise in the default output.
 
@@
 While some of these are inherently due to the likelihood of getting a small p-value with enough attempts despite FDR correction, a larger number of these poor-quality MFs were due to partial integration in which only the tail of a feature was integrated.
-This essentially duplicates the signal of the original peak in later analyses and should be removed.
+This essentially duplicates the signal of the original MF in later analyses and should be removed.
 The real features showed a strong biological trend of high concentration throughout the surface ocean and down through the deep chlorophyll maximum (DCM), with most features equally abundant at 15 meters and this ~110 meter depth feature before dropping off at depth.
@@
 
-The large number of noise peaks present in metabolomics datasets can be controlled using a simple logistic classification model.
+The large number of molecular features due to noise present in metabolomics datasets can be controlled using a simple logistic classification model.
 We trained such a model on two full-labeled open ocean HILIC datasets and found that the best performing parameters in the model for were a custom signal-to-noise metric and a test of similarity to a bell curve.
@@
 Inspection consisted of plotting the raw intensity values against the corrected retention-time values for all data points within the *m/z* x RT bounding box determined by the most extreme values for the given feature.
-For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labeling and to more accurately represent what MS experts typically do when assessing the quality of a given peak (Figure 7).
+For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labeling and to more accurately represent what MS experts typically do when assessing the quality of a given molecular feature (Figure 7).
 We also decided to ignore missing values and linearly interpolate between known data points rather than filling with zeroes.
-These EICs were then shown to an MS expert for classification into one of 4 categories: Good, Bad, Ambiguous, or Stans only if the peak appeared to only show up in the standards.
+These EICs were then shown to an MS expert for classification into one of 4 categories: Good, Bad, Ambiguous, or Stans only if the feature appeared to only show up in the standards.
 A few randomly-chosen features from the manually-assigned Good and Bad classifications are shown in Figure 7.
@@
 Although we requested additional parameters from XCMS with the verboseColumns = TRUE argument, all of the values returned were NAs and could not be used in the model building so were dropped.
-The total number of peaks found was calculated as the number of peaks divided by the total number of files, and the total number of files in which a peak was initially detected was calculated by subtracting the number of NAs in the peak baseline estimate (which is NA if the peak was filled in) from the total number of files and then dividing by the number of files to normalize across datasets.
-This calculation were performed for the Falkor, MESOSCOPE, and CultureData datasets to estimate the proportion of sample and standard files in which a given peak was found, but could not be supplied for the Pttime dataset because no standard files were available.
+The total number of peaks found for a given feature was calculated as the number of peaks divided by the total number of files, and the total number of files in which a peak was initially detected was calculated by subtracting the number of NAs in the peak baseline estimate (which is NA if the peak was filled in) from the total number of files and then dividing by the number of files to normalize across datasets.
+This calculation were performed for the Falkor, MESOSCOPE, and CultureData datasets to estimate the proportion of sample and standard files in which a given molecular feature (MF) was found, but could not be calculated for the Pttime dataset because no standard files were available.
 The sn metric contained a large number of zero values to begin with that then became negative infinities when log-scaled, so we replaced those values with zeroes and functionally equated an sn of 0 with an sn of 1.
@@
 Both the isotope shape similarity and the isotope area correlation were used as metrics in the downstream analysis.
-Peaks for which no isotope signal was detected or had too few scans to calculate the above metrics were imputed with NA values that were again dropped in the calculation of summary statistics.
+Peaks for which no isotope signal was detected or had too few scans to calculate the above metrics were imputed with NA values that were again dropped in the calculation of summary statistics for the molecular feature as a whole.
 
@@
 This comparison was done in both the raw probability space as well as a rank-ordered space to test whether the most extreme likelihood (i.e.
-very best and very worst) peaks were consistently found to be most extreme independently of the actual likelihood predicted.
+very best and very worst) MFs were consistently found to be most extreme independently of the actual likelihood predicted.
 For the raw probability space we compared the predictions using Pearson's correlation coefficient, while Spearman's rank-ordered coefficient was used for the ranked space.
@@
 
-Each of these datasets were only fractionally labeled, with those peaks above the 0.9 likelihood threshold according to the two-parameter model reviewed manually and categorized.
-This stricter threshold was chosen because we felt less comfortable interpreting results based on mass features that were only 50% likely to be real, but did not feel the need to be so strict with this exploratory analysis that we wanted to limit it to 99+% likelihood peaks.
+Each of these datasets were only fractionally labeled, with those MFs above the 0.9 likelihood threshold according to the two-parameter model reviewed manually and categorized.
+This stricter threshold was chosen because we felt less comfortable interpreting results based on mass features that were only 50% likely to be real, but did not feel the need to be so strict with this exploratory analysis that we wanted to limit it to 99+% likelihood MFs.
 
@@
 These univariate tests were then controlled for multiply hypothesis testing using R's `p.adjust` function with method `fdr`.[@Benjamini1995] We also performed post-hoc Dunn tests provided by the `rstatix` package to categorize the response to depth for those mass features for which the KW test was significant, with responses falling into one of the 14 classes possible when permuting the sign and significance of the Dunn test.[@Dunn1964] p-values obtained from the Dunn tests were not FDR controlled because it was used as a categorization tool rather than a null hypothesis test.
-In multivariate space, we used a permutational MANOVA (PERMANOVA) [@Anderson2014] provided by the vegan package's `adonis2` function to test for multivariate differences in structure of the metabolome with depth.[@veganpkg] We ran multiple PERMANOVAs with a different subset of mass features included each time, corresponding to using the output from XCMS directly, likelihood thresholds of 0.01, 0.1, 0.5, 0.9, and finally only those peaks manually annotated as good.
+In multivariate space, we used a permutational MANOVA (PERMANOVA) [@Anderson2014] provided by the vegan package's `adonis2` function to test for multivariate differences in structure of the metabolome with depth.[@veganpkg] We ran multiple PERMANOVAs with a different subset of mass features included each time, corresponding to using the output from XCMS directly, likelihood thresholds of 0.01, 0.1, 0.5, 0.9, and finally only those MFs manually annotated as good.
 

### 2023-06-06 manuscript/manuscript.Rmd
Switched from GPF to GFF and recompiled everything

@@
 
-The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR values in the 5-10% range and 80-90% GPF values implying that a large majority of the good MFs passed the threshold with very little noise included.
-The XCMS metrics performed slightly worse, with FDR values in the 10-15% range and GPF values closer to 75%.
-The two-parameter model performed worst when tested internally, with an FDR of about 20% and %GPF also around 75% (Figure 2).
-However, when the models were trained on a different dataset than the one they were used to predict classifications for, all models had similar performance with %FDR around 10-25 and %GPF around 60-80.
+The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR values in the 5-10% range and 80-90% GFF values implying that a large majority of the good MFs passed the threshold with very little noise included.
+The XCMS metrics performed slightly worse, with FDR values in the 10-15% range and GFF values closer to 75%.
+The two-parameter model performed worst when tested internally, with an FDR of about 20% and GFF also around 75% (Figure 2).
+However, when the models were trained on a different dataset than the one they were used to predict classifications for, all models had similar performance with %FDR around 10-25 and GFF around 60-80.
 The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more MFs recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE (Figure 2).
 
-```{r Fig2 FDR vs GPF cross-train triangle figure feature selection}
-fdr_gpf_cross_train_df <- lst(all_params, raw_data_params, xcms_params) %>%
+```{r Fig2 FDR vs GFF cross-train triangle figure feature selection}
+fdr_gff_cross_train_df <- lst(all_params, raw_data_params, xcms_params) %>%
   imap(function(param_selection, param_name){
@@
   mutate(FDR=FP/(FP+TP)) %>%
-  mutate(GPF=TP/(FN+TP))
+  mutate(GFF=TP/(FN+TP))
 
@@
 train_test_labels <- c("MESO-MESO", "Falkor-Falkor", "MESO-Falkor", "Falkor-MESO")
-fdr_gpf_cross_train_gp <- fdr_gpf_cross_train_df %>%
-  pivot_longer(c(FDR, GPF)) %>%
+fdr_gff_cross_train_gp <- fdr_gff_cross_train_df %>%
+  pivot_longer(c(FDR, GFF)) %>%
   mutate(name=paste0("%", name)) %>%
@@
   mutate(train_test=factor(train_test, levels=train_test_levels, labels=train_test_labels)) %>%
-  mutate(name=factor(name, levels=c("%GPF", "%FDR"), labels=c("GPF", "FDR"))) %>%
+  mutate(name=factor(name, levels=c("%GFF", "%FDR"), labels=c("GFF", "FDR"))) %>%
   ggplot() +
@@
                      aesthetics = c("color", "fill")) +
-  scale_shape_manual(breaks = c("GPF", "FDR"), values = c("\u25B2", "\u25BC")) +
+  scale_shape_manual(breaks = c("GFF", "FDR"), values = c("\u25B2", "\u25BC")) +
   coord_cartesian(ylim = c(0, 1)) +
@@
 
-ggsave(plot = fdr_gpf_cross_train_gp, 
-       filename = "fdr_gpf_cross_train.png", 
+ggsave(plot = fdr_gff_cross_train_gp, 
+       filename = "fdr_gff_cross_train.png", 
        device = "png", path = "manuscript/figures",
@@
 
-![Figure 2: False discovery rate (%FDR) and fraction of good peaks found (GPF) plotted across different subsets of model parameters.
-Lower %FDR indicates a smaller fraction of false positives among those molecular features the model categorized as "Good" using a threshold of 0.5, and higher %GPF indicates a larger fraction of the total good features were found using the same threshold.
+![Figure 2: False discovery rate (%FDR) and fraction of good features found (%GFF) plotted across different subsets of model parameters.
+Lower %FDR indicates a smaller fraction of false positives among those molecular features the model categorized as "Good" using a threshold of 0.5, and higher %GFF indicates a larger fraction of the total good features were found using the same threshold.
 Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel.
-Lines of best fit have been estimated and plotted in black on top of the data points.](figures/fdr_gpf_cross_train.png){width=90%}
+Lines of best fit have been estimated and plotted in black on top of the data points.](figures/fdr_gff_cross_train.png){width=90%}
 
@@
 None of the penalized regression models significantly improved cross-validated performance between the MESOSCOPE and Falkor datasets when measured by both initial performance and the performance drop when applied across datasets.
-All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GPF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent false negative acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp.
+All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GFF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent false negative acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp.
 figure 3).
-The random forest model, interestingly, had perfect predictive capacity when tested internally on the training data (FDR=0%, GPF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp.
+The random forest model, interestingly, had perfect predictive capacity when tested internally on the training data (FDR=0%, GFF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp.
 figure 3).
@@
 In both the Falkor and MESOSCOPE datasets, less than half of the good MFs were labeled as such, with actual values of 39.4% and 26.5%, respectively.
-Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the %GPF but expect it to be in a similar range (Figure 4).
-
-```{r FDR vs GPF with additional datasets}
+Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the %GFF but expect it to be in a similar range (Figure 4).
+
+```{r FDR vs GFF with additional datasets}
 # Currently a fairly simple figure, may need additional data?
@@
   mutate(`FDR`=`%FP`/(`%FP`+`%TP`)) %>%
-  mutate(`GPF`=`%TP`/(`%FN`+`%TP`)) %>%
-  mutate(`GPF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `GPF`))
-
-fdr_gpf_all_datasets_gp <- error_type_all_datasets %>%
-  pivot_longer(c(`FDR`, `GPF`)) %>%
+  mutate(`GFF`=`%TP`/(`%FN`+`%TP`)) %>%
+  mutate(`GFF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `GFF`))
+
+fdr_gff_all_datasets_gp <- error_type_all_datasets %>%
+  pivot_longer(c(`FDR`, `GFF`)) %>%
   drop_na() %>%
@@
     name=="FDR"~ paste0("frac(",`%FP`,",",`%FP`+`%TP`,")"),
-    name=="GPF"~ paste0("frac(",`%TP`,",",`%FN`+`%TP`,")")
+    name=="GFF"~ paste0("frac(",`%TP`,",",`%FN`+`%TP`,")")
   )) %>%
@@
     name=="FDR" ~ -0.3,
-    name=="GPF" ~ 1.3,
+    name=="GFF" ~ 1.3,
   )) %>%
-  mutate(name=factor(name, levels=c("GPF", "FDR"),
-                     labels=c("Fraction of good peaks found (GPF)", 
+  mutate(name=factor(name, levels=c("GFF", "FDR"),
+                     labels=c("Fraction of good features found (GFF)", 
                               "False discovery rate (FDR)"))) %>%
@@
 
-ggsave(plot = fdr_gpf_all_datasets_gp, 
-       filename = "fdr_gpf_all_datasets.png", 
+ggsave(plot = fdr_gff_all_datasets_gp, 
+       filename = "fdr_gff_all_datasets.png", 
        device = "png", path = "manuscript/figures",
@@
 ![Figure 4: False discovery rate and proportion of total good features identified as good by the two-parameter model trained on the combined MESOSCOPE/Falkor dataset and applied to each dataset individually with a threshold of 0.9.
-FDR is calculated by dividing the number of false positives by the total positives produced by the model and GPF is calculated by dividing the number of true positives by the total number of good features as identified manually.
-Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fdr_gpf_all_datasets.png){width=30%}
+FDR is calculated by dividing the number of false positives by the total positives produced by the model and GFF is calculated by dividing the number of true positives by the total number of good features as identified manually.
+Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fdr_gff_all_datasets.png){width=30%}
 
@@
   mutate(filename=paste0(filename, ".mzML")) %>%
-  dplyr::rename(area="IS_area") %>%
+  dplyr::rename(IS_area="area") %>%
   filter(!str_detect(filename, "QC")) %>%
@@
   mutate(error_type=factor(error_type, levels=c("TP", "FP", "FN", "TN"),
-                           labels=c("Good peak found", "Bad peak included",
-                                    "Good peak missed", "Bad peak avoided")))
+                           labels=c("Good MF found", "Bad MF included",
+                                    "Good MF missed", "Bad MF avoided")))
 
@@
              data = clockplot_complete_df) +
-  scale_shape_manual(breaks = c("Good peak found", "Bad peak included",
-                                "Good peak missed", "Bad peak avoided"), 
+  scale_shape_manual(breaks = c("Good MF found", "Bad MF included",
+                                "Good MF missed", "Bad MF avoided"), 
                      values = c(19, 3, 4, 1))
@@
 We used two fully-labeled and two partially-labeled HILIC LC-MS datasets to assess the performance of the XCMS algorithm and construct a robust model of peak quality.
-To measure performance, we used two measurements of success closely tied to intuitive questions about a dataset: the percentage of total good peaks found (GPF, also known as recall or sensitivity) and the percentage of bad molecular features (MFs) included, also known as the false discovery rate (FDR).
+To measure performance, we used two measurements of success closely tied to intuitive questions about a dataset: the percentage of total good features found (GFF, also known as recall or sensitivity) and the percentage of bad molecular features (MFs) included, also known as the false discovery rate (FDR).
 We decided against using the F$_1$ score as an overall summary statistic because false negatives and false positives have very different implications in this context and should be treated separately.
@@
 This model outperforms the previously reported logistic regression model in @Kantz2019 and is highly simplified.
-There, they used a nineteen-parameter multiple logistic regression model and found a maximum performance of 80% GPF and an FDR of 34% on a cross-validated second cohort, similar to our cross-dataset testing.
-Our final two-parameter model also had an 80% GPF at a 0.5 likelihood threshold, but a significantly lower FDR of ~22.9%.
+There, they used a nineteen-parameter multiple logistic regression model and found a maximum performance of 80% GFF and an FDR of 34% on a cross-validated second cohort, similar to our cross-dataset testing.
+Our final two-parameter model also had an 80% GFF at a 0.5 likelihood threshold, but a significantly lower FDR of ~22.9%.
 This increased performance is likely due to our use of the metrics recalculated from the raw data, as their metrics were only calculated from the default MzMine2 peak parameters reported: similar to our XCMS-only model.
@@
 The class imbalance, with mostly poor quality MFs, is partially why we chose to measure precision and recall instead of total accuracy.
-However, precision and recall can also be ambiguous when the positive class is not specified and the raw confusion matrices are unavailable, thus our very precise use of the FDR and GPF metrics as well as providing the raw confusion matrices in Supplemental Table XX.
-@Melnikov2020 reported precision and recall in their presentation of `peakonly`, relative to which we obtained higher accuracy (they report 89% accuracy) but worse GPF and FDR (89% and 3%, respectively, relative to our 77.1% GPF and 19.6% FDR overall).
+However, precision and recall can also be ambiguous when the positive class is not specified and the raw confusion matrices are unavailable, thus our very precise use of the FDR and GFF metrics as well as providing the raw confusion matrices in Supplemental Table XX.
+@Melnikov2020 reported precision and recall in their presentation of `peakonly`, relative to which we obtained higher accuracy (they report 89% accuracy) but worse GFF and FDR (89% and 3%, respectively, relative to our 77.1% GFF and 19.6% FDR overall).
 However, if we report precision and recall with the positive class set to "Bad", essentially trying to predict poor-quality MFs instead of good ones, our precision becomes 96.5% and our recall 95.7% due to the strong prior information about most MFs being bad.
@@
   mutate(feature=factor(feature, levels=c("FT0588", "FT0280"), 
-                        labels=c("Good peak", "Bad peak")))
+                        labels=c("Good MF", "Bad MF")))
 summary_stats <- perf_peak_data %>%
@@
 In each case, we categorized each molecular feature as a true positive (TP) if it was predicted to be Good and was manually classified as Good, a true negative if both predicted and classifed as Bad, a false positive if predicted to be Good but manually classified as Bad, and a false negative if predicted to be Bad but was in fact manually classified as Good.
-This allowed us to additionally define two useful measures of success, the traditionally-defined false discovery rate (FDR, defined as 1-precision or the number of false positives divided by the total number of predicted positives) and the percentage of good peaks found (GPF, also known as the recall or sensitivity and defined as the number of true positives divided by the total number of actual positives).
+This allowed us to additionally define two useful measures of success, the traditionally-defined false discovery rate (FDR, defined as 1-precision or the number of false positives divided by the total number of predicted positives) and the percentage of good features found (GFF, also known as the recall or sensitivity and defined as the number of true positives divided by the total number of actual positives).
 
@@
 
-GPF: Good Peaks Found
+GFF: Good Feature Found
 
@@
 
-fdr_gpf_cross_train_df_fancy <- net_cross_train_df %>%
+fdr_gff_cross_train_df_fancy <- net_cross_train_df %>%
   rbind(rf_cross_train_df) %>%
@@
   mutate(FDR=FP/(FP+TP)) %>%
-  mutate(GPF=TP/(FN+TP)) %>%
+  mutate(GFF=TP/(FN+TP)) %>%
   mutate(cross_type_num=as.numeric(cross_type))
@@
 train_test_labels <- c("MESO-MESO", "Falkor-Falkor", "MESO-Falkor", "Falkor-MESO")
-fdr_gpf_cross_train_gp <- fdr_gpf_cross_train_df_fancy %>%
+fdr_gff_cross_train_gp <- fdr_gff_cross_train_df_fancy %>%
   mutate(train_test=factor(train_test, levels=train_test_levels, labels=train_test_labels)) %>%
@@
                             labels=c("Ridge: α=0", "Elastic: α=0.5", "Lasso: α=1", "Random forest"))) %>%
-  pivot_longer(c(FDR, GPF)) %>%
+  pivot_longer(c(FDR, GFF)) %>%
   ggplot() +
@@
 
-ggsave(plot = fdr_gpf_cross_train_gp, 
-       filename = "regregrf_fdrgpf_cross_train.png", 
+ggsave(plot = fdr_gff_cross_train_gp, 
+       filename = "regregrf_fdrgff_cross_train.png", 
        device = "png", path = "manuscript/supplement",
@@
 
-![Supplemental figure 3: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.](supplement/regregrf_fdrgpf_cross_train.png){width=90%}
+![Supplemental figure 3: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.](supplement/regregrf_fdrgff_cross_train.png){width=90%}
 
@@
                   "Random forest")
-fdr_gpf_individual <- fdr_gpf_cross_train_df %>%
-  bind_rows(fdr_gpf_cross_train_df_fancy %>% dplyr::rename(which_params=which_alpha)) %>%
+fdr_gff_individual <- fdr_gff_cross_train_df %>%
+  bind_rows(fdr_gff_cross_train_df_fancy %>% dplyr::rename(which_params=which_alpha)) %>%
   separate("train_test", into = c("train", "test"), sep = "-") %>%
   select(-`NA`) %>%
-  select(model_type=which_params, train, test, TP, TN, FP, FN, FDR, GPF) %>%
+  select(model_type=which_params, train, test, TP, TN, FP, FN, FDR, GFF) %>%
   mutate(model_type=factor(model_type, levels=model_levels, labels=model_labels)) %>%
   arrange(model_type)
-fdr_gpf_full_model <- bind_rows(lst(FT2040_features, MS3000_features), .id = "dataset") %>%
+fdr_gff_full_model <- bind_rows(lst(FT2040_features, MS3000_features), .id = "dataset") %>%
   mutate(pred_prob=predict(object=both_min_model, newdata = ., type = "response")) %>%
@@
   mutate(FDR=FP/(FP+TP)) %>%
-  mutate(GPF=TP/(FN+TP)) %>%
+  mutate(GFF=TP/(FN+TP)) %>%
   mutate(model_type="Two-parameter model") %>%
@@
   )) %>%
-  select(model_type, train, test, TP, TN, FP, FN, FDR, GPF)
-
-fdr_gpf_strict_all <- bind_rows(lst(FT2040_features, MS3000_features), .id = "dataset") %>%
+  select(model_type, train, test, TP, TN, FP, FN, FDR, GFF)
+
+fdr_gff_strict_all <- bind_rows(lst(FT2040_features, MS3000_features), .id = "dataset") %>%
   mutate(pred_prob=predict(object=both_min_model, newdata = ., type = "response")) %>%
@@
   )) %>%
-  select(model_type, train, test, TP, TN, FP, FN, FDR, GPF)
-
-fdr_gpf_stricter_thresh <- error_type_all_datasets %>%
-  set_names(c("test", "FN", "FP", "TN", "TP", "FDR", "GPF")) %>%
+  select(model_type, train, test, TP, TN, FP, FN, FDR, GFF)
+
+fdr_gff_stricter_thresh <- error_type_all_datasets %>%
+  set_names(c("test", "FN", "FP", "TN", "TP", "FDR", "GFF")) %>%
   mutate(model_type="Both (threshold 0.9)") %>%
   mutate(train="Both") %>%
-  select(model_type, train, test, TP, TN, FP, FN, FDR, GPF)
-
-bind_rows(fdr_gpf_individual, fdr_gpf_full_model, fdr_gpf_stricter_thresh) %>%
+  select(model_type, train, test, TP, TN, FP, FN, FDR, GFF)
+
+bind_rows(fdr_gff_individual, fdr_gff_full_model, fdr_gff_stricter_thresh) %>%
   write_csv("manuscript/supplement/model_confusion_matrices.csv")

### 2023-06-06 manuscript/outline.Rmd
Adding outline back in for figure info

--- 2023-04-25 manuscript/outline.Rmd
@@
     - Do the FDR corrected p-value comparisons with more/less peaks by threshold
-
   - PERMANOVA table
@@
 
-  - Fundamental tradeoff between GPF and FDR should be determined by the scientific question
-  - FDR and GPF rates are very rarely reported but critically important to include
-  - Full model and XCMS model are definitely overfit
-    - Steep performance drop on new data
-    - Disagreement based on training set
-  - Only tested on HILIC data, expectations for RP columns
-  - Using medians/means as summary statistics ignores differences between treatments (i.e.
-lots of blanks blows up the median): maybe better to use max/top3?
-Maybe better to implement independently for each sample class?
-  - XCMS alone does not provide enough information to distinguish good peaks from bad with much confidence
+See the actual manuscript
 
@@
 
-In a Bayesian framework, the base rate (likelihood of being a real peak) is critically important for any downstream analysis.
-
 ## Supplement

### 2023-06-06 manuscript/manuscript.Rmd
Updated Figure 7 with lab comments

@@
   filter(feat_class=="Good") %>%
+  arrange(feature) %>%
+  mutate(feature=case_when(
+    feature=="FT0588"~"DMS-Ac 13C",
+    feature=="FT0677"~"Taurine",
+    TRUE~feature
+  )) %>%
+  mutate(feature=factor(feature, levels=unique(feature))) %>%
   ggplot() +
   geom_line(aes(x=rt, y=int, group=filename, color=depth)) +
-  facet_wrap(~feature, nrow = 1, scales = "free") +
+  ggh4x::facet_grid2(feat_class~feature, scales = "free", 
+                     independent = "y", switch = "y") +
   scale_color_manual(breaks = c("15m", "DCM", "175m"), values = c("#f4bb23", "#028e34", "#0b505c")) +
@@
         legend.position = "none", 
-        panel.grid = element_blank()) +
+        panel.grid = element_blank(),
+        panel.spacing.x = unit(0, "lines")) +
   guides(color=guide_legend(override.aes = list(linewidth=2), title = "Sample depth:  "))
@@
   geom_line(aes(x=rt, y=int, group=filename, color=depth)) +
-  facet_wrap(~feature, nrow = 1, scales = "free") +
+  ggh4x::facet_grid2(feat_class~feature, scales = "free", 
+                     independent = "y", switch = "y") +
   scale_color_manual(breaks = c("15m", "DCM", "175m"), values = c("#f4bb23", "#028e34", "#0b505c")) +
@@
         legend.position = "none",
-        panel.grid = element_blank())
+        panel.grid = element_blank(),
+        panel.spacing.x = unit(0, "lines"))
 chroms_legend <- cowplot::get_legend(good_chroms_gp + theme(legend.position = "bottom"))
@@
 
-![Figure 7: Randomly selected ion chromatograms from both "Good" (top row) and "Bad" (bottom row) manual classifications, colored by the depth from which the biological sample was taken.
-DCM = deep chlorophyll maximum, approximately 110 meters.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
+![Figure 7: Randomly selected ion chromatograms from both "Good" (top row) and "Bad" (bottom row) manual classifications.
+Plots show retention time along the x-axis in a 1 minute window around the center of the feature and show measured intensity on the y.
+Features are from the MESOSCOPE dataset and colored by the depth from which the biological sample was taken.
+DCM = deep chlorophyll maximum, approximately 110 meters.
+Molecular feature ids are provided as the title of each panel, starting with "FT" and followed by 4 digits.
+Two features were annotated using authentic standards run alongside: the 13C isotope of dimethylsulfonioacetate (DMS-Ac) and taurine.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
 

### 2023-06-06 manuscript/outline.Rmd
Updated Figure 7 with lab comments

@@
 DCM = deep chlorophyll maximum, approximately 110 meters.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
-
-Include the descriptor for what "FTXXXX" means.
-
-Include the molecular mass in the feature label? If known, include name?
-
-Put the actual good/bad label in the figure somewhere
-
-What's the importance of the different colors?
-
-Note that these were taken from the MESOSCOPE dataset.
-
-Explain the chromatograms (RT/int/ what "goodness" requires?) explain x/y axis in figure legend, maybe do in bulk across all at once?
-
-Explain that we grabbed the center of the peak +/- 1 (2?) minutes
 

### 2023-06-06 manuscript/manuscript.Rmd
Updated Figure 8 with lab thoughts

@@
   geom_text(aes(x=x_coord, y=Inf, label=single_sd, color=name),
-             data = summary_stats, vjust=3) +
-  annotate("text", x=0.5, y=Inf, vjust=1.5, label="Residual SD:") +
+             data = summary_stats, vjust=2.7) +
+  annotate("text", x=0.5, y=Inf, vjust=1.35, label="Residual SD:") +
   facet_grid(~feature) +

### 2023-06-07 manuscript/manuscript.Rmd
Switching from "Meh" to "Ambiguous"
and polishing up Figure 1

@@
   facet_wrap(~dataset, scales="free_y") +
-  scale_fill_manual(breaks = c("Good", "Bad", "Meh", "Stans only", "Unclassified"),
+  scale_fill_manual(breaks = c("Good", "Bad", "Ambiguous", "Stans only", "Unclassified"),
                     values = c("#f4bb23", "#a41118", "#0b505c", "#028e34", "grey50")) +
@@
                      labels = c(0, 50, 100, 150, axis_breaks["Falkor"])) +
-  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
   annotate("rect", xmin=0, xmax=1, ymin=180, ymax=185, fill="white", color="black") +
@@
                      labels = c(0, 50, 100, 150, 200, axis_breaks["MESOSCOPE"])) +
-  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
-  annotate("rect", xmin=0, xmax=1, ymin=220, ymax=230, fill="white", color="black") +
+  annotate("rect", xmin=0, xmax=1, ymin=226, ymax=232, fill="white", color="black") +
   annotate("rect", xmin=0, xmax=0.125, ymin=240, ymax=Inf, fill="white", color="grey90") +
@@
                      labels = c(0, 50, 100, axis_breaks["CultureData"])) +
-  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
+  geom_vline(xintercept = 0.9, color="black", linewidth=0.5, lty=2) +
   annotate("rect", xmin=0, xmax=1, ymin=110, ymax=113, fill="white", color="black") +
@@
                      labels = c(0,100,200,300,400,axis_breaks["Pttime"])) +
-  geom_vline(xintercept = 0.9, color="black", linewidth=1) +
-  annotate("rect", xmin=0, xmax=1, ymin=470, ymax=485, fill="white", color="black") +
+  geom_vline(xintercept = 0.9, color="black", linewidth=0.5, lty=2) +
+  annotate("rect", xmin=0, xmax=1, ymin=479, ymax=492, fill="white", color="black") +
   annotate("rect", xmin=0, xmax=0.125, ymin=510, ymax=Inf, fill="white", color="grey90") +
@@
 
-![Figure 1: Histograms showing the estimated likelihood of a given mass feature being categorized as "Good" according to the two-parameter logistic model trained on the combined fully-labeled Falkor and MESOSCOPE datasets.
+![Figure 1: Histograms showing the estimated likelihood of a given mass feature being categorized as "Good" according to the two-parameter logistic model trained on the combined fully-labeled Falkor and MESOSCOPE environmental datasets.
 Colors indicate the category in which each feature was manually assigned by an expert, with "Stans only" referring to a good molecular feature that was only visible in the standards run alongside the samples.
-Datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (black vertical line) and were otherwise unclassified.](figures/pred_prob_class_color_hists.png){width=90%}
+Culture datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (black vertical line) and were otherwise unclassified.](figures/pred_prob_class_color_hists.png){width=90%}
 

### 2023-06-07 training_tool.R

@@ -132,7 +132,7 @@ while(TRUE){
     keyinput,
     "Right" = "Bad",
     "Left" = "Good",
-    "Up" = "Meh",
+    "Up" = "Ambiguous",
     "Down" = "Stans only"
   )
   cbind(row_data, feat_class) %>%

### 2023-06-07 feature_extraction.R
Removing depth_diffs calculations

@@ -256,29 +256,29 @@ write.csv(feat_isodata, paste0(output_folder, "feat_isodata.csv"), row.names = F
 
 # Calculate DOE metrics ----
 # library(lmPerm)
-depth_diffs <- peak_data %>%
-  select(feature, filename, into) %>%
-  left_join(file_data) %>%
-  filter(samp_type=="Smp") %>%
-  nest(data=-feature) %>%
-  mutate(t_pval=map_dbl(data, function(x){
-    depthtab <- table(x$depth)
-    if(any(depthtab<2))return(99)
-    if(length(depthtab)<2)return(999)
-    broom::tidy(aov(x$into~x$depth))$p.value[1]
-    # broom::tidy(aovp(x$into~x$depth, settings=FALSE, perm = "Prob",
-    #                  maxIter=1000))$`Pr(Prob)`[1]
-  })) %>%
-  mutate(t_pval=ifelse(t_pval==99, (max(t_pval[t_pval<1])+9)/10, t_pval)) %>%
-  mutate(t_pval=ifelse(t_pval==999, min(t_pval[t_pval<1])/10, t_pval)) %>%
-  mutate(t_pval=log10(t_pval)) %>%
-  mutate(t_pval=log10(t_pval*-1)*-1) %>%
-  select(feature, t_pval) %>%
-  # Cover cases where there's zero data in samples OR blank
-  # 0.9 p-value is high enough to be ignored but not 1 to cause problems
-  right_join(distinct(peak_data, feature)) %>%
-  mutate(smp_to_blk=ifelse(is.na(t_pval), 0.9, t_pval))
-write.csv(depth_diffs, paste0(output_folder, "depth_diffs.csv"), row.names = FALSE)
+# depth_diffs <- peak_data %>%
+#   select(feature, filename, into) %>%
+#   left_join(file_data) %>%
+#   filter(samp_type=="Smp") %>%
+#   nest(data=-feature) %>%
+#   mutate(t_pval=map_dbl(data, function(x){
+#     depthtab <- table(x$depth)
+#     if(any(depthtab<2))return(99)
+#     if(length(depthtab)<2)return(999)
+#     broom::tidy(aov(x$into~x$depth))$p.value[1]
+#     # broom::tidy(aovp(x$into~x$depth, settings=FALSE, perm = "Prob",
+#     #                  maxIter=1000))$`Pr(Prob)`[1]
+#   })) %>%
[…]

### 2023-06-07 manuscript/manuscript.Rmd
Switching from Meh to Ambiguous in figures

@@
   # source("peakpicking_and_prep.R")
-  # source("feature_extraction.R")
+  source("feature_extraction.R")
 }

### 2023-06-07 manuscript/manuscript.Rmd
Switching to lighter/yellower green

@@
                    breaks = c("Same", "Different")) +
-  scale_color_manual(values = c("#0b505c", "#184d1f", "#0d8299", "#028e34"), 
+  scale_color_manual(values = c("#0b505c", "#3A601E", "#0d8299", "#028e34"), 
                      breaks = train_test_labels,

### 2023-06-07 manuscript/manuscript.Rmd
Making edits to Fig 4 (and supp fig)
Thinner line, behind data, added results to caption. Chose to stick with GFF metric.

@@
   mutate(which_params=factor(which_params, 
-                             labels=c("All metrics", "XCMS metrics", "Raw data metrics"),
+                             labels=c("All metrics", "XCMS-only metrics", "Raw data metrics"),
                              levels=c("all_params", "xcms_params", "raw_data_params"))) %>%
@@
   ggplot() +
+  geom_smooth(aes(x=cross_type_num, y=value, group=name), 
+              method = "lm", formula="y~x", color="black", 
+              linewidth=0.5, se=FALSE) +
   geom_point(aes(x=cross_type, y=value, color=train_test, 
                  fill=train_test, shape=name), size=5) +
-  geom_smooth(aes(x=cross_type_num, y=value, group=name), 
-              method = "lm", formula="y~x", color="black", se=FALSE) +
   facet_grid(~which_params) +
@@
 Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel.
-Lines of best fit have been estimated and plotted in black on top of the data points.](figures/fdr_gff_cross_train.png){width=90%}
+Lines of best fit have been estimated and plotted in black behind the data points, with the steepest slopes found in the full and XCMS-only models indicating excessive overfitting on the training data.](figures/fdr_gff_cross_train.png){width=90%}
 
@@
   geom_smooth(aes(x=cross_type_num, y=value, group=name), 
-              method = "lm", formula="y~x", color="black", se=FALSE) +
+              method = "lm", formula="y~x", color="black", 
+              linewidth=0.5, se=FALSE) +
   geom_point(aes(x=cross_type, y=value, color=train_test, 

### 2023-06-07 manuscript/outline.Rmd
Updated Figure 8 with lab thoughts
Switching from "Meh" to "Ambiguous"
and polishing up Figure 1
Making edits to Fig 4 (and supp fig)
Thinner line, behind data, added results to caption. Chose to stick with GFF metric.

@@
 Peak shape similarity was measured with Pearson's correlation coefficient and the noise level is estimated as the standard deviation of the residuals after the raw data is subtracted from the idealized peak.](figures/peakmetrics_singlechrom.png){width=70%}
-
-Good figure, spacing between text and labels and data
 
@@
 
-Explain what the different datasets are?
-
-Make sure the break gap is consistent
-
-Maybe turn into two rows, one zoomed and one unzoomed?
-
-Differentiate the 0.9 line from the axis break line - don't like the vertical black line, looks like data
-
-Convert "meh" to "ambiguous"
-
 ### Model cross-validation with novel datasets
@@
 Lines of best fit have been estimated and plotted in black on top of the data points.](figures/fdr_gpf_cross_train.png){width=70%}
-
-Dark blue and dark green are too similar, need to separate these better somehow.
-
-Include conclusions in figure legend? Something like "slope of the line is much greater in All metrics and XCMS metrics, indicating overfitting"?
-
-Is averaging the two points for the best fit line best? Lines sometimes have different slopes
-
-Is GPF worth using instead of more common jargon like 1-precision
-
-Maybe simpler metric
-
-Line feels a little heavy and too dark, put behind data, maybe make line grey instead of dark black
 

### 2023-06-07 manuscript/manuscript.Rmd
Polishing up Fig 3 and surrounding text

@@
 We found that the predictions made from a Falkor-trained dataset consistently differed from a MESOSCOPE-trained dataset for the full and XCMS-only models.
-In the raw probability space, the two-parameter models had the highest $r$ value of 0.996, while the full models and the XCMS-trained models had $r$ values of 0.856 and 0.896, respectively.
-When compared in ranked space, we found an intensification of the effect described previously, with an even higher $r$ for the two-parameter model of 0.998 but lower $r$ values for the full and XCMS-trained model of 0.818 and 0.849, respectively (Figure 3).
+In the raw probability space, the two-parameter models had the highest $r$ value of 0.996, while the full models and the XCMS-trained models had $r$ values of 0.799 and 0.863, respectively.
+When compared in ranked space, we found an intensification of the effect described previously, with a higher $ρ$ for the two-parameter model of 0.998 but lower $ρ$ values for the full and XCMS-trained model of 0.725 and 0.804, respectively (Figure 3).
 
@@
   summarize(pcor=cor(FT2040_features, MS3000_features, method = "pearson"),
-            scor=cor(FT2040_features, MS3000_features, method = "spearman"))
+            scor=cor(FT2040_features, MS3000_features, method = "spearman")) %>% 
+  mutate(pcor=paste0("r=", signif(pcor, digits = 3))) %>%
+  mutate(scor=paste0("ρ=", signif(scor, digits = 3))) %>%
+  mutate(which_params=factor(
+    which_params, levels=c("all_params", "xcms_params", "raw_data_params"),
+    labels=c("All metrics", "XCMS-only metrics", "Raw data metrics")))
 
@@
     which_params, levels=c("all_params", "xcms_params", "raw_data_params"),
-    labels=c("All metrics", "XCMS metrics", "Raw data metrics"))) %>%
+    labels=c("All metrics", "XCMS-only metrics", "Raw data metrics"))) %>%
   ggplot() +
@@
   geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class), alpha=0.2) +
+  geom_text(aes(label=pcor), x=-Inf, y=Inf, hjust=-0.1, vjust=1.5, data=stepdown_cors) +
   facet_wrap(~which_params, nrow = 1) +
@@
     levels=c("all_params", "xcms_params", "raw_data_params"),
-    labels=c("All metrics", "XCMS metrics", "Raw data metrics"))) %>%
+    labels=c("All metrics", "XCMS-only metrics", "Raw data metrics"))) %>%
   filter(feat_class%in%c("Good", "Bad")) %>%
@@
   geom_point(aes(x=FT2040_features, y=MS3000_features, color=feat_class), alpha=0.2) +
+  geom_text(aes(label=scor), x=-Inf, y=Inf, hjust=-0.1, vjust=1.5, data=stepdown_cors) +
   facet_wrap(~which_params, nrow = 1) +
@@
         strip.background = element_rect(fill="white"),
-        legend.position=c(0.875, 0.25),
+        legend.position=c(0.925, 0.25),
         legend.background = element_blank(),

### 2023-06-07 manuscript/manuscript.Rmd
Updating Figure 4 with lab comments

@@
 
-fdr_gff_all_datasets_gp <- error_type_all_datasets %>%
-  pivot_longer(c(`FDR`, `GFF`)) %>%
-  drop_na() %>%
-  mutate(desc=case_when(
-    is.na(value) ~ NA,
-    name=="FDR"~ paste0("frac(",`%FP`,",",`%FP`+`%TP`,")"),
-    name=="GFF"~ paste0("frac(",`%TP`,",",`%FN`+`%TP`,")")
-  )) %>%
-  mutate(vjust_custom=case_when(
-    is.na(value) ~ NA,
-    name=="FDR" & dataset=="Pttime" ~ 1.3,
-    name=="FDR" ~ -0.3,
-    name=="GFF" ~ 1.3,
-  )) %>%
-  mutate(name=factor(name, levels=c("GFF", "FDR"),
-                     labels=c("Fraction of good features found (GFF)", 
-                              "False discovery rate (FDR)"))) %>%
+fdr_alldatasets_gp <- error_type_all_datasets %>%
+  mutate(FDR_label=paste0("frac(",`%FP`,",",`%FP`+`%TP`,")")) %>%
+  mutate(vjust_custom=ifelse(dataset=="Pttime", 1.3, -0.3)) %>%
   mutate(dataset=factor(dataset, 
@@
                         labels= c("Falkor", "MESO-\nSCOPE", "Culture\nData", "Pttime"))) %>%
-  ggplot(aes(x=dataset, y=value)) +
+  ggplot(aes(x=dataset, y=FDR)) +
   geom_point(size=3) +
-  geom_text(aes(label=desc, vjust=vjust_custom), parse=TRUE) +
-  facet_wrap(~name, scales = "free_y", ncol = 1) +
-  scale_y_continuous(limits = c(0, NA), labels = scales::label_percent()) +
+  geom_text(aes(label=FDR_label, vjust=vjust_custom), parse=TRUE) +
+  scale_y_continuous(limits = c(0, NA), labels = scales::label_percent(),
+                     name = "False discovery rate (FDR)\n") +
   theme_bw() +
-  theme(axis.title = element_blank(),
+  theme(axis.title.x = element_blank(),
         strip.background = element_rect(fill="white"))
+gff_alldatasets_gp <- error_type_all_datasets %>%
+  mutate(GFF_label=paste0("frac(",`%TP`,",",`%FN`+`%TP`,")")) %>%
+  mutate(vjust_custom=1.3) %>%
+  mutate(dataset=factor(dataset, 
+                        levels = c("Falkor", "MESOSCOPE", "CultureData", "Pttime"),
+                        labels= c("Falkor", "MESO-\nSCOPE", "Culture\nData", "Pttime"))) %>%
+  ggplot(aes(x=dataset, y=GFF)) +
+  geom_point(size=3) +
+  geom_text(aes(label=GFF_label, vjust=vjust_custom), parse=TRUE) +
+  scale_y_continuous(limits = c(0, NA), labels = scales::label_percent(),
+                     name = "Fraction of good\nfeatures found (GFF)") +
+  theme_bw() +
+  theme(axis.title.x = element_blank(),
+        strip.background = element_rect(fill="white"),
+        axis.text.x = element_blank())
+fdr_gff_all_datasets_gp <- egg::ggarrange(gff_alldatasets_gp, fdr_alldatasets_gp, ncol = 1)
 
@@
        width = 3, height = 4, units = "in", dpi = 300)
-
-# Confusion matrices for confirmation
-# pred_probs_all
-# pred_probs_all %>%
-#   ggplot(aes(x=error_type, y=n)) +
-#   geom_col(color="#0b505c", fill="#0b505c") +
-#   geom_text(aes(label=n), vjust=-0.2, color="#0b505c") +
-#   facet_wrap(~dataset, nrow=1) +
-#   theme_bw() +
-#   theme(axis.title = element_blank())
-```
-
-![Figure 4: False discovery rate and proportion of total good features identified as good by the two-parameter model trained on the combined MESOSCOPE/Falkor dataset and applied to each dataset individually with a threshold of 0.9.
-FDR is calculated by dividing the number of false positives by the total positives produced by the model and GFF is calculated by dividing the number of true positives by the total number of good features as identified manually.
+```
+
+![Figure 4: False discovery rate and proportion of total good features identified as good by the two-parameter model trained on the combined MESOSCOPE/Falkor dataset and applied to each dataset.
+A stricter likelihood threshold is used here (0.9) than in Figure 2.
+FDR is calculated by dividing the number of false positives by the total positives produced by the model and GFF is calculated by dividing the number of true positives by the total number of good features as identified manually (only possible for fully-labeled datasets).
 Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fdr_gff_all_datasets.png){width=30%}

### 2023-06-07 manuscript/outline.Rmd
Polishing up Fig 3 and surrounding text
Updating Figure 4 with lab comments

@@
 
-Make sure labels are the same as above - probably stick with XCMS for specificity. (Could just go in tables?)
-
-Remember to add Pearson and Spearman correlation coefficients in the corners!
-
-Maybe avoid "rank" phrasing? Instead of best/worst?
-
 ### Error rates on novel data with stricter likelihood threshold
@@
 Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fdr_gpf_all_datasets.png){width=30%}
-
-Emphasize the use of a different threshold here!
-
-Explain why there are no values for GPF in Pttime/CultureData?
-
-Render as a table and see how that looks instead?
-
-Include Pttime definition in the figure caption?
-
-Maybe include the facet captions as the y-axis label instead? Anitra doesn't love how it looks as-is
 

### 2023-06-07 manuscript/manuscript.Rmd
Updated version of Figure 5 with histograms and boxplots

@@
 Of the 182 good MFs, less than a fifth had no trend with depth (44/249) and a majority had unequivocally lowest values in the 175 meter samples (those with 15m/DCM > 175m, 145 features).
-The two-parameter model, when applied with a 50% likelihood threshold, also recovered this general feature distribution and classified many of the features with no significant depth signal as likely to be bad [Figure 7B, currently in Slack?].
+The two-parameter model, when applied with a 50% likelihood threshold, also recovered this general feature distribution and classified many of the features with no significant depth signal as likely to be bad.
 
 Additionally, a large number of features manually identified as bad nonetheless had significant differences with depth.
-This was surprising because we had assumed that bad MFs corresponded to instrument noise, which we did not expect to have any biological trend.
+This was surprising because we had assumed that bad MFs corresponded to instrument or chemical noise, which we did not expect to have any biological trend.
 Further investigation of a few randomly selected bad features with a biological difference revealed the reason behind this: most of those investigated were actually tails of other MFs.
@@
     "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
-    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
-    "Hard to tell")
+    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m")
 clockplot_complete_df <- read_csv(file = "manuscript/MS3000_clockplot_coords.csv") %>%
+  mutate(resp_class=ifelse(resp_class=="Hard to tell", "All equal", resp_class)) %>%
   mutate(resp_class=factor(resp_class, levels=resp_class_levels))
@@
 axis_y_dodge <- c(2, 2, -2)*clock_scale/25
+
+error_order <- c("Good MF found", "Good MF missed", "Bad MF included", 
+                 "Bad MF avoided")
+error_pchs <- c(19, 4, 3, 1)
+error_unicodes <- c("\u25cf", "\u00d7", "\u002B", "\u25e6")
+error_type_labels <- error_order %>%
+  str_replace(" (?=[^ ]*$)", "\n") %>%
+  paste0(" (", error_unicodes, ")")
+# plot(1:4, pch=error_pchs, ylim=c(1, 6), xlim=c(-1.5, 4))
+# points(1:4, 2:5, pch=error_unicodes)
+# text(1:4, 3:6, label=error_order, adj = 1)
+
+clockplot_hist_df <- clockplot_complete_df %>% 
+  with(table(resp_class, error_type)) %>% 
+  as.data.frame() %>%
+  mutate(Freq=ifelse(Freq==1365, 60, Freq)) %>%
+  group_by(error_type) %>%
+  mutate(hjust_custom=ifelse(Freq>max(Freq)*2/3, 1.2, -0.2)) %>%
+  mutate(text_color=ifelse(Freq>max(Freq)*2/3, "white", "black")) %>%
+  ungroup() %>%
+  mutate(freq_label=ifelse(Freq==0, "", Freq)) %>%
+  mutate(freq_label=ifelse(
+    resp_class=="All equal" & error_type=="Bad MF avoided", "1365 \u25ba", freq_label
+  )) %>%
+  mutate(hjust_custom=ifelse(
+    resp_class=="All equal" & error_type=="Bad MF avoided", 1, hjust_custom
+  )) %>%
+  mutate(resp_class=factor(resp_class, levels=rev(levels(resp_class)))) %>%
+  mutate(error_type=factor(error_type, levels=error_order,
+                           labels=error_type_labels))
+clockplot_hist_gp <- clockplot_hist_df %>% 
+  ggplot() +
+  geom_col(aes(y=resp_class, x=Freq, fill=resp_class)) +
+  geom_text(aes(y=resp_class, x=Freq, hjust=hjust_custom, label=freq_label, color=text_color)) +
+  scale_fill_manual(values = color_scale, breaks = names(color_scale), drop=FALSE) +
+  scale_color_identity() +
+  scale_y_discrete(breaks=resp_class_levels) +
+  facet_wrap(~error_type, nrow=1, scales="free_x") +
+  theme_bw() +
+  labs(x=NULL, y=NULL) +
+  theme(legend.position = "none",
+        axis.text.y = ggtext::element_markdown(color = c(
+          "black", color_scale[2:13]
+          ), face = "bold"))
 
@@
            label=c("15m axis", "DCM axis", "175m axis"), 
-           fill="#FFFFFFAA", label.size=0) +
-  coord_fixed(xlim = c(-0.65, 1.3)*clock_scale, 
+           fill="#FFFFFFAA", label.size=0, hjust=c(0.5, 0.9, 0.5)) +
+  coord_fixed(xlim = c(-0.65, 1.05)*clock_scale, 
               ylim = c(-1, 1)*clock_scale, 
@@
   theme_void() +
-  theme(legend.position = "right", 
-        legend.title = element_blank(),
-        plot.background = element_rect(fill="white", color="white")) +
-  guides(colour = guide_legend(override.aes = list(size=4, alpha=1), drop=FALSE)) +
-  aes(x=x, y=y, color=resp_class, label=feature)
-
-clockplot_complete_gp <- clockplot_base_gp +
-  geom_point(aes(shape=error_type), alpha=0.5, size=2,
-             data = clockplot_complete_df) +
-  scale_shape_manual(breaks = c("Good MF found", "Bad MF included",
-                                "Good MF missed", "Bad MF avoided"), 
-                     values = c(19, 3, 4, 1))
-
+  theme(legend.title = element_blank(),
+        legend.position = c(0.75, 0.12),
+        # legend.position="left",
+        # legend.background = element_rect(fill="transparent", color="grey40"),
+        plot.background = element_rect(fill="white", color="transparent"),
+        plot.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "in")) +
+  guides(colour = "none") +
+  aes(x=x, y=y, color=resp_class, label=feature) +
+  geom_point(aes(shape=error_type), alpha=0.5, size=2, data = clockplot_complete_df) +
+  scale_shape_manual(breaks = error_order, values = error_pchs)
+
+norm_area_data <- read_csv("manuscript/MS3000_norm_areas.csv")
+#Palythine = FT1946
+#Missed good MF w DCM minimum = FT1564 or FT1584
+#DMS-Ac = FT0567
+#Arsenobetaine = FT1432
+#Interesting 146.1175 mass = FT1051
+
+box_list <- list(c("FT1946", "FT1584"), c("FT0567", "FT1051")) %>%
+  lapply(function(feat_ids){
+    norm_area_data %>%
+      filter(feature%in%feat_ids) %>%
+      left_join(clockplot_complete_df %>% distinct(feature, resp_class),
+                by="feature") %>%
+      mutate(resp_class=factor(resp_class, levels=resp_class_levels)) %>%
+      arrange(resp_class) %>%
+      left_join(file_data %>% select(filename, depth), by="filename") %>%
+      mutate(depth=factor(depth, levels=rev(levels(depth)))) %>%
+      mutate(feature=case_when(
+        feature=="FT1946"~"Palythine",
+        feature=="FT1432"~"Arsenobetaine",
+        feature=="FT0567"~"DMS-Ac",
+        TRUE~feature
+      )) %>%
+      mutate(feature=factor(feature, levels=unique(feature))) %>%
+      ggplot() +
+      geom_boxplot(aes(x=norm_area, y=depth, color=resp_class)) +
+      facet_wrap(~feature, ncol=1, scales = "free_x") +
+      scale_x_log10(labels=scales::label_log(), breaks=10^(4:10)) +
+      scale_color_manual(values = color_scale, breaks = names(color_scale)) +
+      theme_bw() +
+      theme(legend.position = "none",
+            plot.margin=margin(l = 0, unit = "in")) +
+      labs(x="Peak area", y=NULL)
+})
+
+clockplot_withboxplots <- gridExtra::arrangeGrob(
+  box_list[[1]], clockplot_base_gp, box_list[[2]], 
+  layout_matrix = matrix(c(1,2,2,2,3),nrow = 1)
+)
+clockplot_complete_gp <- gridExtra::arrangeGrob(
+  clockplot_hist_gp, clockplot_withboxplots, 
+  layout_matrix = matrix(c(1,2,2), ncol=1)
+)
 ggsave(plot = clockplot_complete_gp, 
@@
        device = "png", path = "manuscript/figures",
-       width = 6.5, height = 5, units = "in", dpi = 300)
+       width = 6.5, height = 7, units = "in", dpi = 300)
 ```

### 2023-06-07 manuscript/manuscript.Rmd
Removing MAA paragraph

@@
 
-[MAA paragraph?] Several mass features that were found dominantly at the surface were putatively identified as mycosporine-like amino acids (MAAs), which have been shown to have UV photoprotective properties.[@ref] Although we were unable to calculate their exact concentrations due to a lack of authentic standards, peak areas were approximately 100 to 1000 times higher in the 15 meter samples than the values averaged between the DCM and 175 meter samples.
-The abundance of MAAs in this particular category indicates that there may be other MAA molecules with this depth pattern that have not yet been characterized also among these unknowns.
-
 ```{r Fig5 clockplot}

### 2023-06-08 manuscript/manuscript.Rmd
Updated figure legend for Figure 5

@@
 
-![Figure 5: Plot of metabolite response to depth shown across three axes using the rank-normalized median value at each depth as the coordinate for that axis.
+![Figure 5: Plot of metabolite response to depth and the model classification error distribution.
+Barplots at the top show the number of MFs (mass features) in each depth response category and are broken down by the classification error types.
+Compounds were assigned a depth category via Dunn's post-hoc test for significant differences between the sample depths.
+Good MF found = true positive, good MF missed = false negative, bad MF found = false positive, and bad MF avoided = true negative according to a 0.5 likelihood threshold.
+Note that the majority of the features in the bad MF avoided category fell into the "All equal" depth class (1365 MFs) but the x-axis has been truncated at 60 for ease of visualization.
+The boxplots in the bottom illustrate the depth response type for 4 specific categories, with raw peak area plotted on a log scale against the sample depth (DCM = deep chlorophyll maximum, ~110 meters).
+All MFs are shown in the central bottom plot across three axes using the rank-normalized median value at each depth as the coordinate for that axis.
 Each molecular feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile.
-Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the deep chlorophyll maximum; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}
+Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the DCM; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}
+
+[Switch the shapes to make empty circles "good peaks missed"]
 

### 2023-06-08 manuscript/outline.Rmd
Updated version of Figure 5 with histograms and boxplots
Updated figure legend for Figure 5

@@
 
-Make sure I include the threshold used for the above plot (in the legend?)
-
-Make sure I explain the statistical significance and the "Hard to tell" category
-
-See aobut including ONLY the good peaks in a separate plot (inset figure?)
-
-Switch the shapes to make empty circles "good peaks missed"
-
-Add some internal circle to help render information about the overplotting in the center? Give a bar graph breakdown for the  things within the circle?
-
 ```{r manual check on clockplot, eval=FALSE}

### 2023-06-08 manuscript/manuscript.Rmd
Finished polishing figures!

@@
 ```{r PERMANOVA results}
+subset_levels <- c("mat_good", "mat_90", "mat_50", "mat_10", "mat_01", "mat_all")
+subset_labels <- c("Only good MFs", paste("Threshold", c(0.9, 0.5, 0.1, 0.01)), "All MFs")
 thresh_tests <- imap(mat_list, function(mat_given, mat_name){
@@
   `rownames<-`(NULL) %>%
-  column_to_rownames("thresh_name") %>%
-  knitr::kable()
-```
-
-In each test, the permutational p-value obtained was less than 0.001, indicating that the differences between samples due to depth were never likely to be due entirely to chance.
-However, the pseudo-F was much larger with higher thresholds, scaling from around 8.5 when thresholding at a 1% likelihood to 42 when thresholded at a 90% likelihood.
-
-|         |    n|   R2|    F| Pr(>F)|
-|:--------|----:|----:|----:|------:|
-|mat_good |  249| 0.44| 38.0|  0.001|
-|mat_90   |   75| 0.46| 42.0|  0.001|
-|mat_50   |  287| 0.34| 25.3|  0.001|
-|mat_10   |  516| 0.25| 16.0|  0.001|
-|mat_01   | 1129| 0.15|  8.5|  0.001|
-|mat_all  | 2086| 0.10|  5.5|  0.001|
-
-*Table 1: Number of mass features (n), variance explained (R2), pseudo-F statistic (F), and permutational p-value (Pr(>F)) estimates obtained from performing a permutational MANOVA on subsets of the full mass feature selection according to variable likelihood thresholds.
-[Needs cleaning, will fix]*
-
-We also tested the inclusion of all the features identified with XCMS (corresponding to a 0% threshold) and the results when only the manually-identified "Good" features were included.
-The default XCMS output continued the trend observed above, as expected, with the least variance explained and the lowest R$^2$ value.
-Subsetting for the "Good" MFs only, however, did not actually return the highest F-ratio or R$^2$, instead falling between the 50% and 90% thresholds for these two metrics.
-In large part this is due to the much smaller number of features: 249 features were manually labeled as Good, while only 75 exceeded the 90% likelihood threshold.
-
-The relative power of identifying only the very best MFs was also illustrated visually with non-metric multidimensional scaling (NMDS) plots (Figure XX).
-In these common exploratory plots, the MFs with likelihoods above 50% strongly separated by depth while lower thresholds disguised the true signal and had higher stress values.
-Performing an NMDS on the manually-identified Good MFs resulted in output nearly indistinguishable from those of the 90% and 50% thresholds, while the 0.1 and 0.01 thresholds looked almost identical to the same analysis performed on all the XCMS features.
-
-```{r Fig6 NMDS results}
+  mutate(mat_name=factor(thresh_name, levels=subset_levels, labels=subset_labels))
 thresh_NMDS_df <- imap(mat_list, function(mat_given, mat_name){
@@
   mutate(basename=str_extract(filename, "MS\\d+C\\d+.*(?=_[A-C])")) %>%
-  mutate(mat_name=factor(
-    mat_name, levels=c("mat_good", "mat_90", "mat_50", "mat_10", "mat_01", "mat_all"),
-    labels=c("Only good", paste("Threshold", c(0.9, 0.5, 0.1, 0.01)), "All features")
-  )) %>%
-  mutate(stress=format(stress, digits=2))
+  mutate(mat_name=factor(mat_name, levels=subset_levels, labels=subset_labels)) %>%
+  mutate(stress=format(stress, digits=3))
+
+thresh_statistic_df <- thresh_NMDS_df %>%
+  distinct(mat_name, stress) %>%
+  left_join(thresh_tests, by="mat_name") %>%
+  mutate(R2=round(R2*100, digits = 1)) %>%
+  mutate(`F`=round(`F`, digits = 1)) %>%
+  mutate(stress=as.numeric(stress)) %>%
+  select(`Data subset`=mat_name, `# of MFs`=n, `% Var\ Expl.`=R2, `Pseudo-F`=`F`,
+         `2D NMDS\ stress`=stress) %>%
+  arrange(desc(row_number()))
+
+knitr::kable(thresh_statistic_df)
+```
+
+In each test, the permutational p-value obtained was less than 0.001, indicating that the differences between samples due to depth were never likely to be due entirely to chance.
+However, the pseudo-F and was much larger with higher thresholds, scaling from around 8.5 when thresholding at a 1% likelihood to 42 when thresholded at a 90% likelihood (Table 1).
+
+|Data subset    | # of MFs| % Var Expl.| Pseudo-F| 2D NMDS stress|
+|:--------------|--------:|-----------:|--------:|--------------:|
+|All MFs        |     2086|        10.4|      5.5|          0.169|
+|Threshold 0.01 |     1129|        15.2|      8.6|          0.191|
+|Threshold 0.1  |      516|        25.0|     16.0|          0.200|
+|Threshold 0.5  |      287|        34.6|     25.3|          0.180|
+|Threshold 0.9  |       75|        46.7|     42.0|          0.097|
+|Only good MFs  |      249|        44.2|     38.1|          0.142|
+
+*Table 1: Number of mass features, percent variance explained, pseudo-F statistic, and stress values from performing a permutational MANOVA and 2D non-metric multidimensional scaling (NMDS) on subsets of the full mass feature selection according to variable likelihood thresholds.*
+
+We also tested the inclusion of all the features identified with XCMS (corresponding to a 0% threshold) and the results when only the manually-identified "Good" features were included.
+The default XCMS output continued the trend observed above, as expected, with the least variance explained and the lowest R$^2$ value.
+Subsetting for the "Good" MFs only, however, did not actually return the highest F-ratio or R$^2$, instead falling between the 50% and 90% thresholds for these two metrics.
+In large part this is due to the much smaller number of features: 249 features were manually labeled as Good, while only 75 exceeded the 90% likelihood threshold (Table 1).
+
+The relative power of identifying only the very best MFs was also illustrated visually with non-metric multidimensional scaling (NMDS) plots (Figure 6).
+In these common exploratory plots, the MFs with likelihoods above 50% strongly separated by depth while lower thresholds disguised the true signal and had higher stress values.
+Performing an NMDS on the manually-identified Good MFs resulted in output nearly indistinguishable from those of the 90% and 50% thresholds, while the 0.1 and 0.01 thresholds looked almost identical to the same analysis performed on all the XCMS features.
+
+```{r Fig6 NMDS results}
 thresh_NMDS_gp <- thresh_NMDS_df %>%
@@
   geom_polygon(aes(x=MDS1, y=MDS2, fill=depth, color=depth, group=basename), alpha=0.5) +
-  geom_label(aes(label=stress), x=Inf, y=Inf, 
-             label.r = unit(0, "in"), hjust=1, vjust=1,
-             data=thresh_NMDS_df %>% distinct(mat_name, stress)) +
+  geom_label(aes(label=stress), x=Inf, y=Inf,  label.r = unit(0, "in"), 
+             hjust=1, vjust=1, data=distinct(thresh_NMDS_df, mat_name, stress)) +
   facet_wrap(~mat_name, scales = "free", ncol=3) +
@@
                      breaks = c("15m", "DCM", "175m"),
-                     aesthetics = c("color", "fill")) +
+                     aesthetics = c("color", "fill"),
+                     name="Sample\ndepth") +
   theme_bw() +
@@
 
-![Figure 6: NMDS plots of metabolite similarity according to sample depth across multiple likelihood thresholds.
+![Figure 6: 2D non-metric multidimensional scaling (NMDS) plots of metabolite similarity according to sample depth across multiple likelihood thresholds.
 Triplicate samples are represented by the vertices of the triangles and colored by the depth from which they were sampled (DCM = deep chlorophyll maximum, ~110m).
-"Only good" refers to those features manually labeled as "Good".
-NMDS stress values are reported to 3 significant figures in the upper right of the plot.](figures/thresh_NMDS_plots.png){width=90%}
+"Only good MFs" refers to those features manually labeled as "Good".
+NMDS stress values are reported in the upper right corner of each plot.](figures/thresh_NMDS_plots.png){width=90%}
 

### 2023-06-09 manuscript/manuscript.Rmd
Adding discussion about metrics and their relative strengths

@@
   print(paste("Processing", dataset_version))
-  # source("peakpicking_and_prep.R")
+  source("peakpicking_and_prep.R")
   source("feature_extraction.R")
@@
                 "med_cor", "med_SNR", "med_missed_scans", "smp_to_blk", 
-                "smp_to_std", "pooled_cv", "pooled_cv_rob", "shape_cor", 
-                "area_cor")
+                "smp_to_std", "pooled_cv", "pooled_cv_rob", "iso_shape_cor", 
+                "iso_area_cor")
 ```
@@
 In both, 5% of the MFs were unable to be assigned confidently to either "Good" or "Bad" classes and 10% were identified as appearing only in the standards, leaving only ~15% of the features classified as "Good" (16% and 12%, respectively).
+
+Most metrics had reasonably normal distributions after the scaling and normalization described in Methods.
+Visually, the most compelling separations between good and bad MFs were observed in the novel peak shape and SNR metrics, with higher peak shape correlations and larger SNR values corresponding to good MFs and almost complete separation for the peak shape (`med_cor`) metric.
+Peak width and its standard deviation also showed reasonable separation between good and bad MFs (good MFs tended to have low SDs and larger peak widths).
+The isotope shape and area correlations also showed good separation (Supp.
+figure XX).
 
@@
 
+```{r full model estimate csv write}
+set.seed(123)
+
+glm_estimates <- both_datasets %>%
+  select(feat_bin, all_of(all_params)) %>%
+  # mutate(across(-feat_bin, \(x)scale(x)[,1])) %>%
+  glm(formula = feat_bin~., family=binomial) %>%
+  broom::tidy() %>%
+  rename_with(~paste0("GLM.", .x), -term)
+
+glmnet_estimates <- both_datasets %>% 
+  select(all_of(all_params)) %>% 
+  data.matrix() %>%
+  glmnet::cv.glmnet(y=both_datasets$feat_bin, family="binomial", alpha=0.5) %>%
+  coef() %>%
+  as.matrix() %>%
+  as.data.frame() %>%
+  rownames_to_column("term") %>%
+  dplyr::rename(`Elastic.net.estimate`=s1)
+
+rf_estimates <- both_datasets %>% 
+  select(all_of(all_params)) %>%
+  randomForest::randomForest(factor(both_datasets$feat_bin), importance=TRUE) %>%
+  pluck("importance") %>%
+  as.data.frame() %>%
+  rownames_to_column("term") %>%
+  select(term, RF.accuracy.decrease=MeanDecreaseAccuracy,
+         RF.Gini.Decrease=MeanDecreaseGini)
+
+glm_estimates %>%
+  left_join(glmnet_estimates, by="term") %>%
+  left_join(rf_estimates, by="term") %>%
+  write_csv("manuscript/supplement/full_model_estimates.csv")
+```
+
+We explored the relative predictive power of the individual parameters using the full model and found that the predictors least likely to be different from zero due to chance were the mean *m/z* ratio, the novel peak shape correlation metric `med_cor`, and the novel SNR estimate `med_SNR`, all with reported p-values < 10$^{-10}$.
+The novel parameters were then validated using a random forest model that also found them to have the highest importance (Supp table XX).
+
 The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR values in the 5-10% range and 80-90% GFF values implying that a large majority of the good MFs passed the threshold with very little noise included.
@@
 A majority of the time, the estimates from the two models disagreed by more than 2x the standard error of the estimate.
-Some parameters disagreed not only in magnitude but also in sign, with the Falkor-trained full model increasing MF goodness likelihood with larger ppm variation and a wider peak width, while the MESOSCOPE-trained full model had negative estimates for each of these parameters.
+Some parameters disagreed not only in magnitude but also in sign, with the Falkor-trained full model increasing MF goodness likelihood with larger PPM variation and a wider peak width, while the MESOSCOPE-trained full model had negative estimates for each of these parameters.
 Notably, the peak shape and novel SNR parameters used in the two-parameter model were among the most robust to training model variation, potentially explaining the consistency described above (Supp.
-figure 1).
+figure XX).
 
 When testing model stability under a smaller sample size, we found reasonably good convergence in a dataset containing half the mass features with most values falling within two standard errors of the estimate for the XCMS and two-parameter model, while the full model required closer to 80% of the mass features to produce estimates consistent with the original model (Supp.
-figure 2).
+figure XX).
 
@@
 The random forest model, interestingly, had perfect predictive capacity when tested internally on the training data (FDR=0%, GFF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp.
-figure 3).
+figure XX).
 In each case, the performance drop when applied to a novel dataset was more extreme than the simple two-parameter model described above.
@@
 The class imbalance, with mostly poor quality MFs, is partially why we chose to measure precision and recall instead of total accuracy.
-However, precision and recall can also be ambiguous when the positive class is not specified and the raw confusion matrices are unavailable, thus our very precise use of the FDR and GFF metrics as well as providing the raw confusion matrices in Supplemental Table XX.
+However, precision and recall can also be ambiguous when the positive class is not specified and the raw confusion matrices are unavailable, thus our very precise use of the FDR and GFF metrics as well as providing the confusion matrices in Supplemental Table XX.
 @Melnikov2020 reported precision and recall in their presentation of `peakonly`, relative to which we obtained higher accuracy (they report 89% accuracy) but worse GFF and FDR (89% and 3%, respectively, relative to our 77.1% GFF and 19.6% FDR overall).
@@
 
-Our process of feature engineering was divided into three categories of features: those that could be calculated using a single peak trace, those that required access to the entire file, and those that could only be calculated across multiple files.
-[Rewrite without the "three categories" business ugh] For metrics belonging to the first two categories, summary statistics were used (mean/median, standard deviation) to create a single consensus value for the mass feature as a whole.
-Missing values were ignored during these summary statistic calculations by setting na.rm=TRUE.
-Distributions were visually inspected and highly abnormal metrics were transformed using log-scaling if necessary.
-Using a pairs plot, highly correlated (above a Pearson's r ~ 0.9) had one of the correlated metrics removed.
-
-The first category consisted largely of information produced directly by XCMS: the very typical *m/z* ratios, retention times, peak widths, and integrated areas as well as the less common and poorly-documented sn, f, scale, and lmin parameters.
-Although we requested additional parameters from XCMS with the verboseColumns = TRUE argument, all of the values returned were NAs and could not be used in the model building so were dropped.
-The total number of peaks found for a given feature was calculated as the number of peaks divided by the total number of files, and the total number of files in which a peak was initially detected was calculated by subtracting the number of NAs in the peak baseline estimate (which is NA if the peak was filled in) from the total number of files and then dividing by the number of files to normalize across datasets.
-This calculation were performed for the Falkor, MESOSCOPE, and CultureData datasets to estimate the proportion of sample and standard files in which a given molecular feature (MF) was found, but could not be calculated for the Pttime dataset because no standard files were available.
-The sn metric contained a large number of zero values to begin with that then became negative infinities when log-scaled, so we replaced those values with zeroes and functionally equated an sn of 0 with an sn of 1.
-
-We also calculated several metrics from the raw mz/rt/int values.
-Using the R package RaMS we extracted the data points falling within each individual peak's mz and retention time bounding box (values between the XCMS-reported min and max) separately for each file.
+Our process of feature engineering involved querying several MS experts in our lab about their intuition for what they thought best distinguished poor-quality MFs and noise from good ones.
+
+The simplest metrics to calculate were summary statistics of those parameters reported directly by XCMS.
+These features consisted of the mean retention time (RT) of each MF and the standard deviation (SD) within the feature and the mean peak width (calculated by subtracting the max RT from the minimum) and its SD.
+We also calculated the mean *m/z* ratio and the SD in parts-per-million (PPM) by dividing each peak's reported *m/z* ratio by the *m/z* ratio of the feature as a whole, then multiplying by one million.
+Mean peak area was calculated by taking the log$_{10}$ of the individual areas then taking the mean, and the same process (log$_{10}$ then mean) was repeated for the SD of the peak areas.
+XCMS's default signal-to-noise parameter, `sn`, was also summarized in this way, but we only used `sn` values that were greater than or equal to zero and replaced any zeros with ones to avoid negative infinities after taking the log$_{10}$.
+We also used the mean of other parameters reported by XCMS (f, scale, and lmin) as features.
+We additionally calculated several design-of-experiments metrics, using the number of peaks in each feature divided by the total number of files as well as the fraction of files in which a peak initially found by the peakpicker.
+This last metric was further subset into the fraction of samples in which a peak was initially found and the fraction of standards in which a peak was found (for those datasets in which standards were run).
+Finally, the coefficient of variance was estimated for the pooled sample peak areas by dividing the SD of the pooled sample peak areas by the mean of the same and additionally done in a robust way by using the median absolute deviation and median, respectively.
+For all of the above features, missing values were dropped silently from the summary calculations.
+We were unable to use any of the columns produced by enabling the `verboseColumns = TRUE` option in `findChromPeaks` because all of the values returned were NAs.
+
+We also calculated several novel metrics from the raw *m/z*/RT/intensity values.
+Using the R package RaMS we extracted the data points falling within each individual peak's *m/z* and RT bounding box (values between the XCMS-reported min and max) separately for each file.
 The data points were then linearly scaled to fall within the 0-1 range by subtracting the minimum RT and dividing by the maximum RT, then each scaled RT was fit to a beta distribution with α values of 2.5, 3, 4, and 5, and a fixed β value of 5.
 This approach allowed us to approximate a bell curve with increasing degrees of right-skewness and the beta distribution was chosen because it is constrained between 0 and 1 and simple and speedy to generate in R.
-For each α value, Pearson's correlation coefficient was calculated between the beta distribution and the raw data, with the highest value returned as a metric for how peak-shaped the data were (Figure 8).
-The beta distribution with the highest *r* was also then used to estimate the noise level within the peak by also scaling both the beta distribution densities and the raw data intensity values as described above, then subtracting the scaled beta distribution from the scaled intensity values, producing the residuals of the fit (Figure 8).
+For each α value, Pearson's correlation coefficient (*r*) was calculated between the beta distribution and the raw data, with the highest value returned as a metric for how peak-shaped the data were (Figure 8).
+The beta distribution with the highest *r* was also then used to estimate the noise level within the peak by scaling both the beta distribution probability densities and the raw data intensity values as described above, then subtracting the scaled beta distribution from the scaled intensity values, producing the residuals of the fit (Figure 8).
 The signal-to-noise ratio (SNR) was calculated by dividing the maximum original peak height by the standard deviation of the residuals multiplied by the maximum height of the original peak.
 This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional molecular signals.[@Myers2017b] If there were fewer than 5 data points, a missing value was returned and dropped in subsequent summary calculations.
-Accessing the raw data values also allowed us to calculate the proportion of "missed" scans in a peak for which a retention time exists at other masses in the same sample but for which no data was produced at the selected *m/z* ratio.
+Accessing the raw data values also allowed us to calculate the proportion of "missed" scans in a peak for which an RT exists at other masses in the same sample but for which no data was produced at the selected *m/z* ratio, divided by the total number of scans between the min and max RTs.
 
@@
 Normalization was performed by linearly scaling the raw values into the 0-1 range by subtracting the minimum value and dividing by the maximum.
-Peak shape similarity was measured with Pearson's correlation coefficient and the noise level is estimated as the standard deviation of the residuals after the raw data is subtracted from the idealized peak.](figures/peakmetrics_singlechrom.png){width=70%}
-
-We additionally estimated the existence of a $^{13}$C isotope using a similar method to extract the raw mz/rt/int values within the peak bounding box, then searched the same RT values at an *m/z* delta of +1.003355.
+Peak shape similarity was measured with Pearson's correlation coefficient and the noise level is estimated as the standard deviation of the residuals after the raw data is subtracted from the idealized peak.](figures/peakmetrics_singlechrom.png){width=90%}
+
+We additionally estimated the presence or absence of a $^{13}$C isotope using a similar method to extract the raw *m/z*/RT/intensity values within the peak bounding box, then searched the same RT values at an *m/z* delta of +1.003355 ± 4 PPM.
 In places where more than 5 data points existed at both the original mass and the $^{13}$C mass, we again used Pearson's correlation coefficient to estimate the similarity between the two mass traces and used a trapezoidal Riemann sum to estimate the area of the original and isotope peaks.
 The overall feature isotope shape similarity was calculated by taking the median of the correlation coefficients.
-We also calculated the correlation coefficient of the ratio of the $\frac{^{13}C}{^{12}C}$ across multiple files, expecting that a true isotope would have a fixed $\frac{^{13}C}{^{12}C}$ ratio.
+We also calculated the correlation coefficient of the ratio of the $\frac{^{13}C}{^{12}C}$ peak areas across multiple files, expecting that a true isotope would have a fixed $\frac{^{13}C}{^{12}C}$ ratio.
 Both the isotope shape similarity and the isotope area correlation were used as metrics in the downstream analysis.
 Peaks for which no isotope signal was detected or had too few scans to calculate the above metrics were imputed with NA values that were again dropped in the calculation of summary statistics for the molecular feature as a whole.
-
-Finally, we calculated several metrics using a design-of-experiments (DoE) approach for the Falkor and MESOSCOPE datasets which were expected to have a large difference in metabolite composition between those samples collected at the surface (15-25 meters) and those collected deeper in the water column.
-We used ANOVAs to compare the integrated peak area with depth and extracted the p-values of each test, which became an additional potential peak quality metric.
-We also calculated the difference between the mean sample area and the mean blank area and the difference between the mean sample area and the mean area of the standards which also became metrics.
-When there were too many missing values to perform the ANOVA or sample averages, we replaced the calculated statistic with a value a single order of magnitude outside the most extreme value.
+Because these isotope metrics typically had highly skewed distributions with most values very close to one, we normalized them by taking the log$_{10}$ of one minus the value.
+
+Distributions were visually inspected using a pairs plot and highly correlated (above a Pearson's r ~ 0.9) had one of the correlated metrics removed.
 
@@
 
+EIC: Extracted Ion Chromatogram
+
+FDR: False Discovery Rate
+
+GFF: Good Feature Found
+
+HILIC: Hydrophilic Interaction Liquid Chromatography
+
+LC: Liquid Chromatography
+
+MF: Molecular Feature
+
+MS: Mass Spectrometry
+
+PPM: parts-per-million
+
+RT: Retention time
+
 SNR: Signal to Noise Ratio
-
-MF: Molecular Feature
-
-GFF: Good Feature Found
-
-FDR: False Discovery Rate
-
-HILIC: Hydrophilic Interaction Liquid Chromatography
-
-EIC: Extracted Ion Chromatogram
-
-LC: Liquid Chromatography
-
-MS: Mass Spectrometry
 
@@
 ## Supplement
+
+```{r Supp figure param distributions, warning=FALSE}
+both_datasets <- list(MESOSCOPE=MS3000_features, Falkor=FT2040_features) %>%
+  bind_rows(.id = "dataset") %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_bin=ifelse(feat_class=="Good", 1, 0))
+pdf(file = "manuscript/supplement/metric_distributions.pdf", width = 6.5, height = 4)
+both_datasets %>%
+  select(mean_mz:area_cor) %>%
+  colnames() %>%
+  walk(function(chosen_param){
+    hist_gp <- both_datasets %>%
+      ggplot() +
+      geom_hline(yintercept = 0) +
+      geom_histogram(aes(x=get(chosen_param), fill=feat_class), bins=40) +
+      scale_fill_manual(breaks = c("Bad", "Good"), values = c("#a41118", "#f4bb23")) +
+      labs(y="# of mass features", x=chosen_param) +
+      theme_bw() +
+      theme(legend.direction = "horizontal", legend.position = "none") +
+      facet_wrap(~dataset, ncol=1, scales = "free_y")
+    log_gp <- both_datasets %>%
+      ggplot(aes(x=get(chosen_param), y=feat_bin)) +
+      geom_jitter(aes(color=feat_class), alpha=0.2, height = 0.1, width=0) +
+      geom_smooth(method = "glm", method.args=list(family=binomial), formula="y~x", color="black") +
+      scale_color_manual(breaks = c("Bad", "Good"), values = c("#a41118", "#f4bb23")) +
+      scale_y_continuous(breaks = c(0, 1), labels = c("Bad", "Good"), name=NULL) +
+      labs(x=chosen_param) +
+      theme_bw() +
+      theme(legend.position = "none", 
+            axis.text.y = ggtext::element_markdown(
+              color = c("#a41118", "#f4bb23"), face = "bold")) +
+      facet_wrap(~dataset, ncol=1, scales = "free_y")
+    egg::ggarrange(hist_gp, log_gp, nrow=1)
+  }, .progress = TRUE)
+graphics.off()
+```
+
+*Supplemental figure XX: Distribution and single-parameter logistic curves for each metric extracted for model training, shown separately for the MESOSCOPE and Falkor datasets.
+Histograms show the distribution of good and bad mass features by color across the span of the data on the x-axis with the number of MFs in each bin shown on the y-axis.
+Scatterplots show the same x-axis but show the results of a logistic regression on the single parameter, with the line of best fit in black and a ±1 standard error ribbon around it in grey.
+Vertical jittering has been applied when plotting to reduce the number of overlapping points.*
 
@@
   mutate(param_name=str_remove(param_name, "_params")) %>%
+  mutate(param_name=factor(param_name, levels=c("all", "xcms", "raw_data"), 
+                           labels=paste(c("Full", "XCMS-only", "Two-parameter"), 
+                                        "model"))) %>%
+  mutate(train_set=factor(train_set, levels=c("MS3000_features", "FT2040_features", "both_features"),
+                          labels=c("MESOSCOPE", "Falkor", "Combined"))) %>%
   ggplot(aes(x=param_name, color=train_set)) +
@@
   facet_wrap(~term, scales="free_y", ncol = 4) +
+  labs(x="Model type", y="Logistic regression estimate", color="Training set") +
   theme_bw() +
-  theme(legend.position = "top", axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
+  theme(legend.position = "top", 
+        axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
 ggsave(plot = suppfig_model_param_estimates_gp, 
@@
 
-![Supplemental figure 1: Model parameter estimates for each of the metrics in the full model, additionally broken down by their inclusion in the two-parameter (raw_data) and XCMS-exclusively models.
+![Supplemental figure XX: Model parameter estimates for each of the metrics in the full model, additionally broken down by their inclusion in the two-parameter (raw_data) and XCMS-exclusively models.
 Colors correspond to the dataset used to train the logistic regression model, with "both" indicating a combined model using all manually-labeled features across both datasets.
@@
 
-![Supplemental figure 2: Robustness of the two most significant metrics across the full (all_params), XCMS-only (xcms_params), and two-parameter (raw_data_params) models.
+![Supplemental figure XX: Robustness of the two most significant metrics across the full (all_params), XCMS-only (xcms_params), and two-parameter (raw_data_params) models.
 The x-axis corresponds to the fraction of the data used to train the model and the y-coordinate shows the estimated value for the specified term in the subset across 10-fold replicated subsampling.
@@
 
-![Supplemental figure 3: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.](supplement/regregrf_fdrgff_cross_train.png){width=90%}
+![Supplemental figure XX: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.](supplement/regregrf_fdrgff_cross_train.png){width=90%}
 

### 2023-06-09 feature_extraction.R

@@ -246,12 +246,12 @@ write.csv(peak_isodata, file = paste0(output_folder, "peak_isodata.csv"), row.na
 peak_isodata <- read_csv(paste0(output_folder, "peak_isodata.csv"))
 feat_isodata <- peak_isodata %>%
   group_by(feature) %>%
-  summarise(shape_cor=log10(1-median(iso_cor, na.rm=TRUE)), 
-            area_cor=log10(1-cor(init_area, iso_area, use="pairwise"))) %>%
-  mutate(shape_cor=ifelse(is.na(shape_cor), 1, shape_cor)) %>%
-  mutate(area_cor=ifelse(is.na(area_cor), 1, area_cor)) %>%
-  mutate(shape_cor=ifelse(is.infinite(shape_cor), 1, shape_cor)) %>%
-  mutate(area_cor=ifelse(is.infinite(area_cor), 1, area_cor))
+  summarise(iso_shape_cor=log10(1-median(iso_cor, na.rm=TRUE)), 
+            iso_area_cor=log10(1-cor(init_area, iso_area, use="pairwise"))) %>%
+  mutate(iso_shape_cor=ifelse(is.na(iso_shape_cor), 1, iso_shape_cor)) %>%
+  mutate(iso_area_cor=ifelse(is.na(iso_area_cor), 1, iso_area_cor)) %>%
+  mutate(iso_shape_cor=ifelse(is.infinite(iso_shape_cor), 1, iso_shape_cor)) %>%
+  mutate(iso_area_cor=ifelse(is.infinite(iso_area_cor), 1, iso_area_cor))
 write.csv(feat_isodata, paste0(output_folder, "feat_isodata.csv"), row.names = FALSE)
 
 # Calculate DOE metrics ----
@@ -404,7 +404,7 @@ features_extracted %>%
           color=~feat_class, text=~feature,
           type = "scatter3d", mode="markers")
 features_extracted %>%
-  plot_ly(x=~log10(1-shape_cor), y=~med_SNR, z=~log10(1-med_cor), color=~feat_class,
+  plot_ly(x=~log10(1-iso_shape_cor), y=~med_SNR, z=~log10(1-med_cor), color=~feat_class,
           type = "scatter3d", mode="markers")
 
 

### 2023-06-12 manuscript/manuscript.Rmd
Full rerun of entire analysis, more metric discussion

@@
 set.seed(123)
+both_datasets <- list(MESOSCOPE=MS3000_features, Falkor=FT2040_features) %>%
+  bind_rows(.id = "dataset") %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_bin=ifelse(feat_class=="Good", 1, 0))
 
@@
 both_datasets %>%
-  select(mean_mz:area_cor) %>%
+  select(mean_mz:iso_area_cor) %>%
   colnames() %>%

### 2023-06-12 manuscript/manuscript.Rmd
Scattered ce, fixed refs

@@
 In the past two decades, data-driven methods have established workflows for untargeted metabolomics but the imperfect performance of the core peakpicking algorithms continue to require manual oversight and curation.
-This problem has been exacerbated by the increased use of non-traditional chromatography such as hydrophilic interaction which tends to produce noisier peaks.[@Bajad2006; @Myers2017b; @Gika2019]
+This problem has been exacerbated by the increased use of non-traditional chromatography such as hydrophilic interaction which tends to produce noisier peaks [@Bajad2006; @Myers2017b; @Gika2019].
 
 Noisy data and imperfect detection algorithms introduce a tradeoff between false positives (where background instrument or chemical noise is misclassified as biological signal) and false negatives (where real signals are undetected).
-Existing algorithms tend to favor false positives because downstream analyses can always remove erroneous mass features, but false negatives cannot be later recovered.[@Pirttila2022; @Gloaguen2022] However, this approach creates two additional downstream problems.
-First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of molecular features (MFs), a task that scales combinatorially with the number of samples and compounds measured.[@Myers2017a] Second, false negatives cannot be fully accounted for in the way false positives can because false negatives are introduced in every step of metabolomic analyses from sample collection through injection.
+Existing algorithms tend to favor false positives because downstream analyses can always remove erroneous mass features, but false negatives cannot be later recovered [@Pirttila2022; @Gloaguen2022].
+However, this approach creates two downstream problems.
+First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of molecular features (MFs), a task that scales combinatorially with the number of samples and compounds measured [@Myers2017a].
+Second, false negatives cannot be fully accounted for in the way false positives can because false negatives are introduced in every step of metabolomic analyses from sample collection through injection.
 Instead of minimizing false negatives, we believe that emphasis should be placed on allowing the experimenter to set a threshold for the false discovery rate and accept that this will inherently add to the number of MFs already lost in the data collection process.
@@
 Existing peak-detection softwares do not provide a clear way to do this.
-Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each molecular feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew.[@Pirttila2022] None of these parameters answer the critical question about the likelihood that a given feature corresponds to a molecule present in the original sample.
+Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each molecular feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew [@Pirttila2022].
+None of these parameters answer the critical question about the likelihood that a given feature corresponds to a molecule present in the original sample.
 This parameter is crucial for downstream analysis because it represents the base rate for error propagation and acceptable thresholds should vary widely by the particular project's goals.
@@
 Constructing such a single comprehensive metric calibrated to likelihood is also more effective than multiple independent thresholds because it has meaningful units, does not require estimating the relative power of individual metrics, and allows a good MF to compensate for weak performance in one area with strong performance in other metrics, e.g.
-as implemented in @Pirttila2022.
+as implemented in @Pirttila2022 and @Kantz2019.
 
@@
 
-An area particularly ripe for metabolomic analysis is that of the open ocean.[@KidoSoule2015] Low compound and high salt concentrations make this area difficult to study but its vast size and direct effect on the Earth's biogeochemistry make it critical that we understand its dynamics of energy and nutrients on a molecular scale.[@Boysen2018] Metabolites are the currency of chemical exchange both intra- and inter-cellularly, serving as building blocks of larger molecules, regulators of osmotic balance and storage of nutrients, as well as important chemical signals on their own.
-These small molecules serve both as signposts for the complex biological landscape in this highly dynamic region and give a sense of not only who is present but also what ecological roles they're serving and the niches they fill.[@KidoSoule2015, @Boysen2021, @Heal2021]
+An area particularly ripe for metabolomic analysis is that of the open ocean [@KidoSoule2015].
+Low compound and high salt concentrations make this area difficult to study but its vast size and direct effect on the Earth's biogeochemistry make it critical that we understand its dynamics of energy and nutrients on a molecular scale [@Boysen2018].
+Metabolites are the currency of chemical exchange both intra- and inter-cellularly, serving as building blocks of larger molecules, regulators of osmotic balance and storage of nutrients, as well as important chemical signals on their own.
+These small molecules serve both as signposts for the complex biological landscape in this highly dynamic region and give a sense of not only who is present but also what ecological roles they're serving and the niches they fill [@KidoSoule2015; @Boysen2021; @Heal2021].
 
@@
 Colors indicate the category in which each feature was manually assigned by an expert, with "Stans only" referring to a good molecular feature that was only visible in the standards run alongside the samples.
-Culture datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (black vertical line) and were otherwise unclassified.](figures/pred_prob_class_color_hists.png){width=90%}
+Culture datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (dotted black vertical line) according to the final model and were otherwise unclassified.](figures/pred_prob_class_color_hists.png){width=90%}
 
@@
 The two-parameter model performed worst when tested internally, with an FDR of about 20% and GFF also around 75% (Figure 2).
-However, when the models were trained on a different dataset than the one they were used to predict classifications for, all models had similar performance with %FDR around 10-25 and GFF around 60-80.
+However, when the models were trained on a different dataset than the one they were used to predict classifications for, they all had similar performance with FDRs around 10-25 and GFF around 60-80.
 The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more MFs recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE (Figure 2).
@@
 
-![Figure 2: False discovery rate (%FDR) and fraction of good features found (%GFF) plotted across different subsets of model parameters.
-Lower %FDR indicates a smaller fraction of false positives among those molecular features the model categorized as "Good" using a threshold of 0.5, and higher %GFF indicates a larger fraction of the total good features were found using the same threshold.
+![Figure 2: False discovery rate (FDR) and fraction of good features found (GFF) plotted across different subsets of model parameters.
+Lower FDR indicates a smaller fraction of false positives among those molecular features the model categorized as "Good" using a threshold of 0.5, and higher GFF indicates a larger fraction of the total good features were found using the same threshold.
 Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel.
-Lines of best fit have been estimated and plotted in black behind the data points, with the steepest slopes found in the full and XCMS-only models indicating excessive overfitting on the training data.](figures/fdr_gff_cross_train.png){width=90%}
+Lines of best fit have been estimated and plotted in black behind the data points, with the steeper slopes found in the full and XCMS-only models indicating excessive overfitting on the training data.](figures/fdr_gff_cross_train.png){width=90%}
 
@@
 We found that the predictions made from a Falkor-trained dataset consistently differed from a MESOSCOPE-trained dataset for the full and XCMS-only models.
-In the raw probability space, the two-parameter models had the highest $r$ value of 0.996, while the full models and the XCMS-trained models had $r$ values of 0.799 and 0.863, respectively.
-When compared in ranked space, we found an intensification of the effect described previously, with a higher $ρ$ for the two-parameter model of 0.998 but lower $ρ$ values for the full and XCMS-trained model of 0.725 and 0.804, respectively (Figure 3).
+In the raw probability space, the two-parameter models had the highest Pearson correlation coefficient ($r$) value of 0.996, while the full models and the XCMS-trained models had $r$ values of 0.799 and 0.863, respectively.
+When compared in ranked space using Spearman's ranked correlation, we found an intensification of this effect, with a higher $ρ$ for the two-parameter model of 0.998 but lower $ρ$ values for the full and XCMS-trained model of 0.725 and 0.804, respectively (Figure 3).
 
@@
   theme_bw() +
-  theme(legend.position = "none",
-        strip.background = element_rect(fill="white"))
+  theme(legend.position = "none")
 rank_prob_space_gp <- dotplot_stepdown_data %>%
@@
 
-A majority of the time, the estimates from the two models disagreed by more than 2x the standard error of the estimate.
+A majority of the time, the estimates from the two models disagreed by more than two times the standard error of the estimate.
 Some parameters disagreed not only in magnitude but also in sign, with the Falkor-trained full model increasing MF goodness likelihood with larger PPM variation and a wider peak width, while the MESOSCOPE-trained full model had negative estimates for each of these parameters.
@@
 
-When testing model stability under a smaller sample size, we found reasonably good convergence in a dataset containing half the mass features with most values falling within two standard errors of the estimate for the XCMS and two-parameter model, while the full model required closer to 80% of the mass features to produce estimates consistent with the original model (Supp.
+When testing model stability under a smaller sample size, we found reasonably good convergence in a dataset containing half the mass features with most model parameters falling within two standard errors of the estimate for the XCMS and two-parameter model, while the full model required closer to 80% of the mass features to produce estimates consistent with the original model (Supp.
 figure XX).
@@
 All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GFF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent false negative acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp.
-figure 3).
+figure XX).
 The random forest model, interestingly, had perfect predictive capacity when tested internally on the training data (FDR=0%, GFF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp.
@@
 In both the Falkor and MESOSCOPE datasets, less than half of the good MFs were labeled as such, with actual values of 39.4% and 26.5%, respectively.
-Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the %GFF but expect it to be in a similar range (Figure 4).
-
-```{r FDR vs GFF with additional datasets}
+Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the GFF but expect it to be in a similar range (Figure 4).
+
+```{r FDR vs GFF with additional datasets, warning=FALSE}
 # Currently a fairly simple figure, may need additional data?
@@
 gff_alldatasets_gp <- error_type_all_datasets %>%
+  drop_na() %>%
   mutate(GFF_label=paste0("frac(",`%TP`,",",`%FN`+`%TP`,")")) %>%
@@
                      name = "Fraction of good\nfeatures found (GFF)") +
+  scale_x_discrete(drop=FALSE) +
   theme_bw() +
@@
         axis.text.x = element_blank())
-fdr_gff_all_datasets_gp <- egg::ggarrange(gff_alldatasets_gp, fdr_alldatasets_gp, ncol = 1)
+fdr_gff_all_datasets_gp <- egg::ggarrange(gff_alldatasets_gp, fdr_alldatasets_gp, ncol = 1, draw = FALSE)
 
@@
 These features possess an unexpected biological signal that does not track with depth or other common oceanographic parameters, thereby potentially representing an interesting biomarker or pathway that's downregulated in excess of the biomass effect.
-[Maybe better for discussion?]
 
@@
                  "Bad MF avoided")
-error_pchs <- c(19, 4, 3, 1)
-error_unicodes <- c("\u25cf", "\u00d7", "\u002B", "\u25e6")
+error_pchs <- c(19, 1, 3, 4)
+error_unicodes <- c("\u25cf", "\u25e6", "\u002B", "\u00d7")
 error_type_labels <- error_order %>%
@@
 
-knitr::kable(thresh_statistic_df)
+# knitr::kable(thresh_statistic_df)
 ```
@@
 In these common exploratory plots, the MFs with likelihoods above 50% strongly separated by depth while lower thresholds disguised the true signal and had higher stress values.
-Performing an NMDS on the manually-identified Good MFs resulted in output nearly indistinguishable from those of the 90% and 50% thresholds, while the 0.1 and 0.01 thresholds looked almost identical to the same analysis performed on all the XCMS features.
+Performing an NMDS on the manually-identified Good MFs resulted in output nearly indistinguishable from those of the 90% and 50% thresholds.
 
@@
 One of the major ways in which this manuscript differs from prior work is its focus on summary statistics calculated across multiple files.
-Most existing peakpicking literature uses the single-file EIC peak as the core bit of training data, but that approach ignores critical information obtained elsewhere in the MS run that can change the judgement made on a single chromatogram.[@Pirttila2022; @Guo2021; @Muller2020] Features that are high quality are typically represented in multiple files, especially in quality control pooled samples, and a feature that only has a peak in a single file is typically regarded as highly suspect, if not removed entirely.
+Most existing peakpicking literature uses the single-file EIC peak as the core bit of training data, but that approach ignores critical information obtained elsewhere in the MS run that can change the judgement made on a single chromatogram [@Pirttila2022; @Guo2021; @Muller2020].
+Features that are high quality are typically represented in multiple files, especially in quality control pooled samples, and a feature that only has a peak in a single file is typically regarded as highly suspect, if not removed entirely.
 Classifying an entire feature at once not only has the advantage of reducing the amount of manual labor by a factor equal to the number of files in the run (typically 10s-100s smaller) but is also a better representation of the judgement made by an MS expert.
@@
 @Guo2021 presented EVA and reported an accuracy of 90-95%, a range inclusive of our accuracy on both the Falkor (92.1%) and MESOSCOPE (94.4%) datasets when using a likelihood threshold of 0.5.
-However, we note that accuracy alone can be a highly misleading statistic to report when working with unbalanced datasets because very high accuracy can be obtained by simply classifying everything as bad, with a strong incentive to actually *increase* the number of bad MFs detected while doing so.
+However, we note that accuracy alone can be a highly misleading statistic to report when working with unbalanced datasets because very high accuracy can be obtained by simply classifying everything as bad, with a strong incentive to actually *increase* the number of bad MFs initially picked while doing so.
 This strategy, when applied to our data, returned accuracies in the 80-90% range despite being a useless classifier for downstream analysis.
-This is perhaps a facetious argument, but many of the single-parameter logistic models we tested never exceeded a 50% likelihood and essentially functioned as this type of classifier.
 
@@
 The two metrics in the final model were rederived from the raw EIC data because they matched our intuition about what makes an MF look good to an MS expert.
+However, they could be implemented in XCMS directly, removing the need to revisit the raw data.
 It is worth noting that these are very simple metrics and therefore fast to calculate, but we expect that more complicated metrics could perform even better.
@@
 Additionally, the calculation of peak shape using a Pearson's correlation to an idealized curve was not expected to be especially powerful given prior research (e.g.
-@Ipsen2010) and that the centWave algorithm essentially uses this information already during the wavelet fitting, but still proved to be a highly effective parameter.
+@Ipsen2010) and that the centWave algorithm essentially uses this information already during the wavelet fitting, but still proved to be a highly informative parameter.
 This metric could be improved with more careful summary statistics that account for the differences between samples.
@@
 We were surprised at the poor performance of several other metrics.
-The isotope information in particular was expected to be a very strong predictor of MF quality given previous work that uses this metric extensively.[@Libiseller2015; @Treutler2016; @ElAbiead2021] We learned that many noise MFs still have reliable isotopes (perhaps unsurprising, given that the noise is in fact often caused by solvents or contaminants that are still chemical in nature) and that many real MFs are simply too low-intensity to have them in this kind of dilute environmental sample.
+The isotope information in particular was expected to be a very strong predictor of MF quality given previous work that uses this metric extensively [@Libiseller2015; @Treutler2016; @ElAbiead2021].
+We learned that many noise MFs still have reliable isotopes (perhaps unsurprising, given that the noise is in fact often caused by solvents or contaminants that are still chemical in nature) and that many real MFs are simply too low-intensity to have them in this kind of dilute environmental sample.
 
@@
 This result was also reported by @Gloaguen2022 who noted that while the RSD was typically lower for high-quality features there were many noise MFs with low RSDs as well.
-We also note that though they do not report the statistical significance of the RSD shift across feature quality, it doesn't appear to be hugely significant without a very large sample size.
 
@@
 Finally, we were surprised to find essentially no predictive power offered by peak area or intensity, with good MFs distributed almost identically to the bad MFs in this space.
-This cautions strongly against an arbitrarily-decided intensity threshold for winnowing down the number of MFs, in agreement with previous work.[@Houriet2022; @Barupal2021]
+This cautions strongly against an arbitrarily-decided intensity threshold for winnowing down the number of MFs, in agreement with previous work [@Houriet2022; @Barupal2021].
 
@@
 We settled on the highly reduced model of just two parameters because we found that additional parameters often improved performance on the training set but did not do so significantly for the novel datasets where the application of such a model is actually useful (Figure 2).
+The drastic drop in performance on out-of-sample data was particularly concerning because it creates overconfidence in the true level of noise actually ending up in the final dataset.
 One important caveat to note is that for the partially-labelled CultureData and Pttime datasets, there exists an uncontrolled degree of experimenter bias because the MS expert responsible for labeling did know that these MFs were all expected to be good.
 However, given that we do still see poor-quality MFs in this set indicates that this was not an overwhelming bias.
-The drastic drop in performance on out-of-sample data was particularly concerning because it creates overconfidence in the true level of noise actually ending up in the final dataset.
 
 We also found that this reduced two-parameter model was largely independent of the particular training set used, unlike in the more complex models (Figure 3).
-This was true in both absolute likelihood as well as rank-ordered space, a particularly important distinction when one imagines manually labeling "down" the dataset where the researcher desires to start viewing the chromatograms associated with the very best features and eventually reaching a point where enough MFs have been reviewed or bad MFs are consistent enough that they decide to forgo additional labeling.
+This was true in both absolute likelihood as well as rank-ordered space, a particularly important distinction when one imagines manually labeling "down" the dataset where the researcher starts viewing the chromatograms associated with the very best features and eventually reaches a point where enough MFs have been reviewed or bad MFs are frequent enough that they decide to stop.
 
@@
 The real features showed a strong biological trend of high concentration throughout the surface ocean and down through the deep chlorophyll maximum (DCM), with most features equally abundant at 15 meters and this ~110 meter depth feature before dropping off at depth.
-This pattern tracks well with previous reports of biomass from the same sample site as well as earlier literature.[@Barone2022; @Heal2021] Critically, this also highlights the danger of noise MFs when additional normalizations are later applied.
+This pattern tracks well with previous reports of biomass from the same sample site as well as earlier literature [@Barone2022; @Heal2021].
+Critically, this also highlights the danger of noise MFs when additional normalizations are later applied.
 Scaling metabolomic data to biomass measurements is a common technique, and yet here it would have caused an enormous number of false positives that would have appeared to be intriguingly enriched below the DCM.
@@
 
-Culture samples used as the validation sets for this paper have been previously described by [@Durham2022] and on Metabolomics Workbench (Project ID PR001317).
+Culture samples used as the validation sets for this paper have been previously described by @Durham2022 and on Metabolomics Workbench (Project ID PR001317).
 
@@
 
-Extraction of the environmental samples followed a modified Bligh & Dyer approach as detailed in [@Boysen2018].
+Extraction of the environmental samples followed a modified Bligh & Dyer approach as detailed in @Boysen2018.
 Briefly, filters were added to PTFE centrifuge tubes with a 1:1 mix of 100 µm and 400 µm silica beads, approximately 2mL -20 °C Optima-grade DCM, and approximately 3mL -20 °C 1:1 methanol/water solution (both also Optima-grade).
@@
 
-The R package XCMS was used to perform peakpicking, retention time correction, and peak correspondence.[@Smith2006, @Tautenhahn2008] Files were loaded and run separately for each dataset (Falkor, MESOSCOPE, CultureData, and Pttime) using the new "OnDiskMSnExp" infrastructure.
+The R package XCMS was used to perform peakpicking, retention time correction, and peak correspondence [@Smith2006; @Tautenhahn2008].
+Files were loaded and run separately for each dataset using the "OnDiskMSnExp" infrastructure.
 Default parameters for the CentWave peakpicking algorithm were used except for: ppm, which was set to 5; peakwidth, which was widened to 20-80 seconds; prefilter, for which the intensity threshold was raised to $10^6$; and integrate, which was set to 2 instead of 1.
 snthresh was set to zero because there are known issues with background estimation in this algorithm [@Myers2017a], and both verboseColumns and the extendLengthMSW parameter were set to TRUE.
-For retention time correction, the Obiwarp method was used except for the CultureData dataset, which was visually inspected and determined not to require correction.[@Benton2010] For the Obiwarp algorithm, the binsize was reduced to 0.1 but all other parameters were left at their defaults or equivalents.
+For retention time correction, the Obiwarp method was used except for the CultureData dataset, which was visually inspected and determined not to require correction [@Benton2010].
+For the Obiwarp algorithm, the binsize was reduced to 0.1 but all other parameters were left at their defaults or equivalents.
 
@@
 For the Falkor and MESOSCOPE datasets, every molecular feature was inspected, while only those features with a predicted probability of 0.9 or higher according to the two-parameter model produced below were inspected for the CultureData and Pttime datasets.
-Inspection consisted of plotting the raw intensity values against the corrected retention-time values for all data points within the *m/z* x RT bounding box determined by the most extreme values for the given feature.
+Inspection consisted of plotting the raw intensity values against the corrected retention-time values for all data points within the *m/z* by RT bounding box determined by the most extreme values for the given feature.
 For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labeling and to more accurately represent what MS experts typically do when assessing the quality of a given molecular feature (Figure 7).
@@
 
-We also calculated several novel metrics from the raw *m/z*/RT/intensity values.
-Using the R package RaMS we extracted the data points falling within each individual peak's *m/z* and RT bounding box (values between the XCMS-reported min and max) separately for each file.
+We also calculated several novel metrics from the raw *m/z*/RT/intensity values by extracting the data points falling within each individual peak's *m/z* and RT bounding box (values between the XCMS-reported min and max) separately for each file.
 The data points were then linearly scaled to fall within the 0-1 range by subtracting the minimum RT and dividing by the maximum RT, then each scaled RT was fit to a beta distribution with α values of 2.5, 3, 4, and 5, and a fixed β value of 5.
@@
 The signal-to-noise ratio (SNR) was calculated by dividing the maximum original peak height by the standard deviation of the residuals multiplied by the maximum height of the original peak.
-This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional molecular signals.[@Myers2017b] If there were fewer than 5 data points, a missing value was returned and dropped in subsequent summary calculations.
+This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional molecular signals [@Myers2017b].
+If there were fewer than 5 data points, a missing value was returned and dropped in subsequent summary calculations.
 Accessing the raw data values also allowed us to calculate the proportion of "missed" scans in a peak for which an RT exists at other masses in the same sample but for which no data was produced at the selected *m/z* ratio, divided by the total number of scans between the min and max RTs.
@@
 
-In each case, we categorized each molecular feature as a true positive (TP) if it was predicted to be Good and was manually classified as Good, a true negative if both predicted and classifed as Bad, a false positive if predicted to be Good but manually classified as Bad, and a false negative if predicted to be Bad but was in fact manually classified as Good.
+In each case, we categorized each molecular feature as a true positive (TP) if it was predicted to be Good and was manually classified as Good, a true negative if both predicted and classified as Bad, a false positive if predicted to be Good but manually classified as Bad, and a false negative if predicted to be Bad but was in fact manually classified as Good.
 This allowed us to additionally define two useful measures of success, the traditionally-defined false discovery rate (FDR, defined as 1-precision or the number of false positives divided by the total number of predicted positives) and the percentage of good features found (GFF, also known as the recall or sensitivity and defined as the number of true positives divided by the total number of actual positives).
@@
 For the raw probability space we compared the predictions using Pearson's correlation coefficient, while Spearman's rank-ordered coefficient was used for the ranked space.
-We also looked at the estimates produced by these two models and compared them with the combined model trained on both datasets combined to assess the model stability directly.
-
-We also tested the robustness of the model under a smaller training set, emulating a situation in which only a fraction of the data was available or only a portion of the mass features had been labeled.
+We additionally looked at the estimates produced by these two models and compared them with the combined model trained on both datasets combined to assess the model stability directly.
+
+We also measured the robustness of the model under a smaller training set, emulating a situation in which only a fraction of the data was available or only a portion of the mass features had been labeled.
 This allowed us to test the required sample size for the different models, with a larger sample size presumably required for the models with more parameters.
@@
 
-We also tested whether the performance could be improved with elastic net regression or random forest models.
+Finally, we tested whether the performance could be improved with regularized regression or random forest models.
 These models handle correlated variables better than ordinary least squares regression, so we also included several additional implementations of the peak shape and novel SNR parameters when summarizing across multiple files, using a max and a median of the top-three best values rather than just the overall median as well as a log-transformed version of the median peak shape calculated as $median(log_{10}(1-r))$ where $r$ is Pearson's correlation coefficient, as described above (Figure 2).
 Cross-validation was used to select the optimal tuning parameter $\lambda$ with `glmnet` package's `cv.glmnet` for an elastic net penalty (α) of 0, 0.5, and 1.
-Random forests were implemented using the randomForest package with default settings and a factor-type response vector to ensure classification was used rather than regression.
+Random forests were implemented using the randomForest package with default settings and a factor-type response vector to ensure classification was applied rather than regression.
 
@@
 This project dataset consists of *Phaeodactylum tricornutum* cultures collected at a variety of timepoints from both pelleted cells and the released exudate.
-This dataset was chosen because of the similar LC-MS setup used as a benchmark comparison for the performance other labs with similar setups may expect to achieve using the trained model directly.
+This dataset was chosen because of the similar LC-MS setup used as a benchmark for the performance other labs with similar setups may expect to achieve using the trained model directly.
 
@@
 In univariate space, we used nonparametric Kruskal-Wallis analyses of variance to measure the difference between the surface (15m), DCM (~110m), and 175m samples because the metabolite peak areas could not be assumed to be normally distributed.
-These univariate tests were then controlled for multiply hypothesis testing using R's `p.adjust` function with method `fdr`.[@Benjamini1995] We also performed post-hoc Dunn tests provided by the `rstatix` package to categorize the response to depth for those mass features for which the KW test was significant, with responses falling into one of the 14 classes possible when permuting the sign and significance of the Dunn test.[@Dunn1964] p-values obtained from the Dunn tests were not FDR controlled because it was used as a categorization tool rather than a null hypothesis test.
-In multivariate space, we used a permutational MANOVA (PERMANOVA) [@Anderson2014] provided by the vegan package's `adonis2` function to test for multivariate differences in structure of the metabolome with depth.[@veganpkg] We ran multiple PERMANOVAs with a different subset of mass features included each time, corresponding to using the output from XCMS directly, likelihood thresholds of 0.01, 0.1, 0.5, 0.9, and finally only those MFs manually annotated as good.
-
-All analyses were run in R, version 4.2.2, and code is available on the Github at https://github.com/wkumler/MS_metrics.[@R]
+These univariate tests were then controlled for multiply hypothesis testing using R's `p.adjust` function with method `fdr` [@Benjamini1995].
+We also performed post-hoc Dunn tests provided by the `rstatix` package to categorize the response to depth for those mass features for which the KW test was significant, with responses falling into one of the 14 classes possible when permuting the sign and significance of the Dunn test outputs [@Dunn1964].
+p-values obtained from the Dunn tests were not FDR controlled because it was used as a categorization tool rather than a null hypothesis test.
+In multivariate space, we used a permutational MANOVA (PERMANOVA) [@Anderson2014] provided by the vegan package's `adonis2` function to test for multivariate differences in structure of the metabolome with depth [@veganpkg].
+We ran multiple PERMANOVAs with a different subset of mass features included each time, corresponding to using the output from XCMS directly, likelihood thresholds of 0.01, 0.1, 0.5, 0.9, and finally only those MFs manually annotated as good.
+
+All analyses were run in R [@R], version 4.2.2, and code is available on GitHub at https://github.com/wkumler/MS_metrics.
 
@@
     egg::ggarrange(hist_gp, log_gp, nrow=1)
-  }, .progress = TRUE)
+  })
 graphics.off()
@@
 ![Supplemental figure XX: Model parameter estimates for each of the metrics in the full model, additionally broken down by their inclusion in the two-parameter (raw_data) and XCMS-exclusively models.
-Colors correspond to the dataset used to train the logistic regression model, with "both" indicating a combined model using all manually-labeled features across both datasets.
-<Explanations of the actual facet names forthcoming.>](supplement/suppfig_model_param_estimates.png){width=90%}
+Colors correspond to the dataset used to train the logistic regression model, with "both" indicating a combined model using all manually-labeled features across both datasets.](supplement/suppfig_model_param_estimates.png){width=90%}
 

### 2023-06-12 manuscript/manuscript.Rmd
Switching to "Mass" instead of "Molecular" feature terminology

@@
 However, this approach creates two downstream problems.
-First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of molecular features (MFs), a task that scales combinatorially with the number of samples and compounds measured [@Myers2017a].
+First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of mass features (MFs), a task that scales combinatorially with the number of samples and compounds measured [@Myers2017a].
 Second, false negatives cannot be fully accounted for in the way false positives can because false negatives are introduced in every step of metabolomic analyses from sample collection through injection.
@@
 Existing peak-detection softwares do not provide a clear way to do this.
-Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each molecular feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew [@Pirttila2022].
+Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each mass feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew [@Pirttila2022].
 None of these parameters answer the critical question about the likelihood that a given feature corresponds to a molecule present in the original sample.
 This parameter is crucial for downstream analysis because it represents the base rate for error propagation and acceptable thresholds should vary widely by the particular project's goals.
-In an exploratory analysis, any molecular feature more than 50% likely to be real is perhaps worth considering, while in a confirmatory study this threshold may need to be above 99%.
+In an exploratory analysis, any mass feature more than 50% likely to be real is perhaps worth considering, while in a confirmatory study this threshold may need to be above 99%.
 Despite significant effort invested in improving the peakpicking algorithms, very little has been done to quantify the accuracy and precision of their outputs across the wide variety of datasets to which they are applied.
@@
 
-An average of 3,300 molecular features (MFs) were reported by XCMS across the 4 datasets, with the fewest (1,495) in the Falkor data and the most in the Pttime samples (7,781).
+An average of 3,300 mass features (MFs) were reported by XCMS across the 4 datasets, with the fewest (1,495) in the Falkor data and the most in the Pttime samples (7,781).
 In the Falkor and MESOSCOPE datasets that were fully labeled by an MS expert, approximately 70% (69% and 73%, respectively) of the features were given a "Bad" designation, corresponding to noise MFs that the expert would not have included in a downstream analysis.
@@
 ![Figure 1: Histograms showing the estimated likelihood of a given mass feature being categorized as "Good" according to the two-parameter logistic model trained on the combined fully-labeled Falkor and MESOSCOPE environmental datasets.
-Colors indicate the category in which each feature was manually assigned by an expert, with "Stans only" referring to a good molecular feature that was only visible in the standards run alongside the samples.
+Colors indicate the category in which each feature was manually assigned by an expert, with "Stans only" referring to a good mass feature that was only visible in the standards run alongside the samples.
 Culture datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (dotted black vertical line) according to the final model and were otherwise unclassified.](figures/pred_prob_class_color_hists.png){width=90%}
@@
 ![Figure 2: False discovery rate (FDR) and fraction of good features found (GFF) plotted across different subsets of model parameters.
-Lower FDR indicates a smaller fraction of false positives among those molecular features the model categorized as "Good" using a threshold of 0.5, and higher GFF indicates a larger fraction of the total good features were found using the same threshold.
+Lower FDR indicates a smaller fraction of false positives among those mass features the model categorized as "Good" using a threshold of 0.5, and higher GFF indicates a larger fraction of the total good features were found using the same threshold.
 Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel.
@@
 A majority of the features (1,323 of 1,832 total) in the original, non-thresholded MESOSCOPE dataset had no significant trend with depth, with FDR-controlled Kruskal-Wallis p-values exceeding 0.05 (Figure 5).
-The largest category that did have a trend with depth was the 15m = DCM > 175m category, containing 118 molecular features, with largest peak areas distributed evenly between the 15 meter and deep chlorophyll maximum (DCM) samples, while the 175 meter samples had significantly smaller areas.
+The largest category that did have a trend with depth was the 15m = DCM > 175m category, containing 118 mass features, with largest peak areas distributed evenly between the 15 meter and deep chlorophyll maximum (DCM) samples, while the 175 meter samples had significantly smaller areas.
 The similar but statistically distinct categories of 15m > DCM > 175m and DCM > 15m > 175m had 68 and 35 features, respectively, and together indicate that many molecules are highly abundant throughout the surface ocean down to the DCM layer and decrease in concentration thereafter.
@@
 All MFs are shown in the central bottom plot across three axes using the rank-normalized median value at each depth as the coordinate for that axis.
-Each molecular feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile.
+Each mass feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile.
 Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the DCM; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}
@@
 We used two fully-labeled and two partially-labeled HILIC LC-MS datasets to assess the performance of the XCMS algorithm and construct a robust model of peak quality.
-To measure performance, we used two measurements of success closely tied to intuitive questions about a dataset: the percentage of total good features found (GFF, also known as recall or sensitivity) and the percentage of bad molecular features (MFs) included, also known as the false discovery rate (FDR).
+To measure performance, we used two measurements of success closely tied to intuitive questions about a dataset: the percentage of total good features found (GFF, also known as recall or sensitivity) and the percentage of bad mass features (MFs) included, also known as the false discovery rate (FDR).
 We decided against using the F$_1$ score as an overall summary statistic because false negatives and false positives have very different implications in this context and should be treated separately.
@@
 
-We explored several different types of classification models for separating good molecular features (MFs) from bad, with a particular focus on quantifying the likelihood of each class rather than just returning the label.
+We explored several different types of classification models for separating good mass features (MFs) from bad, with a particular focus on quantifying the likelihood of each class rather than just returning the label.
 We found that a simple two-parameter logistic regression model trained on two novel metrics of peak quality had reasonably good performance on the training set and was highly robust when applied to novel datasets.
@@
 This cautions strongly against an arbitrarily-decided intensity threshold for winnowing down the number of MFs, in agreement with previous work [@Houriet2022; @Barupal2021].
+Similarly surprising was the lack of power in the design-of-experiments metrics, although this was less surprising given the number of missing values that were later filled in with an order of magnitude outside the most extreme value (Supp.
+figure XX).
 
@@
 
-The large number of molecular features due to noise present in metabolomics datasets can be controlled using a simple logistic classification model.
+The large number of mass features due to noise present in metabolomics datasets can be controlled using a simple logistic classification model.
 We trained such a model on two full-labeled open ocean HILIC datasets and found that the best performing parameters in the model for were a custom signal-to-noise metric and a test of similarity to a bell curve.
@@
 
-After the full XCMS workflow was completed, the molecular features were visually inspected by a single qualified MS expert.
-For the Falkor and MESOSCOPE datasets, every molecular feature was inspected, while only those features with a predicted probability of 0.9 or higher according to the two-parameter model produced below were inspected for the CultureData and Pttime datasets.
+After the full XCMS workflow was completed, the mass features were visually inspected by a single qualified MS expert.
+For the Falkor and MESOSCOPE datasets, every mass feature was inspected, while only those features with a predicted probability of 0.9 or higher according to the two-parameter model produced below were inspected for the CultureData and Pttime datasets.
 Inspection consisted of plotting the raw intensity values against the corrected retention-time values for all data points within the *m/z* by RT bounding box determined by the most extreme values for the given feature.
-For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labeling and to more accurately represent what MS experts typically do when assessing the quality of a given molecular feature (Figure 7).
+For this step, we decided to plot the entire feature across all files simultaneously rather than viewing each sample individually to both accelerate labeling and to more accurately represent what MS experts typically do when assessing the quality of a given mass feature (Figure 7).
 We also decided to ignore missing values and linearly interpolate between known data points rather than filling with zeroes.
@@
 DCM = deep chlorophyll maximum, approximately 110 meters.
-Molecular feature ids are provided as the title of each panel, starting with "FT" and followed by 4 digits.
-Two features were annotated using authentic standards run alongside: the 13C isotope of dimethylsulfonioacetate (DMS-Ac) and taurine.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
+Mass feature ids are provided as the title of each panel, starting with "FT" and followed by 4 digits except for the two features annotated using authentic standards run alongside: the 13C isotope of dimethylsulfonioacetate (DMS-Ac) and taurine.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
 
@@
 Both the isotope shape similarity and the isotope area correlation were used as metrics in the downstream analysis.
-Peaks for which no isotope signal was detected or had too few scans to calculate the above metrics were imputed with NA values that were again dropped in the calculation of summary statistics for the molecular feature as a whole.
+Peaks for which no isotope signal was detected or had too few scans to calculate the above metrics were imputed with NA values that were again dropped in the calculation of summary statistics for the mass feature as a whole.
 Because these isotope metrics typically had highly skewed distributions with most values very close to one, we normalized them by taking the log$_{10}$ of one minus the value.
@@
 
-In each case, we categorized each molecular feature as a true positive (TP) if it was predicted to be Good and was manually classified as Good, a true negative if both predicted and classified as Bad, a false positive if predicted to be Good but manually classified as Bad, and a false negative if predicted to be Bad but was in fact manually classified as Good.
+In each case, we categorized each mass feature as a true positive (TP) if it was predicted to be Good and was manually classified as Good, a true negative if both predicted and classified as Bad, a false positive if predicted to be Good but manually classified as Bad, and a false negative if predicted to be Bad but was in fact manually classified as Good.
 This allowed us to additionally define two useful measures of success, the traditionally-defined false discovery rate (FDR, defined as 1-precision or the number of false positives divided by the total number of predicted positives) and the percentage of good features found (GFF, also known as the recall or sensitivity and defined as the number of true positives divided by the total number of actual positives).
@@
 This allowed us to test the required sample size for the different models, with a larger sample size presumably required for the models with more parameters.
-Because no parameter was present in all 3 models, we looked at the top 2 most significant parameters from each model: average *m/z* and peak shape for the full model, average peakwidth and the standard deviation in retention time for the XCMS model, and peak shape and SNR for the two-parameter model.
+Because no parameter was present in all 3 models, we looked at the top 2 most significant parameters from each model: average *m/z* and peak shape for the full model, average *m/z* and the standard deviation in retention time for the XCMS model, and peak shape and SNR for the two-parameter model.
 
@@
 
-MF: Molecular Feature
+MF: Mass Feature
 

### 2023-07-17 manuscript/manuscript.Rmd
Fixed supplemental figure numbers

@@
     reference_docx: template.docx
-  pdf_document: default
+  pdf_document:
+    latex_engine: xelatex
 bibliography: refs.bib
@@
 
-[Where to talk about the peakpicking vs integration? Here we just looked at peakpicking]
-
 An area particularly ripe for metabolomic analysis is that of the open ocean [@KidoSoule2015].
@@
 The isotope shape and area correlations also showed good separation (Supp.
-figure XX).
+figure 1).
 
@@
   left_join(rf_estimates, by="term") %>%
-  write_csv("manuscript/supplement/full_model_estimates.csv")
+  write_csv("manuscript/supplement/supp_table_1_full_model_estimates.csv")
 ```
@@
 We explored the relative predictive power of the individual parameters using the full model and found that the predictors least likely to be different from zero due to chance were the mean *m/z* ratio, the novel peak shape correlation metric `med_cor`, and the novel SNR estimate `med_SNR`, all with reported p-values < 10$^{-10}$.
-The novel parameters were then validated using a random forest model that also found them to have the highest importance (Supp table XX).
+The novel parameters were then validated using a random forest model that also found them to have the highest importance (Supp.
+table 1).
 
@@
 Notably, the peak shape and novel SNR parameters used in the two-parameter model were among the most robust to training model variation, potentially explaining the consistency described above (Supp.
-figure XX).
+figure 2).
 
 When testing model stability under a smaller sample size, we found reasonably good convergence in a dataset containing half the mass features with most model parameters falling within two standard errors of the estimate for the XCMS and two-parameter model, while the full model required closer to 80% of the mass features to produce estimates consistent with the original model (Supp.
-figure XX).
+figure 3).
 
@@
 All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GFF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent false negative acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp.
-figure XX).
+figure 4).
 The random forest model, interestingly, had perfect predictive capacity when tested internally on the training data (FDR=0%, GFF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp.
-figure XX).
+figure 4).
 In each case, the performance drop when applied to a novel dataset was more extreme than the simple two-parameter model described above.
@@
 Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the DCM; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}
-
-[Switch the shapes to make empty circles "good peaks missed"]
 
@@
 The logistic regression in particular was favored over the random forest and regularized regression we tried due to their similar performance and increased interpretability (Supp.
-figure XX).
+figure 4).
 
@@
 The class imbalance, with mostly poor quality MFs, is partially why we chose to measure precision and recall instead of total accuracy.
-However, precision and recall can also be ambiguous when the positive class is not specified and the raw confusion matrices are unavailable, thus our very precise use of the FDR and GFF metrics as well as providing the confusion matrices in Supplemental Table XX.
+However, precision and recall can also be ambiguous when the positive class is not specified and the raw confusion matrices are unavailable, thus our very precise use of the FDR and GFF metrics as well as providing the confusion matrices in Supplemental Table 2.
 @Melnikov2020 reported precision and recall in their presentation of `peakonly`, relative to which we obtained higher accuracy (they report 89% accuracy) but worse GFF and FDR (89% and 3%, respectively, relative to our 77.1% GFF and 19.6% FDR overall).
@@
 Similarly surprising was the lack of power in the design-of-experiments metrics, although this was less surprising given the number of missing values that were later filled in with an order of magnitude outside the most extreme value (Supp.
-figure XX).
+figure 1).
 
@@
 
-A final benefit to the reduced model is the smaller training set required to reach stability (Supp figure XX).
+A final benefit to the reduced model is the smaller training set required to reach stability (Supp.
+figure 3).
 This reduced size means that a useful model could be trained using only a fraction of the MFs identified in a sample set and then used to predict the quality for the remainder of the features.
@@
 The signal-to-noise ratio (SNR) was calculated by dividing the maximum original peak height by the standard deviation of the residuals multiplied by the maximum height of the original peak.
-This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional molecular signals [@Myers2017b].
+This method of SNR calculation allowed us to rapidly estimate the noise within the peak itself rather than relying on background estimation using data points outside the peak, which may not exist or may be influenced by additional mass signals [@Myers2017b].
 If there were fewer than 5 data points, a missing value was returned and dropped in subsequent summary calculations.
@@
 ### Authors' contributions
+
+WK performed the analyses and wrote the manuscript.
+BH helped design the metrics and implement the regressions as well as providing support and context for the analysis.
+AI provided funding and support and helped interpret the conclusions.
+
 ### Acknowledgements
+
+The authors would like to acknowledge
+
+[TBD]
 
@@
 
-*Supplemental figure XX: Distribution and single-parameter logistic curves for each metric extracted for model training, shown separately for the MESOSCOPE and Falkor datasets.
+*Supplemental figure 1: Distribution and single-parameter logistic curves for each metric extracted for model training, shown separately for the MESOSCOPE and Falkor datasets.
 Histograms show the distribution of good and bad mass features by color across the span of the data on the x-axis with the number of MFs in each bin shown on the y-axis.
@@
 
-![Supplemental figure XX: Model parameter estimates for each of the metrics in the full model, additionally broken down by their inclusion in the two-parameter (raw_data) and XCMS-exclusively models.
+![Supplemental figure 2: Model parameter estimates for each of the metrics in the full model, additionally broken down by their inclusion in the two-parameter (raw_data) and XCMS-exclusively models.
 Colors correspond to the dataset used to train the logistic regression model, with "both" indicating a combined model using all manually-labeled features across both datasets.](supplement/suppfig_model_param_estimates.png){width=90%}
@@
 
-![Supplemental figure XX: Robustness of the two most significant metrics across the full (all_params), XCMS-only (xcms_params), and two-parameter (raw_data_params) models.
+![Supplemental figure 3: Robustness of the two most significant metrics across the full (all_params), XCMS-only (xcms_params), and two-parameter (raw_data_params) models.
 The x-axis corresponds to the fraction of the data used to train the model and the y-coordinate shows the estimated value for the specified term in the subset across 10-fold replicated subsampling.
@@
 
-![Supplemental figure XX: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.](supplement/regregrf_fdrgff_cross_train.png){width=90%}
+![Supplemental figure 4: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.](supplement/regregrf_fdrgff_cross_train.png){width=90%}
 
@@
 bind_rows(fdr_gff_individual, fdr_gff_full_model, fdr_gff_stricter_thresh) %>%
-  write_csv("manuscript/supplement/model_confusion_matrices.csv")
-```
+  write_csv("manuscript/supplement/supp_table_2_model_confusion_matrices.csv")
+```

### 2023-07-18 MW_upload/MW_formatting.R
Adding files for upload to MW

@@ -1,0 +1,109 @@
+
+library(tidyverse)
+filled_file_metadata <- read_csv("MW_upload/filled_file_metadata.csv")
+untarg_peaks <- read_csv("MW_upload/untarg_peaks.csv")
+untarg_feats <- read_csv("MW_upload/untarg_feats.csv")
+targ_peaks <- read_csv("MW_upload/targ_peaks.csv")
+
+## Falkor ----
+### Metadata
+filled_file_metadata %>%
+  filter(cruise=="FK") %>%
+  filter(polarity=="pos") %>%
+  mutate(Sample_ID=filename) %>%
+  mutate(Raw_file_name=filename) %>%
+  select(Raw_file_name, Sample_ID, station, cast, depth, abs_depth,
+         time, lat, lon, temp, sal, sla_corr, PC_um) %>%
+  write.table("MW_upload/falkor_metadata.txt", sep = "\t", row.names = FALSE)
+
+### Untargeted data
+untarg_peaks %>%
+  filter(cruise=="FK") %>%
+  select(-cruise, -polarity) %>%
+  mutate(norm_area=ifelse(!is.finite(norm_area), NA, norm_area)) %>%
+  pivot_wider(names_from = filename, values_from = norm_area) %>%
+  left_join(all_feats %>% select(feature, mzmed, rtmed)) %>%
+  mutate(mz_rt=paste0(round(mzmed, 5), "_", round(rtmed, 2))) %>%
+  select(mz_rt, starts_with("190715")) %>%
+  write.table("MW_upload/falkor_untargeted.txt", sep = "\t", row.names = FALSE)
+
+### Targeted data
+metabs <- unique(targ_peaks$compound_name)
+metabs_enc <- URLencode(paste(metabs, collapse = "\n"))
+metab_url <- "https://www.metabolomicsworkbench.org/databases/refmet/name_to_refmet.php"
+metab_body <- paste0("metabolite_name=", metabs_enc, "&submit=Submit")
+heads <- c(`Content-Type`= 'application/x-www-form-urlencoded')
+stan_names <- httr::POST(metab_url, body = metab_body, config = httr::add_headers(heads)) %>%
+  httr::content() %>%
+  rvest::html_table() %>%
+  `[[`(2)
+
+targ_peaks %>%
+  filter(str_detect(filename, "190715")) %>%
[…]

### 2023-07-21 MW_upload/MW_formatting.R
Adding all files uploaded to MW

@@ -4,30 +4,8 @@ filled_file_metadata <- read_csv("MW_upload/filled_file_metadata.csv")
 untarg_peaks <- read_csv("MW_upload/untarg_peaks.csv")
 untarg_feats <- read_csv("MW_upload/untarg_feats.csv")
 targ_peaks <- read_csv("MW_upload/targ_peaks.csv")
+clean_stans <- read_csv("MW_upload/clean_stans.csv")
 
-## Falkor ----
-### Metadata
-filled_file_metadata %>%
-  filter(cruise=="FK") %>%
-  filter(polarity=="pos") %>%
-  mutate(Sample_ID=filename) %>%
-  mutate(Raw_file_name=filename) %>%
-  select(Raw_file_name, Sample_ID, station, cast, depth, abs_depth,
-         time, lat, lon, temp, sal, sla_corr, PC_um) %>%
-  write.table("MW_upload/falkor_metadata.txt", sep = "\t", row.names = FALSE)
-
-### Untargeted data
-untarg_peaks %>%
-  filter(cruise=="FK") %>%
-  select(-cruise, -polarity) %>%
-  mutate(norm_area=ifelse(!is.finite(norm_area), NA, norm_area)) %>%
-  pivot_wider(names_from = filename, values_from = norm_area) %>%
-  left_join(all_feats %>% select(feature, mzmed, rtmed)) %>%
-  mutate(mz_rt=paste0(round(mzmed, 5), "_", round(rtmed, 2))) %>%
-  select(mz_rt, starts_with("190715")) %>%
-  write.table("MW_upload/falkor_untargeted.txt", sep = "\t", row.names = FALSE)
-
-### Targeted data
 metabs <- unique(targ_peaks$compound_name)
 metabs_enc <- URLencode(paste(metabs, collapse = "\n"))
 metab_url <- "https://www.metabolomicsworkbench.org/databases/refmet/name_to_refmet.php"
@@ -38,23 +16,268 @@ stan_names <- httr::POST(metab_url, body = metab_body, config = httr::add_header
   rvest::html_table() %>%
   `[[`(2)
 
+
+## Falkor ----
+### Metadata
+filled_file_metadata %>%
+  filter(cruise=="FK") %>%
+  filter(polarity=="pos") %>%
+  mutate(Sample_ID=filename) %>%
+  mutate(Raw_file_name=filename) %>%
+  mutate(samp_type=case_when(
+    samp_type=="Smp"~"Sample",
+    samp_type=="Std"~"Standard mix",
[…]

### 2023-07-21 manuscript/manuscript.Rmd
Continuing use of long parameter names instead of shorthand

@@
                 "iso_area_cor")
+param_longnames <- c(
+  mean_mz = "Mean feature m/z ratio",
+  sd_ppm = "SD of feature m/z ratios in ppm",
+  mean_rt = "Mean feature retention time",
+  sd_rt = "SD of feature retention time",
+  mean_pw = "Mean peak width, in seconds",
+  sd_pw = "SD of peak widths, in seconds",
+  log_mean_area = "Log-10 scaled mean feature area",
+  log_sd_area = "Log-10 scaled SD of feature areas",
+  sn = "XCMS 'sn' metric",
+  f = "XCMS 'f' metric",
+  scale = "XCMS 'scale' metric",
+  lmin = "XCMS 'lmin' metric",
+  feat_npeaks = "Average number of peaks detected per file",
+  n_found = "Proportion of total files in which a peak was found",
+  samps_found = "Proportion of total samples in which a peak was found",
+  stans_found = "Proportion of total standards in which a peak was found",
+  pooled_cv = "Coefficient of variance among pooled sample replicate peak areas",
+  pooled_cv_rob = "Robust CV among pooled sample peak areas\n(median absolute deviation divided by median)",
+  med_SNR = "Median novel SNR metric across a feature",
+  med_cor = "Median peak shape metric across a feature",
+  max_SNR = "Maximum novel SNR metric",
+  max_cor = "Maximum peak shape metric",
+  medtop3_cor = "Median of 3 highest peak shape values",
+  log_med_cor = "Median log-10 scaled (1-peak shape)",
+  med_missed_scans = "Median number of scans with no data",
+  smp_to_blk = "Ratio of average sample peak area to peak area in blank",
+  smp_to_std = "Ratio of average sample peak area to peak area in standards",
+  iso_shape_cor = "Median log-10 scaled 1-correlation of 13C isotope EIC trace to monoisotopic EIC trace",
+  iso_area_cor = "Correlation between 13C isotope area and monoisotopic peak area"
+)
 ```
@@
 glmnet_estimates <- both_datasets %>% 
-  select(all_of(all_params)) %>% 
+  select(all_of(all_params), c("max_SNR", "max_cor", "medtop3_cor", "log_med_cor")) %>% 
   data.matrix() %>%
@@
 rf_estimates <- both_datasets %>% 
-  select(all_of(all_params)) %>%
+  select(all_of(all_params), c("max_SNR", "max_cor", "medtop3_cor", "log_med_cor")) %>% 
   randomForest::randomForest(factor(both_datasets$feat_bin), importance=TRUE) %>%
@@
 
-glm_estimates %>%
+param_longnames %>%
+  as.data.frame() %>%
+  set_names("long_name") %>%
+  rownames_to_column("term") %>%
+  left_join(glm_estimates, by="term") %>%
   left_join(glmnet_estimates, by="term") %>%
@@
 
-We explored the relative predictive power of the individual parameters using the full model and found that the predictors least likely to be different from zero due to chance were the mean *m/z* ratio, the novel peak shape correlation metric `med_cor`, and the novel SNR estimate `med_SNR`, all with reported p-values < 10$^{-10}$.
+We explored the relative predictive power of the individual parameters using the full model and found that the predictors least likely to be different from zero due to chance were the mean *m/z* ratio, our novel peak shape correlation metric, and our novel SNR estimate, all with reported p-values < 10$^{-10}$.
 The novel parameters were then validated using a random forest model that also found them to have the highest importance (Supp.
@@
   mutate(feat_bin=ifelse(feat_class=="Good", 1, 0))
-pdf(file = "manuscript/supplement/metric_distributions.pdf", width = 6.5, height = 4)
+
+pdf(file = "manuscript/supplement/supp_fig_1_metric_distributions.pdf", width = 6.5, height = 4)
 both_datasets %>%
@@
       scale_fill_manual(breaks = c("Bad", "Good"), values = c("#a41118", "#f4bb23")) +
-      labs(y="# of mass features", x=chosen_param) +
+      labs(y="# of mass features", x=param_longnames[chosen_param]) +
       theme_bw() +
-      theme(legend.direction = "horizontal", legend.position = "none") +
-      facet_wrap(~dataset, ncol=1, scales = "free_y")
+      theme(legend.direction = "horizontal", legend.position = "none",
+            axis.title.x = element_text(hjust=0)) +
+      facet_wrap(~dataset, ncol=1, scales = "free_y") +
+      coord_cartesian(clip = "off")
     log_gp <- both_datasets %>%
@@
       scale_y_continuous(breaks = c(0, 1), labels = c("Bad", "Good"), name=NULL) +
-      labs(x=chosen_param) +
+      labs(x=NULL) +
       theme_bw() +
@@
 ggsave(plot = suppfig_model_param_estimates_gp, 
-       filename = "suppfig_model_param_estimates.png", 
+       filename = "supp_fig_2_model_param_estimates.png", 
        device = "png", path = "manuscript/supplement",
@@
 ggsave(plot = sampled_model_est_gp, 
-       filename = "sampled_model_est.png", 
+       filename = "supp_fig_3_sampled_model_est.png", 
        device = "png", path = "manuscript/supplement",
@@
 ggsave(plot = fdr_gff_cross_train_gp, 
-       filename = "regregrf_fdrgff_cross_train.png", 
+       filename = "supp_fig_4_regregrf_fdrgff_cross_train.png", 
        device = "png", path = "manuscript/supplement",

### 2023-07-21 manuscript/manuscript.Rmd
Adding Bai2022 reference

@@
 This increased performance is likely due to our use of the metrics recalculated from the raw data, as their metrics were only calculated from the default MzMine2 peak parameters reported: similar to our XCMS-only model.
+Previous work on a "shape-orientated" algorithm also established the utility of testing the extracted ion chromatogram against a Gaussian shape [@Bai2022].
+There, the use of a Marr wavelet had GFF values in the 98-100% range but very high FDR values of 82-91%, representing a very lenient threshold much closer to the XCMS or ADAP defaults.
 

### 2023-07-21 manuscript/manuscript.Rmd
Adding Anitra and Bryna's comments to text

@@
 We used these datasets to construct a predictive model calibrated to the likelihood that visual inspection by an MS expert would include a given mass feature in the downstream analysis.
-We found that a simple logistic regression model built on two metrics calculated from the raw MS data reduced the fraction of false positives in the analysis from 70-80% down to 1-5% and showed minimal overfitting when applied to novel datasets.
+We implemented two novel peak quality metrics, a custom signal-to-noise metric and a test of similarity to a bell curve, both calculated from the raw data in the extracted ion chromatogram and found that these outperformed existing measurements of peak quality.
+A simple logistic regression model built on two metrics reduced the fraction of false positives in the analysis from 70-80% down to 1-5% and showed minimal overfitting when applied to novel datasets.
 We then explored the implications of this quality thresholding on the conclusions obtained by the downstream analysis and found that while only 10% of the variance in the dataset could be explained by depth in the default output from the peakpicker, approximately 40% of the variance was explained when restricted to high-quality peaks alone.
@@
 
-Noisy data and imperfect detection algorithms introduce a tradeoff between false positives (where background instrument or chemical noise is misclassified as biological signal) and false negatives (where real signals are undetected).
-Existing algorithms tend to favor false positives because downstream analyses can always remove erroneous mass features, but false negatives cannot be later recovered [@Pirttila2022; @Gloaguen2022].
-However, this approach creates two downstream problems.
-First, the abundance of false positives requires more time from the researcher as they manually evaluate a potentially enormous number of mass features (MFs), a task that scales combinatorially with the number of samples and compounds measured [@Myers2017a].
-Second, false negatives cannot be fully accounted for in the way false positives can because false negatives are introduced in every step of metabolomic analyses from sample collection through injection.
-Instead of minimizing false negatives, we believe that emphasis should be placed on allowing the experimenter to set a threshold for the false discovery rate and accept that this will inherently add to the number of MFs already lost in the data collection process.
-
-Existing peak-detection softwares do not provide a clear way to do this.
+Noisy data and imperfect detection algorithms introduce a tradeoff between false positives (where contamination, background instrument or chemical noise is misclassified as biological signal) and false negatives (where real signals are undetected).
+Existing algorithms tend to favor the inclusion of false positives because downstream analyses can always remove erroneous mass features, but false negatives cannot be later recovered [@Pirttila2022; @Gloaguen2022].
+However, this approach requires more time from the researcher as they manually evaluate a potentially enormous number of mass features (MFs), a task that scales combinatorially with the number of samples and compounds measured [@Myers2017a].
+Instead of minimizing false negatives, we believe that emphasis should be placed on allowing the experimenter to set a threshold for the proportion of false positives (the false discovery rate or FDR) and accept that this will inherently add to the number of MFs already lost in the data collection process.
+
+Existing peak-detection softwares do not provide a clear way to exclude false positives in an a consistent and unbiased way.
 Typical outputs consistent across the different implementations consist of the *m/z* ratio, retention time, and area for each mass feature, with some additional useful information occasionally provided such as the peak's signal-to-noise ratio or degree of skew [@Pirttila2022].
@@
 This parameter is crucial for downstream analysis because it represents the base rate for error propagation and acceptable thresholds should vary widely by the particular project's goals.
-In an exploratory analysis, any mass feature more than 50% likely to be real is perhaps worth considering, while in a confirmatory study this threshold may need to be above 99%.
+In an exploratory analysis, any mass feature more than 50% likely to be real is perhaps worth considering, while in a confirmatory study this threshold may need to be above 99% likely to be real.
 Despite significant effort invested in improving the peakpicking algorithms, very little has been done to quantify the accuracy and precision of their outputs across the wide variety of datasets to which they are applied.
@@
 
-An area particularly ripe for metabolomic analysis is that of the open ocean [@KidoSoule2015].
-Low compound and high salt concentrations make this area difficult to study but its vast size and direct effect on the Earth's biogeochemistry make it critical that we understand its dynamics of energy and nutrients on a molecular scale [@Boysen2018].
+An area particularly ripe for improved tools for metabolomic data analysis is that of the open ocean [@KidoSoule2015].
+Low compound and high salt concentrations make metabolomics analyses difficult to study in this area but its vast size and the direct effect of its microbial communities on the Earth's biogeochemistry make it critical that we understand the transformation of energy and nutrients on a molecular scale [@Boysen2018].
 Metabolites are the currency of chemical exchange both intra- and inter-cellularly, serving as building blocks of larger molecules, regulators of osmotic balance and storage of nutrients, as well as important chemical signals on their own.
@@
 In this paper, we use open ocean marine metabolite LC-MS samples to develop and test a variety of chromatographic peak metrics.
-Here, we construct and validate multiple predictive models of MF quality based on metrics both common in the literature and custom implementations we've found useful in our own analysis.
+We construct and validate multiple predictive models of MF quality based on metrics both common in the literature and custom implementations we've found useful in our own analysis.
 This allows us to connect the physical, chemical, and biological measurements taken regularly around the globe to a molecular-scale perspective of particulate organic matter in the ocean by linking the chemical currencies that fuel the planet to the environments in which they're found.
@@
 Most metrics had reasonably normal distributions after the scaling and normalization described in Methods.
-Visually, the most compelling separations between good and bad MFs were observed in the novel peak shape and SNR metrics, with higher peak shape correlations and larger SNR values corresponding to good MFs and almost complete separation for the peak shape (`med_cor`) metric.
+Visually, the most compelling separations between good and bad MFs were observed in our peak shape and novel SNR metrics, with almost complete separation between good and bad peaks provided by the new peak shape metric alone.
 Peak width and its standard deviation also showed reasonable separation between good and bad MFs (good MFs tended to have low SDs and larger peak widths).
@@
 According to all three logistic regression models (see Methods), the majority of MFs were estimated to have a less than 1% chance of being good.
-The full model (containing all possible peak metrics) and the XCMS model (built on only those metrics calculated from the XCMS output) both displayed a strongly bimodal distribution, with a large number of MFs also exceeding a 99% chance of being good, while the two-parameter model (consisting of the novel SNR metric and the peak shape correlation metric) had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Figure 1).
+The full model (containing all evaluated peak metrics) and the XCMS model (built on only those metrics calculated from the XCMS output) both displayed a strongly bimodal distribution, with a large number of MFs also exceeding a 99% chance of being good, while the two-parameter model (consisting of the novel SNR metric and the peak shape correlation metric) had a flatter distribution with fewer high-confidence MF assignments and more intermediate values (Figure 1).
 
@@
 We explored the relative predictive power of the individual parameters using the full model and found that the predictors least likely to be different from zero due to chance were the mean *m/z* ratio, our novel peak shape correlation metric, and our novel SNR estimate, all with reported p-values < 10$^{-10}$.
-The novel parameters were then validated using a random forest model that also found them to have the highest importance (Supp.
+The value of the novel parameters was then validated using a random forest model that also found them to have the highest importance (Supp.
 table 1).
 
-The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR values in the 5-10% range and 80-90% GFF values implying that a large majority of the good MFs passed the threshold with very little noise included.
+The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR (false discovery rate, defined as the number of false positives divided by the total number of positive predictions) values in the 5-10% range and 80-90% GFF (% good features found, defined as the number of true positives divided by the total number of features manually classifed as good) values implying that a large majority of the good MFs passed the threshold with very little noise included.
 The XCMS metrics performed slightly worse, with FDR values in the 10-15% range and GFF values closer to 75%.
@@
 Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel.
-Lines of best fit have been estimated and plotted in black behind the data points, with the steeper slopes found in the full and XCMS-only models indicating excessive overfitting on the training data.](figures/fdr_gff_cross_train.png){width=90%}
+Lines of best fit have been estimated and plotted in black behind the data points, with the steeper slopes found in the full and XCMS-only models indicating overfitting on the training data.](figures/fdr_gff_cross_train.png){width=90%}
 
@@
     which_params, levels=c("all_params", "xcms_params", "raw_data_params"),
-    labels=c("All metrics", "XCMS-only metrics", "Raw data metrics")))
+    labels=c("Full model", "XCMS-only model", "Two-parameter model")))
 
@@
     which_params, levels=c("all_params", "xcms_params", "raw_data_params"),
-    labels=c("All metrics", "XCMS-only metrics", "Raw data metrics"))) %>%
+    labels=c("Full model", "XCMS-only model", "Two-parameter model"))) %>%
   ggplot() +
@@
         # strip.text = element_blank(),
-        strip.background = element_rect(fill="white"),
         legend.position=c(0.925, 0.25),
@@
 None of the penalized regression models significantly improved cross-validated performance between the MESOSCOPE and Falkor datasets when measured by both initial performance and the performance drop when applied across datasets.
-All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GFF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent false negative acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp.
+All three regularized regression models had similar behavior, with ridge regression (α = 0) obtaining the lowest rates for both GFF and FDR, while lasso (α = 1) obtained higher rates for both and represented a less-stringent false negative acceptance.
+As expected, the elastic net (α = 0.5) fell in between the two (Supp.
 figure 4).
@@
 With the stricter threshold, we obtained FDR values consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for Falkor, MESOSCOPE, CultureData, and Pttime respectively (Figure 4).
-Of course, this low error rate required that we miss out on additional potentially valuable features, with only a fraction of the total good MFs making it past this threshold.
+Of course, this low error rate meant that we miss out on additional potentially valuable features, with only a fraction of the total good MFs making it past this threshold.
 In both the Falkor and MESOSCOPE datasets, less than half of the good MFs were labeled as such, with actual values of 39.4% and 26.5%, respectively.
@@
 The largest category that did have a trend with depth was the 15m = DCM > 175m category, containing 118 mass features, with largest peak areas distributed evenly between the 15 meter and deep chlorophyll maximum (DCM) samples, while the 175 meter samples had significantly smaller areas.
-The similar but statistically distinct categories of 15m > DCM > 175m and DCM > 15m > 175m had 68 and 35 features, respectively, and together indicate that many molecules are highly abundant throughout the surface ocean down to the DCM layer and decrease in concentration thereafter.
+The similar but statistically distinct categories of 15m > DCM > 175m and DCM > 15m > 175m had 68 and 35 features, respectively, and together indicate that many molecules are highly abundant throughout the surface ocean down to the DCM layer and decrease in concentration at 175 meters.
 A surprising number of features were also found to have DCM minima (DCM < 15m = 175m, 26 features) or linear increases with depth (15m < DCM < 175m, 12 features) given the few environmental parameters that have these trends (Figure 7).
@@
 Two features of particular interest were those good MFs with a DCM minimum (DCM < 15m = 175m), both of which were missed by the two-parameter model.
-These features possess an unexpected biological signal that does not track with depth or other common oceanographic parameters, thereby potentially representing an interesting biomarker or pathway that's downregulated in excess of the biomass effect.
+These features possess an unexpected biological signal that does not track with depth or other common oceanographic parameters, thereby potentially representing an interesting biomarker that decreases despite an increase in biomass.
 
@@
 Good MF found = true positive, good MF missed = false negative, bad MF found = false positive, and bad MF avoided = true negative according to a 0.5 likelihood threshold.
-Note that the majority of the features in the bad MF avoided category fell into the "All equal" depth class (1365 MFs) but the x-axis has been truncated at 60 for ease of visualization.
+Note that the majority of the features in the bad MF avoided category fell into the "All equal" depth class for which there was no significant differences between the depths (1365 MFs) but the x-axis has been truncated at 60 for ease of visualization.
 The boxplots in the bottom illustrate the depth response type for 4 specific categories, with raw peak area plotted on a log scale against the sample depth (DCM = deep chlorophyll maximum, ~110 meters).
@@
 
-In each test, the permutational p-value obtained was less than 0.001, indicating that the differences between samples due to depth were never likely to be due entirely to chance.
+In each test, the permutational p-value obtained was less than 0.001, indicating that the differences between samples due to depth were unlikely to be due to chance.
 However, the pseudo-F and was much larger with higher thresholds, scaling from around 8.5 when thresholding at a 1% likelihood to 42 when thresholded at a 90% likelihood (Table 1).
@@
 The two metrics in the final model were rederived from the raw EIC data because they matched our intuition about what makes an MF look good to an MS expert.
-However, they could be implemented in XCMS directly, removing the need to revisit the raw data.
-It is worth noting that these are very simple metrics and therefore fast to calculate, but we expect that more complicated metrics could perform even better.
+These are very simple metrics and therefore fast to calculate, but we expect that more complicated metrics could perform even better.
 For example, the method of using the data within the peak boundaries for SNR calculation rather than data outside of them is not known to the authors to be implemented elsewhere but could be further improved by more advanced smoothing methods rather than using the residuals directly.
@@
 Also worth noting is that the calculation of any new metrics such as these that rely on access to the raw data require exact specification of the maximum and minimum *m/z* and retention time for a peak, values that are not always returned by peakpicking algorithms and must be recalculated, as in @Kantz2019.
+To avoid the additional overhead of recalculation and the possiblity of raw data unavailability, we have implemented these metrics during the initial peakpicking step of XCMS in a fork of the GitHub available at https://github.com/wkumler/xcms and have submitted a pull request to implement them into XCMS directly.
 
@@
 The isotope information in particular was expected to be a very strong predictor of MF quality given previous work that uses this metric extensively [@Libiseller2015; @Treutler2016; @ElAbiead2021].
-We learned that many noise MFs still have reliable isotopes (perhaps unsurprising, given that the noise is in fact often caused by solvents or contaminants that are still chemical in nature) and that many real MFs are simply too low-intensity to have them in this kind of dilute environmental sample.
+We learned that many noise MFs still have reliable isotopes (perhaps unsurprising, given that the noise is in fact often caused by solvents or contaminants that are still chemical in nature) and that many real MFs are simply too small (low-intensity) to have detectable isotope peaks in this kind of dilute environmental sample.
 
@@
 The large number of mass features due to noise present in metabolomics datasets can be controlled using a simple logistic classification model.
-We trained such a model on two full-labeled open ocean HILIC datasets and found that the best performing parameters in the model for were a custom signal-to-noise metric and a test of similarity to a bell curve.
+We trained such a model on two full-labeled open ocean HILIC datasets and found that the best performing parameters in the model were a custom signal-to-noise metric and a test of similarity to a bell curve.
 This model showed robustness to overfitting, independence from the training set, and a reduced degree of manual labeling required.
@@
 DCM = deep chlorophyll maximum, approximately 110 meters.
-Mass feature ids are provided as the title of each panel, starting with "FT" and followed by 4 digits except for the two features annotated using authentic standards run alongside: the 13C isotope of dimethylsulfonioacetate (DMS-Ac) and taurine.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
+Mass feature identifications are provided as the title of each panel, starting with "FT" and followed by 4 digits except for the two features annotated using authentic standards run alongside: the 13C isotope of dimethylsulfonioacetate (DMS-Ac) and taurine.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
 
@@
 
-Distributions were visually inspected using a pairs plot and highly correlated (above a Pearson's r ~ 0.9) had one of the correlated metrics removed.
+Distributions were visually inspected using a pairs plot and highly correlated (above a Pearson's r ~ 0.9) metrics had one of the redundant metrics removed.
 
@@
 This project dataset consists of *Phaeodactylum tricornutum* cultures collected at a variety of timepoints from both pelleted cells and the released exudate.
-This dataset was chosen because of the similar LC-MS setup used as a benchmark for the performance other labs with similar setups may expect to achieve using the trained model directly.
+This dataset was chosen because of the similar LC-MS setup used as a benchmark for the performance that other labs with similar setups may expect to achieve using the trained model directly.
 
@@
 The raw mzML files are all available on Metabolomics Workbench.
-Falkor, MESOSCOPE, and CultureData can be found under Project ID [blah] and accessed directly using the Project ID http://dx.doi.org/[blah], while Pttime is located under Project ID PR001317 and can be accessed directly using its Project DOI: http://dx.doi.org/10.21228/M8GH6P.
+The Falkor and MESOSCOPE datasets can be found under project ID PR001738 via http://dx.doi.org/10.21228/M82719.
+The CultureData samples were appended to the previously existing culturing collection, accessible at project ID PR001021 via http://dx.doi.org/10.21228/M8QM5H.
+Pttime is located under Project ID PR001317 and can also be accessed directly using its Project DOI: http://dx.doi.org/10.21228/M8GH6P.
 Code and other raw data is available on the GitHub repository at https://github.com/wkumler/MS_metrics.
@@
 
-[TBD, get from Anitra. eScience?]
+This work was supported by grants from the Simons Foundation (SCOPE Award ID 329108 to AI, SF Award ID 385428 to AI).
 
@@
 
-WK performed the analyses and wrote the manuscript.
+WK extracted the Falkor samples, processed the data, performed the analyses, and wrote the manuscript.
 BH helped design the metrics and implement the regressions as well as providing support and context for the analysis.
-AI provided funding and support and helped interpret the conclusions.
+AI provided funding and data and helped to interpret the conclusions and edit the manuscript.
 
@@
 
-The authors would like to acknowledge
-
-[TBD]
+This work was supported by the University of Washington eScience Institute through their Data Science Incubator program.
+The authors would also like to acknowledge Laura Carlson for her expertise in obtaining the CultureData, MESOSCOPE, and Falkor datasets; Katherine Heal and Bryn Durham for their assistance with culturing and collecting the cultured samples; the captain and crew of the R/Vs *Kilo Moana* and *Falkor*, Wei Qin, Rachel Lundeen, and the SCOPE ops team for collecting and processing the environmental samples; Joshua Sacks, Dave Beck, and the entire eScience incubator team for their feedback on the project development and scope; and Brisson Vanessa and LLNL for making their Pttime dataset available for reuse on Metabolomics Workbench.
+Additionally, the R packages plotly, dbscan, ClusterR, RaMS, and the entire tidyverse were crucial for preliminary data exploration.
 

### 2023-07-21 manuscript/manuscript.Rmd
Changing model names in figure, homogenizing

@@
     levels=c("all_params", "xcms_params", "raw_data_params"),
-    labels=c("All metrics", "XCMS-only metrics", "Raw data metrics"))) %>%
+    labels=c("Full model", "XCMS-only model", "Two-parameter model"))) %>%
   filter(feat_class%in%c("Good", "Bad")) %>%

### 2023-07-21 manuscript/manuscript.Rmd
Renaming figures with numbers for upload

@@
 ggsave(plot = egg_complete, 
-       filename = "manuscript/figures/pred_prob_class_color_hists.png", 
+       filename = "manuscript/figures/fig_1_pred_prob_class_color_hists.png", 
        device = "png",
@@
 ggsave(plot = fdr_gff_cross_train_gp, 
-       filename = "fdr_gff_cross_train.png", 
+       filename = "fig_2_fdr_gff_cross_train.png", 
        device = "png", path = "manuscript/figures",
@@
 ggsave(plot = comb_prob_space_gp, 
-       filename = "comb_prob_space.png", 
+       filename = "fig_3_comb_prob_space.png", 
        device = "png", path = "manuscript/figures",
@@
 ggsave(plot = fdr_gff_all_datasets_gp, 
-       filename = "fdr_gff_all_datasets.png", 
+       filename = "fig_4_fdr_gff_all_datasets.png", 
        device = "png", path = "manuscript/figures",
@@
 ggsave(plot = clockplot_complete_gp, 
-       filename = "clockplot_complete.png", 
+       filename = "fig_5_clockplot_complete.png", 
        device = "png", path = "manuscript/figures",
@@
   theme(aspect.ratio = 1)
-ggsave(plot = thresh_NMDS_gp, filename = "thresh_NMDS_plots.png", 
+ggsave(plot = thresh_NMDS_gp, filename = "fig_6_thresh_NMDS_plots.png", 
        device = "png", path = "manuscript/figures",
@@
 ggsave(plot = comb_cowplot, 
-       filename = "somegood_somebad_depthcolor_chroms.png", 
+       filename = "fig_7_somegood_somebad_depthcolor_chroms.png", 
        device = "png", path = "manuscript/figures",
@@
 ggsave(plot = comb_peakmetrics_gp, 
-       filename = "peakmetrics_singlechrom.png", 
+       filename = "fig_8_peakmetrics_singlechrom.png", 
        device = "png", path = "manuscript/figures",

### 2023-07-22 manuscript/manuscript.Rmd
Switching to metric for figure sizes

--- 2023-07-21 manuscript/manuscript.Rmd
@@
 ggsave(plot = egg_complete, 
-       filename = "manuscript/figures/pred_prob_class_color_hists.png", 
+       filename = "manuscript/figures/fig_1_pred_prob_class_color_hists.png", 
        device = "png",
-       width = 6.5, height = 4, units = "in", dpi = 300)
+       width = 170, height = round((4/6.5)*170), units = "mm", dpi = 300)
 ```
@@
 Colors indicate the category in which each feature was manually assigned by an expert, with "Stans only" referring to a good mass feature that was only visible in the standards run alongside the samples.
-Culture datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (dotted black vertical line) according to the final model and were otherwise unclassified.](figures/pred_prob_class_color_hists.png){width=90%}
+Culture datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (dotted black vertical line) according to the final model and were otherwise unclassified.](figures/fig_1_pred_prob_class_color_hists.png){width=90%}
 
@@
 ggsave(plot = fdr_gff_cross_train_gp, 
-       filename = "fdr_gff_cross_train.png", 
+       filename = "fig_2_fdr_gff_cross_train.png", 
        device = "png", path = "manuscript/figures",
-       width = 6.5, height = 3, units = "in", dpi = 300)
+       width = 170, height = round((3/6.5)*170), units = "mm", dpi = 300)
 ```
@@
 Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel.
-Lines of best fit have been estimated and plotted in black behind the data points, with the steeper slopes found in the full and XCMS-only models indicating overfitting on the training data.](figures/fdr_gff_cross_train.png){width=90%}
+Lines of best fit have been estimated and plotted in black behind the data points, with the steeper slopes found in the full and XCMS-only models indicating overfitting on the training data.](figures/fig_2_fdr_gff_cross_train.png){width=90%}
 
@@
 ggsave(plot = comb_prob_space_gp, 
-       filename = "comb_prob_space.png", 
+       filename = "fig_3_comb_prob_space.png", 
        device = "png", path = "manuscript/figures",
-       width = 6.5, height = 5, units = "in", dpi = 300)
+       width = 170, height = round((5/6.5)*170), units = "mm", dpi = 300)
 ```
@@
 The top row of plots show the exact likelihood predicted by the logistic model across three different subsets of parameters, while the bottom row shows the estimates ranked from least likely to most likely.
-Points are colored by their manually-assigned quality according to an expert.](figures/comb_prob_space.png){width=90%}
+Points are colored by their manually-assigned quality according to an expert.](figures/fig_3_comb_prob_space.png){width=90%}
 
@@
 ggsave(plot = fdr_gff_all_datasets_gp, 
-       filename = "fdr_gff_all_datasets.png", 
+       filename = "fig_4_fdr_gff_all_datasets.png", 
        device = "png", path = "manuscript/figures",
-       width = 3, height = 4, units = "in", dpi = 300)
+       width = 85, height = round((4/6.5)*170), units = "mm", dpi = 300)
 ```
@@
 FDR is calculated by dividing the number of false positives by the total positives produced by the model and GFF is calculated by dividing the number of true positives by the total number of good features as identified manually (only possible for fully-labeled datasets).
-Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fdr_gff_all_datasets.png){width=30%}
+Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fig_4_fdr_gff_all_datasets.png){width=30%}
 
@@
 ggsave(plot = clockplot_complete_gp, 
-       filename = "clockplot_complete.png", 
+       filename = "fig_5_clockplot_complete.png", 
        device = "png", path = "manuscript/figures",
-       width = 6.5, height = 7, units = "in", dpi = 300)
+       width = 170, height = round((7/6.5)*170), units = "mm", dpi = 300)
 ```
@@
 Each mass feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile.
-Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the DCM; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}
+Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the DCM; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/fig_5_clockplot_complete.png){width=90%}
 
@@
   theme(aspect.ratio = 1)
-ggsave(plot = thresh_NMDS_gp, filename = "thresh_NMDS_plots.png", 
+ggsave(plot = thresh_NMDS_gp, filename = "fig_6_thresh_NMDS_plots.png", 
        device = "png", path = "manuscript/figures",
-       width = 6.5, height = 4, units = "in", dpi = 300)
+       width = 170, height = round((4/6.5)*170), units = "mm", dpi = 300)
 ```
@@
 "Only good MFs" refers to those features manually labeled as "Good".
-NMDS stress values are reported in the upper right corner of each plot.](figures/thresh_NMDS_plots.png){width=90%}
+NMDS stress values are reported in the upper right corner of each plot.](figures/fig_6_thresh_NMDS_plots.png){width=90%}
 
@@
 ggsave(plot = comb_cowplot, 
-       filename = "somegood_somebad_depthcolor_chroms.png", 
+       filename = "fig_7_somegood_somebad_depthcolor_chroms.png", 
        device = "png", path = "manuscript/figures",
-       width = 6.5, height = 3, units = "in", dpi = 300)
+       width = 170, height = round((3/6.5)*170), units = "mm", dpi = 300)
 ```
@@
 DCM = deep chlorophyll maximum, approximately 110 meters.
-Mass feature identifications are provided as the title of each panel, starting with "FT" and followed by 4 digits except for the two features annotated using authentic standards run alongside: the 13C isotope of dimethylsulfonioacetate (DMS-Ac) and taurine.](figures/somegood_somebad_depthcolor_chroms.png){width=70%}
+Mass feature identifications are provided as the title of each panel, starting with "FT" and followed by 4 digits except for the two features annotated using authentic standards run alongside: the 13C isotope of dimethylsulfonioacetate (DMS-Ac) and taurine.](figures/fig_7_somegood_somebad_depthcolor_chroms.png){width=70%}
 
@@
 ggsave(plot = comb_peakmetrics_gp, 
-       filename = "peakmetrics_singlechrom.png", 
+       filename = "fig_8_peakmetrics_singlechrom.png", 
        device = "png", path = "manuscript/figures",
-       width = 6.5, height = 4, units = "in", dpi = 300)
+       width = 170, height = round((4/6.5)*170), units = "mm", dpi = 300)
 ```
@@
 Normalization was performed by linearly scaling the raw values into the 0-1 range by subtracting the minimum value and dividing by the maximum.
-Peak shape similarity was measured with Pearson's correlation coefficient and the noise level is estimated as the standard deviation of the residuals after the raw data is subtracted from the idealized peak.](figures/peakmetrics_singlechrom.png){width=90%}
+Peak shape similarity was measured with Pearson's correlation coefficient and the noise level is estimated as the standard deviation of the residuals after the raw data is subtracted from the idealized peak.](figures/fig_8_peakmetrics_singlechrom.png){width=90%}
 
@@
 
-![Supplemental figure 2: Model parameter estimates for each of the metrics in the full model, additionally broken down by their inclusion in the two-parameter (raw_data) and XCMS-exclusively models.
-Colors correspond to the dataset used to train the logistic regression model, with "both" indicating a combined model using all manually-labeled features across both datasets.](supplement/suppfig_model_param_estimates.png){width=90%}
+*Supplemental figure 2: Model parameter estimates for each of the metrics in the full model, additionally broken down by their inclusion in the two-parameter (raw_data) and XCMS-exclusively models.
+Colors correspond to the dataset used to train the logistic regression model, with "both" indicating a combined model using all manually-labeled features across both datasets.*
 
@@
 
-![Supplemental figure 3: Robustness of the two most significant metrics across the full (all_params), XCMS-only (xcms_params), and two-parameter (raw_data_params) models.
+*Supplemental figure 3: Robustness of the two most significant metrics across the full (all_params), XCMS-only (xcms_params), and two-parameter (raw_data_params) models.
 The x-axis corresponds to the fraction of the data used to train the model and the y-coordinate shows the estimated value for the specified term in the subset across 10-fold replicated subsampling.
-The grey bar in the background corresponds to the estimate of the full model +/- 1SE (thinner dark grey bar) and 2SE (thicker light grey bar).](supplement/sampled_model_est.png){width=90%}
+The grey bar in the background corresponds to the estimate of the full model +/- 1SE (thinner dark grey bar) and 2SE (thicker light grey bar).*
 
@@
 
-![Supplemental figure 4: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.](supplement/regregrf_fdrgff_cross_train.png){width=90%}
+*Supplemental figure 4: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.*
 

### 2023-07-22 manuscript/manuscript.Rmd
Removing unnecessary header

@@
 author: "William Kumler, Bryna Hazelton, and Anitra Ingalls"
-date: "`r Sys.Date()`"
 output:
@@
     reference_docx: template.docx
-  pdf_document:
-    latex_engine: xelatex
 bibliography: refs.bib

### 2023-07-27 manuscript/manuscript.Rmd
Updating Anitra's initials and adding Angie to acknowledgements

@@
   metrics in metabolomics'
-author: "William Kumler, Bryna Hazelton, and Anitra Ingalls"
+author: "William Kumler, Bryna Hazelton, and Anitra E. Ingalls"
 output:
@@
 
-This work was supported by grants from the Simons Foundation (SCOPE Award ID 329108 to AI, SF Award ID 385428 to AI).
+This work was supported by grants from the Simons Foundation (SCOPE Award ID 329108 to AEI, SF Award ID 385428 to AEI).
 
@@
 BH helped design the metrics and implement the regressions as well as providing support and context for the analysis.
-AI provided funding and data and helped to interpret the conclusions and edit the manuscript.
+AEI provided funding and data and helped to interpret the conclusions and edit the manuscript.
 
@@
 This work was supported by the University of Washington eScience Institute through their Data Science Incubator program.
-The authors would also like to acknowledge Laura Carlson for her expertise in obtaining the CultureData, MESOSCOPE, and Falkor datasets; Katherine Heal and Bryn Durham for their assistance with culturing and collecting the cultured samples; the captain and crew of the R/Vs *Kilo Moana* and *Falkor*, Wei Qin, Rachel Lundeen, and the SCOPE ops team for collecting and processing the environmental samples; Joshua Sacks, Dave Beck, and the entire eScience incubator team for their feedback on the project development and scope; and Brisson Vanessa and LLNL for making their Pttime dataset available for reuse on Metabolomics Workbench.
+The authors would also like to acknowledge Laura Carlson for her expertise in obtaining the CultureData, MESOSCOPE, and Falkor datasets; Katherine Heal, Angie Boysen, and Bryn Durham for their assistance with culturing and collecting the cultured samples; the captain and crew of the R/Vs *Kilo Moana* and *Falkor*, Wei Qin, Rachel Lundeen, and the SCOPE ops team for collecting and processing the environmental samples; Joshua Sacks, Dave Beck, and the entire eScience incubator team for their feedback on the project development and scope; and Brisson Vanessa and LLNL for making their Pttime dataset available for reuse on Metabolomics Workbench.
 Additionally, the R packages plotly, dbscan, ClusterR, RaMS, and the entire tidyverse were crucial for preliminary data exploration.

### 2023-07-28 manuscript/manuscript.Rmd
Updating to use Bryna's middle initial

@@
   metrics in metabolomics'
-author: "William Kumler, Bryna Hazelton, and Anitra E. Ingalls"
+author: "William Kumler, Bryna J. Hazelton, and Anitra E. Ingalls"
 output:
@@
 WK extracted the Falkor samples, processed the data, performed the analyses, and wrote the manuscript.
-BH helped design the metrics and implement the regressions as well as providing support and context for the analysis.
+BJH helped design the metrics and implement the regressions as well as providing support and context for the analysis.
 AEI provided funding and data and helped to interpret the conclusions and edit the manuscript.

### 2023-07-28 manuscript/manuscript.Rmd
Homogenizing model names between figures

@@
   mutate(which_params=factor(which_params, 
-                             labels=c("All metrics", "XCMS-only metrics", "Raw data metrics"),
+                             labels=c("Full model", "XCMS-only model", "Two-parameter model"),
                              levels=c("all_params", "xcms_params", "raw_data_params"))) %>%

### 2023-07-28 manuscript/manuscript.Rmd
Making figures render larger, oops

@@
 Colors indicate the category in which each feature was manually assigned by an expert, with "Stans only" referring to a good mass feature that was only visible in the standards run alongside the samples.
-Culture datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (dotted black vertical line) according to the final model and were otherwise unclassified.](figures/fig_1_pred_prob_class_color_hists.png){width=90%}
+Culture datasets CultureData and Pttime were manually labeled only for those features with an estimated likelihood above 90% (dotted black vertical line) according to the final model and were otherwise unclassified.](figures/fig_1_pred_prob_class_color_hists.png){width=100%}
 
@@
 Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel.
-Lines of best fit have been estimated and plotted in black behind the data points, with the steeper slopes found in the full and XCMS-only models indicating overfitting on the training data.](figures/fig_2_fdr_gff_cross_train.png){width=90%}
+Lines of best fit have been estimated and plotted in black behind the data points, with the steeper slopes found in the full and XCMS-only models indicating overfitting on the training data.](figures/fig_2_fdr_gff_cross_train.png){width=100%}
 
@@
 The top row of plots show the exact likelihood predicted by the logistic model across three different subsets of parameters, while the bottom row shows the estimates ranked from least likely to most likely.
-Points are colored by their manually-assigned quality according to an expert.](figures/fig_3_comb_prob_space.png){width=90%}
+Points are colored by their manually-assigned quality according to an expert.](figures/fig_3_comb_prob_space.png){width=100%}
 
@@
 FDR is calculated by dividing the number of false positives by the total positives produced by the model and GFF is calculated by dividing the number of true positives by the total number of good features as identified manually (only possible for fully-labeled datasets).
-Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fig_4_fdr_gff_all_datasets.png){width=30%}
+Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fig_4_fdr_gff_all_datasets.png){width=50%}
 
@@
 Each mass feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile.
-Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the DCM; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/fig_5_clockplot_complete.png){width=90%}
+Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the DCM; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/fig_5_clockplot_complete.png){width=100%}
 
@@
 "Only good MFs" refers to those features manually labeled as "Good".
-NMDS stress values are reported in the upper right corner of each plot.](figures/fig_6_thresh_NMDS_plots.png){width=90%}
+NMDS stress values are reported in the upper right corner of each plot.](figures/fig_6_thresh_NMDS_plots.png){width=100%}
 
@@
 DCM = deep chlorophyll maximum, approximately 110 meters.
-Mass feature identifications are provided as the title of each panel, starting with "FT" and followed by 4 digits except for the two features annotated using authentic standards run alongside: the 13C isotope of dimethylsulfonioacetate (DMS-Ac) and taurine.](figures/fig_7_somegood_somebad_depthcolor_chroms.png){width=70%}
+Mass feature identifications are provided as the title of each panel, starting with "FT" and followed by 4 digits except for the two features annotated using authentic standards run alongside: the 13C isotope of dimethylsulfonioacetate (DMS-Ac) and taurine.](figures/fig_7_somegood_somebad_depthcolor_chroms.png){width=100%}
 
@@
 Normalization was performed by linearly scaling the raw values into the 0-1 range by subtracting the minimum value and dividing by the maximum.
-Peak shape similarity was measured with Pearson's correlation coefficient and the noise level is estimated as the standard deviation of the residuals after the raw data is subtracted from the idealized peak.](figures/fig_8_peakmetrics_singlechrom.png){width=90%}
+Peak shape similarity was measured with Pearson's correlation coefficient and the noise level is estimated as the standard deviation of the residuals after the raw data is subtracted from the idealized peak.](figures/fig_8_peakmetrics_singlechrom.png){width=100%}
 

### 2023-07-31 manuscript/manuscript.Rmd
Fixing reference style for BMC

@@
 bibliography: refs.bib
+csl: bmc-bioinformatics.csl
 ---

### 2023-07-31 manuscript/manuscript.Rmd
Moving supplement text into separate doc
Can be copy-pasted into the "label" spot? Currently unclear

@@
 
-## Supplement
+```{r supp figure text}
+supp_fig_1_text <- "Supplemental figure 1: Distribution and single-parameter logistic curves for each metric extracted for model training, shown separately for the MESOSCOPE and Falkor datasets.
+Histograms show the distribution of good and bad mass features by color across the span of the data on the x-axis with the number of MFs in each bin shown on the y-axis.
+Scatterplots show the same x-axis but show the results of a logistic regression on the single parameter, with the line of best fit in black and a ±1 standard error ribbon around it in grey.
+Vertical jittering has been applied when plotting to reduce the number of overlapping points."
+
+supp_fig_2_text <- 'Supplemental figure 2: Model parameter estimates for each of the metrics in the full model, additionally broken down by their inclusion in the two-parameter (raw_data) and XCMS-exclusively models.
+Colors correspond to the dataset used to train the logistic regression model, with "both" indicating a combined model using all manually-labeled features across both datasets.'
+
+supp_fig_3_text <- "Supplemental figure 3: Robustness of the two most significant metrics across the full (all_params), XCMS-only (xcms_params), and two-parameter (raw_data_params) models.
+The x-axis corresponds to the fraction of the data used to train the model and the y-coordinate shows the estimated value for the specified term in the subset across 10-fold replicated subsampling.
+The grey bar in the background corresponds to the estimate of the full model +/- 1SE (thinner dark grey bar) and 2SE (thicker light grey bar)."
+
+supp_fig_4_text <- "Supplemental figure 4: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets."
+
+comb_supp_text <- paste(supp_fig_1_text, supp_fig_2_text, supp_fig_3_text, supp_fig_4_text, sep = "\n\n")
+
+writeLines(comb_supp_text, "manuscript/supplement/supp_text_to_copy_paste.txt")
+```
 
@@
 
-*Supplemental figure 1: Distribution and single-parameter logistic curves for each metric extracted for model training, shown separately for the MESOSCOPE and Falkor datasets.
-Histograms show the distribution of good and bad mass features by color across the span of the data on the x-axis with the number of MFs in each bin shown on the y-axis.
-Scatterplots show the same x-axis but show the results of a logistic regression on the single parameter, with the line of best fit in black and a ±1 standard error ribbon around it in grey.
-Vertical jittering has been applied when plotting to reduce the number of overlapping points.*
-
 ```{r model stability dotplot errorbars supp figure}
@@
 
-*Supplemental figure 2: Model parameter estimates for each of the metrics in the full model, additionally broken down by their inclusion in the two-parameter (raw_data) and XCMS-exclusively models.
-Colors correspond to the dataset used to train the logistic regression model, with "both" indicating a combined model using all manually-labeled features across both datasets.*
-
 ```{r subsampling model stability supp figure, warning=FALSE}
@@
 ```
-
-*Supplemental figure 3: Robustness of the two most significant metrics across the full (all_params), XCMS-only (xcms_params), and two-parameter (raw_data_params) models.
-The x-axis corresponds to the fraction of the data used to train the model and the y-coordinate shows the estimated value for the specified term in the subset across 10-fold replicated subsampling.
-The grey bar in the background corresponds to the estimate of the full model +/- 1SE (thinner dark grey bar) and 2SE (thicker light grey bar).*
 
@@
 
-*Supplemental figure 4: Performance of regularized regression and random forest models on internally (same train-test) and externally (different train-test) validated datasets.*
-
 ```{r confusion matrices for all the models for supplement}

### 2023-07-31 manuscript/manuscript.Rmd
Fixing author affiliations

@@
   metrics in metabolomics'
-author: "William Kumler, Bryna J. Hazelton, and Anitra E. Ingalls"
+author: 
+  - William Kumler$^1$
+  - Bryna J. Hazelton$^{2, 3}$
+  - Anitra E. Ingalls$^1$*
+  - "Corresponding author email address: aingalls@uw.edu"
+  - "Author affiliations: $^1$University of Washington, School of Oceanography, Seattle, WA 98195, USA; $^2$University of Washington, eScience Institute, Seattle, WA 98195, USA; $^3$University of Washington, Department of Physics, Seattle, WA 98195, USA"
 output:

### 2023-08-02 manuscript/manuscript.Rmd
Renaming supp fig legend txt file
Combining supplemental tables into single Excel file
Also contains titles and legends

@@
 
-```{r full model estimate csv write}
-set.seed(123)
-both_datasets <- list(MESOSCOPE=MS3000_features, Falkor=FT2040_features) %>%
-  bind_rows(.id = "dataset") %>%
-  filter(feat_class%in%c("Good", "Bad")) %>%
-  mutate(feat_bin=ifelse(feat_class=="Good", 1, 0))
-
-glm_estimates <- both_datasets %>%
-  select(feat_bin, all_of(all_params)) %>%
-  # mutate(across(-feat_bin, \(x)scale(x)[,1])) %>%
-  glm(formula = feat_bin~., family=binomial) %>%
-  broom::tidy() %>%
-  rename_with(~paste0("GLM.", .x), -term)
-
-glmnet_estimates <- both_datasets %>% 
-  select(all_of(all_params), c("max_SNR", "max_cor", "medtop3_cor", "log_med_cor")) %>% 
-  data.matrix() %>%
-  glmnet::cv.glmnet(y=both_datasets$feat_bin, family="binomial", alpha=0.5) %>%
-  coef() %>%
-  as.matrix() %>%
-  as.data.frame() %>%
-  rownames_to_column("term") %>%
-  dplyr::rename(`Elastic.net.estimate`=s1)
-
-rf_estimates <- both_datasets %>% 
-  select(all_of(all_params), c("max_SNR", "max_cor", "medtop3_cor", "log_med_cor")) %>% 
-  randomForest::randomForest(factor(both_datasets$feat_bin), importance=TRUE) %>%
-  pluck("importance") %>%
-  as.data.frame() %>%
-  rownames_to_column("term") %>%
-  select(term, RF.accuracy.decrease=MeanDecreaseAccuracy,
-         RF.Gini.Decrease=MeanDecreaseGini)
-
-param_longnames %>%
-  as.data.frame() %>%
-  set_names("long_name") %>%
-  rownames_to_column("term") %>%
-  left_join(glm_estimates, by="term") %>%
-  left_join(glmnet_estimates, by="term") %>%
-  left_join(rf_estimates, by="term") %>%
-  write_csv("manuscript/supplement/supp_table_1_full_model_estimates.csv")
-```
-
 We explored the relative predictive power of the individual parameters using the full model and found that the predictors least likely to be different from zero due to chance were the mean *m/z* ratio, our novel peak shape correlation metric, and our novel SNR estimate, all with reported p-values < 10$^{-10}$.
@@
 
-writeLines(comb_supp_text, "manuscript/supplement/supp_text_to_copy_paste.txt")
+writeLines(comb_supp_text, "manuscript/supplement/supp_fig_legends.txt")
 ```
@@
 
-```{r confusion matrices for all the models for supplement}
+```{r supp table Excel write model params and confusion matrices}
+library(openxlsx)
+wb <- createWorkbook(creator = "William Kumler, wkumler@uw.edu")
+addWorksheet(wb, sheetName = "Table titles and legends")
+data.frame("Supplemental table titles and legends"=c(
+  "Supplemental table 1:",
+  "Model parameters and significance according to both logistic and random forest models",
+  "Model outputs from the full (all parameters) model trained on the combined Falkor",
+  "and MESOSCOPE datasets. The parameter name is reported alongside its more detailed",
+  "name, with the logistic model estimate, standard error, test statistic, and p-value",
+  "for each predictive term. We also included the estimate obtained from the elastic",
+  "net model (alpha=0.5) and the accuracy and Gini index decreases per term. NA values",
+  "are reported for several terms that were removed from the non-regularized regression",
+  "because they were too highly correlated with other parameters.",
+  "",
+  "Supplemental table 2:",
+  "Confusion matrices for logistic, regularized, and random forest regressions",
+  "Confusion matrices reported in long format for the different models reported in this",
+  "manuscript. Models were all both internally tested using the same dataset for testing",
+  "and training as well as cross-validated against the other fully labeled dataset. For",
+  "the two-parameter model, we also report the values obtained when training on the two",
+  "datasets combined (train = Both). For each model and test-train set we report the",
+  'absolute number of false positives (manually labeled as "Bad" but model labeled as',
+  '"Good"), false negatives (manually labeled as "Good" but model labeled as "Bad"),',
+  'true positives (both manual and model labelled as "Good"), and true negatives (both',
+  'manual and model labeled as "Bad") as well as the calculated false discovery rate',
+  "(FDR, calculated as FP/(FP+TP)) and the fraction of good features found (GFF,",
+  "calculated as TP/(TP+FN)). Unless otherwise specified, all values reported here",
+  "used a likelihood threshold of 0.5 for categorization."
+), check.names = FALSE) %>%
+  writeData(wb = wb, sheet = "Table titles and legends")
+
+addWorksheet(wb, sheetName = "Supp table 1 Model parameters")
+set.seed(123)
+both_datasets <- list(MESOSCOPE=MS3000_features, Falkor=FT2040_features) %>%
+  bind_rows(.id = "dataset") %>%
+  filter(feat_class%in%c("Good", "Bad")) %>%
+  mutate(feat_bin=ifelse(feat_class=="Good", 1, 0))
+glm_estimates <- both_datasets %>%
+  select(feat_bin, all_of(all_params)) %>%
+  # mutate(across(-feat_bin, \(x)scale(x)[,1])) %>%
+  glm(formula = feat_bin~., family=binomial) %>%
+  broom::tidy() %>%
+  rename_with(~paste0("GLM.", .x), -term)
+glmnet_estimates <- both_datasets %>% 
+  select(all_of(all_params), c("max_SNR", "max_cor", "medtop3_cor", "log_med_cor")) %>% 
+  data.matrix() %>%
+  glmnet::cv.glmnet(y=both_datasets$feat_bin, family="binomial", alpha=0.5) %>%
+  coef() %>%
+  as.matrix() %>%
+  as.data.frame() %>%
+  rownames_to_column("term") %>%
+  dplyr::rename(`Elastic.net.estimate`=s1)
+rf_estimates <- both_datasets %>% 
+  select(all_of(all_params), c("max_SNR", "max_cor", "medtop3_cor", "log_med_cor")) %>% 
+  randomForest::randomForest(factor(both_datasets$feat_bin), importance=TRUE) %>%
+  pluck("importance") %>%
+  as.data.frame() %>%
+  rownames_to_column("term") %>%
+  select(term, RF.accuracy.decrease=MeanDecreaseAccuracy,
+         RF.Gini.Decrease=MeanDecreaseGini)
+param_longnames %>%
+  as.data.frame() %>%
+  set_names("long_name") %>%
+  rownames_to_column("term") %>%
+  left_join(glm_estimates, by="term") %>%
+  left_join(glmnet_estimates, by="term") %>%
+  left_join(rf_estimates, by="term") %>%
+  writeData(wb = wb, sheet = "Supp table 1 Model parameters")
+
+addWorksheet(wb, sheetName = "Supp table 2 Confusion matrices")
 model_levels <- c("all_params", "xcms_params", "raw_data_params", 
@@
   select(model_type, train, test, TP, TN, FP, FN, FDR, GFF)
-
 fdr_gff_strict_all <- bind_rows(lst(FT2040_features, MS3000_features), .id = "dataset") %>%
@@
   select(model_type, train, test, TP, TN, FP, FN, FDR, GFF)
-
 fdr_gff_stricter_thresh <- error_type_all_datasets %>%
@@
   select(model_type, train, test, TP, TN, FP, FN, FDR, GFF)
-
 bind_rows(fdr_gff_individual, fdr_gff_full_model, fdr_gff_stricter_thresh) %>%
-  write_csv("manuscript/supplement/supp_table_2_model_confusion_matrices.csv")
-```
+  writeData(wb = wb, sheet = "Supp table 2 Confusion matrices")
+saveWorkbook(wb, file = "manuscript/supplement/BMC_supp_tables.xlsx", overwrite = TRUE)
+```

### 2023-08-02 manuscript/manuscript.Rmd
Adding country information to author affiliations

@@
   - "Corresponding author email address: aingalls@uw.edu"
-  - "Author affiliations: $^1$University of Washington, School of Oceanography, Seattle, WA 98195, USA; $^2$University of Washington, eScience Institute, Seattle, WA 98195, USA; $^3$University of Washington, Department of Physics, Seattle, WA 98195, USA"
+  - "Author affiliations: $^1$University of Washington, School of Oceanography, Seattle, WA 98195, United States; $^2$University of Washington, eScience Institute, Seattle, WA 98195, United States; $^3$University of Washington, Department of Physics, Seattle, WA 98195, United States"
 output:
